### 2018-02-13 paper/paper.tex
paper outline

\documentclass[twocolumn]{aastex62}

\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex	
\usepackage{xcolor}
\usepackage[sort&compress]{natbib}
\usepackage[hang,flushmargin]{footmisc}
\usepackage[counterclockwise]{rotating}

% units macros
\newcommand{\unit}[1]{\mathrm{#1}}
\newcommand{\km}{\unit{km}}
\newcommand{\m}{\unit{m}}
\newcommand{\s}{\unit{s}}
\newcommand{\kms}{\km\,\s^{-1}}
\newcommand{\ms}{\m\,\s^{-1}}
\newcommand{\ang}{\text{\normalfont\AA}}

% text macros
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\todo}[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\HARPS}{\project{\acronym{HARPS}}}
\newcommand{\HIRES}{\project{\acronym{HIRES}}}
\newcommand{\RV}{\acronym{RV}}
\newcommand{\CRLB}{\acronym{CRLB}}
\newcommand{\wobble}{\texttt{wobble}}

\begin{document}
\graphicspath{ {../figures/} }
\DeclareGraphicsExtensions{.pdf,.eps,.png}

\title{\textsc{\wobble: a data-driven method for precision radial velocities}}

\author[0000-0001-9907-7742]{Megan Bedell}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave., New York, NY 10010, USA}

\author[0000-0003-2866-9403]{David W. Hogg}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave., New York, NY 10010, USA}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
\affiliation{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affiliation{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

\author{Daniel Foreman-Mack\'{e}y}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave., New York, NY 10010, USA}

\author{Ben Montet}
\affiliation{Department of Astronomy and Astrophysics, University of Chicago, 5640 S. Ellis Ave, Chicago, IL 60637, USA}

\correspondingauthor{Megan Bedell}
\email{E-mail: mbedell@flatironinstitute.org}

\begin{abstract}
blah blah blah [mention high-quality telluric spectrum]
\end{abstract}

\section{Introduction}

Precise radial velocity (\RV) measurements are critical to the discovery and characterization of exoplanets. Many dedicated instruments exist. However, pipelines for deriving \RV s from spectra are generally not made public. In this work, we present an open-source code for \RV determination. We use a highly flexible linear model to extract \RV s in a fully data-driven way. %Our method simultaneously models the stellar and telluric spectra.

Overview of previous work on \acronym{EPRV}.

Say something about tellurics.

In this work, we focus on the ultra-stabilized spectrograph case, i.e. no absorption cell. We also assume that multiple epochs of observations exist and that these epochs are spread out across the observing season(s) to enable the stellar spectrum to be disentangled from telluric features. In this sense our pipeline is intended as a post-processing step, not a real-time data reduction service. We revisit these assumptions in Section \ref{s:future}.

\section{Method}

Our fundamental assumptions.

Linear model.

We run on each echelle order independently and then combine them.

\section{Code}

Here's how we implemented the method. blah blah TensorFlow blah.

Here's how you can download and play with the code yourself.\footnote{\url{https://www.github.com/megbedell/wobble}}

\section{Tests}
\subsection{51 Pegasi}

Description of the data, cite ESO archive.

Show our results: we find the planet with the correct mass and orbital parameters. Compare to \HARPS\ pipeline.

Plots of RV vs. time, models of star and tellurics in a few segments of the spectrum.

\subsection{TBD (a quiet M-dwarf?)}

\section{Generalizing \wobble}
\label{s:future}

Revisit the assumptions: single star, no gas cell, known instrumental wavelength solution, no intrinsic spectral variability, no additive terms, many spectra exist. 
In what situations are these assumptions violated, and how would we configure \wobble\ to deal with them?

\bibliographystyle{apj}
\bibliography{}%general,myref,inprep}

\end{document}

### 2018-02-15 paper/paper.tex
adapted abstract and intro from overleaf version

@@
 \newcommand{\m}{\unit{m}}
+\newcommand{\cm}{\unit{cm}}
 \newcommand{\s}{\unit{s}}
-\newcommand{\kms}{\km\,\s^{-1}}
-\newcommand{\ms}{\m\,\s^{-1}}
+\newcommand{\kms}{$\km\,\s^{-1}$}
+\newcommand{\cms}{$\cm\,\s^{-1}$}
+\newcommand{\ms}{$\m\,\s^{-1}$}
 \newcommand{\ang}{\text{\normalfont\AA}}
@@
 \newcommand{\foreign}[1]{\textsl{#1}}
+\newcommand{\code}[1]{\texttt{#1}}
 \newcommand{\HARPS}{\project{\acronym{HARPS}}}
@@
 \newcommand{\RV}{\acronym{RV}}
+\newcommand{\EPRV}{\acronym{EPRV}}
 \newcommand{\CRLB}{\acronym{CRLB}}
-\newcommand{\wobble}{\texttt{wobble}}
+\newcommand{\wobble}{\code{wobble}}
+\newcommand{\TF}{\code{TensorFlow}}
+\newcommand{\python}{\code{python}}
 
@@
 
-\author{Ben Montet}
+\author{Benjamin T. Montet}
 \affiliation{Department of Astronomy and Astrophysics, University of Chicago, 5640 S. Ellis Ave, Chicago, IL 60637, USA}
@@
 \begin{abstract}
-blah blah blah [mention high-quality telluric spectrum]
+% Context
+Telluric absorption features in stellar spectra provide particular challenges to extreme-precision radial velocity (\EPRV) exoplanet surveys. 
+Large amplitude features are often masked in data analysis pipelines, reducing the number of observable spectral features which can be used to measure an \RV\ shift.
+Low-amplitude features that are hard to see even in high signal-to-noise spectra may be an important contributor to the overall \RV\ noise budget.
+% Aims
+Here we propose a data-driven method to simultaneously extract precise \RV s and infer the underlying stellar and telluric spectra using a fully linear model. 
+We implement this method in \wobble, an open-source \python\ package which uses \TF\ in its first non-machine-learning application to astronomical data. 
+In this work, we demonstrate the performance of \wobble\ on archival \HARPS\ spectra.
+% Results
+We recover the canonical exoplanet 51 Pegasi b at improved accuracy relative to the standard \HARPS\ pipeline \RV s, and we achieve a precision of \todo{xx} \cms\ on the \RV\ standard star \todo{yyy}.
+We also present a detailed telluric spectrum derived from these data. 
+Our method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.
 \end{abstract}
@@
 Precise radial velocity (\RV) measurements are critical to the discovery and characterization of exoplanets. 
-Many dedicated instruments exist.
-However, pipelines for deriving \RV s from spectra are generally not made public.
-In this work, we present an open-source code for \RV determination.
+On order of one dozen dedicated spectrographs exist for the purpose of \RV\ planet-hunting, with at least as many more currently under construction \todo{(Wright)}. 
+However, significant challenges exist in deriving precise \RV\ measurements from these spectra. 
+%The current capabilities of extreme-precision radial velocity (\EPRV) instruments do not extend to the 10 \cms\ regime, 
+%In this work, we present an open-source code for \RV\ determination.
 We use a highly flexible linear model to extract \RV s in a fully data-driven way.
@@
 
-Overview of previous work on \acronym{EPRV}.
+\todo{Overview of previous work on \acronym{EPRV}.}
 
-Say something about tellurics.
+One of the primary drivers of the \RV\ noise budget is the incomplete treatment of telluric features in the
+Earth's atmosphere (cite). 
+Often, particular sections of a spectrum that are likely to feature telluric features are identified before the velocity shift of the stellar spectrum is inferred. 
+These regions are then removed from analysis, leaving only telluric-free regions to be analyzed (e.g. cite).
 
-In this work, we focus on the ultra-stabilized spectrograph case, i.e.
-no absorption cell.
-We also assume that multiple epochs of observations exist and that these epochs are spread out across the observing season(s) to enable the stellar spectrum to be disentangled from telluric features.
+Such an approach has two significant issues. 
+The first is that removing sections of the spectrum removes significant regions of the spectrum that can be used to infer the stellar radial velocity. 
+Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are many narrow spectral features that can be used to improve RV precision (cite).
+This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres (cite).
+Eliminating large chunks of these spectra will therefore significantly inhibit our ability to detect planets around M dwarfs through RVs.
+
+Secondly, not all telluric features are obvious. 
+The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred RV at the $\sim 1$ \ms\ level (Cunha14).
+As the locations of these features are not known \textit{a priori} and may not even be apparent in stacked spectra of many observations, these spectral regions cannot be thrown out. 
+Instead, alternative methods to account for these features must be developed and employed in order to mitigate the effect of the Earth's atmosphere on the measured stellar radial velocities.
+
+\todo{Telluric models exist but are bad. Telluric observations can be taken throughout the night but require a large investment of observing time.}
+
+One such alternative approach is the simultaneous modeling of both telluric and spectral features. 
+As the Earth's motion around the barycenter of the solar system induces a Doppler shift considerably larger than both the motion of telluric features and the size of a single pixel on the detector, these two spectra can be disentangled.
+This process is well-established in the analysis of binary star systems through the development of linear models
+(e.g. Simon94) and in a Gaussian process framework (Czekala17).
+In these cases, both spectra are assumed to be unchanging in time, which is a reasonable approximation of a stellar spectrum but not necessarily of the telluric spectrum.
+A more complicated linear model may be useful for separating the stellar and telluric spectra, enabling a reconstruction of both at each epoch.
+
+Here we develop a linear data-driven model to infer a telluric and stellar spectrum and calculate the RVs for both components at each observed epoch. 
+This model requires no prior knowledge of the star or the Earth's atmosphere. 
+As such, it does not yield absolute measurements of \RV s, only highly precise relative measurements between epochs.
+
+In this work, we focus on the ultra-stabilized spectrograph case, i.e. no absorption cell. 
+We also assume that multiple epochs of observations exist and that these epochs are spread out across the observing season(s). 
+This assumption is necessary to enable the disentangling of telluric features from the stellar spectrum. 
 In this sense our pipeline is intended as a post-processing step, not a real-time data reduction service.
-We revisit these assumptions in Section \ref{s:future}.
 
-\section{Method}
+In Section \ref{s:methods}, we outline the model and present an open-source implementation in \python\ and \TF\ called \wobble. 
+In Section \ref{s:results}, we apply our method to \HARPS\ archival data for two target stars, the planet-hosting solar analog 51 Peg and the quiet M dwarf \todo{yyy}, as a demonstration of \wobble's capabilities.
+\todo{We find xyz.} 
+We conclude in Section \ref{s:future} with a detailed look at the limitations of the current implementation and outline ways to adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and multiple-star systems.
 
-Our fundamental assumptions.
+\section{Methods}
+\label{s:methods}
+\subsection{Model}
 
-Linear model.
+\todo{Our fundamental assumptions: single, unchanging stellar spectrum with \RV\ shifts; convolved telluric features which can vary in both \RV\ and shape.
+Instrument-provided wavelength solution which is good.}
 
-We run on each echelle order independently and then combine them.
+Our data are the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed ln(flux) for pixel $m$ of M at epoch $n$ of N. We also have a wavelength solution vector $\lambda$ of length M.
 
-\section{Code}
+Our model is comprised of three major contributions to the observed spectrum: the star, which has a radial velocity about its center of mass $v_{\star}$ combined with motion of the Earth around the solar system barycenter; the atmospheric telluric spectrum, which has an effective radial velocity $v_t$ at each epoch; and the instrument, which has no velocity shift.
+Each of these components may vary with time in their spectral shape.
+We make the assumption that these variations can be captured in a low-dimensional space.
 
-Here's how we implemented the method. blah blah TensorFlow blah.
+Each data column $y_n$ is modeled as:
+$$y_n = P(v_{\star, n}) W_{\star} x_n + P(v_{t, n}) W_{t} z_n + W_{c} u_n + noise,$$
+where $P$ is an interpolation operator which applies a Doppler shift $D_v$ and $W$ is a matrix of ``principal components'' which are weighted by $x_n, z_n, u_n$ to form the spectral contribution at epoch $n$.
+Each $W$ takes the shape $M' \times K$, where $M'$ is the length of a wavelength-space grid $\lambda'$ which may be different from the data wavelength grid $\lambda$ and $K$ is the number of components to be used (where $K_{\star}$, $K_t$, and $K_c$ may be different numbers).
 
-Here's how you can download and play with the code yourself.\footnote{\url{https://www.github.com/megbedell/wobble}}
+\todo{discuss degeneracies in the model}
 
-\section{Tests}
+We build an initial template for both the stellar and telluric spectra by assuming a Doppler shift for each (equal to the barycentric correction and zero, respectively) and calculating the median flux at each model wavelength $\lambda'$ after that shift is applied.
+With these models in hand, we are then able to infer an RV shift for both model components at each epoch.
+We evaluate a likelihood for each epoch such that
+$$ \ln \mathcal{L} = \sum_{m} -0.5 (y_m - y_m')^T C^{-1} (y_m-y_m'),
+$$
+comparing our model $y'$ to the data $y$, with $C$ representing our covariance matrix of the uncertainties on each data point.
+Our model is agnostic to the intrinsic wavelength of any individual stellar or telluric feature, so while we infer the  Doppler shift relative to our model, it does not apply any information about the absolute \RV\ of the star relative to the solar system.
+To mitigate the potential degeneracy of developing a model effectively redshifted relative to all observations, we add (in log space) a Gaussian prior to our likelihood value on the mean \RV\ for all $N$ epochs such that
+$$ P(m|X) = \ln \mathcal{L}  -0.5 * \frac{\sum(v)^2}{(Nl)^2}.
+$$
+\todo{someone should define these variables}
+
+We use the BFGS algorithm %\citep{Broyden70} 
+as packaged in scipy.optimize %citep{Jones01}
+to maximize this likelihood function for our assumed stellar and telluric models iteratively, first calculating the best-fitting stellar \RV shifts holding the telluric shifts constant, then repeating for the telluric spectrum shift holding the stellar \RV s constant.
+After each optimization, we use these velocities to improve our models for the telluric and stellar spectra, employing a gradient descent scheme to modify our templates $m'$ at each wavelength $\lambda'$ using the newly calculated velocities.
+We repeat this procedure, calculating velocities for each epoch and in turn using these velocities to infer an improved model spectrum, until the procedure converges.
+
+\subsection{Implementation}
+
+\todo{Here's how we implemented the method. blah blah TensorFlow blah.}
+
+In its current realization, \wobble\ is optimized for use with data products from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \todo{(Mayor+2003)}.
+Like most \EPRV\ instruments, \HARPS\ is an echelle spectrograph spanning many spectral orders.
+These orders can be considered as independent spectra \todo{for reasons}.
+\todo{We run on each echelle order independently and then combine them like so:}
+
+Our code is open-source and publicly available.\footnote{\url{https://www.github.com/megbedell/wobble}}
+
+\section{Results}
+\label{s:results}
 \subsection{51 Pegasi}
 
-Description of the data, cite ESO archive.
+Description of the data, cite \acronym{ESO} archive.
 

### 2018-02-20 paper/paper.tex
minor edits

@@
 Secondly, not all telluric features are obvious. 
-The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred RV at the $\sim 1$ \ms\ level (Cunha14).
+The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred RV at the $\sim 1$ \ms\ level \todo{(Cunha14)}.
 As the locations of these features are not known \textit{a priori} and may not even be apparent in stacked spectra of many observations, these spectral regions cannot be thrown out. 
@@
 
-We use the BFGS algorithm %\citep{Broyden70} 
-as packaged in scipy.optimize %citep{Jones01}
-to maximize this likelihood function for our assumed stellar and telluric models iteratively, first calculating the best-fitting stellar \RV shifts holding the telluric shifts constant, then repeating for the telluric spectrum shift holding the stellar \RV s constant.
+We maximize this likelihood function for our assumed stellar and telluric models iteratively, first calculating the best-fitting stellar \RV shifts holding the telluric shifts constant, then repeating for the telluric spectrum shift holding the stellar \RV s constant.
 After each optimization, we use these velocities to improve our models for the telluric and stellar spectra, employing a gradient descent scheme to modify our templates $m'$ at each wavelength $\lambda'$ using the newly calculated velocities.
 We repeat this procedure, calculating velocities for each epoch and in turn using these velocities to infer an improved model spectrum, until the procedure converges. 
+\todo{(say something about model linearity and why this iterative procedure is effective)}
 
-\subsection{Implementation}
+\subsection{Code}
 
-\todo{Here's how we implemented the method. blah blah TensorFlow blah.}
+The above-described model can be implemented in a variety of ways.
+We chose to build our code, \wobble, using the \TF\ package \todo{[terminology?]} in \code{python}.
+\TF\ is a \todo{[blah blah technical blah]}.
+While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this use case as well.
+
+\todo{(more details of implementation)}
 
@@
 
-Our code is open-source and publicly available.\footnote{\url{https://www.github.com/megbedell/wobble}}
+Our code is open-source and publicly available on GitHub.\footnote{\url{https://www.github.com/megbedell/wobble}}
 
@@
 
-Description of the data, cite \acronym{ESO} archive.
+\todo{Description of the data, cite \acronym{ESO} archive.}
 
-Show our results: we find the planet with the correct mass and orbital parameters. Compare to \HARPS\ pipeline.
+\todo{Show our results: we find the planet with the correct mass and orbital parameters. Compare to \HARPS\ pipeline.}
 
-Plots of RV vs. time, models of star and tellurics in a few segments of the spectrum.
+\todo{Plots of RV vs. time, models of star and tellurics in a few segments of the spectrum.}
 
 \subsection{TBD (a quiet M-dwarf?)}
+
+\todo{Results \& more plots.}
 
@@
 
-Revisit the assumptions: single star, no gas cell, known instrumental wavelength solution, no intrinsic spectral variability, no additive terms, many spectra exist. 
-In what situations are these assumptions violated, and how would we configure \wobble\ to deal with them?
+\todo{Revisit the assumptions: single star, no gas cell, known instrumental wavelength solution, no intrinsic spectral variability, no additive terms, many spectra exist. 
+In what situations are these assumptions violated, and how would we configure \wobble\ to deal with them?}
 

### 2018-07-12 paper/paper.tex
more intro

@@
 
+\newcommand{\Mdwarf}{Barnard's Star}
+
 \begin{document}
@@
 % Results
-We recover the canonical exoplanet 51 Pegasi b at improved accuracy relative to the standard \HARPS\ pipeline \RV s, and we achieve a precision of \todo{xx} \cms\ on the \RV\ standard star \todo{yyy}.
+We recover the canonical exoplanet 51 Pegasi b at \todo{improved accuracy?} relative to the standard \HARPS\ pipeline \RV s, and we achieve a precision of \todo{xx} \cms\ on the \RV\ standard M dwarf \Mdwarf.
 We also present a detailed telluric spectrum derived from these data. 
@@
 Precise radial velocity (\RV) measurements are critical to the discovery and characterization of exoplanets. 
-On order of one dozen dedicated spectrographs exist for the purpose of \RV\ planet-hunting, with at least as many more currently under construction \todo{(Wright)}. 
+On order of one dozen dedicated spectrographs exist for the purpose of \RV\ planet-hunting, with at least as many more currently under construction \citep{Wright2017}. 
 However, significant challenges exist in deriving precise \RV\ measurements from these spectra. 
@@
 
-\todo{Overview of previous work on \acronym{EPRV}.}
-
 One of the primary drivers of the \RV\ noise budget is the incomplete treatment of telluric features in the
-Earth's atmosphere (cite). 
+Earth's atmosphere \citep{Halverson2016}. 
 Often, particular sections of a spectrum that are likely to feature telluric features are identified before the velocity shift of the stellar spectrum is inferred. 
-These regions are then removed from analysis, leaving only telluric-free regions to be analyzed (e.g. cite).
+These regions are then removed from analysis, leaving only telluric-free regions to be analyzed \citep[e.g.][]{AngladaEscude2012}.
 
@@
 The first is that removing sections of the spectrum removes significant regions of the spectrum that can be used to infer the stellar radial velocity. 
-Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are many narrow spectral features that can be used to improve RV precision (cite).
-This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres (cite).
+Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve RV precision (cite).
+This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres \citep{Figueira2016}.
 Eliminating large chunks of these spectra will therefore significantly inhibit our ability to detect planets around M dwarfs through RVs.
@@
 Secondly, not all telluric features are obvious. 
-The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred RV at the $\sim 1$ \ms\ level \todo{(Cunha14)}.
+The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred RV at the $\sim 1$ \ms\ level \citep{Cunha2014}.
 As the locations of these features are not known \textit{a priori} and may not even be apparent in stacked spectra of many observations, these spectral regions cannot be thrown out. 
@@
 
-\todo{Telluric models exist but are bad. Telluric observations can be taken throughout the night but require a large investment of observing time.}
+One such approach is modeling the telluric spectrum using existing line databases like HITRAN \citep{HITRAN2016}. 
+The telluric model may then be divided out from the observations, assuming the line spread function of the instrument is known \citep[e.g.][]{Seifahrt2010}. 
+This method relies on existing physical knowledge about the Earth's atmosphere and can be fine-tuned using local observatory measurements of e.g. atmospheric water vapor content \citep{Baker2017}. 
+However, line databases are incomplete even in significant absorption features when compared to actual observations and certainly do not include microtellurics, making them poorly suited for extreme precision RV applications \citep{Bertaux2014}.
 
-One such alternative approach is the simultaneous modeling of both telluric and spectral features. 
+Another option is the use of telluric standard observations: a spectrum of a rapidly rotating early-type star, which is virtually featureless due to extreme rotational line broadening, may be used as a telluric model and divided out.
+This approach has the advantage of naturally reproducing the instrumental line profile and current observing conditions if the standard star has a line-of-sight vector sufficiently close to the target and if both observations are taken close together in time.
+For these conditions to be true, though, requires a significant investment of observing time, which planet search programs often cannot afford. 
+Additionally, artifacts may remain near strong telluric features due to the imperfect correction of unresolved features \citep{Bailey2007}.
+
+An alternative approach is the simultaneous modeling of both telluric and spectral features from the data. 
 As the Earth's motion around the barycenter of the solar system induces a Doppler shift considerably larger than both the motion of telluric features and the size of a single pixel on the detector, these two spectra can be disentangled.
 This process is well-established in the analysis of binary star systems through the development of linear models
-(e.g. Simon94) and in a Gaussian process framework (Czekala17).
+\citep[e.g.][]{Simon1994} and in a Gaussian process framework \citep{Czekala2017}.
 In these cases, both spectra are assumed to be unchanging in time, which is a reasonable approximation of a stellar spectrum but not necessarily of the telluric spectrum.
 A more complicated linear model may be useful for separating the stellar and telluric spectra, enabling a reconstruction of both at each epoch.
+\todo{(say something about Artigau2014 here)}
 
-Here we develop a linear data-driven model to infer a telluric and stellar spectrum and calculate the RVs for both components at each observed epoch. 
-This model requires no prior knowledge of the star or the Earth's atmosphere. 
+\todo{[Other works that have done similar stuff -- be sure to cite: Artigau2018, Gao2016]}
+
+Here we develop a linear data-driven model to infer a telluric and stellar spectrum and calculate the stellar RV at each observed epoch. 
+The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
+Our model requires no prior knowledge of the star or the Earth's atmosphere. 
 As such, it does not yield absolute measurements of \RV s, only highly precise relative measurements between epochs.
@@
 In Section \ref{s:methods}, we outline the model and present an open-source implementation in \python\ and \TF\ called \wobble. 
-In Section \ref{s:results}, we apply our method to \HARPS\ archival data for two target stars, the planet-hosting solar analog 51 Peg and the quiet M dwarf \todo{yyy}, as a demonstration of \wobble's capabilities.
+In Section \ref{s:results}, we apply our method to \HARPS\ archival data for two target stars, the planet-hosting solar analog 51 Peg and the quiet M dwarf \Mdwarf, as a demonstration of \wobble's capabilities.
 \todo{We find xyz.} 
@@
 While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this use case as well.
+In particular, the necessary optimizations can be performed with high efficiency by \TF: the below-described analysis of \todo{x} spectra of 51 Peg runs in \todo{y} hours on a standard Macbook laptop.
 
@@
 
-In its current realization, \wobble\ is optimized for use with data products from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \todo{(Mayor+2003)}.
+In its current realization, \wobble\ is optimized for use with data products from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}.
 Like most \EPRV\ instruments, \HARPS\ is an echelle spectrograph spanning many spectral orders.
@@
 
-\subsection{TBD (a quiet M-dwarf?)}
+\subsection{\Mdwarf}
 
 \todo{Results \& more plots.}
+
+Compare to HARPS-TERRA results for this star \citep{AngladaEscude2012}.
 
@@
 
+Intrinsic stellar variability: cite \citet{Davis2017} for PCA approach
+
 \bibliographystyle{apj}
-\bibliography{}%general,myref,inprep}
+\bibliography{paper.bib}%general,myref,inprep}
 

### 2018-07-13 paper/paper.tex
edited equations

@@
 The first is that removing sections of the spectrum removes significant regions of the spectrum that can be used to infer the stellar radial velocity. 
-Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve RV precision (cite).
+Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV\ precision (cite).
 This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres \citep{Figueira2016}.
-Eliminating large chunks of these spectra will therefore significantly inhibit our ability to detect planets around M dwarfs through RVs.
+Eliminating large chunks of these spectra will therefore significantly inhibit our ability to detect planets around M dwarfs through \RV s.
 
 Secondly, not all telluric features are obvious. 
-The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred RV at the $\sim 1$ \ms\ level \citep{Cunha2014}.
+The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred \RV\ at the $\sim 1$ \ms\ level \citep{Cunha2014}.
 As the locations of these features are not known \textit{a priori} and may not even be apparent in stacked spectra of many observations, these spectral regions cannot be thrown out. 
@@
 
-One such approach is modeling the telluric spectrum using existing line databases like HITRAN \citep{HITRAN2016}. 
+One such approach is modeling the telluric spectrum using existing line databases like \acronym{HITRAN} \citep{HITRAN2016}. 
 The telluric model may then be divided out from the observations, assuming the line spread function of the instrument is known \citep[e.g.][]{Seifahrt2010}. 
@@
 Each data column $y_n$ is modeled as:
-$$y_n = P(v_{\star, n}) W_{\star} x_n + P(v_{t, n}) W_{t} z_n + W_{c} u_n + noise,$$
-where $P$ is an interpolation operator which applies a Doppler shift $D_v$ and $W$ is a matrix of ``principal components'' which are weighted by $x_n, z_n, u_n$ to form the spectral contribution at epoch $n$.
+%$$y_n = P(v_{\star, n}) W_{\star} x_n + P(v_{t, n}) W_{t} z_n + W_{c} u_n + noise,$$
+%where $P$ is an interpolation operator which applies a Doppler shift $D_v$ and $W$ is a matrix of ``principal components'' which are weighted by $x_n, z_n, u_n$ to form the spectral contribution at epoch $n$.
 Each $W$ takes the shape $M' \times K$, where $M'$ is the length of a wavelength-space grid $\lambda'$ which may be different from the data wavelength grid $\lambda$ and $K$ is the number of components to be used (where $K_{\star}$, $K_t$, and $K_c$ may be different numbers).
+$$y_n = P(v_{\star, n}) \mu_{\star} + P(v_{t, n}) (  \mu_{t} + W_{t} z_n ) + noise,$$
+where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu$ is the mean spectral template for each component.
+In addition to its mean spectrum $\mu_t$, the telluric component also takes a time-dependent component assembled from two variables: $W$, a matrix of ``principal components,'' is weighted by $z_n$ to form the spectral contribution at epoch $n$.
+Each $\mu$ has length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid; these grids may in principle be different lengths for different spectral components, but for simplicity we use the same grid for both $\mu$.
+Correspondingly, $W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of principle components.
+
+Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
+$$ \log \lambda(v) = \log \lambda_{0} + \log \left(\frac{1 - v/c}{1 + v/c}\right).
+$$
+This shift is applied to the $M'$-length wavelength grid for the template spectrum before interpolating the template to the data space.
 
@@
 We build an initial template for both the stellar and telluric spectra by assuming a Doppler shift for each (equal to the barycentric correction and zero, respectively) and calculating the median flux at each model wavelength $\lambda'$ after that shift is applied.
-With these models in hand, we are then able to infer an RV shift for both model components at each epoch.
+With these models in hand, we are then able to infer an \RV\ shift for both model components at each epoch.
 We evaluate a likelihood for each epoch such that
@@
 
-We maximize this likelihood function for our assumed stellar and telluric models iteratively, first calculating the best-fitting stellar \RV shifts holding the telluric shifts constant, then repeating for the telluric spectrum shift holding the stellar \RV s constant.
+We maximize this likelihood function for our assumed stellar and telluric models iteratively, first calculating the best-fitting stellar \RV\ shifts holding the telluric shifts constant, then repeating for the telluric spectrum shift holding the stellar \RV s constant.
 After each optimization, we use these velocities to improve our models for the telluric and stellar spectra, employing a gradient descent scheme to modify our templates $m'$ at each wavelength $\lambda'$ using the newly calculated velocities.

### 2018-07-13 wobble/make_data_2d.py
saving in case I break everything

@@ -1,7 +1,7 @@
 import numpy as np
 from scipy.io.idl import readsav
 from scipy.interpolate import interp1d
-from harps_hacks import read_harps, rv_model 
+from harps_hacks import read_harps
 import h5py
 import math
 from astropy.io import fits
@@ -14,7 +14,7 @@ def dimensions(instrument):
         M = 4096 # pixels per order
         R = 72 # orders
     else:
-        print "instrument not recognized. valid options are: HARPS"
+        print("instrument not recognized. valid options are: HARPS")
         return
     return M, R
 
@@ -35,7 +35,8 @@ def read_data_from_fits(filelist):
         airms[n] = sp[0].header['HIERARCH ESO TEL AIRM START']
         drifts[n] = sp[0].header['HIERARCH ESO DRS DRIFT SPE RV']  
         
-        spec_file = str.replace(f, 'ccf_G2', 'e2ds')  
+        spec_file = str.replace(f, 'ccf_G2', 'e2ds') 
+        spec_file = str.replace(f, 'ccf_M2', 'e2ds') 
         try:
             wave, spec = read_harps.read_spec_2d(spec_file)
         except:
@@ -74,7 +75,7 @@ def read_data_from_savfile(savfile):
     empty = np.array([], dtype=int)
     for n,(f,b,snr) in enumerate(zip(s.files, s.berv, s.snr)):
             # read in the spectrum
-            spec_file = str.replace(f, 'ccf_G2', 'e2ds')
+            spec_file = str.replace(f, 'ccf_G2', 'e2ds')            
             try:
                 wave, spec = read_harps.read_spec_2d(spec_file)
             except:
@@ -144,23 +145,37 @@ def write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffi
 
 if __name__ == "__main__":
 
-    if False: # HIP54287
+    if False: #HIP54287
         data_dir = "/Users/mbedell/Documents/Research/HARPSTwins/Results/"
         savfile = data_dir+'HIP54287_result.dat'
         ccf_filelist = savfile_to_filelist(savfile)
     
-        if False: # missing wavelength files
+        if False: # check for missing wavelength files
             e2ds_filelist = [str.replace(f, 'ccf_G2', 'e2ds') for f in ccf_filelist]
             missing_files = missing_wavelength_files(e2ds_filelist)
             np.savetxt('missing_files.txt', missing_files, fmt='%s')
[…]

### 2018-07-13 wobbleflow_test.py

@@ -6,7 +6,7 @@ from time import time
 import h5py
 
 if __name__ == "__main__":
-    starname = 'hip54287'
+    starname = '51peg'
     
     if False:
         # quick single-order test
@@ -14,9 +14,10 @@ if __name__ == "__main__":
         model = wobble.Model(data)
         model.add_star(starname)
         model.add_telluric('tellurics', rvs_fixed=True)
-        nll_history, rvs_history, template_history, chis_history = wobble.optimize_order(model, data, 0, 
+        histories = wobble.optimize_order(model, data, 0, 
                 niter=50, output_history=True)
-                
+        (nll_history, rvs_history, template_history, chis_history, \
+            basis_vectors_history, basis_weights_history) = histories        
         assert False
     
     
@@ -28,7 +29,7 @@ if __name__ == "__main__":
     
     model = wobble.Model(data)
     model.add_star(starname)
-    K = 3
+    K = 2
     model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K)
     print(model)
     print("time elapsed: {0:.2f} s".format(time() - start_time))
@@ -39,7 +40,7 @@ if __name__ == "__main__":
     plot_dir = 'results/plots/'
     niter = 80
     for r in range(data.R):
-        if False: # no plots
+        if True: # no plots
             wobble.optimize_order(model, data, r, 
                         niter=niter, output_history=False)
         else: # plots out the wazoo
@@ -80,9 +81,10 @@ if __name__ == "__main__":
         
         # save telluric variability:
         session = wobble.get_session()
-        c = model.components[1]
-        telluric_basis_vectors[r,:,:] = np.copy(session.run(c.basis_vectors[r]))
-        telluric_basis_weights[r,:,:] = np.copy(session.run(c.basis_weights[r]))
+        if K > 0:
+            c = model.components[1]
+            telluric_basis_vectors[r,:,:] = np.copy(session.run(c.basis_vectors[r]))
+            telluric_basis_weights[r,:,:] = np.copy(session.run(c.basis_weights[r]))
                     
         # HACK to save everything else:    
         data.wobble_obj.rvs_star[r] = session.run(model.components[0].rvs_block[r])          
@@ -94,10 +96,11 @@ if __name__ == "__main__":
         session.close()
         print("order {1} saved. time elapsed: {0:.2f} s".format(time() - start_time, r))
                 
-    # save telluric variability:    
[…]

### 2018-09-04 paper/paper.tex
docstrings and paper edits

--- 2018-07-12 paper/paper.tex
@@
-\documentclass[twocolumn]{aastex62}
+\documentclass[modern]{aastex62}
 
-\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex	
-\usepackage{xcolor}
-\usepackage[sort&compress]{natbib}
-\usepackage[hang,flushmargin]{footmisc}
-\usepackage[counterclockwise]{rotating}
-
-% units macros
-\newcommand{\unit}[1]{\mathrm{#1}}
-\newcommand{\km}{\unit{km}}
-\newcommand{\m}{\unit{m}}
-\newcommand{\cm}{\unit{cm}}
-\newcommand{\s}{\unit{s}}
-\newcommand{\kms}{$\km\,\s^{-1}$}
-\newcommand{\cms}{$\cm\,\s^{-1}$}
-\newcommand{\ms}{$\m\,\s^{-1}$}
-\newcommand{\ang}{\text{\normalfont\AA}}
-
-% text macros
-\newcommand{\documentname}{\textsl{Article}}
-\newcommand{\sectionname}{Section}
-\newcommand{\todo}[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
-\newcommand{\acronym}[1]{{\small{#1}}}
-\newcommand{\project}[1]{\textsl{#1}}
-\newcommand{\foreign}[1]{\textsl{#1}}
-\newcommand{\code}[1]{\texttt{#1}}
-\newcommand{\HARPS}{\project{\acronym{HARPS}}}
-\newcommand{\HIRES}{\project{\acronym{HIRES}}}
-\newcommand{\RV}{\acronym{RV}}
-\newcommand{\EPRV}{\acronym{EPRV}}
-\newcommand{\CRLB}{\acronym{CRLB}}
-\newcommand{\wobble}{\code{wobble}}
-\newcommand{\TF}{\code{TensorFlow}}
-\newcommand{\python}{\code{python}}
-
+\include{preamble}
 \newcommand{\Mdwarf}{Barnard's Star}
 
-\begin{document}
-\graphicspath{ {../figures/} }
+\setlength{\parindent}{1.4em} % trust in Hogg
+\begin{document}\sloppy\sloppypar\raggedbottom\frenchspacing % trust in Hogg
+
+\shorttitle{\wobble}
+\shortauthors{Bedell et al.}
+
+\graphicspath{ {figures/} }
 \DeclareGraphicsExtensions{.pdf,.eps,.png}
@@
 \author[0000-0001-9907-7742]{Megan Bedell}
-\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave., New York, NY 10010, USA}
+\affiliation{\flatiron}
 
 \author[0000-0003-2866-9403]{David W. Hogg}
-\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave., New York, NY 10010, USA}
+\affiliation{\flatiron}
 \affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
@@
 \author{Daniel Foreman-Mack\'{e}y}
-\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave., New York, NY 10010, USA}
+\affiliation{\flatiron}
 
 \author{Benjamin T. Montet}
-\affiliation{Department of Astronomy and Astrophysics, University of Chicago, 5640 S. Ellis Ave, Chicago, IL 60637, USA}
+\affiliation{\chicago}
 
@@
 Telluric absorption features in stellar spectra provide particular challenges to extreme-precision radial velocity (\EPRV) exoplanet surveys. 
-Large amplitude features are often masked in data analysis pipelines, reducing the number of observable spectral features which can be used to measure an \RV\ shift.
-Low-amplitude features that are hard to see even in high signal-to-noise spectra may be an important contributor to the overall \RV\ noise budget.
+Large amplitude features are often masked in data analysis pipelines, reducing the number of observable spectral features which can be used to measure an \RV shift.
+Low-amplitude features that are hard to see even in high signal-to-noise spectra may be an important contributor to the overall \RV noise budget.
 % Aims
 Here we propose a data-driven method to simultaneously extract precise \RV s and infer the underlying stellar and telluric spectra using a fully linear model. 
-We implement this method in \wobble, an open-source \python\ package which uses \TF\ in its first non-machine-learning application to astronomical data. 
-In this work, we demonstrate the performance of \wobble\ on archival \HARPS\ spectra.
+We implement this method in \wobble, an open-source \python package which uses \TF in its first non-machine-learning application to astronomical data. 
+In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
 % Results
-We recover the canonical exoplanet 51 Pegasi b at \todo{improved accuracy?} relative to the standard \HARPS\ pipeline \RV s, and we achieve a precision of \todo{xx} \cms\ on the \RV\ standard M dwarf \Mdwarf.
+We recover the canonical exoplanet 51 Pegasi b at \todo{improved accuracy?} relative to the standard \HARPS pipeline, and we achieve a precision of \todo{xx} \cms on the \RV standard M dwarf \Mdwarf. 
 We also present a detailed telluric spectrum derived from these data. 
-Our method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.
+This method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.
 \end{abstract}
@@
 Precise radial velocity (\RV) measurements are critical to the discovery and characterization of exoplanets. 
-On order of one dozen dedicated spectrographs exist for the purpose of \RV\ planet-hunting, with at least as many more currently under construction \citep{Wright2017}. 
-However, significant challenges exist in deriving precise \RV\ measurements from these spectra. 
+On order of one dozen dedicated spectrographs exist for the purpose of \RV planet-hunting, with at least as many more currently under construction \citep{Wright2017}. 
+However, significant challenges exist in deriving precise \RV measurements from these spectra. 
 %The current capabilities of extreme-precision radial velocity (\EPRV) instruments do not extend to the 10 \cms\ regime, 
@@
 
-One of the primary drivers of the \RV\ noise budget is the incomplete treatment of telluric features in the
+One of the primary drivers of the \RV noise budget is the incomplete treatment of telluric features in the
 Earth's atmosphere \citep{Halverson2016}. 
@@
 The first is that removing sections of the spectrum removes significant regions of the spectrum that can be used to infer the stellar radial velocity. 
-Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve RV precision (cite).
+Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV precision (cite).
 This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres \citep{Figueira2016}.
-Eliminating large chunks of these spectra will therefore significantly inhibit our ability to detect planets around M dwarfs through RVs.
+Eliminating large chunks of these spectra will therefore significantly inhibit our ability to detect planets around M dwarfs through \RV s.
 
 Secondly, not all telluric features are obvious. 
-The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred RV at the $\sim 1$ \ms\ level \citep{Cunha2014}.
+The Earth's atmosphere induces many small-amplitude features, often referred to as ``microtellurics,'' which are not obvious by eye but can affect the star's inferred \RV at the $\sim 1$ \ms level \citep{Cunha2014}.
 As the locations of these features are not known \textit{a priori} and may not even be apparent in stacked spectra of many observations, these spectral regions cannot be thrown out. 
@@
 
-In Section \ref{s:methods}, we outline the model and present an open-source implementation in \python\ and \TF\ called \wobble. 
-In Section \ref{s:results}, we apply our method to \HARPS\ archival data for two target stars, the planet-hosting solar analog 51 Peg and the quiet M dwarf \Mdwarf, as a demonstration of \wobble's capabilities.
+In Section \ref{s:methods}, we outline the model and present an open-source implementation in \python and \TF called \wobble. 
+In Section \ref{s:results}, we apply our method to \HARPS archival data for two target stars, the planet-hosting solar analog 51 Peg and the quiet M dwarf \Mdwarf, as a demonstration of \wobble's capabilities.
 \todo{We find xyz.} 
@@
 
-\todo{Our fundamental assumptions: single, unchanging stellar spectrum with \RV\ shifts; convolved telluric features which can vary in both \RV\ and shape.
-Instrument-provided wavelength solution which is good.}
+The model underpinning \wobble is designed to be flexible and easily extensible to a variety of use cases. However, a few rigid assumptions are necessary, and we outline those here.
 
-Our data are the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed ln(flux) for pixel $m$ of M at epoch $n$ of N. We also have a wavelength solution vector $\lambda$ of length M.
+We take data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\log f$ for pixel $m$ of M at epoch $n$ of N.
+We also have a corresponding $M \times N$ matrix of wavelength solutions obtained in the course of instrument calibrations, which we assume to be accurate.
+Again, the entries of this matrix are actually in logarithmic form ($\log \lambda$).
+This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.
 
-Our model is comprised of three major contributions to the observed spectrum: the star, which has a radial velocity about its center of mass $v_{\star}$ combined with motion of the Earth around the solar system barycenter; the atmospheric telluric spectrum, which has an effective radial velocity $v_t$ at each epoch; and the instrument, which has no velocity shift.
-Each of these components may vary with time in their spectral shape.
-We make the assumption that these variations can be captured in a low-dimensional space.
+For practical applications, the number of epochs $N$ must be $\gg 1$.
+If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.
+a single night), it will be nearly impossible to disentangle the components of the model from each other, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
+We will revisit this restriction in Section \label{s:future}.
 
-Each data column $y_n$ is modeled as:
-$$y_n = P(v_{\star, n}) W_{\star} x_n + P(v_{t, n}) W_{t} z_n + W_{c} u_n + noise,$$
-where $P$ is an interpolation operator which applies a Doppler shift $D_v$ and $W$ is a matrix of ``principal components'' which are weighted by $x_n, z_n, u_n$ to form the spectral contribution at epoch $n$.
-Each $W$ takes the shape $M' \times K$, where $M'$ is the length of a wavelength-space grid $\lambda'$ which may be different from the data wavelength grid $\lambda$ and $K$ is the number of components to be used (where $K_{\star}$, $K_t$, and $K_c$ may be different numbers).
+The spectrum model may be composed of an arbitrary number of components convolved together, but for the purposes of this work we consider the two-component case: a single star whose spectrum is convolved with a telluric absorption spectrum.
+The star has a radial velocity about its center of mass $v_{\star}$, which is added to the (known) motion of the Earth around the solar system barycenter (the Barycentric Earth Radial Velocity or \BERV ) to produce the observed stellar \RV\ shift $v_{obs}$.
+The telluric spectrum is fixed at zero velocity shift relative to the observatory rest frame, but its spectral shape may vary in time. 
+We assume that these variations can be captured in a low-dimensional space.
 
-\todo{discuss degeneracies in the model}
+Given these assumptions, we can model each data column $y_n$ as:
+$$y_n = P(v_{\star, n}) \mu_{\star} + \mu_{t} + W_{t} z_n + noise,$$
+where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu$ is the mean spectral template for each component.
+In addition to its mean spectrum $\mu_t$, the telluric component also takes a time-dependent component assembled from two variables: $W$, a matrix of ``basis vectors,'' is weighted by $z_n$ to form the spectral contribution at epoch $n$.
+Each $\mu$ has length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid. 
+Correspondingly, $W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
+For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications.
 
-We build an initial template for both the stellar and telluric spectra by assuming a Doppler shift for each (equal to the barycentric correction and zero, respectively) and calculating the median flux at each model wavelength $\lambda'$ after that shift is applied.
-With these models in hand, we are then able to infer an RV shift for both model components at each epoch.
-We evaluate a likelihood for each epoch such that
-$$ \ln \mathcal{L} = \sum_{m} -0.5 (y_m - y_m')^T C^{-1} (y_m-y_m'),
+Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
+$$ \log \lambda(v) = \log \lambda_{0} + \log \left(\frac{1 - v/c}{1 + v/c}\right).
+$$
+This shift is applied to the $M'$-length wavelength grid for the template spectrum before interpolating the template to the $M$-length data space.
+
+%With these models in hand, we are then able to infer an \RV\ shift for both model components at each epoch.
+We evaluate a likelihood for each epoch
+$$ \ln \mathcal{L_n} = \sum_{m} -0.5 (y_{m,n} - y_{m',n})^T C^{-1} (y_{m,n}-y_{m',n}),
 $$
 comparing our model $y'$ to the data $y$, with $C$ representing our covariance matrix of the uncertainties on each data point.
-Our model is agnostic to the intrinsic wavelength of any individual stellar or telluric feature, so while we infer the  Doppler shift relative to our model, it does not apply any information about the absolute \RV\ of the star relative to the solar system.
-To mitigate the potential degeneracy of developing a model effectively redshifted relative to all observations, we add (in log space) a Gaussian prior to our likelihood value on the mean \RV\ for all $N$ epochs such that
+Our model is agnostic to the intrinsic wavelength of any individual stellar or telluric feature, so while we infer the  Doppler shift relative to our model, it does not apply any information about the absolute \RV of the star relative to the solar system.
+To mitigate the potential degeneracy of developing a model effectively redshifted relative to all observations, we add (in log space) a Gaussian prior to our likelihood value on the mean \RV for all $N$ epochs such that
 $$ P(m|X) = \ln \mathcal{L}  -0.5 * \frac{\sum(v)^2}{(Nl)^2}.
@@
 
-We maximize this likelihood function for our assumed stellar and telluric models iteratively, first calculating the best-fitting stellar \RV shifts holding the telluric shifts constant, then repeating for the telluric spectrum shift holding the stellar \RV s constant.
+We build an initial template for the stellar spectrum by setting $v_{\star} = 0$ for all epochs.
+We can then Doppler shift the spectra by the barycentric velocity \BERV and calculate the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$.
+The telluric template can then be initialized using the residuals after the stellar contribution has been removed, again by binning (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
+Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
+The $K$ highest eigenweights and their corresponding eigenvectors were taken as the basis weights and vectors.
+\todo{(math here?)}
+
+This likelihood function is maximized for our assumed stellar and telluric models iteratively, first calculating the best-fitting stellar \RV\ shifts holding the telluric shifts constant, then repeating for the telluric spectrum shift holding the stellar \RV s constant.
 After each optimization, we use these velocities to improve our models for the telluric and stellar spectra, employing a gradient descent scheme to modify our templates $m'$ at each wavelength $\lambda'$ using the newly calculated velocities.
@@
 The above-described model can be implemented in a variety of ways.
-We chose to build our code, \wobble, using the \TF\ package \todo{[terminology?]} in \code{python}.
-\TF\ is a \todo{[blah blah technical blah]}.
+We chose to build our code, \wobble, using the \TF package \todo{[terminology?]} in \code{python}.
+\TF is a \todo{[blah blah technical blah]}.
 While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this use case as well.
@@
 In its current realization, \wobble\ is optimized for use with data products from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}.
-Like most \EPRV\ instruments, \HARPS\ is an echelle spectrograph spanning many spectral orders.
+Like most \EPRV instruments, \HARPS is an echelle spectrograph spanning many spectral orders.
 These orders can be considered as independent spectra \todo{for reasons}.
@@
 
-\todo{Description of the data, cite \acronym{ESO} archive.}
+\todo{Description of the data, cite \acronym{ESO} archive. Talk about continuum normalization here?}
 
-\todo{Show our results: we find the planet with the correct mass and orbital parameters. Compare to \HARPS\ pipeline.}
+\todo{Show our results: we find the planet with the correct mass and orbital parameters. Compare to \HARPS pipeline.}
 
@@
 \todo{Revisit the assumptions: single star, no gas cell, known instrumental wavelength solution, no intrinsic spectral variability, no additive terms, many spectra exist. 
-In what situations are these assumptions violated, and how would we configure \wobble\ to deal with them?}
+In what situations are these assumptions violated, and how would we configure \wobble to deal with them?}
 

### 2018-09-04 wobble/model.py

@@ -13,6 +13,15 @@ class Model(object):
     """
     Keeps track of all components in the model.
     Model is specific to order `r` of data object `data`.
+    
+    Parameters
+    ----------
+    data : `object`
+        a wobble Data object
+    results: `object`
+        a wobble Results object
+    r : `int`
+        the index of the order to be fit in Data
     """
     def __init__(self, data, results, r):
         self.components = []
@@ -33,6 +42,7 @@ class Model(object):
         return string
 
     def add_component(self, name, starting_rvs, **kwargs):
+        """Add a new Component object to the model."""
         if np.isin(name, self.component_names):
             print("The model already has a component named {0}. Try something else!".format(name))
             return
@@ -43,11 +53,13 @@ class Model(object):
             self.results.add_component(c)
 
     def add_star(self, name, starting_rvs=None, **kwargs):
+        """Add a component with RVs initialized to zero in the barycentric-corrected rest frame."""
         if starting_rvs is None:
             starting_rvs = -1. * np.copy(self.data.bervs) + np.mean(self.data.bervs)
         self.add_component(name, starting_rvs, **kwargs)
 
     def add_telluric(self, name, starting_rvs=None, **kwargs):
+        """Add a component with RVs initialized to zero in the observatory rest frame."""
         if starting_rvs is None:
             starting_rvs = np.zeros(self.data.N)
         kwargs['learning_rate_template'] = kwargs.get('learning_rate_template', 0.1)
@@ -56,6 +68,12 @@ class Model(object):
         self.add_component(name, starting_rvs, **kwargs)
 
     def initialize_templates(self):
+        """Initialize spectral templates for all components. 
+        
+        *NOTE:* this will initialize each subsequent component from the residuals 
+        of the previous, so make sure you have added the components in order of 
+        largest to smallest contribution to the net spectrum.
+        """
         data_xs = self.data.xs[self.r]
         data_ys = np.copy(self.data.ys[self.r])
         data_ivars = self.data.ivars[self.r]
@@ -63,6 +81,7 @@ class Model(object):
             data_ys = c.initialize_template(data_xs, data_ys, data_ivars)
 
     def setup(self):
+        """Initialize component templates and do TensorFlow magic in prep for optimizing"""
         self.initialize_templates()
         self.synth = tf.zeros(np.shape(self.data.xs[self.r]), dtype=T, name='synth')
         for c in self.components:
@@ -99,22 +118,37 @@ class Model(object):
 
[…]

### 2018-09-04 wobble/results.py

@@ -13,14 +13,26 @@ COMPONENT_TF_ATTRS = ['rvs', 'ivars', 'template_xs', 'template_ys', 'basis_vecto
 COMMON_ATTRS = ['R', 'N', 'orders', 'origin_file', 'epochs', 'component_names']
 
 class Results(object):
-    """
-    Stores RV & template results across all orders.
+    """A read/writeable object which stores RV & template results across all orders. 
+    At the end of each Model optimize() call, the associated Results object is 
+    updated with numpy outputs from each optimized TensorFlow variable. 
+    This allows us to clear out the graph and retain the solution.
+    
+    One of the two keywords is required for initialization.
+    
+    Parameters
+    ----------
+    data : `object`
+        a wobble Data object
+    filename : str
+        a file path pointing to a saved Results object (HDF5 format).
     """
     def __init__(self, data=None, filename=None):
         if filename is None:
             self.component_names = []
             self.R = data.R
             self.N = data.N
+            # get everything we'd need to reconstruct the data used:
             self.orders = data.orders
             self.origin_file = data.origin_file
             self.epochs = data.epochs
@@ -29,10 +41,27 @@ class Results(object):
             self.read(filename)
             return
         print("Results: must supply either data or filename keywords.")
-                
+        
+    def __str__(self):
+        string = 'wobble Results object consisting of the following components: '
+        for n in self.component_names:
+            string += '\n{0}: '.format(n)
+        return string    
+                            
     def add_component(self, c):
+        """Initialize a new model component and prepare to save its optimized outputs. 
+        The component name should be consistent across all order models. 
+        
+        Note that if a component name was initialized in the models for 1+ orders but 
+        was not included in all order models, its RV values/uncertainties will be set 
+        to NaNs and all other properties set to 0 for the excluded order(s).
+        
+        Parameters
+        ----------
+        c : a wobble.Model.Component object
+        """
         if np.isin(c.name, self.component_names):
-            print("A component of name {0} has already been added to the results object.".format(c.name))
+            print("Results: A component of name {0} has already been added.".format(c.name))
             return
         self.component_names.append(c.name)
         basename = c.name+'_'
@@ -47,6 +76,12 @@ class Results(object):
             setattr(self, basename+attr, [0 for r in range(self.R)])
                 
     def update(self, c, **kwargs):
[…]

### 2018-09-04 test.py
optimization history plots working again

@@ -13,16 +13,17 @@ if __name__ == "__main__":
     star_reg_file = 'wobble/regularization/{0}_star_K{1}.hdf5'.format(starname, K_star)
     tellurics_reg_file = 'wobble/regularization/{0}_t_K{1}.hdf5'.format(starname, K_t)
     
-    if False:
+    if True:
         # quick test on two orders
         data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=[30,56])
         results = wobble.Results(data=data)
-        model = wobble.Model(data, results, r)
-        model.add_star('star', variable_bases=K_star, 
+        for r in range(data.R):
+            model = wobble.Model(data, results, r)
+            model.add_star('star', variable_bases=K_star, 
                             regularization_par_file=star_reg_file)
-        model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
+            model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
                                 regularization_par_file=tellurics_reg_file)
-        wobble.optimize_order(model, niter=80)
+            wobble.optimize_order(model, niter=80, save_history=True, basename='results/plots_{0}/history'.format(starname))
         assert False
     
     
@@ -121,7 +122,7 @@ if __name__ == "__main__":
     plt.savefig('results/{0}_Kstar{1}_Kt{2}_rvs.png'.format(starname, K_star, K_t))
     plt.close(fig)
     
-    for o in np.arange(R):
+    for o in np.arange(72):
         e = 0
         fig, (ax, ax2) = plt.subplots(2, 1, gridspec_kw = {'height_ratios':[4, 1]}, figsize=(12,5))
         ax.scatter(np.exp(data.xs[o][e]), np.exp(data.ys[o][e]), c='k', alpha=0.6)

### 2018-09-04 wobble/data.py

@@ -8,17 +8,35 @@ class Data(object):
     The data object: contains the spectra and associated data.
     All objects in `data` are numpy arrays or lists of arrays.
     Includes all orders and epochs.
+    
+    Parameters
+    ----------
+    filename : `str`
+        Name of HDF5 file storing the data.
+    filepath : `str` (default `../data/`)
+        Path to append to filename.
+    orders : `list` (default `None`)
+        Indices of spectral orders to read in. If `None` include all. 
+        Even if it's only one index, it must be a list.
+    epochs : `int` or `list` (default `None`)
+        Indices of epochs to read in. If `None` include all.
+    min_flux : `float` (default `1.`)
+        Flux in counts/pixel below which a pixel is masked out.
+    max_norm_flux : `float` (default `2.`)
+        Flux in normalized counts/pixel above which a pixel is masked out.
+    padding : `int` (default `3`)
+        Number of pixels to additionally mask on either side of a bad high pixel.
     """
     def __init__(self, filename, filepath='../data/', 
-                    orders = [30], 
+                    orders = None, 
                     epochs = None,
                     min_flux = 1.,
                     max_norm_flux = 2.,
                     padding = 3):
-        self.R = len(orders) # number of orders to be analyzed
-        self.orders = orders
         self.origin_file = filepath+filename
         with h5py.File(self.origin_file) as f:
+            if orders is None:
+                orders = np.arange(len(f['data']))
             if epochs is None:
                 self.N = len(f['dates']) # all epochs
                 self.epochs = np.arange(self.N)
@@ -37,6 +55,9 @@ class Data(object):
             self.drifts = np.copy(f['drifts'])[self.epochs]
             self.airms = np.copy(f['airms'])[self.epochs]
             
+        self.R = len(orders) # number of orders
+        self.orders = orders # indices of orders in origin_file
+            
         # mask out low pixels:
         for r in range(self.R):
             bad = self.ys[r] < min_flux
@@ -54,13 +75,14 @@ class Data(object):
         for r in range(self.R):
             bad = self.ys[r] > max_norm_flux
             self.ys[r][bad] = 1.
-            for pad in range(padding):
+            for pad in range(padding): # mask out neighbors of high pixels
                 bad = np.logical_or(bad, np.roll(bad, pad+1))
                 bad = np.logical_or(bad, np.roll(bad, -pad-1))
[…]

### 2018-09-04 wobble/history.py

@@ -1,79 +1,49 @@
 import numpy as np
 import matplotlib.pyplot as plt
-import matplotlib
 from matplotlib import animation
-import sys
-import h5py
-import copy
-import tensorflow as tf
-T = tf.float64
 
 from .utils import get_session
 
 class History(object):
     """
-    Information about optimization history of a single order stored in numpy arrays/lists
+    Information about optimization history of a single order stored in numpy arrays/lists.
     """   
-    def __init__(self, model, data, r, niter, filename=None):
-        for c in model.components:
-            assert c.template_exists[r], "ERROR: Cannot initialize History() until templates are initialized."
-        self.nll_history = np.empty(niter)
-        self.rvs_history = [np.empty((niter, data.N)) for c in model.components]
-        self.template_history = [np.empty((niter, int(c.template_ys[r].shape[0]))) for c in model.components]
-        self.basis_vectors_history = [np.empty((niter, c.K, 4096)) for c in model.components] # HACK
-        self.basis_weights_history = [np.empty((niter, data.N, c.K)) for c in model.components]
-        self.chis_history = np.empty((niter, data.N, 4096)) # HACK
+    def __init__(self, model, data, r, niter):
         self.r = r
         self.niter = niter
-        if filename is not None:
-            self.read(filename)
+        self.data = data
+        self.nll_history = np.empty(niter)
+        self.synth_history = np.empty(np.append(niter, np.shape(data.ys[r])))
+        self.rvs_history = [np.empty((niter, data.N)) for c in model.components]
+        self.ivars_history = [np.empty((niter, data.N)) for c in model.components]
+        template_Ms = [int(c.template_ys.shape[0]) for c in model.components] # number of elements in templates
+        self.template_history = [np.empty((niter, template_Ms[i])) for i,c in enumerate(model.components)]
+        self.basis_vectors_history = [np.empty((niter, c.K, template_Ms[i])) for i,c in enumerate(model.components)]
+        self.basis_weights_history = [np.empty((niter, data.N, c.K)) for c in model.components]
+
         
-    def save_iter(self, model, data, i, nll, chis):
+    def save_iter(self, model, i):
         """
-        Save all necessary information at optimization step i
+        Save all necessary information after optimization number i
         """
[…]

### 2018-09-04 wobble/model.py

@@ -126,25 +126,29 @@ class Model(object):
             Number of iterations.
         save_history : `bool` (default `False`)
             If `True`, create a wobble History object to track progress across 
-            iterations. The resulting output will be saved under `[basename]_order[#].hdf5`. 
+            iterations and generate plots.
         basename : `str` (default `wobble`)
-            Path/name to use when saving history. Only accessed if save_history = `True`.
+            Path/name to use when saving plots. Only accessed if save_history = `True`.
         feed_dict : `dict` (default `None`)
             TensorFlow magic; passed to the optimizer.
         """
         # initialize helper classes:
         if save_history:
-            history = History(self, self.data, self.r, niter)
+            history = History(self, self.data, self.r, niter+1)
+            history.save_iter(self, 0)
         # optimize:
         session = get_session()
         for i in tqdm(range(niter), total=niter, miniters=int(niter/10)):
-            if save_history:
-                history.save_iter(model, self.data, i, self.nll, chis)
             session.run(self.updates, feed_dict=feed_dict)
-        # copy over the outputs to Results for safekeeping:
+            if save_history:
+                history.save_iter(self, i+1)
+        # copy over the outputs to Results:
         for c in self.components:
             self.results.update(c)
-        # save history
+        # save optimization plots:
+        if save_history:
+            history.save_plots(basename)
+        
 
 class Component(object):
     """

### 2018-09-04 wobble/results.py

@@ -134,8 +134,8 @@ class Results(object):
         
         Parameters
         ----------
-        component_name : str
-        Name of the model component to use.
+        component_name : `str`
+            Name of the model component to use.
         """
         if not np.isin(component_name, self.component_names):
             print("Results: component name {0} not recognized. Valid options are: {1}".format(component_name, 

### 2018-09-05 regularization.py
smarter treatment of order indexing

@@ -4,7 +4,7 @@ import wobble
 import tensorflow as tf
 from tqdm import tqdm
 import h5py
-import pdb
+import os
 
 __all__ = ["improve_order_regularization", "improve_parameter", "test_regularization_value"]
 
@@ -74,7 +74,7 @@ def improve_order_regularization(r, best_regularization_par,
     tensors_to_tune = [training_model.components[1].L2_template_tensor, training_model.components[0].L2_template_tensor,
                        training_model.components[1].L1_template_tensor, training_model.components[0].L1_template_tensor]
     tensor_names = ['L2_template_tellurics', 'L2_template_star', 'L1_template_tellurics',
-                     'L1_template_star'] # HACK
+                     'L1_template_star'] # HACK - this is needed bc TF appends garbage to the end of the tensor name
     if K_star > 0:
         tensors_to_tune = np.append(tensors_to_tune, [training_model.components[0].L2_basis_vectors_tensor, 
                                                     training_model.components[0].L1_basis_vectors_tensor])
@@ -89,11 +89,13 @@ def improve_order_regularization(r, best_regularization_par,
     for (tensor,name) in zip(tensors_to_tune, tensor_names):
         regularization_dict[tensor] = np.copy(best_regularization_par[name][r_init])
 
-    for (tensor,name) in zip(tensors_to_tune, tensor_names):
+    i = 0 # track order in which parameters are improved
+    for tensor,name in zip(tensors_to_tune, tensor_names):
         if (name[0:2] == "L1" and L1) or (name[0:2] == "L2" and L2):
+            i += 1
             regularization_dict[tensor] = improve_parameter(tensor, training_model, validation_model, 
                                                          regularization_dict, verbose=verbose,
-                                                         plot=plot, basename=basename)
+                                                         plot=plot, basename=basename+'_par{0}'.format(i))
             best_regularization_par[name][r] = np.copy(regularization_dict[tensor])
     
     
@@ -152,7 +154,7 @@ def improve_parameter(par, training_model, validation_model, regularization_dict
         ax.set_xlabel('{0} values'.format(name))
         ax.set_ylabel('NLL')
         fig.tight_layout()
-        plt.savefig('{0}_{1}_nll.png'.format(basename, name))
+        plt.savefig('{0}_nll.png'.format(basename))
[…]

### 2018-09-05 test.py

@@ -4,16 +4,17 @@ import tensorflow as tf
 import wobble
 from time import time
 import h5py
+import os
 
 if __name__ == "__main__":
-    starname = '51peg'
+    starname = 'barnards'
     K_star = 0
     K_t = 3
     
     star_reg_file = 'wobble/regularization/{0}_star_K{1}.hdf5'.format(starname, K_star)
     tellurics_reg_file = 'wobble/regularization/{0}_t_K{1}.hdf5'.format(starname, K_t)
     
-    if True:
+    if False:
         # quick test on two orders
         data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=[30,56])
         results = wobble.Results(data=data)
@@ -28,64 +29,33 @@ if __name__ == "__main__":
     
     
     start_time = time()
-    data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=np.arange(72))
+    orders = np.arange(10,72)
+    data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders)
     results = wobble.Results(data=data)
     
     print("data loaded")
     print("time elapsed: {0:.2f} s".format(time() - start_time))
 
-    plot_dir = 'results/plots/'
+    plot_dir = 'results/plots_{0}_Kstar{1}_Kt{2}/'.format(starname, K_star, K_t)
+    if not os.path.exists(plot_dir):
+        os.makedirs(plot_dir)
     niter = 80
-    if True: # no plots
-        for r in range(data.R):
-            model = wobble.Model(data, results, r)
-            model.add_star('star', variable_bases=K_star, 
-                            regularization_par_file=star_reg_file)
-            model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
-                                regularization_par_file=tellurics_reg_file)
-            print("--- ORDER {0} ---".format(r))
+    plots = True
+    for r,o in enumerate(orders):
+        model = wobble.Model(data, results, r)
+        model.add_star('star', variable_bases=K_star, 
+                        regularization_par_file=star_reg_file)
+        model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
+                            regularization_par_file=tellurics_reg_file)
+        print("--- ORDER {0} ---".format(o))
+        if plots:
[…]

### 2018-09-05 wobble/data.py

@@ -57,7 +57,7 @@ class Data(object):
             
         self.R = len(orders) # number of orders
         self.orders = orders # indices of orders in origin_file
-            
+                    
         # mask out low pixels:
         for r in range(self.R):
             bad = self.ys[r] < min_flux
@@ -66,11 +66,12 @@ class Data(object):
                 bad = np.logical_or(bad, np.roll(bad, pad+1))
                 bad = np.logical_or(bad, np.roll(bad, -pad-1))
             self.ivars[r][bad] = 0.
+            
 
         # log and normalize:
         self.ys = np.log(self.ys) 
         self.continuum_normalize() 
-        
+                
         # mask out high pixels:
         for r in range(self.R):
             bad = self.ys[r] > max_norm_flux
@@ -79,10 +80,14 @@ class Data(object):
                 bad = np.logical_or(bad, np.roll(bad, pad+1))
                 bad = np.logical_or(bad, np.roll(bad, -pad-1))
             self.ivars[r][bad] = 0.
+
         
     def continuum_normalize(self, **kwargs):
         """Continuum-normalize all spectra using a polynomial fit. Takes kwargs of utils.fit_continuum"""
         for r in range(self.R):
             for n in range(self.N):
-                self.ys[r][n] -= fit_continuum(self.xs[r][n], self.ys[r][n], self.ivars[r][n], **kwargs)
+                try:
+                    self.ys[r][n] -= fit_continuum(self.xs[r][n], self.ys[r][n], self.ivars[r][n], **kwargs)
+                except:
+                    print("ERROR: Data: order {0}, epoch {1} could not be continuum normalized!".format(r,n))
                 

### 2018-09-05 wobble/history.py

@@ -8,18 +8,19 @@ class History(object):
     """
     Information about optimization history of a single order stored in numpy arrays/lists.
     """   
-    def __init__(self, model, data, r, niter):
-        self.r = r
+    def __init__(self, model, niter):
+        self.r = model.r
+        self.order = model.order
         self.niter = niter
-        self.data = data
+        self.data = model.data
         self.nll_history = np.empty(niter)
-        self.synth_history = np.empty(np.append(niter, np.shape(data.ys[r])))
-        self.rvs_history = [np.empty((niter, data.N)) for c in model.components]
-        self.ivars_history = [np.empty((niter, data.N)) for c in model.components]
+        self.synth_history = np.empty(np.append(niter, np.shape(self.data.ys[r])))
+        self.rvs_history = [np.empty((niter, self.data.N)) for c in model.components]
+        self.ivars_history = [np.empty((niter, self.data.N)) for c in model.components]
         template_Ms = [int(c.template_ys.shape[0]) for c in model.components] # number of elements in templates
         self.template_history = [np.empty((niter, template_Ms[i])) for i,c in enumerate(model.components)]
         self.basis_vectors_history = [np.empty((niter, c.K, template_Ms[i])) for i,c in enumerate(model.components)]
-        self.basis_weights_history = [np.empty((niter, data.N, c.K)) for c in model.components]
+        self.basis_weights_history = [np.empty((niter, self.data.N, c.K)) for c in model.components]
 
         
     def save_iter(self, model, i):
@@ -100,6 +101,10 @@ class History(object):
         """
         xs = np.exp(self.template_xs[ind])
         ys = np.exp(self.template_history[ind])
+        # set up ylims if not otherwise specified:
+        y_pad = 0.1
+        ylims = (np.min(ys)-y_pad, min([np.max(ys)+y_pad, 10.])) # hard upper limit to prevent inf
+        kwargs['ylims'] = kwargs.get('ylims', ylims)
         return self.plot(xs, ys, linestyle='line', **kwargs)
         
     def plot_synth(self, e, **kwargs):
@@ -110,23 +115,27 @@ class History(object):
         xs = np.exp(self.data.xs[self.r][e])
         ys = np.exp(self.synth_history[:,e,:])
[…]

### 2018-09-05 wobble/model.py

@@ -28,10 +28,11 @@ class Model(object):
         self.component_names = []
         self.data = data
         self.results = results
-        self.r = r
+        self.r = r # order index
+        self.order = data.orders[r] # order number
 
     def __str__(self):
-        string = 'Model for order {0} consisting of the following components: '.format(self.data.orders[self.r])
+        string = 'Model for order {0} consisting of the following components: '.format(self.order)
         for c in self.components:
             string += '\n{0}: '.format(c.name)
             if c.rvs_fixed:
@@ -134,7 +135,7 @@ class Model(object):
         """
         # initialize helper classes:
         if save_history:
-            history = History(self, self.data, self.r, niter+1)
+            history = History(self, niter+1)
             history.save_iter(self, 0)
         # optimize:
         session = get_session()

### 2018-09-05 wobble/data.py
throw out uninformative epochs and/or orders

@@ -1,5 +1,6 @@
 import numpy as np
 import h5py
+import pdb
 
 from .utils import fit_continuum
 
@@ -34,6 +35,48 @@ class Data(object):
                     max_norm_flux = 2.,
                     padding = 3):
         self.origin_file = filepath+filename
+        self.read_data(orders=orders, epochs=epochs)
+        
+        orders = np.asarray(self.orders)
+        epochs = np.asarray(self.epochs)
+        min_snr = 10.
+        chis = np.asarray(self.fluxes) * np.sqrt(np.asarray(self.ivars))
+        snrs_by_epoch = np.sqrt(np.nanmean(chis, axis=(0,2)))
+        epochs_to_cut = snrs_by_epoch < min_snr
+        if np.sum(epochs_to_cut) > 0:
+            epochs = epochs[~epochs_to_cut]
+            self.read_data(orders=orders, epochs=epochs) # overwrite with new data
+        chis = np.asarray(self.fluxes) * np.sqrt(np.asarray(self.ivars))
+        snrs_by_order = np.sqrt(np.nanmean(chis, axis=(1,2)))
+        orders_to_cut = snrs_by_order < min_snr
+        if np.sum(orders_to_cut) > 0:
+            orders = orders[~orders_to_cut]
+            self.read_data(orders=orders, epochs=epochs) # overwrite with new data
+                           
+        # mask out low pixels:
+        for r in range(self.R):
+            bad = self.fluxes[r] < min_flux
+            self.fluxes[r][bad] = min_flux
+            for pad in range(padding): # mask out neighbors of low pixels
+                bad = np.logical_or(bad, np.roll(bad, pad+1))
+                bad = np.logical_or(bad, np.roll(bad, -pad-1))
+            self.ivars[r][bad] = 0.
+            
+        # log and normalize:
+        self.ys = np.log(self.fluxes) 
+        self.continuum_normalize() 
+                
+        # mask out high pixels:
+        for r in range(self.R):
+            bad = self.ys[r] > max_norm_flux
+            self.ys[r][bad] = 1.
+            for pad in range(padding): # mask out neighbors of high pixels
+                bad = np.logical_or(bad, np.roll(bad, pad+1))
+                bad = np.logical_or(bad, np.roll(bad, -pad-1))
+            self.ivars[r][bad] = 0.
+            
+    def read_data(self, orders = None, epochs = None):
+        """Read origin file and set up data attributes from it"""
         with h5py.File(self.origin_file) as f:
             if orders is None:
[…]

### 2018-09-07 regularization.py
regularization more parallelization-friendly

@@ -13,7 +13,7 @@ def get_name_from_tensor(tensor):
     # NOTE - does not handle '_2' type additions!
     return str.split(tensor.name, ':')[0]
 
-def improve_order_regularization(r, best_regularization_par,
+def improve_order_regularization(r, o, star_filename, tellurics_filename,
                                  training_data, training_results,
                                  validation_data, validation_results,
                                  verbose=True, plot=False, basename='', 
@@ -21,14 +21,14 @@ def improve_order_regularization(r, best_regularization_par,
     """
     Use a validation scheme to determine the best regularization parameters for 
     all model components in a given order r.
-    Update dict best_regularization_parameters and return it.
+    Update files at star_filename, tellurics_filename and return it.
     """
     
     training_model = wobble.Model(training_data, training_results, r)
     training_model.add_star('star', variable_bases=K_star)
     training_model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t)
     training_model.setup()
-    training_model.optimize(niter=0)
+    training_model.optimize(niter=0, verbose=verbose)
     
     if plot:
         n = 0
@@ -73,35 +73,49 @@ def improve_order_regularization(r, best_regularization_par,
     # the order in which these are defined will determine the order in which they are optimized:
     tensors_to_tune = [training_model.components[1].L2_template_tensor, training_model.components[0].L2_template_tensor,
                        training_model.components[1].L1_template_tensor, training_model.components[0].L1_template_tensor]
-    tensor_names = ['L2_template_tellurics', 'L2_template_star', 'L1_template_tellurics',
-                     'L1_template_star'] # HACK - this is needed bc TF appends garbage to the end of the tensor name
+    tensor_names = ['L2_template', 'L2_template', 'L1_template',
+                     'L1_template'] # HACK - this is needed bc TF appends garbage to the end of the tensor name
+    tensor_components = ['tellurics', 'star', 'tellurics', 'star'] # HACK
     if K_star > 0:
         tensors_to_tune = np.append(tensors_to_tune, [training_model.components[0].L2_basis_vectors_tensor, 
                                                     training_model.components[0].L1_basis_vectors_tensor])
-        tensor_names = np.append(tensor_names, ['L2_basis_vectors_star', 'L1_basis_vectors_star'])
[…]

### 2018-09-07 test.py

@@ -29,8 +29,9 @@ if __name__ == "__main__":
     
     
     start_time = time()
-    orders = np.arange(10,72)
+    orders = np.arange(72)
     data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders)
+    orders = np.copy(data.orders)
     results = wobble.Results(data=data)
     
     print("data loaded")

### 2018-09-07 wobble/model.py

@@ -118,7 +118,7 @@ class Model(object):
         session.run(tf.global_variables_initializer())
 
     def optimize(self, niter=100, save_history=False, basename='wobble',
-                 feed_dict=None):
+                 feed_dict=None, verbose=True):
         """Optimize the model!
             
         Parameters
@@ -139,7 +139,11 @@ class Model(object):
             history.save_iter(self, 0)
         # optimize:
         session = get_session()
-        for i in tqdm(range(niter), total=niter, miniters=int(niter/10)):
+        if verbose:
+            iterator = tqdm(range(niter), total=niter, miniters=int(niter/10))
+        else:
+            iterator = range(niter)
+        for i in iterator:
             session.run(self.updates, feed_dict=feed_dict)
             if save_history:
                 history.save_iter(self, i+1)
@@ -170,8 +174,10 @@ class Component(object):
         self.learning_rate_rvs = learning_rate_rvs
         self.learning_rate_template = learning_rate_template
         self.learning_rate_basis = learning_rate_basis
-        regularization_par = ['L1_template', 'L2_template', 
-                              'L1_basis_vectors', 'L2_basis_vectors', 'L2_basis_weights']
+        regularization_par = ['L1_template', 'L2_template']
+        if self.K > 0:
+            regularization_par = np.append(regularization_par, ['L1_basis_vectors', 'L2_basis_vectors', 'L2_basis_weights'])
+        self.regularization_par = regularization_par
         for par in regularization_par:
             setattr(self, par, eval(par)) # set to input values/defaults
         if regularization_par_file is not None:
@@ -180,7 +186,7 @@ class Component(object):
                     for par in regularization_par:
                         setattr(self, par, np.copy(f[par][r])) # overwrite with value from file
             except:
-                print('Regularization parameter file {0} not recognized; using default parameters.'.format(regularization_par_file))
+                print('Regularization parameter file {0} not recognized; using keywords instead.'.format(regularization_par_file))
         self.starting_rvs = starting_rvs
         self.template_xs = template_xs
 
@@ -197,7 +203,7 @@ class Component(object):
         self.data_xs = tf.constant(data.xs[r], dtype=T, name='data_xs_'+self.name)
 
         # Set up the regularization
-        for name in ['L1_template', 'L2_template', 'L1_basis_vectors', 'L2_basis_vectors', 'L2_basis_weights']:
+        for name in self.regularization_par:
[…]

### 2018-09-07 wobble/results.py

@@ -7,9 +7,10 @@ T = tf.float64
 from .utils import get_session
 
 COMPONENT_NP_ATTRS = ['K', 'r', 'rvs_fixed', 'scale_by_airmass', 'learning_rate_rvs', 'learning_rate_template', 
-                      'learning_rate_basis', 'L1_template', 'L2_template', 'L1_basis_vectors', 
-                      'L2_basis_vectors', 'L2_basis_weights']
-COMPONENT_TF_ATTRS = ['rvs', 'ivars', 'template_xs', 'template_ys', 'basis_vectors', 'basis_weights']
+                      'L1_template', 'L2_template']
+OPT_COMPONENT_NP_ATTRS = ['learning_rate_basis', 'L1_basis_vectors', 'L2_basis_vectors', 'L2_basis_weights']
+COMPONENT_TF_ATTRS = ['rvs', 'ivars', 'template_xs', 'template_ys']
+OPT_COMPONENT_TF_ATTRS = ['basis_vectors', 'basis_weights']
 COMMON_ATTRS = ['R', 'N', 'orders', 'origin_file', 'epochs', 'component_names']
 
 class Results(object):
@@ -69,10 +70,14 @@ class Results(object):
         setattr(self, basename+'ivars', np.empty((self.R,self.N)) + np.nan)
         setattr(self, basename+'template_xs', [0 for r in range(self.R)])
         setattr(self, basename+'template_ys', [0 for r in range(self.R)])
-        setattr(self, basename+'basis_vectors', [0 for r in range(self.R)])
-        setattr(self, basename+'basis_weights', [0 for r in range(self.R)])
+        if c.K > 0:
+            setattr(self, basename+'basis_vectors', [0 for r in range(self.R)])
+            setattr(self, basename+'basis_weights', [0 for r in range(self.R)])
         setattr(self, basename+'ys_predicted', [0 for r in range(self.R)])
-        for attr in COMPONENT_NP_ATTRS:
+        attrs = COMPONENT_NP_ATTRS
+        if c.K > 0:
+            attrs = np.append(attrs, OPT_COMPONENT_NP_ATTRS)
+        for attr in attrs:
             setattr(self, basename+attr, [0 for r in range(self.R)])
                 
     def update(self, c, **kwargs):
@@ -83,15 +88,18 @@ class Results(object):
         c : a wobble.Model.Component object
         """
         basename = c.name+'_'
[…]

### 2018-09-10 regularization.py
plotting bug

@@ -343,7 +343,7 @@ if __name__ == "__main__":
         for key in list(f.keys()):
             fig = plt.figure()
             ax = fig.add_subplot(111)
-            ax.plot(orders, f[key][orders], 'o')
+            ax.plot(orders, np.array(f[key])[orders], 'o')
             ax.set_yscale('log')
             ax.set_xlabel('Order #')
             ax.set_ylabel(key)
@@ -356,7 +356,7 @@ if __name__ == "__main__":
         for key in list(f.keys()):
             fig = plt.figure()
             ax = fig.add_subplot(111)
-            ax.plot(orders, f[key][orders], 'o')
+            ax.plot(orders, np.array(f[key])[orders], 'o')
             ax.set_yscale('log')
             ax.set_xlabel('Order #')
             ax.set_ylabel(key)

### 2018-09-10 paper/paper.tex
writing

@@
 
-The model underpinning \wobble is designed to be flexible and easily extensible to a variety of use cases. However, a few rigid assumptions are necessary, and we outline those here.
+The model underpinning \wobble is designed to be flexible and easily extensible to a variety of situations. 
+However, a few rigid assumptions are necessary, and we outline those here.
 
-We take data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\log f$ for pixel $m$ of M at epoch $n$ of N.
+We take data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
 We also have a corresponding $M \times N$ matrix of wavelength solutions obtained in the course of instrument calibrations, which we assume to be accurate.
-Again, the entries of this matrix are actually in logarithmic form ($\log \lambda$).
+Again, the entries of this matrix are in logarithmic form ($\ln \lambda$). 
 This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.
@@
 
-Given these assumptions, we can model each data column $y_n$ as:
-$$y_n = P(v_{\star, n}) \mu_{\star} + \mu_{t} + W_{t} z_n + noise,$$
-where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu$ is the mean spectral template for each component.
+Given these assumptions, we can model each data column $y_n$ as the sum of the stellar and telluric spectra:
+\begin{equation}
+y_n = y_{\star, n} + y_{t, n} + noise.
+\end{equation}
+
+The stellar spectrum contribution is: 
+\begin{equation}
+y_{\star, n} = P(v_{obs, n}) \mu_{\star},
+\end{equation}
+where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid.
+Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
+\begin{equation}
+ \ln \lambda(v) = \ln \lambda_{0} + \ln \left(\frac{1 - v/c}{1 + v/c}\right).
+\end{equation}
+The $P$ operator applies this term to the $M'$-length wavelength grid for the template spectrum before interpolating the template to the $M$-length data space.
+
+The telluric spectrum contribution is:
+\begin{equation}
+y_{t, n} =  A_n(\mu_{t} + W_{t} z_n)
+\end{equation}
 In addition to its mean spectrum $\mu_t$, the telluric component also takes a time-dependent component assembled from two variables: $W$, a matrix of ``basis vectors,'' is weighted by $z_n$ to form the spectral contribution at epoch $n$.
-Each $\mu$ has length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid. 
-Correspondingly, $W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
+$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
 For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
+The net telluric spectrum (mean + time-variable components) is weighted by the air mass at the time of observation $A_n$, a known quantity.
 
-Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
-$$ \log \lambda(v) = \log \lambda_{0} + \log \left(\frac{1 - v/c}{1 + v/c}\right).
-$$
-This shift is applied to the $M'$-length wavelength grid for the template spectrum before interpolating the template to the $M$-length data space.
+We evaluate a likelihood for each epoch
+\begin{equation}
+\ln \mathcal{L}_n = \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C^{-1} (y_{m,n}-y'_{m,n}),
+\end{equation}
+comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties on each data point.
+Our model is agnostic to the intrinsic wavelength of any individual stellar or telluric feature, so while we infer the Doppler shift relative to our model, it does not apply any information about the absolute \RV of the star relative to the solar system.
+%To mitigate the potential degeneracy of developing a model effectively redshifted relative to all observations, we add (in log space) a Gaussian prior to our likelihood value on the mean \RV for all $N$ epochs such that
+%$$ P(m|X) = \ln \mathcal{L}  -0.5 * \frac{\sum(v)^2}{(Nl)^2}.
+%$$
+%\todo{someone should define these variables}
 
-%With these models in hand, we are then able to infer an \RV\ shift for both model components at each epoch.
-We evaluate a likelihood for each epoch
-$$ \ln \mathcal{L_n} = \sum_{m} -0.5 (y_{m,n} - y_{m',n})^T C^{-1} (y_{m,n}-y_{m',n}),
-$$
-comparing our model $y'$ to the data $y$, with $C$ representing our covariance matrix of the uncertainties on each data point.
-Our model is agnostic to the intrinsic wavelength of any individual stellar or telluric feature, so while we infer the  Doppler shift relative to our model, it does not apply any information about the absolute \RV of the star relative to the solar system.
-To mitigate the potential degeneracy of developing a model effectively redshifted relative to all observations, we add (in log space) a Gaussian prior to our likelihood value on the mean \RV for all $N$ epochs such that
-$$ P(m|X) = \ln \mathcal{L}  -0.5 * \frac{\sum(v)^2}{(Nl)^2}.
-$$
-\todo{someone should define these variables}
+The number of free parameters in this model is large: we must optimize every grid point in the mean spectral templates and telluric basis vectors along with the stellar \RV s and the telluric basis weights.
+While in principle degeneracies between the stellar and telluric spectra should not exist since the star will always vary in \RV by at least \BERV \todo{(uhhh is this true??)}, in practice the complexity of the optimization problem makes parameter degeneracy a significant concern \todo{(total BS)}.
+We use a combination of L1 and L2 regularization to mitigate this issue. 
+\todo{define these, discuss their purpose}
 
-We build an initial template for the stellar spectrum by setting $v_{\star} = 0$ for all epochs.
-We can then Doppler shift the spectra by the barycentric velocity \BERV and calculate the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$.
-The telluric template can then be initialized using the residuals after the stellar contribution has been removed, again by binning (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
+The final likelihood to be optimized is then:
+\begin{equation}
+\begin{split}
+\ln \mathcal{L} = & \sum_{n} \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C^{-1} (y_{m,n}-y'_{m,n})  \\
+ & + \alpha_{L1, \mu_{\star}}  \sum_{m'} | \mu_{{\star}} | + \alpha_{L2, \mu_{\star}}  \sum_{m'} \mu_{{\star}}^2 \\
+ & + \alpha_{L1, \mu_t}  \sum_{m'} | \mu_{t} |  + \alpha_{L2, \mu_t}  \sum_{m'} \mu_{t}^2 \\
+ & + \alpha_{L1, W_t}  \sum_{k} \sum_{m'} | W_{t} |  + \alpha_{L2, W_t}  \sum_{k} \sum_{m'} W_{t}^2 + \alpha_{L2, z_t} \sum_{k} \sum_{n} z_{t}^2 ,
+\end{split}
+\end{equation}
+where $\alpha_{L2, z_t}$ is set to 1.0 and all other regularization amplitudes are set by a validation procedure \todo{as described below}.
+
+\subsection{Optimizing the Model}
+
+We build an initial template for the stellar spectrum by setting $v_{obs} = v_{BERV}$ for all epochs.
+We can then calculate the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$.
+The telluric template is then initialized using the residuals after the stellar contribution has been removed, again by binning (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
 Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
-The $K$ highest eigenweights and their corresponding eigenvectors were taken as the basis weights and vectors.
+The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors.
 \todo{(math here?)}
@@
 
-The above-described model can be implemented in a variety of ways.
-We chose to build our code, \wobble, using the \TF package \todo{[terminology?]} in \code{python}.
+The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, using the \TF package \todo{[terminology?]} in \code{python}. 
 \TF is a \todo{[blah blah technical blah]}. 
-While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this use case as well.
-In particular, the necessary optimizations can be performed with high efficiency by \TF: the below-described analysis of \todo{x} spectra of 51 Peg runs in \todo{y} hours on a standard Macbook laptop.
+While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
+The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of \todo{x} spectra of 51 Peg runs in \todo{y} minutes on a standard Mac desktop.
 

### 2018-09-10 wobble/make_data.py

@@ -47,7 +47,7 @@ def read_data_from_fits(filelist):
         for r in range(R):
             data[r][n,:] = spec[r,:]
             ivars[r][n,:] = np.zeros(M) + snrs[r]**2
-            xs[r][n,:] = wave[r,:]  
+            xs[r][n,:] = wave[r,:] 
             
     for r in range(R):
         data[r] = np.delete(data[r], empty, axis=0)
@@ -107,7 +107,7 @@ def write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffi
 
 if __name__ == "__main__":
         
-    if True: #51 Peg
+    if False: #51 Peg
         ccf_filelist = np.genfromtxt('51peg_ccf_filelist.txt', dtype=None, encoding=None)
         data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
         hdffile = '../data/51peg_e2ds.hdf5'

### 2018-09-12 wobble/history.py
bug fixes

@@ -14,7 +14,7 @@ class History(object):
         self.niter = niter
         self.data = model.data
         self.nll_history = np.empty(niter)
-        self.synth_history = np.empty(np.append(niter, np.shape(self.data.ys[r])))
+        self.synth_history = np.empty(np.append(niter, np.shape(self.data.ys[self.r])))
         self.rvs_history = [np.empty((niter, self.data.N)) for c in model.components]
         self.ivars_history = [np.empty((niter, self.data.N)) for c in model.components]
         template_Ms = [int(c.template_ys.shape[0]) for c in model.components] # number of elements in templates
@@ -37,8 +37,8 @@ class History(object):
             self.rvs_history[j][i,:] = np.copy(session.run(c.rvs))  
             self.ivars_history[j][i,:] = np.copy(session.run(c.ivars))  
             if c.K > 0:
-                self.basis_vectors_history[j][i,:,:] = np.copy(session.run(c.basis_vectors[self.r])) 
-                self.basis_weights_history[j][i,:,:] = np.copy(session.run(c.basis_weights[self.r]))
+                self.basis_vectors_history[j][i,:,:] = np.copy(session.run(c.basis_vectors)) 
+                self.basis_weights_history[j][i,:,:] = np.copy(session.run(c.basis_weights))
         if i == 0:
             self.template_xs = [session.run(c.template_xs) for c in model.components]
             

### 2018-09-12 wobble/results.py

@@ -115,8 +115,8 @@ class Results(object):
                 setattr(self, basename+'ys_predicted', [0 for r in range(self.R)])
                 all_order_attrs.append(basename+'ys_predicted')
                 attrs = np.append(COMPONENT_NP_ATTRS, COMPONENT_TF_ATTRS)
-                if np.copy(f[basename+'K']) > 0:
-                    attrs = np.append(attrs, [OPT_COMPONENT_NP_ATTRS, OPT_COMPONENT_TF_ATTRS])
+                if np.any(np.array(f[basename+'K']) > 0):
+                    attrs = np.append(attrs, np.append(OPT_COMPONENT_NP_ATTRS, OPT_COMPONENT_TF_ATTRS))
                 for attr in attrs:
                     setattr(self, basename+attr, [0 for r in range(self.R)])
                     all_order_attrs.append(basename+attr)
@@ -134,8 +134,8 @@ class Results(object):
                 for n in self.component_names:
                     g.create_dataset(n+'_ys_predicted', data=getattr(self, n+'_ys_predicted')[r])
                     attrs = np.append(COMPONENT_NP_ATTRS, COMPONENT_TF_ATTRS)
-                    if getattr(self, n+'_K') > 0:
-                        attrs = np.append(attrs, [OPT_COMPONENT_NP_ATTRS, OPT_COMPONENT_TF_ATTRS])
+                    if np.any(np.array(getattr(self, n+'_K')) > 0):
+                        attrs = np.append(attrs, np.append(OPT_COMPONENT_NP_ATTRS, OPT_COMPONENT_TF_ATTRS))
                     for attr in attrs:
                         g.create_dataset(n+'_'+attr, data=getattr(self, n+'_'+attr)[r])
             self.component_names = [a.encode('utf8') for a in self.component_names] # h5py workaround

### 2018-10-09 test.py
better learning rates

@@ -7,12 +7,13 @@ import h5py
 import os
 
 if __name__ == "__main__":
-    starname = 'barnards'
+    starname = '51peg'
     K_star = 0
     K_t = 0    
-    niter = 200 # for optimization
+    niter = 150 # for optimization
     plots = True
     epochs = [0, 50] # to plot
+    movies = False
     
     star_reg_file = 'wobble/regularization/{0}_star_K{1}.hdf5'.format(starname, K_star)
     tellurics_reg_file = 'wobble/regularization/{0}_t_K{1}.hdf5'.format(starname, K_t)
@@ -25,10 +26,13 @@ if __name__ == "__main__":
         for r in range(data.R):
             model = wobble.Model(data, results, r)
             model.add_star('star', variable_bases=K_star, 
-                            regularization_par_file=star_reg_file)
+                            regularization_par_file=star_reg_file,
+                            learning_rate_template=0.01, learning_rate_rvs=1.)
             model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
-                                regularization_par_file=tellurics_reg_file)
-            wobble.optimize_order(model, niter=80, save_history=False)
+                                regularization_par_file=tellurics_reg_file,
+                                learning_rate_template=0.01)
+            wobble.optimize_order(model, niter=niter, save_history=True, uncertainties=False,
+                                  basename='results/test', epochs=epochs, movies=movies)
         results.write('results/test_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t))
         assert False
     
@@ -40,22 +44,28 @@ if __name__ == "__main__":
     results = wobble.Results(data=data)
     
     print("data loaded")
-    print("time elapsed: {0:.2f} s".format(time() - start_time))
+    print("time elapsed: {0:.2f} min".format((time() - start_time)/60.0))
+    elapsed_time = time() - start_time
+    
 
     if plots:
         print("plots will be saved under directory: {0}".format(plot_dir))
         if not os.path.exists(plot_dir):
             os.makedirs(plot_dir)
+    star_learning_rate = 0.1
+    telluric_learning_rate = 0.01
     for r,o in enumerate(orders):
         model = wobble.Model(data, results, r)
         model.add_star('star', variable_bases=K_star, 
-                        regularization_par_file=star_reg_file)
+                        regularization_par_file=star_reg_file, 
+                        learning_rate_template=star_learning_rate)
[…]

### 2018-10-09 wobble/history.py

@@ -133,12 +133,12 @@ class History(object):
         synths = np.exp(self.synth_history[:,e,:])
         data = np.exp(self.data.ys[self.r][e])
         data_mask = self.data.ivars[self.r][e] <= 1.e-8
-        resids = np.exp(self.synth_history[:,e,:] - np.repeat([self.data.ys[self.r][e]], self.niter, axis=0))
+        resids =  np.exp(np.repeat([self.data.ys[self.r][e]], self.niter, axis=0)) - np.exp(self.synth_history[:,e,:])
         # set up ylims if not otherwise specified:
         y_pad = 0.1
-        ylims = (np.min(synths)-y_pad, min([np.max(synths)+y_pad, 10.])) # hard upper limit to prevent inf
-        kwargs['ylims'] = kwargs.get('ylims', ylims)
-        kwargs['ylims2'] = kwargs.get('ylims', [-0.1, 0.1])
+        yl = (np.min(synths)-y_pad, min([np.max(synths)+y_pad, 10.])) # hard upper limit to prevent inf
+        ylims = kwargs.get('ylims', yl)
+        ylims2 = kwargs.get('ylims2', [-0.25, 0.25])
         if nframes is None:
             nframes = self.niter
         fig, (ax, ax2) = plt.subplots(2, 1, gridspec_kw = {'height_ratios':[4, 1]}, figsize=(12,5))
@@ -150,21 +150,19 @@ class History(object):
         return ani  
         
                 
-    def save_plots(self, basename, epochs=[0,50]):
+    def save_plots(self, basename, epochs=[0,50], movies=True):
         plt.scatter(np.arange(len(self.nll_history)), self.nll_history)
         ax = plt.gca()
         ax.set_yscale('log')
         plt.savefig(basename+'_order{0}_nll.png'.format(self.order))   
-        plt.clf()
-        
-        rvs_ani_star = self.plot_rvs(0, compare_to_pipeline=True)
-        rvs_ani_star.save(basename+'_order{0}_rvs_star.mp4'.format(self.order), fps=30, extra_args=['-vcodec', 'libx264'])
-        
-        template_ani_star = self.plot_template(0, nframes=50)
-        template_ani_star.save(basename+'_order{0}_template_star.mp4'.format(self.order), fps=30, extra_args=['-vcodec', 'libx264'])
-        template_ani_t = self.plot_template(1, nframes=50)
[…]

### 2018-10-09 wobble/model.py

@@ -120,7 +120,7 @@ class Model(object):
         session.run(tf.global_variables_initializer())
 
     def optimize(self, niter=100, save_history=False, basename='wobble',
-                 feed_dict=None, verbose=True, uncertainties=True):
+                 feed_dict=None, verbose=True, uncertainties=True, **kwargs):
         """Optimize the model!
             
         Parameters
@@ -161,7 +161,7 @@ class Model(object):
             self.results.update(c)
         # save optimization plots:
         if save_history:
-            history.save_plots(basename)
+            history.save_plots(basename, **kwargs)
             
     def estimate_uncertainties(self, verbose=True):
         """Estimate uncertainties using the second derivative of the likelihood. 
@@ -204,7 +204,7 @@ class Component(object):
     """
     def __init__(self, name, r, starting_rvs, regularization_par_file=None,
                  L1_template=0., L2_template=0., L1_basis_vectors=0.,
-                 L2_basis_vectors=0., L2_basis_weights=1., learning_rate_rvs=10.,
+                 L2_basis_vectors=0., L2_basis_weights=1., learning_rate_rvs=1.,
                  learning_rate_template=0.01, learning_rate_basis=0.01,
                  rvs_fixed=False, variable_bases=0, scale_by_airmass=False,
                  template_xs=None, initialize_at_zero=False):

### 2018-10-09 wobble/results.py

@@ -110,7 +110,7 @@ class Results(object):
             getattr(self, basename+attr)[c.r] = session.run(getattr(c,attr), **kwargs)
                 
     def read(self, filename):
-        """Write to HDF5 file."""
+        """Read from HDF5 file."""
         print("Results: reading from {0}".format(filename))
         with h5py.File(filename,'r') as f:
             for attr in COMMON_ATTRS:
@@ -145,7 +145,7 @@ class Results(object):
                 
                     
     def write(self, filename):
-        """Read from HDF5 file."""
+        """Write to HDF5 file."""
         print("Results: writing to {0}".format(filename))
         with h5py.File(filename,'w') as f:            
             for r in range(self.R):

### 2018-10-09 paper/paper.tex

@@
 % Aims
-Here we propose a data-driven method to simultaneously extract precise \RV s and infer the underlying stellar and telluric spectra using a fully linear model. 
+Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a fully linear model. 
 We implement this method in \wobble, an open-source \python package which uses \TF in its first non-machine-learning application to astronomical data. 
@@
 This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres \citep{Figueira2016}.
-Eliminating large chunks of these spectra will therefore significantly inhibit our ability to detect planets around M dwarfs through \RV s.
+Eliminating large chunks of these spectra will therefore significantly inhibit our ability to detect planets around M dwarfs through \RVs.
 
@@
 Our model requires no prior knowledge of the star or the Earth's atmosphere. 
-As such, it does not yield absolute measurements of \RV s, only highly precise relative measurements between epochs.
+As such, it does not yield absolute measurements of \RVs, only highly precise relative measurements between epochs.
 
@@
 a single night), it will be nearly impossible to disentangle the components of the model from each other, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
-We will revisit this restriction in Section \label{s:future}.
+We will revisit this restriction in Section \ref{s:future}.
 
 The spectrum model may be composed of an arbitrary number of components convolved together, but for the purposes of this work we consider the two-component case: a single star whose spectrum is convolved with a telluric absorption spectrum.
-The star has a radial velocity about its center of mass $v_{\star}$, which is added to the (known) motion of the Earth around the solar system barycenter (the Barycentric Earth Radial Velocity or \BERV ) to produce the observed stellar \RV\ shift $v_{obs}$.
+The star has a radial velocity about its center of mass $v_{\star}$, which is added to the (known) motion of the Earth around the solar system barycenter (the Barycentric Earth Radial Velocity or \BERV ) to produce the observed stellar \RV shift $v_{obs}$.
 The telluric spectrum is fixed at zero velocity shift relative to the observatory rest frame, but its spectral shape may vary in time. 
 We assume that these variations can be captured in a low-dimensional space.
+
+Finally, we assume that, when the stellar and telluric components are properly modeled, only Gaussian noise remains. 
+This assumption will be revisited in Section \ref{s:future}.
 
@@
 \end{equation}
-The $P$ operator applies this term to the $M'$-length wavelength grid for the template spectrum before interpolating the template to the $M$-length data space.
+The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
 
@@
 \end{equation}
-In addition to its mean spectrum $\mu_t$, the telluric component also takes a time-dependent component assembled from two variables: $W$, a matrix of ``basis vectors,'' is weighted by $z_n$ to form the spectral contribution at epoch $n$.
+In addition to its mean spectrum $\mu_t$, the telluric component also takes a time-dependent component assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
 $W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
@@
 
-The number of free parameters in this model is large: we must optimize every grid point in the mean spectral templates and telluric basis vectors along with the stellar \RV s and the telluric basis weights.
-While in principle degeneracies between the stellar and telluric spectra should not exist since the star will always vary in \RV by at least \BERV \todo{(uhhh is this true??)}, in practice the complexity of the optimization problem makes parameter degeneracy a significant concern \todo{(total BS)}.
-We use a combination of L1 and L2 regularization to mitigate this issue. 
-\todo{define these, discuss their purpose}
+\subsection{Regularization}
 
-The final likelihood to be optimized is then:
+The number of free parameters in this model is large: we must optimize every grid point in the mean spectral templates and telluric basis vectors along with the stellar \RV and the telluric basis weights for each epoch.
+While in principle the stellar and telluric spectra should not be degenerate except in the unlikely scenario that the stellar \RV is exactly equal and opposite to the \BERV, in practice such degeneracies do arise.
+Model overfitting can introduce noise in one of the templates, which then propagates to the other template in an attempt to counterbalance, pulling the overall optimization off course to a local minimum.
+The end result is then a large amount of noise or unphysical trends in the stellar and telluric spectral templates.
+
+Using our expectations for the spectra, we can apply a combination of L1 and L2 regularization to the spectral templates to mitigate this issue. 
+Regularization is a commonly used technique in machine learning, where large numbers of free parameters are standard. 
+It is equivalent to applying a prior to the parameters which pushes them toward zero. 
+In $\ln$ flux space the spectral continuum is at zero, making this a good prior assumption.
+
+L1 normalization adds a term to the log-likelihood that takes the form:
+$$ L1(p, \lambda) = \lambda \sum_{i} | p_{i} | ,$$
+where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.
+
+Similarly, L2 normalization adds a term of the form:
+$$ L2(p, \lambda) = \lambda \sum_{i} p_{i}^2 .$$
+
+The effectiveness of the regularization depends sensitively on the value of $\lambda$ used: if $\lambda$ is too high, real features will be lost as the parameters are forced to zero, whereas setting $\lambda$ too low will make the regularization ineffective, leaving the model vulnerable to overfitting.
+
+We set regularization amplitudes for \wobble using a cross-validation scheme. 
+In brief, we set aside 12.5\% of the total epochs as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set). 
+The resulting best-fit spectral templates and basis vectors are taken as fixed and the time-dependent terms only (\RVs and basis weights) are optimized for the validation epochs. 
+The $\chi^2$ for the validation epochs can then be adopted as a goodness-of-fit measurement for the $\lambda$ value, and the procedure is repeated for different $\lambda$s to choose the best regularization amplitude.
+
+With regularization included, the final model likelihood is:
 \begin{equation}
@@
 \ln \mathcal{L} = & \sum_{n} \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C^{-1} (y_{m,n}-y'_{m,n})  \\
- & + \alpha_{L1, \mu_{\star}}  \sum_{m'} | \mu_{{\star}} | + \alpha_{L2, \mu_{\star}}  \sum_{m'} \mu_{{\star}}^2 \\
- & + \alpha_{L1, \mu_t}  \sum_{m'} | \mu_{t} |  + \alpha_{L2, \mu_t}  \sum_{m'} \mu_{t}^2 \\
- & + \alpha_{L1, W_t}  \sum_{k} \sum_{m'} | W_{t} |  + \alpha_{L2, W_t}  \sum_{k} \sum_{m'} W_{t}^2 + \alpha_{L2, z_t} \sum_{k} \sum_{n} z_{t}^2 ,
+ & + L1(\mu_{\star}, \lambda_{L1, \mu_{\star}}) + L2(\mu_{\star}, \lambda_{L2, \mu_{\star}}) + L1(\mu_{t},  \lambda_{L1, \mu_{t}}) + L2(\mu_{t},  \lambda_{L2, \mu_{t}}) \\
+ & + L1(W_t, \lambda_{L1, W_t}) + L2(W_t, \lambda_{L1, W_t}) + L2(z_t, 1.0) ,
 \end{split}
 \end{equation}
-where $\alpha_{L2, z_t}$ is set to 1.0 and all other regularization amplitudes are set by a validation procedure \todo{as described below}.
+where the regularization amplitude for the basis weights is arbitrarily set to 1.0 and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
+\todo{(todo: explain why regularization on basis weights is needed)}
+
+In practice, the most effective regularization amplitudes can vary by orders of magnitude depending on the characteristics of the spectral region being used. 
+For instance, both $\lambda_{L1, \mu_{t}}$ and $\lambda_{L2, \mu_{t}}$ need to be much higher when the spectra in question span relatively blue wavelengths with few actual telluric features to fit than they do for the near-infrared wavelength regions with strong telluric features.
+Similarly, the regularization amplitudes for the star tend to be higher for the G dwarf than they are for the M dwarf, which has far more absorption features in its spectra. 
+Setting the regularization amplitudes is therefore a crucial first step when applying the \wobble model to data.
 
@@
 
-We build an initial template for the stellar spectrum by setting $v_{obs} = v_{BERV}$ for all epochs.
-We can then calculate the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$.
-The telluric template is then initialized using the residuals after the stellar contribution has been removed, again by binning (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
+Another key step towards running \wobble on data is the initialization of parameters. 
+To begin with, we assume that the star is stationary, e.g. $v_{obs} = v_{BERV}$ at all epochs. 
+We can then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
+The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
 Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
-The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors.
-\todo{(math here?)}
+The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors. \todo{(math here?)}
 
@@
 
-\subsection{Code}
+\subsection{Combining Spectral Orders}
 
-The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, using the \TF package \todo{[terminology?]} in \code{python}. 
+Most \EPRV instruments are echelle spectrographs spanning many orders. 
+These orders can be considered as independent spectra \todo{for reasons}. 
+\todo{We run on each echelle order independently and then combine them like so:}
+
+\subsection{Implementation}
+
+The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF. 
 \TF is a \todo{[blah blah technical blah]}. 
@@
 
-In its current realization, \wobble\ is optimized for use with data products from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}. 
-Like most \EPRV instruments, \HARPS is an echelle spectrograph spanning many spectral orders. 
-These orders can be considered as independent spectra \todo{for reasons}. 
-\todo{We run on each echelle order independently and then combine them like so:}
-
 Our code is open-source and publicly available on GitHub.\footnote{\url{https://www.github.com/megbedell/wobble}}
@@
 \subsection{51 Pegasi}
+
+We tested \wobble on spectra from the public archives of the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}.
 

### 2018-10-10 wobble/model.py
more plots, fixed bug

@@ -295,7 +295,7 @@ class Component(object):
             # initialize basis components
             resids = np.empty((len(self.starting_rvs),len(template_ys)))
             for n in range(len(self.starting_rvs)):
-                resids[n] = np.interp(template_xs, shifted_xs[n], data_ys[n]) - template_ys
+                resids[n] = np.interp(self.template_xs, shifted_xs[n], data_ys[n]) - template_ys
             u,s,v = np.linalg.svd(resids, compute_uv=True, full_matrices=False)
             basis_vectors = v[:self.K,:] # eigenspectra (K x M)
             basis_weights = u[:, :self.K] * s[None, :self.K] # weights (N x K)

### 2018-10-11 wobble/model.py
attempting to fit continuum

@@ -68,6 +68,17 @@ class Model(object):
         kwargs['rvs_fixed'] = kwargs.get('rvs_fixed', True)
         #kwargs['initialize_at_zero'] = kwargs.get('initialize_at_zero', True)
         self.add_component(name, starting_rvs, **kwargs)
+        
+    def add_continuum(self, degree, **kwargs):
+        if np.isin("continuuum", self.component_names):
+            print("The model already has a continuum component.")
+            return
+        c = Continuum(self.r, self.data.N, degree, **kwargs)
+        self.components.append(c)
+        self.component_names.append(c.name)
+        if not np.isin(c.name, self.results.component_names):
+            self.results.add_component(c)
+        
 
     def initialize_templates(self):
         """Initialize spectral templates for all components. 
@@ -306,4 +317,27 @@ class Component(object):
         for n in range(len(self.starting_rvs)):
             data_resids[n] -= np.interp(shifted_xs[n], self.template_xs, full_template[n])
         return data_resids
+        
+class Continuum(Component):
+    """
+    Polynomial continuum component which is modeled in data space
+    """
+    def __init__(self, r, N, degree, **kwargs):
+        Component.__init__(self, 'continuum', r, np.zeros(N),
+                 rvs_fixed=True, variable_bases=0, scale_by_airmass=False, **kwargs)
+        self.degree = degree
+        
+    def setup(self, data, r):
+        self.template_xs = tf.constant(self.wavelength_matrix, dtype=T, name='wavelength_matrix_'+self.name)
+        self.template_ys = tf.Variable(self.weights, dtype=T, name='weights_'+self.name) # HACK to play well with Results
+        self.rvs = tf.constant(self.starting_rvs, dtype=T, name='rvs_'+self.name) # HACK to play well with Results
+        self.synth = tf.matmul(self.template_ys, self.template_xs)
+        self.nll = tf.constant(0., dtype=T) # no regularization
+        
+    def initialize_template(self, data_xs, data_ys, data_ivars):
+        assert np.all(data_xs[0] == data_xs[1]), "Continuum failed: wavelength grid must be constant in time"
+        wavelength_vector = (data_xs[0] - np.mean(data_xs[0]))/(np.max(data_xs[0]) - np.min(data_xs[0]))
+        self.wavelength_matrix = np.array([wavelength_vector**d for d in range(1,self.degree)]) # D_degrees x M_pixels
+        self.weights = np.zeros((self.N, self.degree-1)) # N_epochs x D_degrees
+        return data_ys
 

### 2018-10-22 regularization.py
minor edits

@@ -297,10 +297,10 @@ def plot_pars_from_file(filename, basename, orders=np.arange(72)):
 
         
 if __name__ == "__main__":
-    starname = 'barnards'
+    starname = '51peg'
     orders = np.arange(72)
     K_star = 0
-    K_t = 0
+    K_t = 3
     initialize_at_zero = False # toggle telluric template initialization
     plot = True
     verbose = True
@@ -330,7 +330,7 @@ if __name__ == "__main__":
             f.create_dataset('L2_template', data=np.zeros(R)+1.e6)
             if K_t > 0:
                 f.create_dataset('L1_basis_vectors', data=np.zeros(R)+1.e3)
-                f.create_dataset('L2_basis_vectors', data=np.zeros(R)+1.e9)
+                f.create_dataset('L2_basis_vectors', data=np.zeros(R)+1.e8)
                 f.create_dataset('L2_basis_weights', data=np.ones(R)) # never tuned, just need to pass to wobble
 
     # set up training & validation data sets:

### 2018-10-23 paper/paper.tex
writing

@@
 \include{preamble}
-\newcommand{\Mdwarf}{Barnard's Star}
+\newcommand{\Mdwarf}{Barnard's Star\xspace}
 
@@
 % Aims
-Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a fully linear model. 
+Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a simple linear model. 
 We implement this method in \wobble, an open-source \python package which uses \TF in its first non-machine-learning application to astronomical data. 
@@
 We recover the canonical exoplanet 51 Pegasi b at \todo{improved accuracy?} relative to the standard \HARPS pipeline, and we achieve a precision of \todo{xx} \cms on the \RV standard M dwarf \Mdwarf. 
-We also present a detailed telluric spectrum derived from these data. 
+We also present detailed time-variable telluric spectra derived from these data. 
 This method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.
@@
 Often, particular sections of a spectrum that are likely to feature telluric features are identified before the velocity shift of the stellar spectrum is inferred. 
-These regions are then removed from analysis, leaving only telluric-free regions to be analyzed \citep[e.g.][]{AngladaEscude2012}.
+These regions are then removed from analysis, leaving only seemingly telluric-free regions to be analyzed \citep[e.g.][]{AngladaEscude2012}.
 
 Such an approach has two significant issues. 
-The first is that removing sections of the spectrum removes significant regions of the spectrum that can be used to infer the stellar radial velocity. 
+The first is that removing sections of the spectrum can remove significant amounts of information about the star, lowering the precision at which we can measure the stellar radial velocity. 
 Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV precision (cite). 
@@
 As the Earth's motion around the barycenter of the solar system induces a Doppler shift considerably larger than both the motion of telluric features and the size of a single pixel on the detector, these two spectra can be disentangled.
-This process is well-established in the analysis of binary star systems through the development of linear models
-\citep[e.g.][]{Simon1994} and in a Gaussian process framework \citep{Czekala2017}.
+This process is well-established in the analysis of binary star systems through the development of linear models \citep[e.g.][]{Simon1994} and in a Gaussian process framework \citep{Czekala2017}.
 In these cases, both spectra are assumed to be unchanging in time, which is a reasonable approximation of a stellar spectrum but not necessarily of the telluric spectrum.
@@
 \label{s:methods}
-\subsection{Model}
+\subsection{Model Assumptions}
+\label{s:assumptions}
 
@@
 
-We take data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
-We also have a corresponding $M \times N$ matrix of wavelength solutions obtained in the course of instrument calibrations, which we assume to be accurate.
+First, we assume that the wavelength calibration and spectral extraction of the instrument are perfect: that is, we begin at the stage of having 1D extracted spectra and corresponding wavelength grids in hand and we do not model any corrections to the wavelength solution.
+
+We assume that the spectra can be modeled as the product of a finite number of components. 
+For the cases shown in this work, two components are used: a stellar spectrum which is invariant in shape but may be Doppler-shifted, and a telluric spectrum which is fixed to the observatory rest frame but varies in shape.
+We choose to work in log(flux) space so that the data to be modeled are simply a sum of the component spectra.
+
+For the case of the telluric component, whose spectrum is allowed to vary with time, we assume this spectral variability is low-dimensional. 
+This assumption is physically motivated by the fact that a relatively small number of molecular species contribute to the telluric absorption spectrum. 
+It is also needed in a practical sense, since every additional dimension over which the telluric spectrum can vary adds several thousand more free parameters to the model.
+
+We assume that the stellar spectrum is invariant with time. 
+This assumption will be revisited in Section \ref{s:future}.
+
+We assume that both the stellar and telluric spectra are approximately located at zero in logarithmic flux with small deviations due to absorption lines. 
+As a result of this assumption, we are able to apply L1 and L2 regularization to the spectral templates. 
+Regularization is a commonly used technique in machine learning, where large numbers of free parameters are standard. 
+It is equivalent to applying a prior to the parameters which pushes them toward zero in the absence of strong evidence otherwise from the data. 
+The strength of the regularization may be tuned to suit the data at hand through a cross-validation scheme: for example, the best-suited regularization in the bluest spectral orders may be much stronger for the telluric spectrum, where few features are present, than it is for the star, which generally has a dense forest of spectral lines.
+The exact implementation and validation of regularization in this model is further described in Section \ref{s:model-eqns}.
+
+Aside from this regularization to push spectra to zero in the low-\SNR regime, the model makes no assumptions about the shape of the stellar or telluric spectra. 
+We solve for spectral templates for each component as a series of pixels with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing.
+In addition to keeping this model simple, linear, and largely convex, this means that no physical knowledge about the object being observed is needed to extract its \RVs.
+
+%For practical applications, the number of epochs $N$ must be $\gg 1$. 
+%If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.
+a single night), it will be nearly impossible to disentangle the components of the model from each other, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
+%We will revisit this restriction in Section \ref{s:future}.
+
+Finally, we assume that, when the stellar and telluric components are properly optimized, only Gaussian noise remains. 
+This assumption will be revisited in Section \ref{s:future}.
+
+\subsection{Model Specification}
+\label{s:model-eqns}
+
+We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
+We also have a corresponding $M \times N$ matrix of wavelength solutions. 
 Again, the entries of this matrix are in logarithmic form ($\ln \lambda$). 
@@
 
-For practical applications, the number of epochs $N$ must be $\gg 1$. 
-If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.
-a single night), it will be nearly impossible to disentangle the components of the model from each other, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
-We will revisit this restriction in Section \ref{s:future}.
-
-The spectrum model may be composed of an arbitrary number of components convolved together, but for the purposes of this work we consider the two-component case: a single star whose spectrum is convolved with a telluric absorption spectrum.
-The star has a radial velocity about its center of mass $v_{\star}$, which is added to the (known) motion of the Earth around the solar system barycenter (the Barycentric Earth Radial Velocity or \BERV ) to produce the observed stellar \RV shift $v_{obs}$.
-The telluric spectrum is fixed at zero velocity shift relative to the observatory rest frame, but its spectral shape may vary in time. 
-We assume that these variations can be captured in a low-dimensional space.
-
-Finally, we assume that, when the stellar and telluric components are properly modeled, only Gaussian noise remains. 
-This assumption will be revisited in Section \ref{s:future}.
-
-Given these assumptions, we can model each data column $y_n$ as the sum of the stellar and telluric spectra:
+We model each data column $y_n$ as the sum of the stellar and telluric spectra:
 \begin{equation}
@@
 \begin{equation}
- \ln \lambda(v) = \ln \lambda_{0} + \ln \left(\frac{1 - v/c}{1 + v/c}\right).
+ \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
 \end{equation}
 The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
+
+The apparent stellar \RV, $v_{obs, n}$, is a sum of the star's actual velocity about its center of mass ($v_{\star, n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV$_{n}$ ), the latter of which is known.
 
@@
 
-We evaluate a likelihood for each epoch
+We evaluate a log-likelihood for each epoch
 \begin{equation}
@@
 comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties on each data point.
-Our model is agnostic to the intrinsic wavelength of any individual stellar or telluric feature, so while we infer the Doppler shift relative to our model, it does not apply any information about the absolute \RV of the star relative to the solar system.
-%To mitigate the potential degeneracy of developing a model effectively redshifted relative to all observations, we add (in log space) a Gaussian prior to our likelihood value on the mean \RV for all $N$ epochs such that
-%$$ P(m|X) = \ln \mathcal{L}  -0.5 * \frac{\sum(v)^2}{(Nl)^2}.
-%$$
-%\todo{someone should define these variables}
-
-\subsection{Regularization}
 
 The number of free parameters in this model is large: we must optimize every grid point in the mean spectral templates and telluric basis vectors along with the stellar \RV and the telluric basis weights for each epoch.
-While in principle the stellar and telluric spectra should not be degenerate except in the unlikely scenario that the stellar \RV is exactly equal and opposite to the \BERV, in practice such degeneracies do arise.
-Model overfitting can introduce noise in one of the templates, which then propagates to the other template in an attempt to counterbalance, pulling the overall optimization off course to a local minimum.
-The end result is then a large amount of noise or unphysical trends in the stellar and telluric spectral templates.
-
-Using our expectations for the spectra, we can apply a combination of L1 and L2 regularization to the spectral templates to mitigate this issue. 
-Regularization is a commonly used technique in machine learning, where large numbers of free parameters are standard. 
-It is equivalent to applying a prior to the parameters which pushes them toward zero. 
-In $\ln$ flux space the spectral continuum is at zero, making this a good prior assumption.
+We deal with potential over-fitting issues by applying L1 and L2 regularization to the spectral templates, as discussed in Section \ref{s:assumptions}.
 
@@
 The $\chi^2$ for the validation epochs can then be adopted as a goodness-of-fit measurement for the $\lambda$ value, and the procedure is repeated for different $\lambda$s to choose the best regularization amplitude.
+Because we have multiple regularization terms and multiple amplitudes to set, we begin by hand-setting all amplitudes to a reasonable starting guess and optimize each amplitude sequentially with cross-validation.
 
@@
 
-In practice, the most effective regularization amplitudes can vary by orders of magnitude depending on the characteristics of the spectral region being used. 
-For instance, both $\lambda_{L1, \mu_{t}}$ and $\lambda_{L2, \mu_{t}}$ need to be much higher when the spectra in question span relatively blue wavelengths with few actual telluric features to fit than they do for the near-infrared wavelength regions with strong telluric features.
-Similarly, the regularization amplitudes for the star tend to be higher for the G dwarf than they are for the M dwarf, which has far more absorption features in its spectra. 
-Setting the regularization amplitudes is therefore a crucial first step when applying the \wobble model to data.
+%In practice, the most effective regularization amplitudes can vary by orders of magnitude depending on the characteristics of the spectral region being used. 
+%For instance, both $\lambda_{L1, \mu_{t}}$ and $\lambda_{L2, \mu_{t}}$ need to be much higher when the spectra in question span relatively blue wavelengths with few actual telluric features to fit than they do for the near-infrared wavelength regions with strong telluric features.
+%Similarly, the regularization amplitudes for the star tend to be higher for the G dwarf than they are for the M dwarf, which has far more absorption features in its spectra. 
+%Setting the regularization amplitudes is therefore a crucial first step when applying the \wobble model to data.
 
 \subsection{Optimizing the Model}
+\label{s:optimizing}
 
-Another key step towards running \wobble on data is the initialization of parameters. 
-To begin with, we assume that the star is stationary, e.g. $v_{obs} = v_{BERV}$ at all epochs. 
-We can then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
+As an initial guess, we set the star to be stationary, e.g. $v_{obs} = v_{BERV}$ at all epochs. 
+We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
 The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
 Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
-The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors. \todo{(math here?)}
+The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors. 
+%\todo{(math here?)}
 
-This likelihood function is maximized for our assumed stellar and telluric models iteratively, first calculating the best-fitting stellar \RV\ shifts holding the telluric shifts constant, then repeating for the telluric spectrum shift holding the stellar \RV s constant.
-After each optimization, we use these velocities to improve our models for the telluric and stellar spectra, employing a gradient descent scheme to modify our templates $m'$ at each wavelength $\lambda'$ using the newly calculated velocities.
-We repeat this procedure, calculating velocities for each epoch and in turn using these velocities to infer an improved model spectrum, until the procedure converges. 
-\todo{(say something about model linearity and why this iterative procedure is effective)}
+Once all parameters are initialized, we optimize them iteratively. 
+The likelihood function is maximized first by varying the stellar and telluric templates (including the telluric basis vectors). 
+This step is a convex optimization, meaning that a global optimum for the template parameters should be reached under the condition of fixed time-dependent variables. 
+Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to again maximize the likelihood function. 
+\todo{(is this convex too?)}
+We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
+
+\todo{(say something about calculating uncertainties here?)}
 
 \subsection{Combining Spectral Orders}
+\label{s:combining-orders}
 
 Most \EPRV instruments are echelle spectrographs spanning many orders. 
-These orders can be considered as independent spectra \todo{for reasons}. 
-\todo{We run on each echelle order independently and then combine them like so:}
+These orders are often treated as independent spectra when extracting \RVs, and the \RVs for each order are then combined in some manner to get a final time series.
+
+We follow this precedent and optimize the \wobble model individually for each order. 
+While some \RV precision gain may be possible by inferring a global Doppler shift from all orders simultaneously, it is not clear that the gain would be worth the added computational expense, so we leave this for future work.
+
+After obtaining an $N$-epoch set of observed stellar \RVs from each of the $R$ spectral orders, we combine them by modeling each \RV $v_{n,r}$ as a combination of time-dependent stellar \RV, order-dependent \RV offset, and a Gaussian noise term:
+\begin{equation}
+v_{n,r} = v_n + v_r + \mathcal{N}(0,\,\sigma_{n,r}^{2} + \delta_r^2 )\,,
+\end{equation}
+where $v_n$ and $v_r$ are the characteristic \RVs at epoch $n$ and order $r$, $\sigma_{n,r}$ is the estimated measurement uncertainty on $v_{n,r}$, and $\delta_r$ is an additional jitter term specific to order $r$.
+
+\todo{(To do: fill in details: prior on $<v_r>$, optimization over $\delta_r$.)}
 
 \subsection{Implementation}
+\label{s:implementation}
 
@@
 While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
-The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of \todo{x} spectra of 51 Peg runs in \todo{y} minutes on a standard Mac desktop.
+The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of \todo{80} spectra of 51 Peg runs in \todo{30} minutes on a standard Mac desktop.
 
@@
 
-\section{Results}
+\section{Application to \HARPS Data}
 \label{s:results}
+
+For the purposes of this work, we chose to test the performance of \wobble on archival spectra from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}. 
+\HARPS has operated continuously since 2003 and as such has an extensive catalog of publicly available data. 
+Furthermore, its excellent instrumental stability and careful calibration make the data ideally suited to our method, which relies on having an accurate wavelength solution for every spectrum.
+
+All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra. 
+These data are blocks of 72 order by 4096 pixel extracted 1D spectra. 
+The airmass of the observation and the calculated \BERV are both provided in the \acronym{FITS} header of the spectral data file. 
+We obtained wavelength solutions for each spectrum from the dedicated \HARPS calibration archive, also maintained by \acronym{ESO}.
+
+Before running the above-described model optimization on the data, we first mask out any unreliably measured spectral regions and do a continuum normalization on each spectrum. 
+This masking is done by setting the inverse variance on the masked data point to zero, effectively removing it from the fit. 
+In brief, we mask pixels whose extracted flux is below zero as well as regions at the edges of the spectral orders where the local \SNR falls below 5. 
+After masking these data, we convert the flux to log space and continuum normalize by fitting a polynomial to an asymmetrically clipped subset of the data.
+
+Below we describe the results of applying this procedure and optimizing the \wobble model using \HARPS data for stars of different types: a G dwarf with a known planet and an \RV-quiet M dwarf.
+
 \subsection{51 Pegasi}
 
-We tested \wobble on spectra from the public archives of the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}.
+51 Pegasi is a Sun-like star hosting a planet with an orbital period of 4 days and a mass of 0.5 Jupiter masses \citep{Mayor1995}. 
+Its canonical status as the first exoplanet discovered means that large amounts of data exist for this system. 
+In particular, archival \HARPS data exist mainly from efforts to observe reflected-light spectra of the planet \citep{Martins2015}.
 
-\todo{Description of the data, cite \acronym{ESO} archive. Talk about continuum normalization here?}
+We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.
 
-\todo{Show our results: we find the planet with the correct mass and orbital parameters. Compare to \HARPS pipeline.}
+\wobble spectral fit in Figure \ref{fig:51peg_spectrum}. It's sweet!
 
-\todo{Plots of RV vs. time, models of star and tellurics in a few segments of the spectrum.}
+Planet recovered in \RV in Figure \ref{fig:51peg_planet}.
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{51peg_multispectrum}
+\caption{Three example echelle orders at a randomly chosen epoch from the \HARPS observations of 51 Pegasi.
+Data are shown as dark circles, while the \wobble predictions for the stellar and telluric spectral contributions at this epoch are red and blue lines, respectively.
+Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered.
+Residuals after subtracting both star and telluric models are shown below each spectrum.}
+\label{fig:51peg_spectrum}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=4in]{51peg_planet}
+\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered in the \wobble\ \RVs.
+While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns \todo{(add numbers)}.
+\todo{check on: The subtle trends seen near secondary eclipse (phase 0.75) are suggestive of contributions by reflected light from the planetary dayside.}}
+\label{fig:51peg_planet}
+\end{figure}
 
@@
 
-\todo{Results \& more plots.}
+\Mdwarf is one of the nearest stars to us at a distance of less than 2 parsecs. 
+It is also the star with the highest known proper motion. 
+Its trajectory translates to a projected secular change in \RV of approximately 4.5 $\mathrm{m}~\mathrm{s}^{-1}~\mathrm{yr}^{-1}$ \citep{Kurster2003}. 
+This trend has been observed in some data sets, including \acronym{UVES} observations and the \HARPS spectra as analyzed by the standard instrument pipeline, but is not seen in others such as data from \acronym{HIRES} and the same \HARPS spectra as analyzed by the \HARPS-TERRA pipeline \citep{Kurster2003, Bonfils2013, Choi2013, AngladaEscude2012}.
+With or without a secular trend, no planets have been found around \Mdwarf, despite considerable amounts of \RV monitoring.
 
-Compare to HARPS-TERRA results for this star \citep{AngladaEscude2012}.
+Testing \wobble on the \Mdwarf data allows us to evaluate its performance in a significantly different regime: the mid-M dwarf spectra are far denser in spectral features than a Sun-like star, and individual observations are at a much lower \SNR than the 51 Peg observations used above, although the total number of spectra is greater.
+
+Spectral fit in Figure \ref{fig:barnards_spectrum}.
+
+Comparison to spectral model prediction in Figure \ref{fig:barnards_model}: even in the blue where individual spectra are hugely noisy, we are seeing real features in the composite spectral template.
+This is a huge improvement over most existing M dwarf data and may be highly valuable for spectral characterization.
+
+RVs in Figure \ref{fig:barnards_rvs}.
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{barnards_multispectrum}
+\caption{Example echelle orders at a randomly chosen epoch from the \HARPS observations of Barnard's Star, plotted as in Figure \ref{fig:51peg_spectrum}.
+\todo{(to do: weaken telluric regularization to try and get missing line at 6576 \ang)}}
+\label{fig:barnards_spectrum}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=6in]{barnards_model}
+\caption{A subset of the \Mdwarf spectrum including the sodium doublet at 4456-7 \ang.
+The spectral template derived by \wobble is shown as a red line, while a theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as a blue line.
+The \wobble model has been Doppler-shifted to match the rest-frame line positions.}
+\label{fig:barnards_model}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{barnards_rvs}
+\caption{Radial velocity measurements for \Mdwarf from \wobble (black) and from the \HARPS closed-source pipeline (red). \todo{(Say more about expected/observed secular evolution)}}
+\label{fig:barnards_rvs}
+\end{figure}
+
+\subsection{Telluric Features}
+
+Comparison between 51 Peg and \Mdwarf telluric fits in Figure \ref{fig:telluric_comparison}.
+No constraints used to make them look the same, but they do!
+In principle all stars could have a shared telluric model and this would give the Best Telluric Spectrum Ever (TM).
+
+Comparison to telluric standard star in Figure \ref{fig:telluric_standard}. Higher SNR with no observational overhead!
+
+Sample of telluric variability in Figure \ref{fig:telluric_basis}. To do: figure out what these lines are, do the ones that vary together share a common species?
+
+To do: show that basis weights correlate with things like season or humidity.
+
+\begin{figure}
+\centering
+\includegraphics[width=6.5in]{telluric_comparison}
+\caption{Fits to a 15-\ang region with substantial telluric contamination for 51 Peg (left) and \Mdwarf (right).
+Data and \wobble model fits are shown in the upper panels, while residuals to the fits are shown in the lower panels.
+Both observations shown here were taken at an airmass of $\sim1.8$.
+Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, suggesting that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
+\label{fig:telluric_comparison}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{telluric_standard}
+\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (blue) and the spectrum observed with a telluric standard star (black).}
+\label{fig:telluric_standard}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{telluric_basis2}
+\caption{Example of the mean telluric spectral template in a small wavelength region (upper panel) and the three basis vectors used to capture the time-variable contributions to the spectrum, staggered by an arbitrary offset for display (lower panel).}
+\label{fig:telluric_basis}
+\end{figure}
 
@@
 
-\todo{Revisit the assumptions: single star, no gas cell, known instrumental wavelength solution, no intrinsic spectral variability, no additive terms, many spectra exist. 
-In what situations are these assumptions violated, and how would we configure \wobble to deal with them?}
+Applicability of \wobble as shown here to current and near-future instruments.
+
+Revisit the assumptions: single star, no gas cell, known instrumental wavelength solution, no intrinsic spectral variability, no additive terms, many spectra exist. 
+In what situations are these assumptions violated, and how would we configure \wobble to deal with them?
 

### 2018-10-23 wobble/make_data.py

@@ -71,7 +71,8 @@ def savfile_to_filelist(savfile, destination_dir='../data/'):
     # copies CCF + E2DS files to destination_dir and returns a list of the CCFs
     s = readsav(savfile)
     filelist = []
-    for f in s.files:
+    files = [f.decode('utf8') for f in s.files]
+    for f in files:
         shutil.copy2(f, destination_dir)
         spec_file = str.replace(f, 'ccf_G2', 'e2ds')
         shutil.copy2(spec_file, destination_dir)
@@ -109,13 +110,13 @@ def write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffi
 
 if __name__ == "__main__":
         
-    if True: #51 Peg
+    if False: #51 Peg
         ccf_filelist = np.genfromtxt('51peg_ccf_filelist.txt', dtype=None, encoding=None)
         data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
         hdffile = '../data/51peg_e2ds.hdf5'
         write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffile)
         
-    if True: #Barnard's Star
+    if False: #Barnard's Star
         ccf_filelist = glob.glob('/Users/mbedell/python/wobble/data/barnards/HARPS*ccf_M2_A.fits')
         
         if False: # check for missing wavelength files
@@ -126,4 +127,10 @@ if __name__ == "__main__":
         data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
 
         hdffile = '../data/barnards_e2ds.hdf5'
-        write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffile)                  
+        write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffile)     
+        
+    if True: #HIP11915
+        ccf_filelist = glob.glob('/Users/mbedell/python/wobble/data/hip11915/HARPS*ccf_G2_A.fits')
+        data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
+        hdffile = '../data/hip11915_e2ds.hdf5'
+        write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffile)

### 2018-11-01 paper/paper.tex
adding orcid
more writing

@@
 
-\author{Daniel Foreman-Mack\'{e}y}
+\author{Daniel Foreman-Mackey}
 \affiliation{\flatiron}
 
-\author{Benjamin T. Montet}
+\author[0000-0001-7516-8308]{Benjamin T. Montet}
+\altaffiliation{Sagan Fellow}
 \affiliation{\chicago}
@@
 \begin{equation}
-\ln \mathcal{L}_n = \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C^{-1} (y_{m,n}-y'_{m,n}),
+\ln \mathcal{L}_n = \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C_{m,n}^{-1} (y_{m,n}-y'_{m,n}),
 \end{equation}
@@
 \begin{split}
-\ln \mathcal{L} = & \sum_{n} \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C^{-1} (y_{m,n}-y'_{m,n})  \\
+\ln \mathcal{L} = & \sum_{n} \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C_{m,n}^{-1} (y_{m,n}-y'_{m,n})  \\
  & + L1(\mu_{\star}, \lambda_{L1, \mu_{\star}}) + L2(\mu_{\star}, \lambda_{L2, \mu_{\star}}) + L1(\mu_{t},  \lambda_{L1, \mu_{t}}) + L2(\mu_{t},  \lambda_{L2, \mu_{t}}) \\
@@
 This step is a convex optimization, meaning that a global optimum for the template parameters should be reached under the condition of fixed time-dependent variables. 
-Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to again maximize the likelihood function. 
+Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to maximize the likelihood function again. 
 \todo{(is this convex too?)}
@@
 
-\todo{(say something about calculating uncertainties here?)}
+The inverse variance uncertainty on the parameters is calculated as the Hessian, or the second derivative of the likelihood with respect to each parameter. \todo{(check phrasing with DFM)}
 
@@
 While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
-The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of \todo{80} spectra of 51 Peg runs in \todo{30} minutes on a standard Mac desktop.
+The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 spectra of 51 Peg runs in \todo{30} minutes on a standard Mac desktop.
 
@@
 In brief, we mask pixels whose extracted flux is below zero as well as regions at the edges of the spectral orders where the local \SNR falls below 5.
+\todo{(add some equations for variance estimation and error conversion to log?)}
 After masking these data, we convert the flux to log space and continuum normalize by fitting a polynomial to an asymmetrically clipped subset of the data.
@@
 
-51 Pegasi is a Sun-like star hosting a planet with an orbital period of 4 days and a mass of 0.5 Jupiter masses \citep{Mayor1995}. 
+For \wobble's first test, we chose the first known exoplanet host: 51 Pegasi. 
+This target is a Sun-like star hosting a planet with an orbital period of 4 days and a mass of 0.5 Jupiter masses \citep{Mayor1995}. 
 Its canonical status as the first exoplanet discovered means that large amounts of data exist for this system. 
@@
 
-We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.
+We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell329968 and mbedell331551.}
+The 91 publicly available spectra in the \HARPS archive are largely concentrated on a few nights of intensive observing, but these nights are sufficiently spread out throughout the year to give a wide enough range in \BERV to disentangle the stellar and telluric spectra.
 
-\wobble spectral fit in Figure \ref{fig:51peg_spectrum}. It's sweet!
+Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the correct period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}).
+\todo{(numbers here, address the structure in residuals)}
 
-Planet recovered in \RV in Figure \ref{fig:51peg_planet}.
-
-\begin{figure}
-\centering
-\includegraphics[width=5in]{51peg_multispectrum}
-\caption{Three example echelle orders at a randomly chosen epoch from the \HARPS observations of 51 Pegasi.
-Data are shown as dark circles, while the \wobble predictions for the stellar and telluric spectral contributions at this epoch are red and blue lines, respectively.
-Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered.
-Residuals after subtracting both star and telluric models are shown below each spectrum.}
-\label{fig:51peg_spectrum}
-\end{figure}
+Fits to an individual spectrum are shown in Figure \ref{fig:51peg_spectrum}. 
+We emphasize that no \textit{a priori} information on e.g. expected spectral line positions and depths were used. 
+The current realization of the \wobble algorithm treats each pixel of the template spectra as independent and has no line shape parameterization included. 
+Nevertheless, the optimized stellar templates clearly reproduce the expected appearance of the spectra in a variety of regimes, from a crowded-line region to a sparser, continuum-dominated region to an extremely strong absorption line like the H$\alpha$ feature.
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=5.3in]{51peg_multispectrum}
+\caption{Three example echelle orders at a randomly chosen epoch from the \HARPS observations of 51 Pegasi.
+Data are shown as dark circles, while the \wobble predictions for the stellar and telluric spectral contributions at this epoch are red and blue lines, respectively.
+Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered.
+Residuals after subtracting both star and telluric models are shown below each spectrum.}
+\label{fig:51peg_spectrum}
+\end{figure}
+
 \subsection{\Mdwarf}
+
+Next, we tested \wobble with 306 epochs of \Mdwarf spectra.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell359445 and mbedell359501.} 
+\todo{(say brief things about data quality)}
 
@@
 Its trajectory translates to a projected secular change in \RV of approximately 4.5 $\mathrm{m}~\mathrm{s}^{-1}~\mathrm{yr}^{-1}$ \citep{Kurster2003}. 
-This trend has been observed in some data sets, including \acronym{UVES} observations and the \HARPS spectra as analyzed by the standard instrument pipeline, but is not seen in others such as data from \acronym{HIRES} and the same \HARPS spectra as analyzed by the \HARPS-TERRA pipeline \citep{Kurster2003, Bonfils2013, Choi2013, AngladaEscude2012}.
-With or without a secular trend, no planets have been found around \Mdwarf, despite considerable amounts of \RV monitoring.
-
-Testing \wobble on the \Mdwarf data allows us to evaluate its performance in a significantly different regime: the mid-M dwarf spectra are far denser in spectral features than a Sun-like star, and individual observations are at a much lower \SNR than the 51 Peg observations used above, although the total number of spectra is greater.
-
-Spectral fit in Figure \ref{fig:barnards_spectrum}.
-
-Comparison to spectral model prediction in Figure \ref{fig:barnards_model}: even in the blue where individual spectra are hugely noisy, we are seeing real features in the composite spectral template.
-This is a huge improvement over most existing M dwarf data and may be highly valuable for spectral characterization.
-
-RVs in Figure \ref{fig:barnards_rvs}.
+This trend has been observed in some data sets, including long-term \HARPS and \acronym{HIRES} observations, although \acronym{UVES} \RVs were inconsistent with the predicted linear slope \citep{Kurster2003, Bonfils2013, Choi2013, Montet2014}.
+No planets have been found around \Mdwarf to date despite considerable amounts of \RV monitoring and little stellar activity has been observed, making it a good quiet star test case for \wobble.
 
@@
 \centering
-\includegraphics[width=5in]{barnards_multispectrum}
+\includegraphics[width=5in]{barnards_rvs}
+\caption{Radial velocity measurements for \Mdwarf from \wobble (black) and from the standard \HARPS pipeline (red).
+The median \RV has been subtracted from each data set.
+The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid blue line (top panel).
+Residuals away from the predicted trend are plotted in the lower panel.}
+\label{fig:barnards_rvs}
+\end{figure}
+
+The \RVs found by \wobble are in excellent agreement with the predicted secular motion (Figure \ref{fig:barnards_rvs}). 
+Using \gaia parallax and proper motion measurements and following the calculations outlined in \citet{Kurster2003}, we find a secular trend with a slope of 4.53 $\mathrm{m}~\mathrm{s}^{-1}~\mathrm{yr}^{-1}$, deviating from this linearity by less than 1 $\mathrm{cm}~\mathrm{s}^{-1}~\mathrm{yr}^{-2}$ during the decade of \HARPS observations \citep{gaia2016, gaia2018}.
+
+After subtracting the \RVs as predicted by the equations of motion at each epoch, the residuals have low dispersion, as expected for a quiet star with no known planets. 
+The \acronym{RMS} scatter among \wobble\ \RVs is \todo{1.2} \ms. 
+This compares favorably to the scatter of \todo{1.3} \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates.
+This implies that the deviations from average among \Mdwarf\ \RVs are a physical effect, rather than being a white-noise-driven scatter. 
+A complete analysis of the source of these \RV deviations -- whether it be planets, stellar variability, or imperfect instrumental calibration -- is beyond the scope of this work, but in Section \ref{s:future} we will return to address possibilities for the treatment of stellar activity with \wobble.
+
+Testing \wobble on the \Mdwarf data allows us to evaluate its performance in a significantly different regime: the mid-M dwarf spectra are far denser in spectral features than a Sun-like star, and individual observations are at a much lower \SNR than the 51 Peg observations used above, although the total number of spectra is greater.
+The resulting spectral fit is shown in Figure \ref{fig:barnards_spectrum}.
+
+One consequence of working in the low-\SNR regime is that the power to resolve very small telluric features is reduced. 
+To keep the template spectra from overfitting the noise, the regularization amplitudes on the templates must be raised by several orders of magnitude relative to what was optimal for the 51 Peg fit above, particularly in orders where no strong spectral features are available.
+This has the side effect of flattening out weak features if no strong lines are present in a given order. 
+As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the top panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum} \todo{-- To do: check if there's a better example order}).
+In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}). 
+Nevertheless, we caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum at lower signal \todo{(\SNR $\lessapprox 100$ pixel$^{-1}$)}.
+
+Despite the extreme noise at the bluer end of the wavelength range, where \Mdwarf is very faint, \wobble successfully retrieves a template spectrum that appears consistent with general expectations for a mid-M dwarf through most of the \HARPS wavelength range.
+As a sanity check, we compare the \wobble stellar template model with a high-resolution \PHOENIX model at \Mdwarf's previously measured spectral parameters \citep[\teff = 3200 K, \logg = 5.0, and \mh = $-0.5$;][and references therein]{Husser2013, Artigau2018}.
+Even in the bluest regions, the features inferred by \wobble compare well with the \PHOENIX predictions in broad strokes (Figure \ref{fig:barnards_model}). 
+\todo{(check in about opacity binning affecting the smaller features)}
+
+These results emphasize the potential value of \wobble data products for spectral characterization and model testing, particularly in the case of faint stars and M dwarfs, for which many spectra must be combined to get a reasonably detailed composite spectrum.
+The \wobble algorithm is a simple yet robust method for doing such multi-spectra stacking while preserving telluric-contaminated regions and marginalizing over unknown \RV shifts.
+
+\begin{figure}
+\centering
+\includegraphics[width=5.5in]{barnards_multispectrum}
 \caption{Example echelle orders at a randomly chosen epoch from the \HARPS observations of Barnard's Star, plotted as in Figure \ref{fig:51peg_spectrum}.
@@
 \label{fig:barnards_model}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=5in]{barnards_rvs}
-\caption{Radial velocity measurements for \Mdwarf from \wobble (black) and from the \HARPS closed-source pipeline (red). \todo{(Say more about expected/observed secular evolution)}}
-\label{fig:barnards_rvs}
 \end{figure}
@@
 \includegraphics[width=5in]{telluric_standard}
-\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (blue) and the spectrum observed with a telluric standard star (black).}
+\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (blue) and the spectrum observed with a telluric standard star (black) around an H$_2$O absorption band.}
 \label{fig:telluric_standard}
@@
 
-\section{Generalizing \wobble}
+\section{Limitations and Future Extensions of \wobble}
 \label{s:future}
 
-Applicability of \wobble as shown here to current and near-future instruments.
+\begin{itemize}
 
-Revisit the assumptions: single star, no gas cell, known instrumental wavelength solution, no intrinsic spectral variability, no additive terms, many spectra exist. 
-In what situations are these assumptions violated, and how would we configure \wobble to deal with them?
+\item How much data is needed? Experiment with sub-sampling 51 Peg to fewer epochs. Talk more about low-SNR limitations. In some regime, wobble is just not the right tool for the job.
 
-Intrinsic stellar variability: cite \citet{Davis2017} for PCA approach
+\item What about gas cell instruments? Describe necessary modifications.
+
+\item Revisit assumption about good calibration: using tellurics to modify wavelength calibration? (there are relevant citations here) Spectral variability due to LSF changing?
+
+\item Spectral continuum!!!
+
+\item Intrinsic stellar variability: cite \citet{Davis2017} for PCA approach.
+Some absorption lines are more sensitive to the effects of stellar activity than others (Dumusque).
+If these fall into a small number of categories, a PCA-like dimensionality reduction should do well.
+Address pros and cons of introducing Gaussian processes to parameterize line profiles.
+
+\end{itemize}
+
+\section{Conclusion}
+\label{s:conclusion}
+
+Extracting precise \RVs is important and will become more critical as new instruments come online. 
+...
+
+\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Andreas Seifahrt, and Julian Sturmer for helpful discussions.
+
+This research has made use of the services of the ESO Science Archive Facility.
+
+This work has made use of data from the European Space Agency (ESA) mission
+{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
+Data Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
+has been provided by national institutions, in particular the institutions
+participating in the {\it Gaia} Multilateral Agreement.}
 

### 2018-11-01 regularization.py
more verbose

@@ -182,7 +182,7 @@ def improve_parameter(par, training_model, validation_model, regularization_dict
 def test_regularization_value(par, val, training_model, validation_model, regularization_dict, 
                               validation_data, validation_results, 
                               plot=False, verbose=True, basename='', 
-                              training_niter=100, validation_niter=1000):
+                              training_niter=200, validation_niter=1000):
     '''
     Try setting regularization parameter `par` to value `val`; return goodness metric `nll`.
     '''
@@ -297,8 +297,8 @@ def plot_pars_from_file(filename, basename, orders=np.arange(72)):
 
         
 if __name__ == "__main__":
-    starname = '51peg'
-    orders = np.arange(72)
+    starname = 'HD189733'
+    orders = np.arange(70,72)
     K_star = 0
     K_t = 3
     initialize_at_zero = False # toggle telluric template initialization
@@ -334,14 +334,14 @@ if __name__ == "__main__":
                 f.create_dataset('L2_basis_weights', data=np.ones(R)) # never tuned, just need to pass to wobble
 
     # set up training & validation data sets:
-    data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders) # to get N_epochs    
+    data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders, min_snr=3) # to get N_epochs    
     validation_epochs = np.random.choice(data.N, data.N//8, replace=False) # 12.5% of epochs will be validation set
     training_epochs = np.delete(np.arange(data.N), validation_epochs)
     training_data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders, 
-                        epochs=training_epochs)
+                        epochs=training_epochs, min_snr=3)
     training_results = wobble.Results(training_data)
     validation_data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=training_data.orders, 
-                          epochs=validation_epochs)
+                          epochs=validation_epochs, min_snr=1) # HACK
     validation_results = wobble.Results(validation_data)
     assert len(training_data.orders) == len(validation_data.orders), "Number of orders used is not the same between training and validation data."
     orders = training_data.orders

### 2018-11-01 test.py

@@ -7,12 +7,12 @@ import h5py
 import os
 
 if __name__ == "__main__":
-    starname = '51peg'
+    starname = 'HD189733'
     K_star = 0
-    K_t = 0    
+    K_t = 3    
     niter = 150 # for optimization
     plots = True
-    epochs = [0, 50] # to plot
+    epochs = [0, 80] # to plot
     movies = False
     
     star_reg_file = 'wobble/regularization/{0}_star_K{1}.hdf5'.format(starname, K_star)
@@ -127,6 +127,18 @@ if __name__ == "__main__":
         fig.tight_layout()
         fig.subplots_adjust(hspace=0.05)
         plt.savefig(plot_dir+'results_rvs.png')
+        plt.close(fig)
+        
+        fig, (ax, ax2) = plt.subplots(2, 1, gridspec_kw = {'height_ratios':[3, 1]})
+        ax.scatter(data.dates % 2.21857312, data.pipeline_rvs + data.bervs, c='r', label='DRS', alpha=0.7)
+        ax.scatter(data.dates % 2.21857312, results.star_time_rvs + data.bervs, c='k', label='wobble', alpha=0.7)
+        ax.legend()
+        ax.set_xticklabels([])
+        ax2.scatter(data.dates % 2.21857312, results.star_time_rvs - data.pipeline_rvs, c='k')
+        ax2.set_ylabel('Phase-folded Date')
+        fig.tight_layout()
+        fig.subplots_adjust(hspace=0.05)
+        plt.savefig(plot_dir+'results_rvs_phased.png')
         plt.close(fig)
         
     print("total runtime:{0:.2f} minutes".format((time() - start_time)/60.0))

### 2018-11-08 paper/paper.tex
writing progress

@@
 \include{preamble}
-\newcommand{\Mdwarf}{Barnard's Star\xspace}
+\newcommand{\Mdwarf}{Barnard's Star\xspace} % or should it be Barnard's star? or GJ 699?
 
@@
 
+\author{Rodrigo Luger}
+\affiliation{\flatiron}
+
 \correspondingauthor{Megan Bedell}
@@
 % Results
-We recover the canonical exoplanet 51 Pegasi b at \todo{improved accuracy?} relative to the standard \HARPS pipeline, and we achieve a precision of \todo{xx} \cms on the \RV standard M dwarf \Mdwarf. 
+We recover the canonical exoplanet 51 Pegasi b at \todo{improved accuracy?} relative to the standard \HARPS pipeline, and we achieve an \RV stability of \todo{1.2} \ms on the M dwarf \Mdwarf. 
 We also present detailed time-variable telluric spectra derived from these data. 
@@
 where the regularization amplitude for the basis weights is arbitrarily set to 1.0 and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
-\todo{(todo: explain why regularization on basis weights is needed)}
+\todo{(HOGG: explain why regularization on basis weights is needed)}
 
@@
 Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to maximize the likelihood function again. 
-\todo{(is this convex too?)}
+\todo{(HOGG/DFM: is this convex too?)}
 We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
 
-The inverse variance uncertainty on the parameters is calculated as the Hessian, or the second derivative of the likelihood with respect to each parameter. \todo{(check phrasing with DFM)}
+The inverse variance uncertainty on the parameters is calculated as the Hessian, or the second derivative of the likelihood with respect to each parameter. \todo{(DFM: help me rephrase if needed?)}
 
@@
 \subsection{\Mdwarf}
+\label{s:Mdwarf}
 
@@
 In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}). 
-Nevertheless, we caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum at lower signal \todo{(\SNR $\lessapprox 100$ pixel$^{-1}$)}.
+However, telluric variability was not able to be resolved in the overwhelming majority of orders \todo{(check on this and whether RVs are affected by K setting)}. 
+We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations \todo{(\SNR $\lessapprox 100$ pixel$^{-1}$)}.
 
@@
 
-\subsection{Telluric Features}
+\subsection{HD 189733}
 
-Comparison between 51 Peg and \Mdwarf telluric fits in Figure \ref{fig:telluric_comparison}.
-No constraints used to make them look the same, but they do!
-In principle all stars could have a shared telluric model and this would give the Best Telluric Spectrum Ever (TM).
+While the majority of \EPRV measurements are made for the purposes of observing stellar reflex motion due to planetary orbits, several other applications to exoplanet characterization exist for such high-precision time-series spectra.
+One of these uses is measuring the \RM, in which the apparent stellar \RV is observed during planetary transit as a way of mapping the stellar surface and learning the spin-orbit inclination of the system \citep[e.g.][]{Queloz2000, Winn2005}.
+While the \RM technically manifests in the stellar spectrum as a distortion to the line profile rather than as a true Doppler shift, it is typically detected using the same methods as a standard \RV analysis.
 
-Comparison to telluric standard star in Figure \ref{fig:telluric_standard}. Higher SNR with no observational overhead!
+We apply \wobble to a data set consisting of a single transit of the hot Jupiter HD 189733b to test its sensitivity to stellar line asymmetries as well as its performance in the regime of a single night's observations.
+The observations in question consist of a consecutive series of \todo{x} spectra, each with an \SNR $\sim 90$, taken on the night of \todo{August 28, 2007}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers \todo{x}.}
 
-Sample of telluric variability in Figure \ref{fig:telluric_basis}. To do: figure out what these lines are, do the ones that vary together share a common species?
+\todo{(discuss fixing the telluric spectrum)}
 
-To do: show that basis weights correlate with things like season or humidity.
+The resulting \RV signal retrieved by \wobble is shown in Figure \todo{x}. 
+We fit the signal using \starry \citep{Luger2018}. \todo{(Rodrigo, write this please!)} 
+\todo{(comparison of retrieved parameters to previous work here)}
+
+These results confirm that the \wobble method measures line asymmetries as \RV shifts in the same manner as traditional \RV analysis techniques. 
+While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{}. 
+The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified in a relatively straightforward manner to fit line asymmetries and true Doppler shifts separately.
+We discuss this prospect further in Section \ref{s:future}.
+
+\section{Telluric Features}
+
+In the above analyses, telluric spectra were inferred independently for the 51 Peg and \Mdwarf data sets. 
+However, in principle the telluric features should be common to all \HARPS spectra.
+
+We show a comparison between 51 Peg and \Mdwarf telluric fits in Figure \ref{fig:telluric_comparison}. 
+Although they were fit using entirely different data sets with dissimilar stellar spectra, the resulting telluric fits are extremely similar. 
+This is generally true for all wavelength regions in which tellurics are detected. 
+The only regime in which the comparison fails is for low-\SNR spectral orders of the \Mdwarf data, where the telluric spectra are featureless due to the strong regularization, as discussed in Section \ref{s:Mdwarf}.
+
+The physically motivated expectation that all observations should share a common telluric component could be built into the \wobble model. 
+This assumption is used for the fit to HD 189733, where we fixed the telluric spectrum to use the model inferred from 51 Peg. 
+A better, albeit more computationally expensive, approach would be to fit multiple stars simultaneously so that their shared telluric spectrum can be inferred using all the available data. 
+Given enough spectra, this approach should yield an incredibly detailed model of telluric features and their time variability. 
+We leave such an effort to future work.
+
+As another sanity check on our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{(details)}. 
+As shown in Figure \ref{fig:telluric_standard}, the spectra agree well. 
+Moreover, the composite telluric spectrum inferred from the time series of 51 Peg observations is at a much higher \SNR than the single shot telluric standard spectrum, so that low-amplitude lines which border on statistical insignificance in the standard star show up clearly in the \wobble results.
+Unlike a traditional telluric standard observation, no overhead time is required to produce these results.
+
+Of course, one strength of a telluric standard star is that its spectrum can capture the true telluric spectrum at any given moment in time, while the telluric template determined by \wobble is time-averaged and may not perfectly capture the telluric absorption lines in any given observation.
+For this reason, we included time-variable components in the model. 
+A physical interpretation of these components should be possible.
+
+An example of the inferred variability from the 51 Peg analysis is shown in Figure \ref{fig:telluric_basis}. 
+There is clearly some coherent structure in the variations. 
+\todo{(TO DO: PCA on time series of standard star observations??)}
+
+\todo{(maybe show a plot of basis weights vs. time?)}
 
@@
 \caption{Fits to a 15-\ang region with substantial telluric contamination for 51 Peg (left) and \Mdwarf (right).
-Data and \wobble model fits are shown in the upper panels, while residuals to the fits are shown in the lower panels.
+The upper panels show the \wobble best-fit telluric spectrum (blue line), best-fit stellar spectrum (red line), data (black points), and the sum of the telluric and stellar model predictions (black line).
+Residuals are shown in the lower panels.
 Both observations shown here were taken at an airmass of $\sim1.8$.
@@
 \includegraphics[width=5in]{telluric_standard}
-\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (blue) and the spectrum observed with a telluric standard star (black) around an H$_2$O absorption band.}
+\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (blue) and the spectrum observed with a telluric standard star (black) around an H$_2$O absorption band.
+\todo{TO DO: fix up continuum normalization}}
 \label{fig:telluric_standard}
@@
 
-\section{Limitations and Future Extensions of \wobble}
+\section{Limitations and Extensibility of \wobble}
 \label{s:future}
@@
 
-\item How much data is needed? Experiment with sub-sampling 51 Peg to fewer epochs. Talk more about low-SNR limitations. In some regime, wobble is just not the right tool for the job.
+\item How much data is needed?
+Experiment with sub-sampling 51 Peg to fewer epochs.
+\wobble does have many free parameters and therefore it's not going to work well for very few epochs.
+Also telluric sampling is key.
+in some regime \wobble is just not the right tool for the job.
 
-\item What about gas cell instruments? Describe necessary modifications.
+\item Talk more about low-SNR limitations: what we did is a hack and it should be possible to do much better with better continuum normalization.
 
-\item Revisit assumption about good calibration: using tellurics to modify wavelength calibration? (there are relevant citations here) Spectral variability due to LSF changing?
+\item Regularization framework could be improved: we chose it because it is convex and represents our beliefs but it's the most difficult and unwieldy part of \wobble.
+
+\item Spectral representation: ours sucks.
+Could go to higher-order interpolation scheme like cubic split OR smooth non-parametric rep like a GP OR something that incorporates more physics.
+Nothing about wobble depends on the spectral representation and it should be simple to swap out for something better provided it can be implemented in \TF.
+
+\item We treat the orders separately: we mostly did this to break up the computation, but this could be changed straightforwardly.
+Doing so would take on additional computational cost, but we are in \TF so it shouldn't hurt too much.
+
+\item What about gas cell instruments? Describe necessary modifications. Of course, gas cells exist because calibration isn't perfect...
+
+\item Revisit assumption about good calibration: using tellurics to modify wavelength calibration?
+(there are relevant citations here) Spectral variability due to LSF changing?
+Additional parameters to the P operators, non-convex optimization
 
 \item Spectral continuum!!!
+All methods are questionable at low SNR and in the absence of true continuum.
+Right solution is to simultaneously fit continuum with other model components, but this introduces degeneracies and must be handled carefully.
 
@@
 Address pros and cons of introducing Gaussian processes to parameterize line profiles.
+
+\item Maybe regularize (or put priors on) the \RVs themselves and force them to sum of Keplerians for the purpose of breaking degeneracies with stellar variability.
+
+\item 1D extracted spectra.
+Our P operators are interpolators onto 1D data but they could be 2D interpolators!
+Little about the model equations would change, but implementation would become much trickier.
 
@@
 Extracting precise \RVs is important and will become more critical as new instruments come online.
+
+Fundamental data-driven technique - requires no physics knowledge which makes it a great tool for expanding \RV searches.  
+Open-source, can be adapted for use with any current or future \EPRV instrument
 ...
 
-\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Andreas Seifahrt, and Julian Sturmer for helpful discussions.
+Here's what this enables.
+
+\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions.
 

### 2018-11-13 paper/paper.tex
more intro

@@
 % Context
-Telluric absorption features in stellar spectra provide particular challenges to extreme-precision radial velocity (\EPRV) exoplanet surveys. 
-Large amplitude features are often masked in data analysis pipelines, reducing the number of observable spectral features which can be used to measure an \RV shift.
-Low-amplitude features that are hard to see even in high signal-to-noise spectra may be an important contributor to the overall \RV noise budget.
+Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally informative \RVs from spectra. 
+Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features, which cannot be perfectly modeled. 
+Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
+In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself. 
+%In both cases, the full data set of all observed spectra for a given target generally contains more complete information about the underlying 
+%In both cases, using the full data set of all observed spectra for a given target will yield more precise constraints on the underlying spectra and their \RVs than a traditional analysis.
+%\todo{(reword this to be less controversial?)}
+%Telluric absorption features in stellar spectra provide particular challenges to extreme-precision radial velocity (\EPRV) exoplanet surveys. 
+%Large amplitude features are often masked in data analysis pipelines, reducing the number of observable spectral features which can be used to measure an \RV shift.
+%Low-amplitude features that are hard to see even in high signal-to-noise spectra may be an important contributor to the overall \RV noise budget.
 % Aims
@@
 % Results
-We recover the canonical exoplanet 51 Pegasi b at \todo{improved accuracy?} relative to the standard \HARPS pipeline, and we achieve an \RV stability of \todo{1.2} \ms on the M dwarf \Mdwarf. 
-We also present detailed time-variable telluric spectra derived from these data. 
-This method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.
+We recover the canonical exoplanet 51 Pegasi b, detect the secular \RV evolution of the M dwarf \Mdwarf, and retrieve the \RM for the Hot Jupiter HD 189733b. 
+We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data. 
+%This method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.
 \end{abstract}
@@
 
-Such an approach has two significant issues. 
+This method has two significant issues. 
 The first is that removing sections of the spectrum can remove significant amounts of information about the star, lowering the precision at which we can measure the stellar radial velocity. 
-Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV precision (cite). 
+Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV precision. 
 This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres \citep{Figueira2016}. 
@@
 As the locations of these features are not known \textit{a priori} and may not even be apparent in stacked spectra of many observations, these spectral regions cannot be thrown out. 
-Instead, alternative methods to account for these features must be developed and employed in order to mitigate the effect of the Earth's atmosphere on the measured stellar radial velocities.
+Instead, alternative methods to account for these features as a part of the model must be developed and employed in order to mitigate the effect of the Earth's atmosphere on the measured stellar radial velocities.
 
@@
 In these cases, both spectra are assumed to be unchanging in time, which is a reasonable approximation of a stellar spectrum but not necessarily of the telluric spectrum.
-A more complicated linear model may be useful for separating the stellar and telluric spectra, enabling a reconstruction of both at each epoch.
-\todo{(say something about Artigau2014 here)}
+A more complicated model with time variability in the telluric spectrum may provide a more accurate fit. 
+Work by \citet{Artigau2014} demonstrates that a Principle Component Analysis (\acronym{PCA}) approach is an effective way of parameterizing telluric spectral variability as a low-dimensional model derived from observational data.
+\citet{Artigau2014} use a library of telluric standard observations, which requires a significant investment of observing time to build up, but in principle such a data-driven model should be possible to derive from typical stellar observations provided enough epochs of data are available.
 
-\todo{[Other works that have done similar stuff -- be sure to cite: Artigau2018, Gao2016]}
+Just as imperfect telluric modeling can be a noise source in \EPRV analyses, the choice of stellar template can also be a major source of error. 
+For stabilized, non-gas-cell \RV spectrographs, a standard approach has been to adopt a quasi-binary mask consisting of weighted top-hat functions at the expected locations of informative stellar absorption lines and cross-correlate this mask with the observed spectrum \citep[e.g.][]{Baranne1979, Pepe2002}.
+This approach is limited by the accuracy of the mask, and since most masks are built for a broad category of spectral type rather than customized for the individual star in question, it is unlikely that this techniques retrieve maximally precise \RVs.
+Deriving a custom spectral template by stacking all spectra iteratively as the \RVs are determined was shown to be a superior approach for stars with complex spectra by \citet{AngladaEscude2012}.
 
-Here we develop a linear data-driven model to infer a telluric and stellar spectrum and calculate the stellar RV at each observed epoch. 
+Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is a single observation taken for the purpose \citep[e.g.][]{Butler1996}.
+Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
+\todo{(explain how this differs from what we'll do)}
+
+\todo{(talk about lack of open-source code and difficulty of modifying \RV pipelines here?)}
+
+Here we develop a linear data-driven model to infer the telluric and stellar spectra and calculate the stellar RV at each observed epoch. 
 The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
@@
 
-In this work, we focus on the ultra-stabilized spectrograph case, i.e. no absorption cell. 
+In this work, we focus on the ultra-stabilized spectrograph case, i.e. a reliable instrumental calibration and no absorption cell. 
 We also assume that multiple epochs of observations exist and that these epochs are spread out across the observing season(s). 
 This assumption is necessary to enable the disentangling of telluric features from the stellar spectrum. 
-In this sense our pipeline is intended as a post-processing step, not a real-time data reduction service.
+In this sense our method is intended as a post-processing step, not a real-time data reduction service. 
+However, the implementation that we present here is designed for flexibility and easy extensibility, and we discuss potential ways to overcome these limitations.
 
-In Section \ref{s:methods}, we outline the model and present an open-source implementation in \python and \TF called \wobble. 
-In Section \ref{s:results}, we apply our method to \HARPS archival data for two target stars, the planet-hosting solar analog 51 Peg and the quiet M dwarf \Mdwarf, as a demonstration of \wobble's capabilities.
-\todo{We find xyz.} 
-We conclude in Section \ref{s:future} with a detailed look at the limitations of the current implementation and outline ways to adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and multiple-star systems.
+In Section \ref{s:methods}, we outline the model and its key underlying assumptions. 
+We present an open-source implementation of this method in \python and \TF called \wobble. 
+In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
+We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation.
+We outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data.
+Finally, we conclude with a brief summary in Section \ref{s:conclusion}.
 
@@
 
-%For practical applications, the number of epochs $N$ must be $\gg 1$. 
-%If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.
+For practical applications, the number of epochs $N$ must be large and span a significant fraction of the observing season. 
+If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.
 a single night), it will be nearly impossible to disentangle the components of the model from each other, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
-%We will revisit this restriction in Section \ref{s:future}.
+We will revisit this restriction in Section \ref{s:future}.
 
@@
 The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors.
-%\todo{(math here?)}
 
@@
 Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to maximize the likelihood function again. 
-\todo{(HOGG/DFM: is this convex too?)}
+\todo{(HOGG: how could this be made convex too?)}
 We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
@@
 The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF. 
-\TF is a \todo{[blah blah technical blah]}. 
+\TF is a \todo{DFM: [blah blah technical blah]}. 
 While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
@@
 
-\todo{(more details of implementation)}
-
-Our code is open-source and publicly available on GitHub.\footnote{\url{https://www.github.com/megbedell/wobble}}
+Our code made open-source under the \acronym{MIT} license and is publicly available on GitHub.\footnote{\url{https://www.github.com/megbedell/wobble}}
 
@@
 We obtained wavelength solutions for each spectrum from the dedicated \HARPS calibration archive, also maintained by \acronym{ESO}. 
+\todo{Uncertainty estimates for each pixel were calculated as... [HOGG, help me fill this in - relation to \SNR and conversion to log]}
 
@@
 In brief, we mask pixels whose extracted flux is below zero as well as regions at the edges of the spectral orders where the local \SNR falls below 5. 
-\todo{(add some equations for variance estimation and error conversion to log?)}
 After masking these data, we convert the flux to log space and continuum normalize by fitting a polynomial to an asymmetrically clipped subset of the data.
 
-Below we describe the results of applying this procedure and optimizing the \wobble model using \HARPS data for stars of different types: a G dwarf with a known planet and an \RV-quiet M dwarf.
+\todo{The final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off a drift term specified in the data header, consistent with the standard \HARPS pipeline.}
+
+Below we describe the results of applying this procedure and optimizing the \wobble model using \HARPS data for stars of different types: a G dwarf with a known planet, an \RV-quiet M dwarf, and an early-K dwarf undergoing a planetary transit event.
 
@@
 We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell329968 and mbedell331551.}
-The 91 publicly available spectra in the \HARPS archive are largely concentrated on a few nights of intensive observing, but these nights are sufficiently spread out throughout the year to give a wide enough range in \BERV to disentangle the stellar and telluric spectra.
+The 91 publicly available spectra in the \HARPS archive are largely concentrated on a few nights of intensive observing, but these nights are sufficiently spread out throughout the year for a wide enough range in \BERV to disentangle the stellar and telluric spectra.
 
@@
 Next, we tested \wobble with 306 epochs of \Mdwarf spectra.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell359445 and mbedell359501.} 
-\todo{(say brief things about data quality)}
+\todo{(say brief things about data quality - it's low \SNR)}
 
@@
 The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid blue line (top panel).
-Residuals away from the predicted trend are plotted in the lower panel.}
+Residuals away from the predicted trend are plotted in the lower panel.
+\todo{(TO DO: add error bars)}}
 \label{fig:barnards_rvs}
@@
 After subtracting the \RVs as predicted by the equations of motion at each epoch, the residuals have low dispersion, as expected for a quiet star with no known planets. 
-The \acronym{RMS} scatter among \wobble\ \RVs is \todo{1.2} \ms. 
-This compares favorably to the scatter of \todo{1.3} \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates.
-This implies that the deviations from average among \Mdwarf\ \RVs are a physical effect, rather than being a white-noise-driven scatter. 
-A complete analysis of the source of these \RV deviations -- whether it be planets, stellar variability, or imperfect instrumental calibration -- is beyond the scope of this work, but in Section \ref{s:future} we will return to address possibilities for the treatment of stellar activity with \wobble.
+The \acronym{RMS} scatter among \wobble\ \RVs is 2.0 \ms. 
+This compares favorably to the scatter of 2.5 \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates.
+It is possible that these deviations are a real physical effect in the stellar spectrum, but we caution that approximately \ms-level errors are likely being introduced by the barycentric correction. 
+Currently we assume that the \BERV provided by the native \HARPS pipeline is correct, but this will not hold true if the sky coordinates entered by the observer deviate from the actual on-sky location of the target.
+This is especially likely to be the case for high proper motion targets like \Mdwarf. 
+Indeed, we do note a strong peak at year-long periods in a Lomb-Scargle periodogram of the residual \RVs. 
+Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble. \todo{(TO DO: recalculate \BERV ourselves)}
 
@@
 The resulting spectral fit is shown in Figure \ref{fig:barnards_spectrum}.
-
-One consequence of working in the low-\SNR regime is that the power to resolve very small telluric features is reduced. 
-To keep the template spectra from overfitting the noise, the regularization amplitudes on the templates must be raised by several orders of magnitude relative to what was optimal for the 51 Peg fit above, particularly in orders where no strong spectral features are available.
-This has the side effect of flattening out weak features if no strong lines are present in a given order. 
-As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the top panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum} \todo{-- To do: check if there's a better example order}).
-In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}). 
-However, telluric variability was not able to be resolved in the overwhelming majority of orders \todo{(check on this and whether RVs are affected by K setting)}. 
-We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations \todo{(\SNR $\lessapprox 100$ pixel$^{-1}$)}.
-
-Despite the extreme noise at the bluer end of the wavelength range, where \Mdwarf is very faint, \wobble successfully retrieves a template spectrum that appears consistent with general expectations for a mid-M dwarf through most of the \HARPS wavelength range.
-As a sanity check, we compare the \wobble stellar template model with a high-resolution \PHOENIX model at \Mdwarf's previously measured spectral parameters \citep[\teff = 3200 K, \logg = 5.0, and \mh = $-0.5$;][and references therein]{Husser2013, Artigau2018}.
-Even in the bluest regions, the features inferred by \wobble compare well with the \PHOENIX predictions in broad strokes (Figure \ref{fig:barnards_model}). 
-\todo{(check in about opacity binning affecting the smaller features)}
-
-These results emphasize the potential value of \wobble data products for spectral characterization and model testing, particularly in the case of faint stars and M dwarfs, for which many spectra must be combined to get a reasonably detailed composite spectrum.
-The \wobble algorithm is a simple yet robust method for doing such multi-spectra stacking while preserving telluric-contaminated regions and marginalizing over unknown \RV shifts.
 
@@
 
+One consequence of working in the low-\SNR regime is that the power to resolve very small telluric features is reduced. 
+To keep the template spectra from overfitting the noise, the regularization amplitudes on the templates must be raised by several orders of magnitude relative to what was optimal for the 51 Peg fit above, particularly in orders where no strong spectral features are available.
+This has the side effect of flattening out weak features if no strong lines are present in a given order. 
+As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the top panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum} \todo{-- To do: check if there's a better example order}).
+In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}). 
+However, telluric variability was not able to be resolved in the overwhelming majority of orders \todo{(check on this and whether RVs are affected by K setting)}. 
+We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations \todo{(\SNR $\lessapprox 100$ pixel$^{-1}$)}.
+
+Despite the extreme noise at the bluer end of the wavelength range, where \Mdwarf is very faint, \wobble successfully retrieves a template spectrum that appears consistent with general expectations for a mid-M dwarf through most of the \HARPS wavelength range.
+As a sanity check, we compare the \wobble stellar template model with a high-resolution \PHOENIX model at \Mdwarf's previously measured spectral parameters \citep[\teff = 3200 K, \logg = 5.0, and \mh = $-0.5$;][and references therein]{Husser2013, Artigau2018}.
+Even in the bluest regions, the features inferred by \wobble compare well with the \PHOENIX predictions in broad strokes (Figure \ref{fig:barnards_model}). 
+\todo{(check in about opacity binning affecting the smaller features)}
+
+These results emphasize the potential value of \wobble data products for spectral characterization and model testing, particularly in the case of faint stars and M dwarfs, for which many spectra must be combined to get a reasonably detailed composite spectrum.
+The \wobble algorithm is a simple yet robust method for doing such multi-spectra stacking while preserving telluric-contaminated regions and marginalizing over unknown \RV shifts.
+
 \subsection{HD 189733}
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{hd189_rm}
+\caption{\RVs measured by \wobble for a single night of observations of the Hot Jupiter host HD 189733. The \RM is clearly seen as the planet transits the star. \todo{[blah blah starry blah]}}
+\label{fig:hd189_rm}
+\end{figure}
 
@@
 One of these uses is measuring the \RM, in which the apparent stellar \RV is observed during planetary transit as a way of mapping the stellar surface and learning the spin-orbit inclination of the system \citep[e.g.][]{Queloz2000, Winn2005}.
-While the \RM technically manifests in the stellar spectrum as a distortion to the line profile rather than as a true Doppler shift, it is typically detected using the same methods as a standard \RV analysis.
+While the \RM technically manifests in the stellar spectrum as a distortion to the line profile rather than a true Doppler shift, it is typically detected using similar methods to a standard \RV analysis.
 
 We apply \wobble to a data set consisting of a single transit of the hot Jupiter HD 189733b to test its sensitivity to stellar line asymmetries as well as its performance in the regime of a single night's observations.
-The observations in question consist of a consecutive series of \todo{x} spectra, each with an \SNR $\sim 90$, taken on the night of \todo{August 28, 2007}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers \todo{x}.}
+The observations in question consist of a consecutive series of 40 spectra, each with an \SNR $\sim 90$, taken on the night of August 28, 2007.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers \todo{x}.}
 
-\todo{(discuss fixing the telluric spectrum)}
+Because all observations came from a single night, there will be no significant shift of the stellar spectrum with respect to the telluric spectrum and the power of \wobble to disentangle the two is severely limited.
+For this reason, we fixed the telluric spectrum to a constant template. 
+We derived this template by running \wobble on the 51 Peg data with non-time-variable tellurics and adopting the resulting high-quality time-invariant telluric spectrum. 
+The stellar spectrum of HD 189733 and its \RVs were left as free parameters.
 
-The resulting \RV signal retrieved by \wobble is shown in Figure \todo{x}. 
-We fit the signal using \starry \citep{Luger2018}. \todo{(Rodrigo, write this please!)} 
+The resulting \RV signal retrieved by \wobble is shown in Figure \ref{fig:hd189_rm}. 
+We fit the signal using \starry \citep{Luger2018}. \todo{(TO DO: fill in with Rodrigo)} 
 \todo{(comparison of retrieved parameters to previous work here)}
@@
 The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified in a relatively straightforward manner to fit line asymmetries and true Doppler shifts separately.
+This would help to disentangle signals caused by photospheric features like starspots from the \RV signal due to reflex motion in a planetary system. 
 We discuss this prospect further in Section \ref{s:future}.
@@
 \section{Telluric Features}
+\label{s:tellurics}
 
 In the above analyses, telluric spectra were inferred independently for the 51 Peg and \Mdwarf data sets. 
-However, in principle the telluric features should be common to all \HARPS spectra.
+However, in principle the telluric features should be common to all \HARPS spectra, an assumption that we used for the case of HD 189733.
 
 We show a comparison between 51 Peg and \Mdwarf telluric fits in Figure \ref{fig:telluric_comparison}. 
-Although they were fit using entirely different data sets with dissimilar stellar spectra, the resulting telluric fits are extremely similar. 
+Although they were fit using entirely different data sets with dissimilar stellar spectra, the resulting telluric fits are indeed extremely similar. 
 This is generally true for all wavelength regions in which tellurics are detected. 
@@
 The physically motivated expectation that all observations should share a common telluric component could be built into the \wobble model. 
-This assumption is used for the fit to HD 189733, where we fixed the telluric spectrum to use the model inferred from 51 Peg. 
-A better, albeit more computationally expensive, approach would be to fit multiple stars simultaneously so that their shared telluric spectrum can be inferred using all the available data. 
+For the fit to HD 189733, for example, we fixed the telluric spectrum to use the model inferred from the better-studied star 51 Peg. 
+A more robust, albeit more computationally expensive, approach would be to fit many stars simultaneously so that their shared telluric spectrum can be inferred using all the available data. 
 Given enough spectra, this approach should yield an incredibly detailed model of telluric features and their time variability. 
@@
 
-As another sanity check on our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{(details)}. 
+As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{HR 3090 (+details)}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:385782 and }
 As shown in Figure \ref{fig:telluric_standard}, the spectra agree well. 
@@
 An example of the inferred variability from the 51 Peg analysis is shown in Figure \ref{fig:telluric_basis}. 
-There is clearly some coherent structure in the variations. 
-\todo{(TO DO: PCA on time series of standard star observations??)}
-
-\todo{(maybe show a plot of basis weights vs. time?)}
+There is clearly coherent structure in the variations. 
+As we might expect, many of the lines co-vary with each other: as all known telluric lines in this region arise from H$_2$O, this likely corresponds to time-variable levels of atmospheric water vapor content.
+In general, the basis vectors are suggestive of Doppler shifts, perhaps due to wind speeds in different layers of the atmosphere contributing individually to the overall line shapes. 
+These hypotheses could be tested in the future by seeking correlations between basis weights and externally-measured weather diagnostics. 
+\todo{(is this enough? also show that basis weights change seasonally-ish?)}
 
@@
 Residuals are shown in the lower panels.
-Both observations shown here were taken at an airmass of $\sim1.8$.
-Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, suggesting that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
+Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
 \label{fig:telluric_comparison}
@@
 \includegraphics[width=5in]{telluric_standard}
-\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (blue) and the spectrum observed with a telluric standard star (black) around an H$_2$O absorption band.
-\todo{TO DO: fix up continuum normalization}}
+\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (upper panel) and the spectrum observed with a telluric standard star (lower panel) around an H$_2$O absorption band.
+The \wobble method delivers a very high \SNR telluric spectrum with no overhead observing time required.}
 \label{fig:telluric_standard}
@@
 
-\begin{itemize}
+In this work we have tested \wobble exclusively using \HARPS data, but vast amounts of data from other \EPRV instruments exist and may benefit from the techniques used here. 
+We now turn our discussion to ways in which other data sets may violate the assumptions made in this work and the corresponding modifications to \wobble that would enable it to operate more effectively in these regimes.
+For simplicity, we break this discussion down into three categories: inherent limitations of \wobble; assumptions that could be eliminated; and general changes that may improve the \RV precision achieved by \wobble.
 
-\item How much data is needed?
-Experiment with sub-sampling 51 Peg to fewer epochs.
+\subsection{Inherent Limitations}
+
+limitations on number of spectra: Experiment with sub-sampling 51 Peg to fewer epochs.
 \wobble does have many free parameters and therefore it's not going to work well for very few epochs.
@@
 
-\item Talk more about low-SNR limitations: what we did is a hack and it should be possible to do much better with better continuum normalization.
+low-SNR limitations: what we did is a hack and it should be possible to do much better with better continuum normalization. but still you'll eventually hit a failure mode.
+
+spectrograph resolution: will this work if lines are undersampled?
+
+\subsection{Revisiting Assumptions}
+
+single star: this actually isn't a true assumption, it's trivial to add an extra star, or even a planetary component, to the model.
+
+no gas cell: should be as simple as adding an extra component (Of course, gas cells exist because calibration isn't perfect...)
+
+reliable instrument calibration: use tellurics to modify wavelength calibration?
+(there are relevant citations here) Spectral variability due to LSF changing?
+Additional parameters to the P operators, non-convex optimization
+
+low-dimensional telluric variability: it's possible this isn't true in the current parameterization. what if we work with residuals away from a physical model of the tellurics? or reparameterize lines
+
+constant stellar spectrum: cite \citet{Davis2017} for PCA approach.
+Some absorption lines are more sensitive to the effects of stellar activity than others (Dumusque).
+If these fall into a small number of categories, a PCA-like dimensionality reduction should do well.
+Address pros and cons of introducing Gaussian processes to parameterize line profiles.
+deal with this similarly to the LSF changing?
+also, maybe we need to regularize (or put priors on) the \RVs themselves and force them to sum of Keplerians for the purpose of breaking degeneracies with stellar variability.
+
+\subsection{General Improvements}
+
+\begin{itemize}
 
@@
 
-\item We treat the orders separately: we mostly did this to break up the computation, but this could be changed straightforwardly.
-Doing so would take on additional computational cost, but we are in \TF so it shouldn't hurt too much.
-
-\item What about gas cell instruments? Describe necessary modifications. Of course, gas cells exist because calibration isn't perfect...
-
-\item Revisit assumption about good calibration: using tellurics to modify wavelength calibration?
-(there are relevant citations here) Spectral variability due to LSF changing?
-Additional parameters to the P operators, non-convex optimization
-
 \item Spectral continuum!!!
@@
 
-\item Intrinsic stellar variability: cite \citet{Davis2017} for PCA approach.
-Some absorption lines are more sensitive to the effects of stellar activity than others (Dumusque).
-If these fall into a small number of categories, a PCA-like dimensionality reduction should do well.
-Address pros and cons of introducing Gaussian processes to parameterize line profiles.
+\item We treat the orders separately: we mostly did this to break up the computation, but this could be changed straightforwardly.
+Doing so would take on additional computational cost, but we are in \TF so it shouldn't hurt too much.
 
-\item Maybe regularize (or put priors on) the \RVs themselves and force them to sum of Keplerians for the purpose of breaking degeneracies with stellar variability.
+\item We could fit all stars simultaneously, which would give a much better telluric template
 
@@
 
-\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions.
+\todo{[TO DO: add software citations]}
+
+\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Melissa Ness, Hans-Walter Rix, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions.
 

### 2018-11-13 wobble/data.py

@@ -100,6 +100,7 @@ class Data(object):
             self.bervs = np.copy(f['bervs'])[self.epochs]
             self.drifts = np.copy(f['drifts'])[self.epochs]
             self.airms = np.copy(f['airms'])[self.epochs]
+            self.filelist = [a.decode('utf8') for a in np.copy(f['filelist'])[self.epochs]]
             self.R = len(orders) # number of orders
             self.orders = orders # indices of orders in origin_file
             self.ivars = [self.fluxes[i]**2 * self.flux_ivars[i] for i in range(self.R)] # ivars for log(fluxes)

### 2018-11-13 wobble/make_data.py

@@ -18,7 +18,7 @@ def dimensions(instrument):
         return
     return M, R
 
-def read_data_from_fits(filelist):
+def read_data_from_fits(filelist, e2ds=False):
     # input : a list of CCF filenames
     N = len(filelist)  # number of epochs    
     M, R = dimensions('HARPS')
@@ -28,15 +28,18 @@ def read_data_from_fits(filelist):
     empty = np.array([], dtype=int)
     pipeline_rvs, dates, bervs, airms, drifts = np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N)
     for n,f in enumerate(filelist):
+        print(f)
         sp = fits.open(f)
-        pipeline_rvs[n] = sp[0].header['HIERARCH ESO DRS CCF RVC'] * 1.e3 # m/s
+        if not e2ds:
+            pipeline_rvs[n] = sp[0].header['HIERARCH ESO DRS CCF RVC'] * 1.e3 # m/s
+            drifts[n] = sp[0].header['HIERARCH ESO DRS DRIFT SPE RV']  
         dates[n] = sp[0].header['HIERARCH ESO DRS BJD']        
         bervs[n] = sp[0].header['HIERARCH ESO DRS BERV'] * 1.e3 # m/s
         airms[n] = sp[0].header['HIERARCH ESO TEL AIRM START']
-        drifts[n] = sp[0].header['HIERARCH ESO DRS DRIFT SPE RV']  
         
         spec_file = str.replace(f, 'ccf_G2', 'e2ds') 
         spec_file = str.replace(spec_file, 'ccf_M2', 'e2ds') 
+        spec_file = str.replace(spec_file, 'ccf_K5', 'e2ds') 
         try:
             wave, spec = read_harps.read_spec_2d(spec_file)
         except:
@@ -95,7 +98,7 @@ def missing_wavelength_files(filelist):
     return np.unique(missing_files)
     
     
-def write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffile):
+def write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, filenames, hdffile):
     h = h5py.File(hdffile, 'w')
     dset = h.create_dataset('data', data=data)
     dset = h.create_dataset('ivars', data=ivars)
@@ -105,6 +108,8 @@ def write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, hdffi
     dset = h.create_dataset('bervs', data=bervs)
[…]

### 2018-11-19 paper/paper.tex
more discussion

@@
 % Context
-Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally informative \RVs from spectra. 
+Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra. 
 Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features, which cannot be perfectly modeled. 
@@
 % Aims
-Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a simple linear model. 
+Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a linear model (in the log flux). 
+The model employs a convex objective and convex regularization to keep the optimization of the spectral components fast. 
 We implement this method in \wobble, an open-source \python package which uses \TF in its first non-machine-learning application to astronomical data. 
@@
 A more complicated model with time variability in the telluric spectrum may provide a more accurate fit. 
-Work by \citet{Artigau2014} demonstrates that a Principle Component Analysis (\acronym{PCA}) approach is an effective way of parameterizing telluric spectral variability as a low-dimensional model derived from observational data.
-\citet{Artigau2014} use a library of telluric standard observations, which requires a significant investment of observing time to build up, but in principle such a data-driven model should be possible to derive from typical stellar observations provided enough epochs of data are available.
+Work by \citet{Artigau2014} demonstrates that a Principal Component Analysis (\acronym{PCA}) approach is an effective way of parameterizing telluric spectral variability as a low-dimensional model derived from observational data.
+\citet{Artigau2014} use a library of telluric standard observations, which requires a significant investment of observing time to build up, but in principle such a data-driven model should be possible to derive from typical stellar observations if the same star is observed many times at different barycentric \RV shifts.
 
@@
 Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
-\todo{(explain how this differs from what we'll do)}
+Such data-driven templates have the benefit of being both customized to the star in question and high-\SNR without requiring observing overhead. 
+%\todo{(explain how this differs from what we'll do - v similar in spirit although ours is a full forward model with stellar spectrum as a latent variable)}
 
-\todo{(talk about lack of open-source code and difficulty of modifying \RV pipelines here?)}
+%\todo{(talk about lack of open-source code and difficulty of modifying \RV pipelines here?)}
 
-Here we develop a linear data-driven model to infer the telluric and stellar spectra and calculate the stellar RV at each observed epoch. 
+Here we develop a data-driven model to infer the telluric and stellar spectra and calculate the stellar RV at each observed epoch. 
 The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
@@
 As such, it does not yield absolute measurements of \RVs, only highly precise relative measurements between epochs.
+
+Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
+In this sense, ours is a true forward model.
 
@@
 The model underpinning \wobble is designed to be flexible and easily extensible to a variety of situations. 
-However, a few rigid assumptions are necessary, and we outline those here.
+However, a few assumptions are made in this work to simplify the implementation, and we outline those here. 
+Many of these assumptions could be eliminated with relatively straightforward modifications to the method. 
+We revisit these in Section \ref{s:future}.
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=6.5in]{telluric_comparison}
+\caption{Fits to a 15-\ang region with substantial telluric contamination for 51 Peg (left) and \Mdwarf (right).
+The upper panels show the \wobble best-fit telluric spectrum (blue line), best-fit stellar spectrum (red line), data (black points), and the sum of the telluric and stellar model predictions (black line).
+Residuals are shown in the lower panels.
+Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
+\label{fig:telluric_comparison}
+\end{figure}
+
 In the above analyses, telluric spectra were inferred independently for the 51 Peg and \Mdwarf data sets. 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=5in]{telluric_standard}
+\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (upper panel) and the spectrum observed with a telluric standard star (lower panel) around an H$_2$O absorption band.
+The \wobble method delivers a very high \SNR telluric spectrum with no overhead observing time required.}
+\label{fig:telluric_standard}
+\end{figure}
+
 As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{HR 3090 (+details)}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:385782 and }
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=5in]{telluric_basis2}
+\caption{Example of the mean telluric spectral template in a small wavelength region (upper panel) and the three basis vectors used to capture the time-variable contributions to the spectrum, staggered by an arbitrary offset for display (lower panel).}
+\label{fig:telluric_basis}
+\end{figure}
+
 An example of the inferred variability from the 51 Peg analysis is shown in Figure \ref{fig:telluric_basis}. 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=6.5in]{telluric_comparison}
-\caption{Fits to a 15-\ang region with substantial telluric contamination for 51 Peg (left) and \Mdwarf (right).
-The upper panels show the \wobble best-fit telluric spectrum (blue line), best-fit stellar spectrum (red line), data (black points), and the sum of the telluric and stellar model predictions (black line).
-Residuals are shown in the lower panels.
-Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
-\label{fig:telluric_comparison}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=5in]{telluric_standard}
-\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (upper panel) and the spectrum observed with a telluric standard star (lower panel) around an H$_2$O absorption band.
-The \wobble method delivers a very high \SNR telluric spectrum with no overhead observing time required.}
-\label{fig:telluric_standard}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=5in]{telluric_basis2}
-\caption{Example of the mean telluric spectral template in a small wavelength region (upper panel) and the three basis vectors used to capture the time-variable contributions to the spectrum, staggered by an arbitrary offset for display (lower panel).}
-\label{fig:telluric_basis}
-\end{figure}
-
 \section{Limitations and Extensibility of \wobble}
@@
 
-In this work we have tested \wobble exclusively using \HARPS data, but vast amounts of data from other \EPRV instruments exist and may benefit from the techniques used here. 
+In this work we have tested \wobble exclusively using \HARPS data, which allowed us to make relatively strict assumptions about the data quality. 
+We have also made a number of restrictive assumptions about the underlying physical model, for example taking each astronomical source to be a single star. 
+However, there are many other \EPRV data sets which violate some of these assumptions but could nevertheless benefit from the analysis techniques used by \wobble.
+
 We now turn our discussion to ways in which other data sets may violate the assumptions made in this work and the corresponding modifications to \wobble that would enable it to operate more effectively in these regimes.
-For simplicity, we break this discussion down into three categories: inherent limitations of \wobble; assumptions that could be eliminated; and general changes that may improve the \RV precision achieved by \wobble.
+For simplicity, we break this discussion down into two categories: changes to the quality or type of spectroscopic data used and changes to the model. 
+We additionally outline some general changes or enhancements that could be made to potentially improve the \RV precision achieved by \wobble.
 
-\subsection{Inherent Limitations}
+\subsection{Changes to the Data}
 
-limitations on number of spectra: Experiment with sub-sampling 51 Peg to fewer epochs.
-\wobble does have many free parameters and therefore it's not going to work well for very few epochs.
-Also telluric sampling is key.
-in some regime \wobble is just not the right tool for the job.
+In general, the data used in this work are high quality in terms of \SNR, spectral resolution, and quantity. 
+As demonstrated for the case of \Mdwarf, \wobble is limited in its operability when the data are at low \SNR. 
+\wobble implicitly assumes Gaussian noise in $\ln$ flux, which is not strictly accurate. 
+This may become a substantial issue at low \SNR, making \wobble a poor tool choice for accurate modeling of very low \SNR spectra. 
+\todo{(HOGG: maybe say something about how we could mitigate this with changes to continuum handling?)}
 
-low-SNR limitations: what we did is a hack and it should be possible to do much better with better continuum normalization. but still you'll eventually hit a failure mode.
+Our method may also fail at low spectral resolution. \todo{(what will happen for undersampled lines?)}
 
-spectrograph resolution: will this work if lines are undersampled?
+Along with minimum requirements on the spectral data quality, \wobble has some limitations driven by the number of spectra available and the sampling of those spectra. 
+Some minimum quantity of spectra are necessarily simply because the spectral templates contribute many free parameters, deriving up the data points needed for inference. 
+The sampling of those spectra are also important because the stellar lines must undergo substantial \RV variations to disentangle them from the telluric features, which is only feasibly achieved by observing throughout the season to take advantage of the $\sim 30$ \kms \BERV shift.
+This requirement could be partially mitigated by modeling multiple stars with a shared telluric spectral template, which we discuss in Section \ref{s:improvements}. 
+However, in general the \wobble method fundamentally requires a substantial number of high quality spectra to perform reliably.
 
-\subsection{Revisiting Assumptions}
+Another key assumption made in this work that there is no gas cell represented in the data. 
+While this is true for \HARPS, other instruments like \acronym{HIRES}, \acronym{APF}, and \acronym{PFS} include a gas cell set in optical path to imprint its absorption spectrum on the observed spectra \todo{(cite)}.
+Applying \wobble to spectra that include the imprint of a gas cell is technically trivial. 
+The gas absorption lines are fixed features at the observatory rest frame which multiply into the spectrum. 
+This makes them nearly indistinguishable from the mean telluric spectrum, and they should in fact be absorbed into this component. 
+If the observations in question cover a range of different airmasses, it may be necessary to explicitly add a third model component to represent the gas cell; this component would be identical to the tellurics without the airmass scaling.
 
-single star: this actually isn't a true assumption, it's trivial to add an extra star, or even a planetary component, to the model.
+Of course, absorption cells in general exist because the instantaneous calibration of the spectrograph in question is not reliable at the required \RV precision level. 
+This means that a gas cell instrument will generally violate other, more critical assumptions about the data quality.
 
-no gas cell: should be as simple as adding an extra component (Of course, gas cells exist because calibration isn't perfect...)
+It is likely that for such an instrument, our assumption that the wavelength solution is perfect does not hold. 
+One simple way to deal with this would be to allow the observatory rest frame component(s) of the model to change in \RV. 
+The stellar \RV measurement would then be differential with respect to the effective spectrograph \RV. 
+\todo{(TO DO: add some citations to using tellurics for this)}
 
-reliable instrument calibration: use tellurics to modify wavelength calibration?
-(there are relevant citations here) Spectral variability due to LSF changing?
-Additional parameters to the P operators, non-convex optimization
+Spectral variability due to LSF changing?
+Additional parameters to the P operators, this would go into the non-convex optimization part (ask Hogg to throw in some words earlier on about location vs linear parameters)
+
+\subsection{Changes to the Model}
+
+The test data used in this work were all instances of a single, bright star, so that only one stellar component was necessary to model the astronomical source. 
+Other data sets may require a more complex model, including additional stars, planetary spectra, or non-negligible sky background. 
+There are some trade-offs associated with adding these components to the model, which we discuss here.
+%Some or all of these additional components may also vary with time. 
+%Here we briefly discuss how such modifications might be made to the \wobble model to make it applicable to a wider variety of data.
+
+One important feature of \wobble is that it models all fluxes in the log-space. 
+When the model consists solely of multiplicative components, as is the case for a single star + telluric absorption, this choice makes the optimization convex at fixed \RV and computationally tractable \todo{(HOGG: edit plz)}.
+If we chose instead to convert the $y$ data vectors to linear fluxes and use a model consisting only of additive components, the same equations and algorithms would apply. 
+Example applications of such a model might be a multiple star system or a faint star with substantial sky background present in the observed spectra. 
+If, however, the model needed to incorporate both additive and multiplicative components, major structural changes would be necessary and much of the optimization speed would be lost. 
+This means that at present, additive components may only be easily modeled by either fixing or neglecting the multiplicative components. 
+\todo{(HOGG: say things about hacks)}
+
+Another way in which the \wobble model may fail for some data sets is in its assumptions about spectral variability.
 
@@
 
-\subsection{General Improvements}
+\subsection{General Changes}
+\label{s:improvements}
 

### 2018-11-26 paper/paper.tex
more discussion

@@
 The stellar \RV measurement would then be differential with respect to the effective spectrograph \RV. 
+This effective spectrograph \RV may vary in different spectral orders or smaller wavelength regions. \todo{(HOGG: say a thing)} 
 \todo{(TO DO: add some citations to using tellurics for this)}
@@
 Additional parameters to the P operators, this would go into the non-convex optimization part (ask Hogg to throw in some words earlier on about location vs linear parameters)
+In principle, if the changes to the LSF were well-understood they could be hard-coded into the P operators. 
+Realistically speaking, if LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.
+This leads us into the territory of making more fundamental changes to the \wobble model.
 
@@
 There are some trade-offs associated with adding these components to the model, which we discuss here.
-%Some or all of these additional components may also vary with time. 
-%Here we briefly discuss how such modifications might be made to the \wobble model to make it applicable to a wider variety of data.
 
@@
 \todo{(HOGG: say things about hacks)}
+
+One extra model component that is at present both simpler to implement and perhaps more important to the results than additive components is the spectral continuum. 
+We have assumed that our simple continuum normalization is good, but this is certainly not true at a detailed level.
 
@@
 
-\subsection{General Changes}
+\todo{HOGG: Bayesian version of \wobble?}
+
+\subsection{Basic Adjustments}
 \label{s:improvements}
 
+Even working within \wobble's current assumptions and running on \HARPS data, there are potential changes to be made to the implementation of \wobble that may improve its performance. 
+Many of these are quite basic and straightforward changes to be made within the existing framework of \wobble. 
+We discuss some of potential adjustments here.
+
+Perhaps the most obviously sub-optimal part of \wobble is its reliance on regularization. 
+We chose this framework because it keeps the model convex and its implicit assumptions (that the spectra are a flat continuum by default) closely mirror our physical understanding. 
+However, setting the regularization amplitudes optimally is by far the computationally slowest and most unwieldy part of applying \wobble to a new data set. 
+In principle, as data sets get larger, regularization becomes less necessary. 
+One way to remove the need for telluric regularization might be to run on a data set that is composed of many different stars, each with their own individual spectra and \RVs but with a shared telluric component, so that the data informing the telluric lines is sufficiently large.
+This could be done with relatively trivial changes to the \wobble code, although it would require strong computational resources.
+
+One choice that \wobble makes differently from other data-driven \RV codes is the way we represent the underlying spectral components. 
+We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated by the P operator. 
+One possible improvement would be to go to a higher-order interpolation scheme. 
+Another approach would be to take a non-parameteric approach like a Gaussian process (as \code{STARFISH} uses). 
+A third approach would be to use a model that incorporates more physics: this could range from a model that is parameterized in terms of absorption line profile parameters to a complex model including various atomic and molecular physics of the lines.
+A sensible way to implement this within the existing \wobble framework might be to add a physically motivated mean spectral template and optimize $y_{\star, n}$ vectors of residuals away from this template.
+In any case, the general approach of \wobble does not depend intrinsically on the spectral representation, and swapping out the current representation for something else is simple, provided that the representative function can be implemented in \TF.
+
 \begin{itemize}
-
-\item Regularization framework could be improved: we chose it because it is convex and represents our beliefs but it's the most difficult and unwieldy part of \wobble.
-
-\item Spectral representation: ours sucks.
-Could go to higher-order interpolation scheme like cubic split OR smooth non-parametric rep like a GP OR something that incorporates more physics.
-Nothing about wobble depends on the spectral representation and it should be simple to swap out for something better provided it can be implemented in \TF.
 
@@
 Doing so would take on additional computational cost, but we are in \TF so it shouldn't hurt too much.
-
-\item We could fit all stars simultaneously, which would give a much better telluric template
 

### 2018-11-28 paper/paper.tex
adding orcids

@@
 
-\author{Daniel Foreman-Mackey}
+\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
 \affiliation{\flatiron}
@@
 
-\author{Rodrigo Luger}
+\author[0000-0002-0296-3826]{Rodrigo Luger}
 \affiliation{\flatiron}
@@
 Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra.
-Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features, which cannot be perfectly modeled. 
+Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features, which are not perfectly known.
 Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations.
 In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself.
-%In both cases, the full data set of all observed spectra for a given target generally contains more complete information about the underlying 
-%In both cases, using the full data set of all observed spectra for a given target will yield more precise constraints on the underlying spectra and their \RVs than a traditional analysis.
-%\todo{(reword this to be less controversial?)}
-%Telluric absorption features in stellar spectra provide particular challenges to extreme-precision radial velocity (\EPRV) exoplanet surveys. 
-%Large amplitude features are often masked in data analysis pipelines, reducing the number of observable spectral features which can be used to measure an \RV shift.
-%Low-amplitude features that are hard to see even in high signal-to-noise spectra may be an important contributor to the overall \RV noise budget.
 % Aims
-Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a linear model (in the log flux). 
+Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a linear model (in the log of flux).
 The model employs a convex objective and convex regularization to keep the optimization of the spectral components fast.
-We implement this method in \wobble, an open-source \python package which uses \TF in its first non-machine-learning application to astronomical data. 
+We implement this method in \wobble, an open-source \python package which uses \TF in its first non-neural-network application to astronomical data.
 In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
@@
 We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data.
-%This method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.
 \end{abstract}
+
+\keywords{atmospheric effects, methods: data analysis, planets and satellites: detection, stars: individual (51 Pegasi, Barnard's Star, HD 189733), techniques: radial velocities}
 
@@
 Precise radial velocity (\RV) measurements are critical to the discovery and characterization of exoplanets.
-On order of one dozen dedicated spectrographs exist for the purpose of \RV planet-hunting, with at least as many more currently under construction \citep{Wright2017}. 
+On the order of one dozen dedicated spectrographs exist for the purpose of \RV planet-hunting, with at least as many more currently under construction \citep{Wright2017}.
 However, significant challenges exist in deriving precise \RV measurements from these spectra.
@@
 
-One of the primary drivers of the \RV noise budget is the incomplete treatment of telluric features in the
+One of the primary contributors to the noise budget in \RV measurements is the incomplete treatment of telluric features in the
 Earth's atmosphere \citep{Halverson2016}.
@@
 An alternative approach is the simultaneous modeling of both telluric and spectral features from the data.
-As the Earth's motion around the barycenter of the solar system induces a Doppler shift considerably larger than both the motion of telluric features and the size of a single pixel on the detector, these two spectra can be disentangled.
+As the Earth's motion around the barycenter of the Solar System induces a Doppler shift considerably larger than both the motion of telluric features and the size of a single pixel on the detector, these two spectra can be disentangled.
 This process is well-established in the analysis of binary star systems through the development of linear models \citep[e.g.][]{Simon1994} and in a Gaussian process framework \citep{Czekala2017}.
@@
 For stabilized, non-gas-cell \RV spectrographs, a standard approach has been to adopt a quasi-binary mask consisting of weighted top-hat functions at the expected locations of informative stellar absorption lines and cross-correlate this mask with the observed spectrum \citep[e.g.][]{Baranne1979, Pepe2002}.
-This approach is limited by the accuracy of the mask, and since most masks are built for a broad category of spectral type rather than customized for the individual star in question, it is unlikely that this techniques retrieve maximally precise \RVs.
-Deriving a custom spectral template by stacking all spectra iteratively as the \RVs are determined was shown to be a superior approach for stars with complex spectra by \citet{AngladaEscude2012}.
+This approach is limited by the accuracy of the mask, and since most masks are built for a broad category of spectral type rather than customized for the individual star in question, it is unlikely that this technique retrieves maximally precise \RVs.
+Deriving a custom spectral template by stacking all spectra iteratively as the \RVs are determined has been shown to be a superior approach for stars with complex spectra \citep{AngladaEscude2012}.
 
@@
 
-Here we develop a data-driven model to infer the telluric and stellar spectra and calculate the stellar RV at each observed epoch. 
+Here we develop a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch.
 The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data.
@@
 In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
+We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}.
 We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation.
@@
 For the case of the telluric component, whose spectrum is allowed to vary with time, we assume this spectral variability is low-dimensional.
-This assumption is physically motivated by the fact that a relatively small number of molecular species contribute to the telluric absorption spectrum. 
+This assumption is physically motivated in the sense that that a relatively small number of molecular species contribute to the telluric absorption spectrum.
 It is also needed in a practical sense, since every additional dimension over which the telluric spectrum can vary adds several thousand more free parameters to the model.
@@
 We assume that the stellar spectrum is invariant with time.
-This assumption will be revisited in Section \ref{s:future}.
+This assumption does not hold true in detail and will be revisited in Section \ref{s:future}.
 
-We assume that both the stellar and telluric spectra are approximately located at zero in logarithmic flux with small deviations due to absorption lines. 
+We assume that both the stellar and telluric spectra are approximately located at zero in logarithmic flux (or unity in linear) with small deviations due to absorption lines.
 As a result of this assumption, we are able to apply L1 and L2 regularization to the spectral templates.
@@
 
-Aside from this regularization to push spectra to zero in the low-\SNR regime, the model makes no assumptions about the shape of the stellar or telluric spectra. 
-We solve for spectral templates for each component as a series of pixels with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing.
+Aside from this regularization to push model components to zero in the low-\SNR regime, the model makes no assumptions about the shape of the stellar or telluric spectra.
+We solve for spectral templates for each component as a series of control points with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing.
 In addition to keeping this model simple, linear, and largely convex, this means that no physical knowledge about the object being observed is needed to extract its \RVs.
@@
 If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.
-a single night), it will be nearly impossible to disentangle the components of the model from each other, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
+a single night), it will be nearly impossible to disentangle the components of the model, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
 We will revisit this restriction in Section \ref{s:future}.
@@
 Finally, we assume that, when the stellar and telluric components are properly optimized, only Gaussian noise remains.
-This assumption will be revisited in Section \ref{s:future}.
 
@@
 \end{equation}
-where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid.
+where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid of control points.
 Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
@@
 
-The apparent stellar \RV, $v_{obs, n}$, is a sum of the star's actual velocity about its center of mass ($v_{\star, n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV$_{n}$ ), the latter of which is known.
+The apparent stellar \RV, $v_{obs, n}$, is a sum of the star's actual velocity about its center of mass ($v_{\star, n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
 
@@
 \begin{equation}
-y_{t, n} =  A_n(\mu_{t} + W_{t} z_n)
+y_{t, n} =  A_n(\mu_{t} + W_{t} z_n).
 \end{equation}
-In addition to its mean spectrum $\mu_t$, the telluric component also takes a time-dependent component assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
+In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
 $W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used.
 For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications.
-The net telluric spectrum (mean + time-variable components) is weighted by the air mass at the time of observation $A_n$, a known quantity.
+The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
 
@@
 \end{equation}
-comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties on each data point.
+comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.
 
@@
 
-L1 normalization adds a term to the log-likelihood that takes the form:
+L1 normalization adds a term to the log-likelihood that takes the form \todo{(TO DO: change notation to match literature)}:
 $$ L1(p, \lambda) = \lambda \sum_{i} | p_{i} | ,$$
@@
 
-As an initial guess, we set the star to be stationary, e.g. $v_{obs} = v_{BERV}$ at all epochs. 
+As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs.
 We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$.
@@
 
-\todo{(To do: fill in details: prior on $<v_r>$, optimization over $\delta_r$.)}
+\todo{(To do: fill in details: prior on $<v_r>$, optimization over $\delta_r$ with $v_n$ and $v_r$ analytically solved for.)}
 
@@
 The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF.
-\TF is a \todo{DFM: [blah blah technical blah]}. 
-While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
+\TF is a \todo{(DFM: write something technical here please)}.
+While it has primarily been used in the context of deep learning in the astronomical literature, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
 The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 spectra of 51 Peg runs in \todo{30} minutes on a standard Mac desktop.
@@
 
-\todo{The final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off a drift term specified in the data header, consistent with the standard \HARPS pipeline.}
+The final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off a drift term specified in the data header, consistent with the standard \HARPS pipeline.
+\todo{(TO DO: make sure this is implemented!)}
 
@@
 
-We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell329968 and mbedell331551.}
+We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:329968 and mbedell:331551.}
 The 91 publicly available spectra in the \HARPS archive are largely concentrated on a few nights of intensive observing, but these nights are sufficiently spread out throughout the year for a wide enough range in \BERV to disentangle the stellar and telluric spectra.
+Their \SNR is generally high, ranging from 100 up to 300 pix$^{-1}$ at the central wavelength regions of \HARPS.
 
@@
 We emphasize that no \textit{a priori} information on e.g. expected spectral line positions and depths were used.
-The current realization of the \wobble algorithm treats each pixel of the template spectra as independent and has no line shape parameterization included. 
+The current realization of the \wobble algorithm treats each control point of the template spectra as independent and has no line shape parameterization included.
 Nevertheless, the optimized stellar templates clearly reproduce the expected appearance of the spectra in a variety of regimes, from a crowded-line region to a sparser, continuum-dominated region to an extremely strong absorption line like the H$\alpha$ feature.
@@
 
-Next, we tested \wobble with 306 epochs of \Mdwarf spectra.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell359445 and mbedell359501.} 
-\todo{(say brief things about data quality - it's low \SNR)}
+Next, we tested \wobble with 306 epochs of \Mdwarf spectra.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:359445 and mbedell:359501.}
+\Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50.
+In practice, this means that the 7 bluest echelle orders were consistently masked and dropped from the fit in accordance with the \SNR criteria outlined previously.
+Regardless, a large amount of data remained, allowing us to test \wobble's performance on observations with substantially different properties from the previous case.
 
@@
 This trend has been observed in some data sets, including long-term \HARPS and \acronym{HIRES} observations, although \acronym{UVES} \RVs were inconsistent with the predicted linear slope \citep{Kurster2003, Bonfils2013, Choi2013, Montet2014}.
-No planets have been found around \Mdwarf to date despite considerable amounts of \RV monitoring and little stellar activity has been observed, making it a good quiet star test case for \wobble.
+Aside from this linear trend, \Mdwarf is commonly used as an \RV standard M dwarf because little stellar activity has been observed and no planets discovered until very recently despite considerable \RV monitoring \citep{Ribas2018}.
 
@@
 Residuals away from the predicted trend are plotted in the lower panel.
-\todo{(TO DO: add error bars)}}
+\todo{(TO DO: add error bars and change color scheme)}}
 \label{fig:barnards_rvs}
@@
 \includegraphics[width=5.5in]{barnards_multispectrum}
-\caption{Example echelle orders at a randomly chosen epoch from the \HARPS observations of Barnard's Star, plotted as in Figure \ref{fig:51peg_spectrum}.
-\todo{(to do: weaken telluric regularization to try and get missing line at 6576 \ang)}}
+\caption{Example echelle orders at a randomly chosen epoch from the \HARPS observations of Barnard's Star, plotted as in Figure \ref{fig:51peg_spectrum}.}
 \label{fig:barnards_spectrum}
@@
 The spectral template derived by \wobble is shown as a red line, while a theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as a blue line.
-The \wobble model has been Doppler-shifted to match the rest-frame line positions.}
+The \wobble model has been Doppler-shifted to match the rest-frame line positions.
+\todo{(FIX COLORS!)}}
 \label{fig:barnards_model}
@@
 In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}).
-However, telluric variability was not able to be resolved in the overwhelming majority of orders \todo{(check on this and whether RVs are affected by K setting)}. 
-We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations \todo{(\SNR $\lessapprox 100$ pixel$^{-1}$)}.
+However, telluric variability was not able to be resolved in the overwhelming majority of orders; in fact, the results presented here were obtained with the tellurics model set to $K=0$ (no variability included).
+We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations \SNR $\lessapprox 50$ pixel$^{-1}$).
 
@@
 Even in the bluest regions, the features inferred by \wobble compare well with the \PHOENIX predictions in broad strokes (Figure \ref{fig:barnards_model}).
-\todo{(check in about opacity binning affecting the smaller features)}
+\todo{(TO DO: ask Travis Barman about opacity binning affecting the detailed features)}
 
@@
 The resulting \RV signal retrieved by \wobble is shown in Figure \ref{fig:hd189_rm}.
-We fit the signal using \starry \citep{Luger2018}. \todo{(TO DO: fill in with Rodrigo)} 
+We fit the signal using \starry \citep{Luger2018}. \todo{(RODRIGO: fill in a bit about methods)}
 \todo{(comparison of retrieved parameters to previous work here)}
@@
 While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{}.
-The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified in a relatively straightforward manner to fit line asymmetries and true Doppler shifts separately.
+The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified to fit line asymmetries and true Doppler shifts as separate parameters.
 This would help to disentangle signals caused by photospheric features like starspots from the \RV signal due to reflex motion in a planetary system.
@@
 Residuals are shown in the lower panels.
-Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
+Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.
+\todo{(TO DO: add y-label, fix centering)}}
 \label{fig:telluric_comparison}
@@
 \caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (upper panel) and the spectrum observed with a telluric standard star (lower panel) around an H$_2$O absorption band.
+No telluric standard star data was used in determining the \wobble telluric template.
 The \wobble method delivers a very high \SNR telluric spectrum with no overhead observing time required.}
@@
 
-As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{HR 3090 (+details)}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:385782 and }
+As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{HR 3090 (+details)}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:385782 and \todo{xxx}.}
 As shown in Figure \ref{fig:telluric_standard}, the spectra agree well.
@@
 These hypotheses could be tested in the future by seeking correlations between basis weights and externally-measured weather diagnostics.
-\todo{(is this enough? also show that basis weights change seasonally-ish?)}
+%\todo{(is this enough? also show that basis weights change seasonally-ish?)}
 
@@
 \subsection{Changes to the Data}
+\label{s:data-changes}
 
@@
 As demonstrated for the case of \Mdwarf, \wobble is limited in its operability when the data are at low \SNR.
-\wobble implicitly assumes Gaussian noise in $\ln$ flux, which is not strictly accurate. 
+\wobble implicitly assumes Gaussian noise in logarithmic flux, which is not strictly accurate.
 This may become a substantial issue at low \SNR, making \wobble a poor tool choice for accurate modeling of very low \SNR spectra.
@@
 
-Our method may also fail at low spectral resolution. \todo{(what will happen for undersampled lines?)}
+Our method may also fail at low spectral resolution.
+In particular, if the stellar spectrum does not shift with respect to the telluric spectrum by at least a resolution element, it will become extremely challenging to disentangle the two components.
+\todo{(HOGG: say other things about undersampled lines?)}
 
@@
 
-Spectral variability due to LSF changing?
+%Spectral variability due to LSF changing?
 Additional parameters to the P operators, this would go into the non-convex optimization part (ask Hogg to throw in some words earlier on about location vs linear parameters)
-In principle, if the changes to the LSF were well-understood they could be hard-coded into the P operators. 
-Realistically speaking, if LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.
+Another way a less well-calibrated instrument may violate \wobble's assumptions is in the non-negligible variability of all spectral lines due to changes in the instrumental line spread function (\LSF).
+In principle, if the changes to the \LSF were well-understood they could be hard-coded into the P operators.
+Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.
 This leads us into the territory of making more fundamental changes to the \wobble model.
@@
 \subsection{Changes to the Model}
+\label{s:model-changes}
 
@@
 Other data sets may require a more complex model, including additional stars, planetary spectra, or non-negligible sky background.
+They may also require more complex treatment of spectral variability.
 There are some trade-offs associated with adding these components to the model, which we discuss here.
@@
 This means that at present, additive components may only be easily modeled by either fixing or neglecting the multiplicative components.
-\todo{(HOGG: say things about hacks)}
+\todo{(HOGG: say things about hacks - maybe in a footnote even?)}
 
@@
 We have assumed that our simple continuum normalization is good, but this is certainly not true at a detailed level.
+To some extent, this wrongness is absorbed into the telluric spectrum, as both the tellurics and the continuum can be treated as observatory rest frame features.
+However, since the continuum and tellurics vary through time in different ways, the continuum should in principle be its own independent model component.
+This component will be highly degenerate with the tellurics, but a clever representation could circumvent this: for example, the continuum spectrum could be placed on an extremely sparse wavelength grid to ensure that very localized effects like absorption lines are not included.
 
-Another way in which the \wobble model may fail for some data sets is in its assumptions about spectral variability.
+Another way in which the \wobble model may fail for some regimes is in its treatment of spectral variability.
 
-low-dimensional telluric variability: it's possible this isn't true in the current parameterization. what if we work with residuals away from a physical model of the tellurics? or reparameterize lines
+We have assumed that the telluric spectrum is variable in a low-dimensional sense only.
+While this assumption has a basis in the physical understanding of telluric lines, which are produced by only a small number of molecular species, it is not guaranteed that this reasoning translates into low variability in the current parameterization.
+We may instead wish to change the spectral representation so that additional physics is included, as discussed further in Section \ref{s:improvements}.
+Alternatively, we could keep the current representation but fit the telluric model to the residuals away from some physics-based model.
 
-constant stellar spectrum: cite \citet{Davis2017} for PCA approach.
-Some absorption lines are more sensitive to the effects of stellar activity than others (Dumusque).
+We have also assumed that the stellar spectrum is constant, which is almost certainly false.
+A variety of physical effects cause changes in the spectrum including stellar oscillations, convection, and activity in the form of starspots or plages.
+\wobble can model stellar variability trivially by making the stellar component of the model take a similar form to the tellurics, where a set of basis vectors and weights are inferred from the data to capture variability.
+This \PCA-like approach may be quite effective \citep{Davis2017}.
+It does, however, introduce potential degeneracies between the inferred stellar variability and the Doppler shift.
+To avoid this degeneracy, we might instead represent the stellar spectrum with an explicit parameterization of the line profile and restrict variability to the line depths, widths, and skews; such a change would fall along the same lines as the global \LSF parameterization proposed in Section \ref{s:data-changes}.
+
+%Some absorption lines are more sensitive to the effects of stellar activity than others (Dumusque).
 If these fall into a small number of categories, a PCA-like dimensionality reduction should do well.
@@
 
-\todo{HOGG: Bayesian version of \wobble?}
+\todo{(HOGG: Bayesian version of \wobble?)}
 
@@
 
-Perhaps the most obviously sub-optimal part of \wobble is its reliance on regularization. 
+Perhaps the most obvious shortcoming of \wobble is its reliance on regularization.
 We chose this framework because it keeps the model convex and its implicit assumptions (that the spectra are a flat continuum by default) closely mirror our physical understanding.
@@
 One possible improvement would be to go to a higher-order interpolation scheme.
-Another approach would be to take a non-parameteric approach like a Gaussian process (as \code{STARFISH} uses). 
-A third approach would be to use a model that incorporates more physics: this could range from a model that is parameterized in terms of absorption line profile parameters to a complex model including various atomic and molecular physics of the lines.
+Another approach would be a non-parameteric method like a Gaussian process (as in the \code{STARFISH} code from \citet{Czekala2017}).
+A third approach would be to use a model that incorporates more physics: this could range from a model with fixed lines that may vary only in terms of common line profile parameters to a complex model including various atomic and molecular physics of the lines.
 A sensible way to implement this within the existing \wobble framework might be to add a physically motivated mean spectral template and optimize $y_{\star, n}$ vectors of residuals away from this template.
@@
 
-\begin{itemize}
+Another choice we made which is almost certainly suboptimal is the treatment of each spectral order in the echelle data as an independent spectrum with its own stellar \RV.
+In reality, every \RV should be informed by all spectral orders simultaneously.
+Implementing this would take on some extra computational cost, but being in \TF will help to mitigate this.
 
-\item Spectral continuum!!!
-All methods are questionable at low SNR and in the absence of true continuum.
-Right solution is to simultaneously fit continuum with other model components, but this introduces degeneracies and must be handled carefully.
+Finally, a change that is quite major but worth stating regardless is the possibility of fitting the data directly in the 2D domain.
+Widely used methods of extracting 1D spectra almost certainly sacrifice some degree of spectral information \citep{Bolton2010}.
+In the future, we may wish to move to fitting the data directly without extracting.
+The equations for \wobble would change very little in this regime: essentially, the P operator would become a 2D interpolator.
+However, implementation of this model would become much more complicated, and extra variable components like sky background and spectrograph dispersion functions would need to be introduced.
+While this implementation is far beyond the scope of current work, we nevertheless emphasize that the \wobble model is a useful starting framework in which to work towards this goal.
 
-\item We treat the orders separately: we mostly did this to break up the computation, but this could be changed straightforwardly.
-Doing so would take on additional computational cost, but we are in \TF so it shouldn't hurt too much.
-
-\item 1D extracted spectra.
-Our P operators are interpolators onto 1D data but they could be 2D interpolators!
-Little about the model equations would change, but implementation would become much trickier.
-
-\end{itemize}
+\vspace{2cm}
 
@@
 
-Extracting precise \RVs is important and will become more critical as new instruments come online.
+Extremely precise \RV measurements are a critical part of modern exoplanet detection and characterization.
+Many dedicated \EPRV instruments exist and more are being built with the goal of achieving better than \ms precision over long timescales.
+Beyond the hardware challenges of building such instruments, software challenges exist as well: extracting maximally precise \RVs from high-quality spectra is not a trivial task.
+Moreover, most existing \EPRV pipelines are closed-source, proprietary, and built for use with a specific instrument in mind.
+While traditional \RV extraction methods are demonstrably suboptimal when it comes to treatment of telluric features or use of a fixed stellar template spectrum, experimenting with changes to the existing pipelines is generally not possible.
 
-Fundamental data-driven technique - requires no physics knowledge which makes it a great tool for expanding \RV searches.  
-Open-source, can be adapted for use with any current or future \EPRV instrument
-...
+In this work, we have proposed a simple linear model for simultaneously inferring stellar spectra, telluric spectra, and \RVs from the data.
+Our model does not rely on physical knowledge of the star or the Earth's atmosphere.
+We implement this method in \wobble, an open-source code designed to be extensible and adaptable for a variety of data sets.
 
-Here's what this enables.
+By running the \wobble algorithm on archival \HARPS spectra for a variety of stars, we have demonstrated its basic capabilities.
+The \RVs achieved through this method are comparably precise to those obtained with the closed-source \HARPS pipeline.
+Furthermore, the stellar and telluric spectra inferred purely from the data are high-quality and accurate.
+We have shown that consistent telluric features are found from independent data sets with extremely different properties.
 
-\todo{[TO DO: add software citations]}
+The data-driven methods used in \wobble are highly promising for the treatment of microtelluric lines and other deviations from spectral models.
+This makes the approach particularly valuable for upcoming infrared surveys targeting M dwarfs, where telluric features are non-negligible and stellar models often fall short of matching the data.
+It also eliminates costs in observational overhead for obtaining telluric standard spectra or accurate stellar templates.
 
-\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Melissa Ness, Hans-Walter Rix, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions.
+We make \wobble freely available to the \RV community to be adapted for use with various instruments and data sets.
+We highlight the versatility of the algorithm in Section \ref{s:future} with suggested changes and improvements for applications to different regimes.
+It is our hope that openly shared development of next-generation \RV analysis techniques will enable the community as a whole to make full use of the information-rich data being produced by current and future \EPRV surveys.
+
+\software{\todo{TO DO! --} astropy, numpy, matplotlib, scipy, tensorflow}
+
+\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions.
+We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.
 

### 2018-11-28 paper/paper.tex
adding software citations
fixing likelihood notation

@@
 Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
-In this sense, ours is a true forward model.
+%In this sense, ours is a true forward model.  -- this content-less (DFM)
 
@@
 In Section \ref{s:methods}, we outline the model and its key underlying assumptions.
-We present an open-source implementation of this method in \python and \TF called \wobble.
+We present an open-source implementation of this method in \python and \TF
+\citep{Abadi15} called \wobble.
 In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
@@
 
-We assume that the spectra can be modeled as the product of a finite number of components.
+We assume that the spectra can be modeled as the product of a finite and fixed number of components.
 For the cases shown in this work, two components are used: a stellar spectrum which is invariant in shape but may be Doppler-shifted, and a telluric spectrum which is fixed to the observatory rest frame but varies in shape.
@@
 \begin{equation}
-y_n = y_{\star, n} + y_{t, n} + noise.
+y_n = y_{n}^\star + y_{n}^t + noise.
 \end{equation}
@@
 \begin{equation}
-y_{\star, n} = P(v_{obs, n}) \mu_{\star},
+y_{n}^\star = P(v_{obs, n}) \mu_{\star},
 \end{equation}
@@
 \begin{equation}
-y_{t, n} =  A_n(\mu_{t} + W_{t} z_n).
+y_{n}^t =  A_n(\mu_{t} + W_{t} z_n).
 \end{equation}
@@
 \begin{equation}
-\ln \mathcal{L}_n = \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C_{m,n}^{-1} (y_{m,n}-y'_{m,n}),
+\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n}),
 \end{equation}
@@
 L1 normalization adds a term to the log-likelihood that takes the form \todo{(TO DO: change notation to match literature)}:
-$$ L1(p, \lambda) = \lambda \sum_{i} | p_{i} | ,$$
+$$ L1(p, \lambda) = -\lambda \sum_{i} | p_{i} | ,$$
 where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.
@@
 Similarly, L2 normalization adds a term of the form:
-$$ L2(p, \lambda) = \lambda \sum_{i} p_{i}^2 .$$
+$$ L2(p, \lambda) = -\lambda \sum_{i} p_{i}^2 .$$
 
@@
 
-\software{\todo{TO DO! --} astropy, numpy, matplotlib, scipy, tensorflow}
+\software{%
+    astropy \citep{Astropy13, Astropy18},
+    numpy \citep{Van-Der-Walt:2011},
+    matplotlib \citep{Hunter:2007},
+    scipy \citep{Jones:2001},
+    tensorflow \citep{Abadi15}
+}
 

### 2018-11-28 paper/paper.tex
some questions

@@
 We solve for spectral templates for each component as a series of control points with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing.
-In addition to keeping this model simple, linear, and largely convex, this means that no physical knowledge about the object being observed is needed to extract its \RVs.
+In addition to keeping this model simple, linear, and largely convex
+(\todo{MB -- what does ``largely'' mean here?}), this means that no physical knowledge about the object being observed is needed to extract its \RVs.
 
-For practical applications, the number of epochs $N$ must be large and span a significant fraction of the observing season.
+For practical applications, the number of epochs $N$ must be large (\todo{MB
+-- how large? 12? 50? 100? Probably good to give some sense}) and span a significant fraction of the observing season.
 If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.

### 2018-11-28 paper/paper.tex
linear operator

@@
 \end{equation}
-where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid of control points.
+where $\mu_{\star}$ is a spectral template with length $M'$ corresponding to
+the length of an arbitrarily-chosen wavelength grid of control points, and
+$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
+interpolates $\mu_\star$ into the observatory rest frame.
+\todo{MB -- feel free to cut this:} $P$ could be represented as a rectangular
+$M \times M'$ sparse matrix, but the dot product can be implemented
+efficiently using linear interpolation instead of instantiating the full matrix.
 Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:

### 2018-11-28 paper/paper.tex
hessian

@@
 interpolates $\mu_\star$ into the observatory rest frame.
-\todo{MB -- feel free to cut this:} $P$ could be represented as a rectangular
-$M \times M'$ sparse matrix, but the dot product can be implemented
-efficiently using linear interpolation instead of instantiating the full matrix.
 Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
@@
 The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
+\todo{MB -- feel free to cut this:} $P$ could be represented as a rectangular
+$M \times M'$ sparse matrix, but the dot product can be implemented
+efficiently using linear interpolation instead of instantiating the full matrix.
 
@@
 We set regularization amplitudes for \wobble using a cross-validation scheme.
-In brief, we set aside 12.5\% of the total epochs as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set).
+In brief, we randomly select 12.5\% of the total epochs to set aside as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set).
 The resulting best-fit spectral templates and basis vectors are taken as fixed and the time-dependent terms only (\RVs and basis weights) are optimized for the validation epochs.
@@
 
-The inverse variance uncertainty on the parameters is calculated as the Hessian, or the second derivative of the likelihood with respect to each parameter. \todo{(DFM: help me rephrase if needed?)}
+After convergence, we estimate the uncertainties on the parameters by
+approximating the likelihood function near its maximum as Gaussian where the
+covariance matrix is the negative inverse Hessian (or second derivative
+matrix) of the log likelihood function with respect to the parameters.
 

### 2018-11-28 paper/paper.tex
words about tf

@@
 The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF.
-\TF is a \todo{(DFM: write something technical here please)}.
-While it has primarily been used in the context of deep learning in the astronomical literature, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
+\TF is a model building framework that has been primarily designed for machine
+learning applications, but at its core \TF is fundamentally a collection of
+highly optimized routines for doing linear algebra and efficiently computing the
+derivatives of these models with respect to large numbers of parameters.
+\wobble can be represented in this framework so we have re-purposed \TF for our
+needs.
+By using this framework, we benefit from the high performance and scalability
+of the implementation, as well as the algorithms implemented within \TF for
+fitting large numbers of parameters to large datasets.
+%While it has primarily been used in the context of deep learning in the astronomical literature, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
 The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 spectra of 51 Peg runs in \todo{30} minutes on a standard Mac desktop.
 
-Our code made open-source under the \acronym{MIT} license and is publicly available on GitHub.\footnote{\url{https://www.github.com/megbedell/wobble}}
+Our code is made open-source under the \acronym{MIT} license and is publicly available on GitHub.\footnote{\url{https://www.github.com/megbedell/wobble}}
 

### 2018-11-28 paper/paper.tex
questions

@@
 In addition to keeping this model simple, linear, and largely convex
-(\todo{MB -- what does ``largely'' mean here?}), this means that no physical knowledge about the object being observed is needed to extract its \RVs.
+(\todo{DFM says: what does ``largely'' mean here?}), this means that no physical knowledge about the object being observed is needed to extract its \RVs.
 
@@
 The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
-\todo{MB -- feel free to cut this:} $P$ could be represented as a rectangular
+\todo{DFM says: feel free to cut this:} $P$ could be represented as a rectangular
 $M \times M'$ sparse matrix, but the dot product can be implemented
@@
 \todo{(TO DO: ask Travis Barman about opacity binning affecting the detailed features)}
+\todo{DFM says: when I look at figure 5, I feel like the two spectra really don't
+look similar. I think this could use some more discussion about why we find
+this figure satisfying.}
 
@@
 Example applications of such a model might be a multiple star system or a faint star with substantial sky background present in the observed spectra.
-If, however, the model needed to incorporate both additive and multiplicative components, major structural changes would be necessary and much of the optimization speed would be lost.
+If, however, the model needed to incorporate both additive and multiplicative
+components, major structural changes would be necessary and much of the
+optimization speed would be lost \todo{DFM says: this is not obvious to me.
+We're not currently dependent on anything being convex. I don't think we
+should say this without trying it...}.
 This means that at present, additive components may only be easily modeled by either fixing or neglecting the multiplicative components.

### 2018-11-28 paper/paper.tex
reverting changes

--- 2018-11-28 paper/paper.tex
@@
 
-\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
+\author{Daniel Foreman-Mackey}
 \affiliation{\flatiron}
@@
 
-\author[0000-0002-0296-3826]{Rodrigo Luger}
+\author{Rodrigo Luger}
 \affiliation{\flatiron}
@@
 Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
-%In this sense, ours is a true forward model.  -- this content-less (DFM)
+In this sense, ours is a true forward model.
 
@@
 In Section \ref{s:methods}, we outline the model and its key underlying assumptions. 
-We present an open-source implementation of this method in \python and \TF
-\citep{Abadi15} called \wobble.
+We present an open-source implementation of this method in \python and \TF called \wobble. 
 In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
@@
 
-We assume that the spectra can be modeled as the product of a finite and fixed number of components.
+We assume that the spectra can be modeled as the product of a finite number of components. 
 For the cases shown in this work, two components are used: a stellar spectrum which is invariant in shape but may be Doppler-shifted, and a telluric spectrum which is fixed to the observatory rest frame but varies in shape.
@@
 We solve for spectral templates for each component as a series of control points with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing.
-In addition to keeping this model simple, linear, and largely convex
-(\todo{MB -- what does ``largely'' mean here?}), this means that no physical knowledge about the object being observed is needed to extract its \RVs.
+In addition to keeping this model simple, linear, and largely convex, this means that no physical knowledge about the object being observed is needed to extract its \RVs.
 
-For practical applications, the number of epochs $N$ must be large (\todo{MB
--- how large? 12? 50? 100? Probably good to give some sense}) and span a significant fraction of the observing season.
+For practical applications, the number of epochs $N$ must be large and span a significant fraction of the observing season. 
 If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.
@@
 \begin{equation}
-y_n = y_{n}^\star + y_{n}^t + noise.
+y_n = y_{\star, n} + y_{t, n} + noise.
 \end{equation}
@@
 \begin{equation}
-y_{n}^\star = P(v_{obs, n}) \mu_{\star},
+y_{\star, n} = P(v_{obs, n}) \mu_{\star},
 \end{equation}
-where $\mu_{\star}$ is a spectral template with length $M'$ corresponding to
-the length of an arbitrarily-chosen wavelength grid of control points, and
-$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
-interpolates $\mu_\star$ into the observatory rest frame.
+where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid of control points.
 Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
@@
 The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
-\todo{MB -- feel free to cut this:} $P$ could be represented as a rectangular
-$M \times M'$ sparse matrix, but the dot product can be implemented
-efficiently using linear interpolation instead of instantiating the full matrix.
 
@@
 \begin{equation}
-y_{n}^t =  A_n(\mu_{t} + W_{t} z_n).
+y_{t, n} =  A_n(\mu_{t} + W_{t} z_n).
 \end{equation}
@@
 \begin{equation}
-\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n}),
+\ln \mathcal{L}_n = \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C_{m,n}^{-1} (y_{m,n}-y'_{m,n}),
 \end{equation}
@@
 L1 normalization adds a term to the log-likelihood that takes the form \todo{(TO DO: change notation to match literature)}:
-$$ L1(p, \lambda) = -\lambda \sum_{i} | p_{i} | ,$$
+$$ L1(p, \lambda) = \lambda \sum_{i} | p_{i} | ,$$
 where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.
@@
 Similarly, L2 normalization adds a term of the form:
-$$ L2(p, \lambda) = -\lambda \sum_{i} p_{i}^2 .$$
+$$ L2(p, \lambda) = \lambda \sum_{i} p_{i}^2 .$$
 
@@
 We set regularization amplitudes for \wobble using a cross-validation scheme. 
-In brief, we randomly select 12.5\% of the total epochs to set aside as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set).
+In brief, we set aside 12.5\% of the total epochs as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set). 
 The resulting best-fit spectral templates and basis vectors are taken as fixed and the time-dependent terms only (\RVs and basis weights) are optimized for the validation epochs. 
@@
 
-After convergence, we estimate the uncertainties on the parameters by
-approximating the likelihood function near its maximum as Gaussian where the
-covariance matrix is the negative inverse Hessian (or second derivative
-matrix) of the log likelihood function with respect to the parameters.
+The inverse variance uncertainty on the parameters is calculated as the Hessian, or the second derivative of the likelihood with respect to each parameter. \todo{(DFM: help me rephrase if needed?)}
 
@@
 
-\software{%
-    astropy \citep{Astropy13, Astropy18},
-    numpy \citep{Van-Der-Walt:2011},
-    matplotlib \citep{Hunter:2007},
-    scipy \citep{Jones:2001},
-    tensorflow \citep{Abadi15}
-}
+\software{\todo{TO DO! --} astropy, numpy, matplotlib, scipy, tensorflow}
 

### 2018-11-28 HD189733_script.py
capability of fixing tellurics template for HD189

@@ -1,0 +1,140 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import tensorflow as tf
+import wobble
+from time import time
+import h5py
+import os
+
+if __name__ == "__main__":
+    starname = 'HD189733'
+    K_star = 0
+    K_t = 0    
+    niter = 100 # for optimization
+    plots = True
+    epochs = [0, 20] # to plot
+    movies = False
+    
+    plot_dir = 'results/plots_{0}_Kstar{1}_Kt{2}/'.format(starname, K_star, K_t)
+    
+    print("running wobble on star {0} with K_star = {1}, K_t = {2}".format(starname, K_star, K_t))
+    start_time = time()
+    orders = np.arange(72)
+    e = [ 0,  1,  6,  7,  9, 17, 18, 19, 21, 23, 24, 26, 30, 33, 34, 35, 36,
+       37, 38, 40, 42, 43, 44, 45, 46, 47, 48, 49, 50, 51, 53, 55, 56, 61,
+       66, 69, 70, 72, 73, 75] # night of August 28, 2007
+    data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders, epochs=e)
+    orders = np.copy(data.orders)
+    results = wobble.Results(data=data)
+    
+    results_51peg = wobble.Results(filename='/Users/mbedell/python/wobble/results/results_51peg_Kstar0_Kt0.hdf5')
+    
+    print("data loaded")
+    print("time elapsed: {0:.2f} min".format((time() - start_time)/60.0))
+    elapsed_time = time() - start_time
+    
+
+    if plots:
+        print("plots will be saved under directory: {0}".format(plot_dir))
+        if not os.path.exists(plot_dir):
+            os.makedirs(plot_dir)
+    star_learning_rate = 0.1
+    telluric_learning_rate = 0.01
+    for r,o in enumerate(orders):
+        model = wobble.Model(data, results, r)
+        model.add_star('star', variable_bases=K_star, 
+                        regularization_par_file=None, 
+                        learning_rate_template=star_learning_rate)
+        model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
+                            learning_rate_template=telluric_learning_rate,
+                            template_fixed=True, template_xs=results_51peg.tellurics_template_xs[o],
[…]

### 2018-11-28 regularization.py

@@ -11,6 +11,7 @@ __all__ = ["improve_order_regularization", "improve_parameter", "test_regulariza
 def get_name_from_tensor(tensor):
     # hacky method to get rid of characters TF adds to the variable names
     # NOTE - does not handle '_2' type additions!
+    # also won't work if you put colons in your variable names but why would you do that?
     return str.split(tensor.name, ':')[0]
 
 def improve_order_regularization(r, o, star_filename, tellurics_filename,
@@ -18,17 +19,24 @@ def improve_order_regularization(r, o, star_filename, tellurics_filename,
                                  validation_data, validation_results,
                                  verbose=True, plot=False, basename='', 
                                  K_star=0, K_t=0, L1=True, L2=True,
-                                 initialize_at_zero=False): 
+                                 tellurics_template_fixed=False): 
     """
     Use a validation scheme to determine the best regularization parameters for 
     all model components in a given order r.
-    Update files at star_filename, tellurics_filename and return it.
+    Update files at star_filename, tellurics_filename with the best parameters.
     """
     
     training_model = wobble.Model(training_data, training_results, r)
     training_model.add_star('star', variable_bases=K_star)
-    training_model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
-                                initialize_at_zero=initialize_at_zero)
+    if tellurics_template_fixed: # hackity hack hack
+        results_51peg = wobble.Results(filename='/Users/mbedell/python/wobble/results/results_51peg_Kstar0_Kt0.hdf5')
+        template_xs = np.copy(results_51peg.tellurics_template_xs[o])
+        template_ys = np.copy(results_51peg.tellurics_template_ys[o])
+        training_model.add_telluric('tellurics', rvs_fixed=True, template_fixed=True, 
+                                    variable_bases=K_t, template_xs=template_xs,
+                                    template_ys=template_ys)
+    else:
+        training_model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t)
     training_model.setup()
     training_model.optimize(niter=0, verbose=verbose, uncertainties=False)
     
@@ -42,9 +50,13 @@ def improve_order_regularization(r, o, star_filename, tellurics_filename,
[…]

### 2018-11-28 script.py

@@ -7,12 +7,12 @@ import h5py
 import os
 
 if __name__ == "__main__":
-    starname = 'HD189733'
+    starname = 'barnards'
     K_star = 0
-    K_t = 3    
+    K_t = 0    
     niter = 150 # for optimization
     plots = True
-    epochs = [0, 80] # to plot
+    epochs = [0, 50] # to plot
     movies = False
     
     star_reg_file = 'wobble/regularization/{0}_star_K{1}.hdf5'.format(starname, K_star)
@@ -40,6 +40,10 @@ if __name__ == "__main__":
     start_time = time()
     orders = np.arange(72)
     data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders)
+    if True: # reload data and remove all post-upgrade spectra
+        upgrade = 2457174.5 # June 2015
+        e = data.epochs[data.dates < upgrade]
+        data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders, epochs=e)
     orders = np.copy(data.orders)
     results = wobble.Results(data=data)
     
@@ -65,7 +69,16 @@ if __name__ == "__main__":
         print("--- ORDER {0} ---".format(o))
         if plots:
             wobble.optimize_order(model, niter=niter, save_history=True, 
-                                  basename=plot_dir+'history', epochs=epochs, movies=movies)            
+                                  basename=plot_dir+'history', epochs=epochs, movies=movies) 
+            fig, ax = plt.subplots(1, 1, figsize=(8,5))
+            ax.plot(data.dates, results.star_rvs[r] + data.bervs - np.mean(results.star_rvs[r] + data.bervs), 
+                    'k.', alpha=0.8)
+            ax.plot(data.dates, data.pipeline_rvs + data.bervs - np.mean(data.pipeline_rvs + data.bervs), 
+                    'r.', alpha=0.5)   
+            ax.set_ylabel('RV (m/s)', fontsize=14)     
+            ax.set_xlabel('BJD', fontsize=14)   
+            plt.savefig(plot_dir+'results_rvs_o{0}.png'.format(o))
+            plt.close(fig)           
             for e in epochs:
                 fig, (ax, ax2) = plt.subplots(2, 1, gridspec_kw = {'height_ratios':[4, 1]}, figsize=(12,5))
[…]

### 2018-11-28 wobble/model.py

@@ -110,9 +110,10 @@ class Model(object):
         # Set up optimizers
         self.updates = []
         for c in self.components:
-            c.opt_template = tf.train.AdamOptimizer(c.learning_rate_template).minimize(self.nll,
+            if not c.template_fixed:
+                c.opt_template = tf.train.AdamOptimizer(c.learning_rate_template).minimize(self.nll,
                             var_list=[c.template_ys])
-            self.updates.append(c.opt_template)
+                self.updates.append(c.opt_template)
             if not c.rvs_fixed:
                 c.dnll_dv = tf.gradients(self.nll, c.rvs)
                 c.opt_rvs = tf.train.AdamOptimizer(c.learning_rate_rvs).minimize(self.nll,
@@ -217,13 +218,15 @@ class Component(object):
                  L1_template=0., L2_template=0., L1_basis_vectors=0.,
                  L2_basis_vectors=0., L2_basis_weights=1., learning_rate_rvs=1.,
                  learning_rate_template=0.01, learning_rate_basis=0.01,
-                 rvs_fixed=False, variable_bases=0, scale_by_airmass=False,
-                 template_xs=None, initialize_at_zero=False):
+                 rvs_fixed=False, template_fixed=False, variable_bases=0, 
+                 scale_by_airmass=False,
+                 template_xs=None, template_ys=None, initialize_at_zero=False):
         self.name = name
         self.r = r
         self.K = variable_bases # number of variable basis vectors
         self.N = len(starting_rvs)
         self.rvs_fixed = rvs_fixed
+        self.template_fixed = template_fixed
         self.scale_by_airmass = scale_by_airmass
         self.learning_rate_rvs = learning_rate_rvs
         self.learning_rate_template = learning_rate_template
@@ -244,6 +247,7 @@ class Component(object):
         self.starting_rvs = starting_rvs
         self.ivars_rvs = np.zeros_like(starting_rvs) + 10. # will be overwritten
         self.template_xs = template_xs
+        self.template_ys = template_ys
         self.initialize_at_zero = initialize_at_zero # initialize template ys to flat continuum
 
     def setup(self, data, r):
@@ -295,18 +299,19 @@ class Component(object):
             self.template_xs = np.arange(np.min(shifted_xs)-tiny*dx,
                                    np.max(shifted_xs)+tiny*dx, dx)
                                    
-        if self.initialize_at_zero:
-            template_ys = np.zeros_like(self.template_xs)
[…]

### 2018-11-28 wobble/results.py

@@ -13,7 +13,7 @@ COMPONENT_NP_ATTRS = ['K', 'r', 'rvs_fixed', 'ivars_rvs', 'scale_by_airmass', 'l
 OPT_COMPONENT_NP_ATTRS = ['learning_rate_basis', 'L1_basis_vectors', 'L2_basis_vectors', 'L2_basis_weights'] # it's ok if these don't exist
 COMPONENT_TF_ATTRS = ['rvs', 'template_xs', 'template_ys']
 OPT_COMPONENT_TF_ATTRS = ['basis_vectors', 'basis_weights'] # only present if component K > 0
-POST_COMPONENT_ATTRS = ['time_rvs', 'order_rvs', 'order_sigmas'] # these won't exist until post-processing
+POST_COMPONENT_ATTRS = ['time_rvs', 'time_sigmas', 'order_rvs', 'order_sigmas'] # these won't exist until post-processing
 
 class Results(object):
     """A read/writeable object which stores RV & template results across all orders. 

### 2018-11-29 paper/paper.tex
dfm changes without whitespace

--- 2018-11-28 paper/paper.tex
@@
 Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra. 
-Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features, which are not perfectly known.
+Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features which are not perfectly known. 
 Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
@@
 We implement this method in \wobble, an open-source \python package which uses \TF in its first non-neural-network application to astronomical data. 
+\todo{is this true? See Erdmann et al. 1807.08734}
 In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
@@
 The first is that removing sections of the spectrum can remove significant amounts of information about the star, lowering the precision at which we can measure the stellar radial velocity. 
-Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV precision.
+Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV precision \citep{Bottom2013}
 This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres \citep{Figueira2016}. 
@@
 a single night), it will be nearly impossible to disentangle the components of the model, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
+\todo{We should think about whether this is the optimal way to phrase this.
+A short period SB1 would have no problem disentangling telluric features from spectral features, for example!}
 We will revisit this restriction in Section \ref{s:future}.
@@
 After subtracting the \RVs as predicted by the equations of motion at each epoch, the residuals have low dispersion, as expected for a quiet star with no known planets.
+\todo{what about a quiet star with no known planets with K > X m/s?}
 The \acronym{RMS} scatter among \wobble\ \RVs is 2.0 \ms. 
@@
 Indeed, we do note a strong peak at year-long periods in a Lomb-Scargle periodogram of the residual \RVs. 
-Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble. \todo{(TO DO: recalculate \BERV ourselves)}
+Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble.
+\todo{(TO DO: recalculate \BERV ourselves)} \todo{How does HARPS calculate the BERVs?
+If there is something like cirrus, where the apparent brightness of the star varies over the observation, it is plausible that the flux-weighted observation midpoint and the actual time midpoint are different enough to cause a m/s difference in the observations, too.
+Does HARPS have a way of accounting for this?
+(HIRES does estimate this, and they have shorter exposures on average).}
 
@@
 While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{}.
+\todo{citep}
 The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified to fit line asymmetries and true Doppler shifts as separate parameters. 
@@
 In principle, if the changes to the \LSF were well-understood they could be hard-coded into the P operators.
+\todo{This is the first time you've referred to ``p operators'' using that phrasing.
+Maybe include a callback here to equation 2 so the reader is reminded what you're talking about?}
 Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.
@@
 To avoid this degeneracy, we might instead represent the stellar spectrum with an explicit parameterization of the line profile and restrict variability to the line depths, widths, and skews; such a change would fall along the same lines as the global \LSF parameterization proposed in Section \ref{s:data-changes}.
+\todo{Is it worth citing dumusque 2018 here or elsewhere?}
 
@@
 Widely used methods of extracting 1D spectra almost certainly sacrifice some degree of spectral information \citep{Bolton2010}.
+\todo{Maybe this was a recently fixed change, but the pdf you sent didn't show this reference properly}
 In the future, we may wish to move to fitting the data directly without extracting. 

### 2018-11-29 paper/paper.tex
making some suggested changes

--- 2018-11-28 paper/paper.tex
@@
 The model employs a convex objective and convex regularization to keep the optimization of the spectral components fast. 
-We implement this method in \wobble, an open-source \python package which uses \TF in its first non-neural-network application to astronomical data.
+We implement this method in \wobble, an open-source \python package which uses \TF in one of its first non-neural-network applications to astronomical data. 
 In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
@@
 For the case of the telluric component, whose spectrum is allowed to vary with time, we assume this spectral variability is low-dimensional. 
-This assumption is physically motivated in the sense that that a relatively small number of molecular species contribute to the telluric absorption spectrum.
+This assumption is physically motivated in the sense that a relatively small number of molecular species contribute to the telluric absorption spectrum. 
 It is also needed in a practical sense, since every additional dimension over which the telluric spectrum can vary adds several thousand more free parameters to the model.
@@
 We solve for spectral templates for each component as a series of control points with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing.
-In addition to keeping this model simple, linear, and largely convex
-(\todo{DFM says: what does ``largely'' mean here?}), this means that no physical knowledge about the object being observed is needed to extract its \RVs.
+In addition to keeping this model simple and linear, this means that no physical knowledge about the object being observed is needed to extract its \RVs.
 
-For practical applications, the number of epochs $N$ must be large (\todo{MB
--- how large? 12? 50? 100? Probably good to give some sense}) and span a significant fraction of the observing season.
-If the number of spectra is small or if all spectra are taken over a very small range of time (e.g.
-a single night), it will be nearly impossible to disentangle the components of the model, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
-We will revisit this restriction in Section \ref{s:future}.
+For practical applications, the number of epochs $N$ must be large ($N \gg N_{components}$). 
+This is a fundamental restriction rooted in the fact that every spectrum being modeled introduces a large number of free parameters to the model. 
+In most cases, these epochs will also need to span a significant fraction of the observing season. 
+In order to disentangle the stellar and telluric spectral components, the spectra must undergo Doppler shifts with respect to each other that are at least as large as a resolution element of the spectrograph.
+For stars which do not physically undergo large \RV shifts over very short timescales, this means that it is necessary to observe the star over a significant fraction of the year to take advantage of the changes in the Earth's projected motion.
+We will comment on ways to overcome this restriction in Section \ref{s:future}.
 
@@
 y_{n}^\star = P(v_{obs, n}) \mu_{\star},
+\label{eqn:star}
 \end{equation}
@@
 The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
-\todo{DFM says: feel free to cut this:} $P$ could be represented as a rectangular
-$M \times M'$ sparse matrix, but the dot product can be implemented
-efficiently using linear interpolation instead of instantiating the full matrix.
+$P$ could be represented as a rectangular $M \times M'$ sparse matrix, but the dot product can be implemented efficiently using linear interpolation instead of instantiating the full matrix.
 
@@
 
-After subtracting the \RVs as predicted by the equations of motion at each epoch, the residuals have low dispersion, as expected for a quiet star with no known planets.
+After subtracting the secular \RV trend, the residuals have low dispersion, as expected for a quiet star with no planetary signals above $K \sim 1.2~\ms$ \citep{Choi2013, Ribas2018}. 
 The \acronym{RMS} scatter among \wobble\ \RVs is 2.0 \ms. 
@@
 Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble. \todo{(TO DO: recalculate \BERV ourselves)} 
+%\todo{How does HARPS calculate the BERVs?
+If there is something like cirrus, where the apparent brightness of the star varies over the observation, it is plausible that the flux-weighted observation midpoint and the actual time midpoint are different enough to cause a m/s difference in the observations, too.
+Does HARPS have a way of accounting for this?
+(HIRES does estimate this, and they have shorter exposures on average).}
 
@@
 However, telluric variability was not able to be resolved in the overwhelming majority of orders; in fact, the results presented here were obtained with the tellurics model set to $K=0$ (no variability included).
-We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations \SNR $\lessapprox 50$ pixel$^{-1}$).
+We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations (\SNR $\lessapprox 50$ pixel$^{-1}$).
 
@@
 As a sanity check, we compare the \wobble stellar template model with a high-resolution \PHOENIX model at \Mdwarf's previously measured spectral parameters \citep[\teff = 3200 K, \logg = 5.0, and \mh = $-0.5$;][and references therein]{Husser2013, Artigau2018}.
-Even in the bluest regions, the features inferred by \wobble compare well with the \PHOENIX predictions in broad strokes (Figure \ref{fig:barnards_model}).
-\todo{(TO DO: ask Travis Barman about opacity binning affecting the detailed features)}
-\todo{DFM says: when I look at figure 5, I feel like the two spectra really don't
-look similar. I think this could use some more discussion about why we find
-this figure satisfying.}
+Even in the bluest regions, the placements and relative strengths of the large-amplitude features inferred by \wobble generally compare well with the \PHOENIX predictions (Figure \ref{fig:barnards_model}).
+The \wobble model for \Mdwarf diverges from the theoretical models in the smaller absorption lines and in the degree of broadening for strong lines.  
+\todo{(TO DO: are these known model weaknesses? ask Travis Barman about opacity binning affecting the detailed features)}
 
@@
 Along with minimum requirements on the spectral data quality, \wobble has some limitations driven by the number of spectra available and the sampling of those spectra. 
-Some minimum quantity of spectra are necessarily simply because the spectral templates contribute many free parameters, deriving up the data points needed for inference.
-The sampling of those spectra are also important because the stellar lines must undergo substantial \RV variations to disentangle them from the telluric features, which is only feasibly achieved by observing throughout the season to take advantage of the $\sim 30$ \kms \BERV shift.
+Some minimum quantity of spectra are necessarily simply because the spectral templates contribute many free parameters, driving up quantity of data needed for inference. 
+The sampling of those spectra are also important because the stellar lines must undergo substantial \RV variations to disentangle them from the telluric features, which for most targets can only be achieved by observing throughout the year to take advantage of the $\sim 30$ \kms~\BERV shift.
 This requirement could be partially mitigated by modeling multiple stars with a shared telluric spectral template, which we discuss in Section \ref{s:improvements}. 
-However, in general the \wobble method fundamentally requires a substantial number of high quality spectra to perform reliably.
+However, in general the \wobble method fundamentally requires a substantial number ($N \gtrsim 10$) of high quality spectra to perform reliably.
 
@@
 Another way a less well-calibrated instrument may violate \wobble's assumptions is in the non-negligible variability of all spectral lines due to changes in the instrumental line spread function (\LSF).
-In principle, if the changes to the \LSF were well-understood they could be hard-coded into the P operators.
+In principle, if the changes to the \LSF were well-understood they could be hard-coded into linear operators like the $P$ operator used in Equation \ref{eqn:star}, leaving the essential equations of \wobble largely unchanged.
 Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.
@@
 If, however, the model needed to incorporate both additive and multiplicative
-components, major structural changes would be necessary and much of the
-optimization speed would be lost \todo{DFM says: this is not obvious to me.
-We're not currently dependent on anything being convex. I don't think we
-should say this without trying it...}.
-This means that at present, additive components may only be easily modeled by either fixing or neglecting the multiplicative components.
-\todo{(HOGG: say things about hacks - maybe in a footnote even?)}
+components as free variables, larger structural changes would be needed in the code and the convexity of the template optimization would be lost. 
+Further tests would be needed to determine whether this loss of convexity would slow down the optimization.
 
-One extra model component that is at present both simpler to implement and perhaps more important to the results than additive components is the spectral continuum.
+One extra model component that is at present both simpler to implement and perhaps more important to the results than generic additive components is the spectral continuum. 
 We have assumed that our simple continuum normalization is good, but this is certainly not true at a detailed level. 
@@
 \wobble can model stellar variability trivially by making the stellar component of the model take a similar form to the tellurics, where a set of basis vectors and weights are inferred from the data to capture variability.
-This \PCA-like approach may be quite effective \citep{Davis2017}.
+This \PCA-like approach may be quite effective, since stellar features with similar line formation physics likely co-vary in the spectrum \citep{Davis2017, Dumusque2018}. 
 It does, however, introduce potential degeneracies between the inferred stellar variability and the Doppler shift. 
 To avoid this degeneracy, we might instead represent the stellar spectrum with an explicit parameterization of the line profile and restrict variability to the line depths, widths, and skews; such a change would fall along the same lines as the global \LSF parameterization proposed in Section \ref{s:data-changes}.
-
-%Some absorption lines are more sensitive to the effects of stellar activity than others (Dumusque).
-If these fall into a small number of categories, a PCA-like dimensionality reduction should do well.
-Address pros and cons of introducing Gaussian processes to parameterize line profiles.
-deal with this similarly to the LSF changing?
-also, maybe we need to regularize (or put priors on) the \RVs themselves and force them to sum of Keplerians for the purpose of breaking degeneracies with stellar variability.
 
@@
 One choice that \wobble makes differently from other data-driven \RV codes is the way we represent the underlying spectral components. 
-We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated by the P operator.
+We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated by the $P$ operator. 
 One possible improvement would be to go to a higher-order interpolation scheme. 
-Another approach would be a non-parameteric method like a Gaussian process (as in the \code{STARFISH} code from \citet{Czekala2017}).
+Another approach would be a non-parameteric method like a Gaussian process \citep[as in][]{Czekala2017}. 
 A third approach would be to use a model that incorporates more physics: this could range from a model with fixed lines that may vary only in terms of common line profile parameters to a complex model including various atomic and molecular physics of the lines.
@@
 In the future, we may wish to move to fitting the data directly without extracting. 
-The equations for \wobble would change very little in this regime: essentially, the P operator would become a 2D interpolator.
+The equations for \wobble would change very little in this regime: essentially, the $P$ operator in Equation \ref{eqn:star} would become a 2D interpolator. 
 However, implementation of this model would become much more complicated, and extra variable components like sky background and spectrograph dispersion functions would need to be introduced. 
 While this implementation is far beyond the scope of current work, we nevertheless emphasize that the \wobble model is a useful starting framework in which to work towards this goal.
-
-\vspace{2cm}
 

### 2018-11-29 paper/paper.tex
notation changes and more citations

@@
 The first is that removing sections of the spectrum can remove significant amounts of information about the star, lowering the precision at which we can measure the stellar radial velocity. 
-Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV precision \citep{Bottom2013}
+Many of the regions of significant telluric absorption lie in the red-optical and near-infrared, where there are abundant narrow spectral features that can be used to improve \RV precision \citep{Bottom2013}.
 This is especially true for M dwarfs, which peak in emitted energy at $\approx 1 \mu$m and have many narrow molecular absorption features in their photospheres \citep{Figueira2016}. 
@@
 This approach is limited by the accuracy of the mask, and since most masks are built for a broad category of spectral type rather than customized for the individual star in question, it is unlikely that this technique retrieves maximally precise \RVs.
-Deriving a custom spectral template by stacking all spectra iteratively as the \RVs are determined has been shown to be a superior approach for stars with complex spectra \citep{AngladaEscude2012}.
+Deriving a custom spectral template by stacking all spectra iteratively as the \RVs are determined has been shown to be a superior approach for stars with complex spectra \citep{AngladaEscude2012, Zechmeister2018}.
 
@@
 \citep{Abadi15} called \wobble.
-In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
+In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the canonical planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
 We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}.
@@
 \ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n}),
+\label{eqn:lnlike}
 \end{equation}
@@
 
-L1 normalization adds a term to the log-likelihood that takes the form \todo{(TO DO: change notation to match literature)}:
-$$ L1(p, \lambda) = -\lambda \sum_{i} | p_{i} | ,$$
+L1 normalization adds a term to the log-likelihood that takes the form:
+\begin{equation}
+\Lone{\lambda}{p} \equiv -\lambda \sum_{i} | p_{i} | ,
+\end{equation}
 where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.
@@
 Similarly, L2 normalization adds a term of the form:
-$$ L2(p, \lambda) = -\lambda \sum_{i} p_{i}^2 .$$
+\begin{equation}
+\Ltwo{\lambda}{p} \equiv -\lambda \sum_{i} p_{i}^2 .
+\end{equation}
 
@@
 
-With regularization included, the final model likelihood is:
+With regularization included, the final model likelihood to be optimized is:
 \begin{equation}
 \begin{split}
-\ln \mathcal{L} = & \sum_{n} \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C_{m,n}^{-1} (y_{m,n}-y'_{m,n})  \\
- & + L1(\mu_{\star}, \lambda_{L1, \mu_{\star}}) + L2(\mu_{\star}, \lambda_{L2, \mu_{\star}}) + L1(\mu_{t},  \lambda_{L1, \mu_{t}}) + L2(\mu_{t},  \lambda_{L2, \mu_{t}}) \\
- & + L1(W_t, \lambda_{L1, W_t}) + L2(W_t, \lambda_{L1, W_t}) + L2(z_t, 1.0) ,
+\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}} \\
+ &  + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}}  + \Lone{\lambda_5}{W_t}  + \Ltwo{\lambda_6}{W_t} + \Lone{1.0}{z_t} ,
 \end{split}

### 2018-12-03 paper/paper.tex
getting rid of two hoggs

@@
 \end{equation}
-where the regularization amplitude for the basis weights is arbitrarily set to 1.0 and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
-\todo{(HOGG: explain why regularization on basis weights is needed)}
+\todp{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!)
+where the regularization amplitude on the basis weights is arbitrarily set to unity, and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
+The basis-weight regularization amplitude can be set arbitrarily to unity because there is a perfect degeneracy between amplifying the basis vectors and attenuating the basis weights.
+This regularization choice breaks that perfect degeneracy.
+The basis weights are regularized because we want the mean telluric spectrum to do most of the telluric work, and the variable basis to only pick up what it needs to,
 
@@
 Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to maximize the likelihood function again. 
-\todo{(HOGG: how could this be made convex too?)}
+Technically the velocities are location parameters and their optimization is not convex.
+However, once the velocities are known to a small fraction of a pixel (which they usually are, in practice, in advance of running \wobble), the linearized problem becomes convex at each iteration.
 We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
+This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iteration to convergence.
 
@@
 All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra. 
-These data are blocks of 72 order by 4096 pixel extracted 1D spectra. 
+These data come as extracted 1D spectra, in blocks of 72 orders by 4096 pixels per order.
 The airmass of the observation and the calculated \BERV are both provided in the \acronym{FITS} header of the spectral data file. 

### 2018-12-03 paper/paper.tex
typos

@@
 \end{equation}
-\todp{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!)
+\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
 where the regularization amplitude on the basis weights is arbitrarily set to unity, and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.

### 2018-12-04 docs/conf.py
attempting to add docs

@@ -1,0 +1,185 @@
+# -*- coding: utf-8 -*-
+#
+# Configuration file for the Sphinx documentation builder.
+#
+# This file does only contain a selection of the most common options. For a
+# full list see the documentation:
+# http://www.sphinx-doc.org/en/master/config
+
+# -- Path setup --------------------------------------------------------------
+
+# If extensions (or modules to document with autodoc) are in another directory,
+# add these directories to sys.path here. If the directory is relative to the
+# documentation root, use os.path.abspath to make it absolute, like shown here.
+#
+import os
+import sys
+sys.path.insert(0, os.path.abspath('.'))
+sys.path.insert(0, os.path.abspath('../wobble'))
+sys.path.insert(0, os.path.abspath('../'))
+
+# -- Project information -----------------------------------------------------
+
+project = 'wobble'
+copyright = '2018, Megan Bedell'
+author = 'Megan Bedell'
+
+# The short X.Y version
+version = ''
+# The full version, including alpha/beta/rc tags
+release = '0.0.1'
+
+
+# -- General configuration ---------------------------------------------------
+
+# If your documentation needs a minimal Sphinx version, state it here.
+#
+# needs_sphinx = '1.0'
+
+# Add any Sphinx extension module names here, as strings. They can be
+# extensions coming with Sphinx (named 'sphinx.ext.*') or your custom
+# ones.
+extensions = [
+    'sphinx.ext.autodoc',
+    'sphinx.ext.coverage',
+    'sphinx.ext.githubpages',
+    'sphinx.ext.napoleon',
+]
+
+napoleon_google_docstring = False
+napoleon_use_param = False
+napoleon_use_ivar = True
+
+# Add any paths that contain templates here, relative to this directory.
+templates_path = ['_templates']
+
+# The suffix(es) of source filenames.
+# You can specify multiple suffix as a list of string:
+#
+# source_suffix = ['.rst', '.md']
+source_suffix = '.rst'
+
+# The master toctree document.
+master_doc = 'index'
+
+# The language for content autogenerated by Sphinx. Refer to documentation
+# for a list of supported languages.
+#
+# This is also used if you do content translation via gettext catalogs.
+# Usually you set "language" from the command line for these cases.
+language = None
+
+# List of patterns, relative to source directory, that match files and
+# directories to ignore when looking for source files.
+# This pattern also affects html_static_path and html_extra_path.
+exclude_patterns = ['_build', 'Thumbs.db', '.DS_Store']
+
[…]

### 2018-12-04 paper/paper.tex

@@
 \HARPS has operated continuously since 2003 and as such has an extensive catalog of publicly available data. 
-Furthermore, its excellent instrumental stability and careful calibration make the data ideally suited to our method, which relies on having an accurate wavelength solution for every spectrum.
+Furthermore, its excellent instrumental stability and precise calibration make the data ideally suited to our method, which relies on having an accurate wavelength solution for every spectrum.
 
@@
 We obtained wavelength solutions for each spectrum from the dedicated \HARPS calibration archive, also maintained by \acronym{ESO}.
-\todo{Uncertainty estimates for each pixel were calculated as... [HOGG, help me fill this in - relation to \SNR and conversion to log]}
 
@@
 In brief, we mask pixels whose extracted flux is below zero as well as regions at the edges of the spectral orders where the local \SNR falls below 5. 
-After masking these data, we convert the flux to log space and continuum normalize by fitting a polynomial to an asymmetrically clipped subset of the data.
+After masking these data, we convert the flux to log space and continuum normalize by fitting and subtracting a polynomial to an asymmetrically clipped subset of the data.
+
+To each pixel in the normalized, logarithmic spectrum, we assign an uncertainty variance (squared error).
+Technically this uncertainty is assumed to be Gaussian in the log space, which is an incorrect approximation, but not very far off for high signal-to-noise data.
+\todo{BEDELL: Check this:}
+For each order we assign to each log flux an uncertainty variance that is the inverse of the mean signal-to-noise-squared of the pixels in that order (as reported by the \HARPS pipelines), scaled up or down by the raw linear flux observed in that pixel in the data prior to logging and normalizing.
+This is the best Poisson estimate we can make given that we are working in the log space, and we don't have individual-pixel uncertainty estimates.
+Technically a small bias is introduced by taking the logarithm of the flux, but this bias is very small at the relevant signal-to-noise ratios; furthermore, the small bias acts in the flux direction, not the radial-velocity direction.
 
@@
 In general, the data used in this work are high quality in terms of \SNR, spectral resolution, and quantity. 
-As demonstrated for the case of \Mdwarf, \wobble is limited in its operability when the data are at low \SNR. 
+As demonstrated for the case of \Mdwarf, \wobble is limited in its operability when the data are low in \SNR. 
 \wobble implicitly assumes Gaussian noise in logarithmic flux, which is not strictly accurate. 
-This may become a substantial issue at low \SNR, making \wobble a poor tool choice for accurate modeling of very low \SNR spectra. 
-\todo{(HOGG: maybe say something about how we could mitigate this with changes to continuum handling?)}
+This could become a substantial issue at low \SNR, making \wobble a poor tool choice for accurate modeling of very low \SNR spectra.
+At this point, it is also the case that our continuum-normalization method also does poorly at low \SNR, so there are various reasons \wobble is not optimal for faint sources or very short exposure times.
+However, the fact that the model is successful in the form of a linear additive model in the log space means that it could probably be converted to work as a multiplicative model in the linear space; that conversion and investigation is beyond the scope of the current work, which is optimized for performance on typical \HARPS observing campaigns.
 
 Our method may also fail at low spectral resolution. 
-In particular, if the stellar spectrum does not shift with respect to the telluric spectrum by at least a resolution element, it will become extremely challenging to disentangle the two components. 
-\todo{(HOGG: say other things about undersampled lines?)}
+In particular, if the stellar spectrum does not shift with respect to the telluric spectrum by at least a resolution element (naively this is expected to become a problem at resolutions below $10^4$), it could become challenging to disentangle the two components.
+That said, there still is a causal difference in the data space between signals that are fixed with respect to the star and signals that are fixed with respect to the atmosphere; in principle it is a question not purely of resolution but a combination of resolution and \SNR.
+Naively, at low resolution, the model is expected to work well when the total signal-to-noise in the data set (greatly) exceeds the dimensionless ratio of the line width to the radial-velocity variation.
 
@@
 
-It is likely that for such an instrument, our assumption that the wavelength solution is perfect does not hold. 
+It is likely that for such an instrument, our assumption that the wavelength solution is perfect does not hold; hence the cell. 
 One simple way to deal with this would be to allow the observatory rest frame component(s) of the model to change in \RV. 
 The stellar \RV measurement would then be differential with respect to the effective spectrograph \RV. 
-This effective spectrograph \RV may vary in different spectral orders or smaller wavelength regions. \todo{(HOGG: say a thing)} 
+This effective spectrograph \RV may vary in different spectral orders or smaller wavelength regions.
+The most extreme extension would be to include the full wavelength solution as model parameters, and optimize them along with the spectral components and radial velocities.
 \todo{(TO DO: add some citations to using tellurics for this)}

### 2018-12-04 wobble/__init__.py
cleaning up docs

@@ -1,5 +1,5 @@
 name = "wobble"
-from .utils import get_session, doppler
+from .utils import bin_data, get_session, doppler, fit_continuum
 from .interp import interp
 from .data import Data
 from .results import Results

### 2018-12-10 wobble/data.py
figure changes, filelist optional in Data

@@ -100,7 +100,10 @@ class Data(object):
             self.bervs = np.copy(f['bervs'])[self.epochs]
             self.drifts = np.copy(f['drifts'])[self.epochs]
             self.airms = np.copy(f['airms'])[self.epochs]
-            self.filelist = [a.decode('utf8') for a in np.copy(f['filelist'])[self.epochs]]
+            try:
+                self.filelist = [a.decode('utf8') for a in np.copy(f['filelist'])[self.epochs]]
+            except KeyError:
+                print("WARNING: data missing filelist.")
             self.R = len(orders) # number of orders
             self.orders = orders # indices of orders in origin_file
             self.ivars = [self.fluxes[i]**2 * self.flux_ivars[i] for i in range(self.R)] # ivars for log(fluxes)

### 2018-12-10 wobble/make_data.py

@@ -135,7 +135,7 @@ if __name__ == "__main__":
         hdffile = '../data/wasp4_e2ds.hdf5'
         write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, ccf_filelist, hdffile)
         
-    if True: #Barnard's Star
+    if False: #Barnard's Star
         ccf_filelist = glob.glob('/Users/mbedell/python/wobble/data/barnards/HARPS*ccf_M2_A.fits')
         
         if False: # check for missing wavelength files
@@ -148,9 +148,9 @@ if __name__ == "__main__":
         hdffile = '../data/barnards_e2ds.hdf5'
         write_data(data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts, ccf_filelist, hdffile)     
         
-    if False:
-        ccf_filelist = glob.glob('/Users/mbedell/python/wobble/data/HD4308/HARPS*ccf_G2_A.fits')
-        if True: # check for missing wavelength files
+    if True:
+        ccf_filelist = glob.glob('/Users/mbedell/python/wobble/data/HD189733/HARPS*ccf_G2_A.fits')
+        if False: # check for missing wavelength files
             missing_files = missing_wavelength_files(ccf_filelist)
             np.savetxt('missing_files.txt', missing_files, fmt='%s')
             print('{0} missing wavelength files'.format(len(missing_files)))

### 2018-12-10 wobble/make_data.py
saving pipeline uncertainties

@@ -26,12 +26,13 @@ def read_data_from_fits(filelist, e2ds=False):
     ivars = [np.zeros((N,M)) for r in range(R)]
     xs = [np.zeros((N,M)) for r in range(R)]
     empty = np.array([], dtype=int)
-    pipeline_rvs, dates, bervs, airms, drifts = np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N)
+    pipeline_rvs, pipeline_sigs, dates, bervs, airms, drifts = np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N)
     for n,f in enumerate(filelist):
         print(f)
         sp = fits.open(f)
         if not e2ds:
             pipeline_rvs[n] = sp[0].header['HIERARCH ESO DRS CCF RVC'] * 1.e3 # m/s
+            pipeline_sigs[n] = sp[0].header['HIERARCH ESO DRS CCF NOISE'] * 1.e3 # m/s
             drifts[n] = sp[0].header['HIERARCH ESO DRS DRIFT SPE RV']  
         dates[n] = sp[0].header['HIERARCH ESO DRS BJD']        
         bervs[n] = sp[0].header['HIERARCH ESO DRS BERV'] * 1.e3 # m/s
@@ -59,8 +60,9 @@ def read_data_from_fits(filelist, e2ds=False):
         xs[r] = np.delete(xs[r], empty, axis=0)
     
     pipeline_rvs = np.delete(pipeline_rvs, empty)
-    bervs = np.delete(bervs, empty)
+    pipeline_sigs = np.delete(pipeline_sigs, empty)
     dates = np.delete(dates, empty)
+    bervs = np.delete(bervs, empty)
     airms = np.delete(airms, empty)
     drifts = np.delete(drifts, empty)
     
@@ -68,7 +70,7 @@ def read_data_from_fits(filelist, e2ds=False):
     pipeline_rvs -= bervs  
     pipeline_rvs -= np.mean(pipeline_rvs)
         
-    return data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts
+    return data, ivars, xs, pipeline_rvs, pipeline_sigs, dates, bervs, airms, drifts
     
 def savfile_to_filelist(savfile, destination_dir='../data/'):
     # copies CCF + E2DS files to destination_dir and returns a list of the CCFs
[…]

### 2018-12-11 scripts/make_data.py
moving around scripts

@@ -28,7 +28,6 @@ def read_data_from_fits(filelist, e2ds=False):
     empty = np.array([], dtype=int)
     pipeline_rvs, pipeline_sigs, dates, bervs, airms, drifts = np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N)
     for n,f in enumerate(filelist):
-        print(f)
         sp = fits.open(f)
         if not e2ds:
             pipeline_rvs[n] = sp[0].header['HIERARCH ESO DRS CCF RVC'] * 1.e3 # m/s
@@ -120,7 +119,7 @@ if __name__ == "__main__":
         
     if True: #51 Peg
         ccf_filelist = glob.glob('/Users/mbedell/python/wobble/data/51peg/HARPS*ccf_G2_A.fits')
-        data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
+        data, ivars, xs, pipeline_rvs, pipeline_sigs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
         hdffile = '../data/51peg_e2ds.hdf5'
         write_data(data, ivars, xs, pipeline_rvs, pipeline_sigs, dates, bervs, airms, drifts, ccf_filelist, hdffile)
         
@@ -132,7 +131,7 @@ if __name__ == "__main__":
             np.savetxt('missing_files.txt', missing_files, fmt='%s')
             print('{0} missing wavelength files for Barnard\'s Star'.format(len(missing_files)))
             
-        data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
+        data, ivars, xs, pipeline_rvs, pipeline_sigs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
         hdffile = '../data/barnards_e2ds.hdf5'
         write_data(data, ivars, xs, pipeline_rvs, pipeline_sigs, dates, bervs, airms, drifts, ccf_filelist, hdffile)     
         
@@ -142,16 +141,16 @@ if __name__ == "__main__":
             missing_files = missing_wavelength_files(ccf_filelist)
             np.savetxt('missing_files.txt', missing_files, fmt='%s')
             print('{0} missing wavelength files'.format(len(missing_files)))
-        data, ivars, xs, pipeline_rvs, dates, bervs, airms, drifts = read_data_from_fits(ccf_filelist)
[…]

### 2018-12-11 scripts/script_51peg.py

@@ -7,27 +7,40 @@ import h5py
 import os
 
 if __name__ == "__main__":
-    starname = 'HD189733'
+    starname = '51peg'
     K_star = 0
-    K_t = 0    
-    niter = 100 # for optimization
+    K_t = 3    
+    niter = 150 # for optimization
     plots = True
-    epochs = [0, 20] # to plot
+    epochs = [0, 50] # to plot
     movies = False
     
-    plot_dir = 'results/plots_{0}_Kstar{1}_Kt{2}/'.format(starname, K_star, K_t)
+    star_reg_file = '../wobble/regularization/{0}_star_K{1}.hdf5'.format(starname, K_star)
+    tellurics_reg_file = '../wobble/regularization/{0}_t_K{1}.hdf5'.format(starname, K_t)
+    plot_dir = '../results/plots_{0}_Kstar{1}_Kt{2}/'.format(starname, K_star, K_t)
+    
+    if False:
+        # quick test on two orders
+        data = wobble.Data(starname+'_e2ds.hdf5', filepath='../data/', orders=[30,56])
+        results = wobble.Results(data=data)
+        for r in range(data.R):
+            model = wobble.Model(data, results, r)
+            model.add_star('star', variable_bases=K_star, 
+                            regularization_par_file=star_reg_file,
+                            learning_rate_template=0.01, learning_rate_rvs=1.)
+            model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
+                                regularization_par_file=tellurics_reg_file,
+                                learning_rate_template=0.01)
+            wobble.optimize_order(model, niter=niter, save_history=True, uncertainties=False,
+                                  basename='results/test', epochs=epochs, movies=movies)
+        results.write('results/test_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t))
+        assert False
     
     print("running wobble on star {0} with K_star = {1}, K_t = {2}".format(starname, K_star, K_t))
     start_time = time()
     orders = np.arange(72)
-    e = [ 0,  1,  6,  7,  9, 17, 18, 19, 21, 23, 24, 26, 30, 33, 34, 35, 36,
-       37, 38, 40, 42, 43, 44, 45, 46, 47, 48, 49, 50, 51, 53, 55, 56, 61,
-       66, 69, 70, 72, 73, 75] # night of August 28, 2007
[…]

### 2018-12-11 scripts/script_HD189733.py

@@ -7,45 +7,30 @@ import h5py
 import os
 
 if __name__ == "__main__":
-    starname = 'barnards'
+    starname = 'HD189733'
     K_star = 0
     K_t = 0    
-    niter = 150 # for optimization
+    niter = 100 # for optimization
     plots = True
-    epochs = [0, 50] # to plot
+    epochs = [0, 20] # to plot
     movies = False
     
-    star_reg_file = 'wobble/regularization/{0}_star_K{1}.hdf5'.format(starname, K_star)
-    tellurics_reg_file = 'wobble/regularization/{0}_t_K{1}.hdf5'.format(starname, K_t)
-    plot_dir = 'results/plots_{0}_Kstar{1}_Kt{2}/'.format(starname, K_star, K_t)
-    
-    if False:
-        # quick test on two orders
-        data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=[30,56])
-        results = wobble.Results(data=data)
-        for r in range(data.R):
-            model = wobble.Model(data, results, r)
-            model.add_star('star', variable_bases=K_star, 
-                            regularization_par_file=star_reg_file,
-                            learning_rate_template=0.01, learning_rate_rvs=1.)
-            model.add_telluric('tellurics', rvs_fixed=True, variable_bases=K_t, 
-                                regularization_par_file=tellurics_reg_file,
-                                learning_rate_template=0.01)
-            wobble.optimize_order(model, niter=niter, save_history=True, uncertainties=False,
-                                  basename='results/test', epochs=epochs, movies=movies)
-        results.write('results/test_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t))
-        assert False
+    plot_dir = '../results/plots_{0}_Kstar{1}_Kt{2}/'.format(starname, K_star, K_t)
     
     print("running wobble on star {0} with K_star = {1}, K_t = {2}".format(starname, K_star, K_t))
     start_time = time()
     orders = np.arange(72)
-    data = wobble.Data(starname+'_e2ds.hdf5', filepath='data/', orders=orders)
-    if True: # reload data and remove all post-upgrade spectra
-        upgrade = 2457174.5 # June 2015
-        e = data.epochs[data.dates < upgrade]
[…]

### 2018-12-11 docs/conf.py
Update conf.py

@@ -29,6 +29,15 @@ version = ''
 # The full version, including alpha/beta/rc tags
 release = '0.0.1'
 
+# Mock imports
+autodoc_mock_imports = [
+    "numpy",
+    "scipy",
+    "astropy",
+    "pymc3",
+    "theano",
+    "tensorflow"
+]
 
 # -- General configuration ---------------------------------------------------
 

### 2018-12-11 docs/conf.py
Update conf.py

@@ -35,8 +35,7 @@ autodoc_mock_imports = [
     "scipy",
     "astropy",
     "pymc3",
-    "theano",
-    "tensorflow"
+    "theano"
 ]
 
 # -- General configuration ---------------------------------------------------

### 2018-12-11 docs/conf.py
Rtd patch (#40)
* Update requirements.txt

* Update conf.py

* Update setup.py

@@ -35,7 +35,8 @@ autodoc_mock_imports = [
     "scipy",
     "astropy",
     "pymc3",
-    "theano"
+    "theano",
+    "tensorflow"
 ]
 
 # -- General configuration ---------------------------------------------------

### 2018-12-11 setup.py

@@ -2,11 +2,18 @@
 
 import os
 import sys
-import tensorflow as tf
 from setuptools import setup, Extension
 
-compile_flags = tf.sysconfig.get_compile_flags()
-link_flags = tf.sysconfig.get_link_flags()
+# Read the docs hack
+on_rtd = os.environ.get('READTHEDOCS', None) == 'True'
+if on_rtd:
+    compile_flags = []
+    link_flags = []
+else:
+    import tensorflow as tf
+    compile_flags = tf.sysconfig.get_compile_flags()
+    link_flags = tf.sysconfig.get_link_flags()
+    
 compile_flags += ["-std=c++11"]
 if sys.platform == "darwin":
     compile_flags += ["-mmacosx-version-min=10.9"]

### 2018-12-11 setup.py
Update setup.py

@@ -4,34 +4,38 @@ import os
 import sys
 from setuptools import setup, Extension
 
-# Read the docs hack
 on_rtd = os.environ.get('READTHEDOCS', None) == 'True'
 if on_rtd:
-    compile_flags = []
-    link_flags = []
+    
+    # Skip installing the C extension on ReadTheDocs
+    extensions = []
+    packages = ["wobble"]
+    
 else:
+    
     import tensorflow as tf
     compile_flags = tf.sysconfig.get_compile_flags()
     link_flags = tf.sysconfig.get_link_flags()
     
-compile_flags += ["-std=c++11"]
-if sys.platform == "darwin":
-    compile_flags += ["-mmacosx-version-min=10.9"]
+    compile_flags += ["-std=c++11"]
+    if sys.platform == "darwin":
+        compile_flags += ["-mmacosx-version-min=10.9"]
 
-
-extensions = [
-    Extension(
-        "wobble.interp.interp_op",
-        sources=[
-            "wobble/interp/interp_op.cc",
-            "wobble/interp/interp_rev_op.cc",
-        ],
-        include_dirs=["wobble/interp"],
-        language="c++",
-        extra_compile_args=compile_flags,
-        extra_link_args=link_flags,
-    ),
-]
+    extensions = [
+        Extension(
+            "wobble.interp.interp_op",
+            sources=[
+                "wobble/interp/interp_op.cc",
+                "wobble/interp/interp_rev_op.cc",
+            ],
+            include_dirs=["wobble/interp"],
+            language="c++",
+            extra_compile_args=compile_flags,
+            extra_link_args=link_flags,
+        ),
+    ]
+    
+    packages = ["wobble", "wobble.interp"]
 
 setup(
     name="wobble",
@@ -40,7 +44,7 @@ setup(
     author="Megan Bedell",
     author_email="mbedell@flatironinstitute.org",
     description="precise radial velocities with tellurics",
-    packages=["wobble", "wobble.interp"],
+    packages=packages,
     ext_modules=extensions,
     zip_safe=True,
 )

### 2018-12-11 setup.py
Rtd patch (#42)
* Update requirements.txt

* Update conf.py

* Update setup.py

* Update setup.py

@@ -4,34 +4,37 @@ import os
 import sys
 from setuptools import setup, Extension
 
-# Read the docs hack
-on_rtd = os.environ.get('READTHEDOCS', None) == 'True'
-if on_rtd:
-    compile_flags = []
-    link_flags = []
+
+if os.environ.get('READTHEDOCS', None) == 'True':
+    
+    # Skip installing the C extension on ReadTheDocs
+    extensions = []
+    packages = ["wobble"]
+    
 else:
+
     import tensorflow as tf
     compile_flags = tf.sysconfig.get_compile_flags()
     link_flags = tf.sysconfig.get_link_flags()
+    compile_flags += ["-std=c++11"]
+    if sys.platform == "darwin":
+        compile_flags += ["-mmacosx-version-min=10.9"]
+
+    extensions = [
+        Extension(
+            "wobble.interp.interp_op",
+            sources=[
+                "wobble/interp/interp_op.cc",
+                "wobble/interp/interp_rev_op.cc",
+            ],
+            include_dirs=["wobble/interp"],
+            language="c++",
+            extra_compile_args=compile_flags,
+            extra_link_args=link_flags,
+        ),
+    ]
     
-compile_flags += ["-std=c++11"]
-if sys.platform == "darwin":
-    compile_flags += ["-mmacosx-version-min=10.9"]
-
-
-extensions = [
-    Extension(
-        "wobble.interp.interp_op",
-        sources=[
-            "wobble/interp/interp_op.cc",
-            "wobble/interp/interp_rev_op.cc",
-        ],
-        include_dirs=["wobble/interp"],
-        language="c++",
-        extra_compile_args=compile_flags,
-        extra_link_args=link_flags,
-    ),
-]
+    packages = ["wobble", "wobble.interp"]
 
 setup(
     name="wobble",
@@ -40,7 +43,7 @@ setup(
     author="Megan Bedell",
     author_email="mbedell@flatironinstitute.org",
     description="precise radial velocities with tellurics",
-    packages=["wobble", "wobble.interp"],
+    packages=packages,
     ext_modules=extensions,
     zip_safe=True,
 )

### 2018-12-11 docs/conf.py
Update conf.py

@@ -36,7 +36,8 @@ autodoc_mock_imports = [
     "astropy",
     "pymc3",
     "theano",
-    "tensorflow"
+    "tensorflow",
+    "h5py"
 ]
 
 # -- General configuration ---------------------------------------------------

### 2018-12-11 docs/conf.py
Update conf.py

@@ -37,7 +37,8 @@ autodoc_mock_imports = [
     "pymc3",
     "theano",
     "tensorflow",
-    "h5py"
+    "h5py",
+    "tqdm"
 ]
 
 # -- General configuration ---------------------------------------------------

### 2018-12-11 docs/conf.py
Update conf.py

@@ -38,7 +38,9 @@ autodoc_mock_imports = [
     "theano",
     "tensorflow",
     "h5py",
-    "tqdm"
+    "tqdm",
+    "scipy",
+    "matplotlib"
 ]
 
 # -- General configuration ---------------------------------------------------

### 2018-12-11 paper/paper.tex
figure changes, filelist optional in Data
logo and minor edits

@@
 \end{equation}
-\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
+%\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
 where the regularization amplitude on the basis weights is arbitrarily set to unity, and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
@@
 This regularization choice breaks that perfect degeneracy.
-The basis weights are regularized because we want the mean telluric spectrum to do most of the telluric work, and the variable basis to only pick up what it needs to,
-
-%In practice, the most effective regularization amplitudes can vary by orders of magnitude depending on the characteristics of the spectral region being used. 
-%For instance, both $\lambda_{L1, \mu_{t}}$ and $\lambda_{L2, \mu_{t}}$ need to be much higher when the spectra in question span relatively blue wavelengths with few actual telluric features to fit than they do for the near-infrared wavelength regions with strong telluric features.
-%Similarly, the regularization amplitudes for the star tend to be higher for the G dwarf than they are for the M dwarf, which has far more absorption features in its spectra. 
-%Setting the regularization amplitudes is therefore a crucial first step when applying the \wobble model to data.
+The basis weights are regularized to encourage our desired outcome where the mean telluric spectrum contains as much telluric information as possible, while the variable basis picks up only the necessary time-variable changes.
 
@@
 Technically this uncertainty is assumed to be Gaussian in the log space, which is an incorrect approximation, but not very far off for high signal-to-noise data. 
-\todo{BEDELL: Check this:}
 For each order we assign to each log flux an uncertainty variance that is the inverse of the mean signal-to-noise-squared of the pixels in that order (as reported by the \HARPS pipelines), scaled up or down by the raw linear flux observed in that pixel in the data prior to logging and normalizing.
@@
 
-The final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off a drift term specified in the data header, consistent with the standard \HARPS pipeline.
-\todo{(TO DO: make sure this is implemented!)}
+If possible, the final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off an \RV drift term specified in the data header, consistent with the standard \HARPS pipeline.
+This drift is calculated from the simultaneous reference lamp and is therefore only available in certain observing modes. 
+Its contribution to the \RV solution is consistently below 1 \ms, supporting our assumption that the wavelength solution provided by the \HARPS pipeline is generally accurate.
 
@@
 \caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered in the \wobble\ \RVs.
-While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns \todo{(add numbers)}.
-\todo{check on: The subtle trends seen near secondary eclipse (phase 0.75) are suggestive of contributions by reflected light from the planetary dayside.}}
+While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns \todo{(add numbers)}.}
 \label{fig:51peg_planet}
@@
 \includegraphics[width=5in]{barnards_rvs}
-\caption{Radial velocity measurements for \Mdwarf from \wobble (black) and from the standard \HARPS pipeline (red).
+\caption{Radial velocity measurements for \Mdwarf from \wobble (black circles) and from the standard \HARPS pipeline (purple squares).
 The median \RV has been subtracted from each data set.
 The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid blue line (top panel).
-Residuals away from the predicted trend are plotted in the lower panel.
-\todo{(TO DO: add error bars and change color scheme)}}
+Residuals away from the predicted trend are plotted in the lower panel.}
 \label{fig:barnards_rvs}
@@
 After subtracting the secular \RV trend, the residuals have low dispersion, as expected for a quiet star with no planetary signals above $K \sim 1.2~\ms$ \citep{Choi2013, Ribas2018}. 
-The \acronym{RMS} scatter among \wobble\ \RVs is 2.0 \ms. 
-This compares favorably to the scatter of 2.5 \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates.
+The \acronym{RMS} scatter among \wobble\ \RVs is \todo{2.0} \ms. 
+This compares favorably to the scatter of \todo{2.5} \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates.
 It is possible that these deviations are a real physical effect in the stellar spectrum, but we caution that approximately \ms-level errors are likely being introduced by the barycentric correction. 
@@
 Indeed, we do note a strong peak at year-long periods in a Lomb-Scargle periodogram of the residual \RVs. 
-Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble. \todo{(TO DO: recalculate \BERV ourselves)} 
-%\todo{How does HARPS calculate the BERVs?
-If there is something like cirrus, where the apparent brightness of the star varies over the observation, it is plausible that the flux-weighted observation midpoint and the actual time midpoint are different enough to cause a m/s difference in the observations, too.
-Does HARPS have a way of accounting for this?
-(HIRES does estimate this, and they have shorter exposures on average).}
+Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble.
 
@@
 \caption{A subset of the \Mdwarf spectrum including the sodium doublet at 4456-7 \ang.
-The spectral template derived by \wobble is shown as a red line, while a theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as a blue line.
-The \wobble model has been Doppler-shifted to match the rest-frame line positions.
-\todo{(FIX COLORS!)}}
+Data for the highest-\SNR single observation are shown as black points, while the spectral template derived by \wobble from the full data set is shown as a red line.
+The theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as a purple line.
+The \wobble model and data have been Doppler-shifted by eye to match the rest-frame line positions, since \wobble does not deliver an absolute \RV.
+The data-driven \wobble model of the stellar spectrum shares general strong line locations and relative strengths with the \PHOENIX model, but in detail it delivers a much more accurate fit to the data than the theoretical M dwarf model achieves.}
 \label{fig:barnards_model}
@@
 
-%Spectral variability due to LSF changing?
-Additional parameters to the P operators, this would go into the non-convex optimization part (ask Hogg to throw in some words earlier on about location vs linear parameters)
 Another way a less well-calibrated instrument may violate \wobble's assumptions is in the non-negligible variability of all spectral lines due to changes in the instrumental line spread function (\LSF).
@@
 One important feature of \wobble is that it models all fluxes in the log-space. 
-When the model consists solely of multiplicative components, as is the case for a single star + telluric absorption, this choice makes the optimization convex at fixed \RV and computationally tractable \todo{(HOGG: edit plz)}.
+When the model consists solely of multiplicative components, as is the case for a single star + telluric absorption, this choice makes the optimization convex at fixed \RV. 
 If we chose instead to convert the $y$ data vectors to linear fluxes and use a model consisting only of additive components, the same equations and algorithms would apply. 
@@
 To avoid this degeneracy, we might instead represent the stellar spectrum with an explicit parameterization of the line profile and restrict variability to the line depths, widths, and skews; such a change would fall along the same lines as the global \LSF parameterization proposed in Section \ref{s:data-changes}.
-
-\todo{(HOGG: Bayesian version of \wobble?)}
 

### 2018-12-11 scripts/make_data.py
tweaks to data format

@@ -26,13 +26,18 @@ def read_data_from_fits(filelist, e2ds=False):
     ivars = [np.zeros((N,M)) for r in range(R)]
     xs = [np.zeros((N,M)) for r in range(R)]
     empty = np.array([], dtype=int)
-    pipeline_rvs, pipeline_sigs, dates, bervs, airms, drifts = np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N)
+    pipeline_rvs, pipeline_sigmas, dates, bervs, airms, drifts = np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N), np.zeros(N)
     for n,f in enumerate(filelist):
         sp = fits.open(f)
         if not e2ds:
-            pipeline_rvs[n] = sp[0].header['HIERARCH ESO DRS CCF RVC'] * 1.e3 # m/s
-            pipeline_sigs[n] = sp[0].header['HIERARCH ESO DRS CCF NOISE'] * 1.e3 # m/s
-            drifts[n] = sp[0].header['HIERARCH ESO DRS DRIFT SPE RV']  
+            try:
+                pipeline_rvs[n] = sp[0].header['HIERARCH ESO DRS CCF RVC'] * 1.e3 # m/s
+                pipeline_sigmas[n] = sp[0].header['HIERARCH ESO DRS CCF NOISE'] * 1.e3 # m/s
+                drifts[n] = sp[0].header['HIERARCH ESO DRS DRIFT SPE RV']  
+            except KeyError:
+                print("WARNING: {0} does not appear to be a stellar CCF file. Skipping this one.".format(f))
+                empty = np.append(empty, n)
+                continue
         dates[n] = sp[0].header['HIERARCH ESO DRS BJD']        
         bervs[n] = sp[0].header['HIERARCH ESO DRS BERV'] * 1.e3 # m/s
         airms[n] = sp[0].header['HIERARCH ESO TEL AIRM START']
@@ -59,7 +64,7 @@ def read_data_from_fits(filelist, e2ds=False):
         xs[r] = np.delete(xs[r], empty, axis=0)
     
     pipeline_rvs = np.delete(pipeline_rvs, empty)
-    pipeline_sigs = np.delete(pipeline_sigs, empty)
+    pipeline_sigmas = np.delete(pipeline_sigmas, empty)
     dates = np.delete(dates, empty)
     bervs = np.delete(bervs, empty)
     airms = np.delete(airms, empty)
[…]

### 2018-12-11 scripts/script_HD189733.py

@@ -120,4 +120,16 @@ if __name__ == "__main__":
         plt.savefig(plot_dir+'results_rvs.png')
         plt.close(fig)
         
+        fig, (ax, ax2) = plt.subplots(2, 1, gridspec_kw = {'height_ratios':[3, 1]})
+        ax.scatter(data.dates % 2.21857312, data.pipeline_rvs + data.bervs - np.mean(data.pipeline_rvs + data.bervs), c='r', label='DRS', alpha=0.7)
+        ax.scatter(data.dates % 2.21857312, results.star_time_rvs + data.bervs - np.mean(results.star_time_rvs + data.bervs), c='k', label='wobble', alpha=0.7)
+        ax.legend()
+        ax.set_xticklabels([])
+        ax2.scatter(data.dates % 2.21857312, results.star_time_rvs - data.pipeline_rvs, c='k')
+        ax2.set_ylabel('Phase-folded Date')
+        fig.tight_layout()
+        fig.subplots_adjust(hspace=0.05)
+        plt.savefig(plot_dir+'results_rvs_phased.png')
+        plt.close(fig)
+        
     print("total runtime:{0:.2f} minutes".format((time() - start_time)/60.0))

### 2018-12-11 wobble/data.py

@@ -96,14 +96,12 @@ class Data(object):
             self.xs = [np.log(f['xs'][i][self.epochs,:]) for i in orders]
             self.flux_ivars = [f['ivars'][i][self.epochs,:] for i in orders] # ivars for linear fluxes
             self.pipeline_rvs = np.copy(f['pipeline_rvs'])[self.epochs]
+            self.pipeline_sigmas = np.copy(f['pipeline_sigmas'])[self.epochs]
             self.dates = np.copy(f['dates'])[self.epochs]
             self.bervs = np.copy(f['bervs'])[self.epochs]
             self.drifts = np.copy(f['drifts'])[self.epochs]
             self.airms = np.copy(f['airms'])[self.epochs]
-            try:
-                self.filelist = [a.decode('utf8') for a in np.copy(f['filelist'])[self.epochs]]
-            except KeyError:
-                print("WARNING: data missing filelist.")
+            self.filelist = [a.decode('utf8') for a in np.copy(f['filelist'])[self.epochs]]
             self.R = len(orders) # number of orders
             self.orders = orders # indices of orders in origin_file
             self.ivars = [self.fluxes[i]**2 * self.flux_ivars[i] for i in range(self.R)] # ivars for log(fluxes)

### 2018-12-11 wobble/results.py

@@ -7,7 +7,7 @@ T = tf.float64
 from .utils import get_session
 
 COMMON_ATTRS = ['R', 'N', 'orders', 'origin_file', 'epochs', 'component_names', 
-                'bervs', 'pipeline_rvs', 'drifts', 'dates', 'airms']
+                'bervs', 'pipeline_rvs', 'pipeline_sigmas', 'drifts', 'dates', 'airms'] # common across all orders
 COMPONENT_NP_ATTRS = ['K', 'r', 'rvs_fixed', 'ivars_rvs', 'scale_by_airmass', 'learning_rate_rvs', 
                       'learning_rate_template', 'L1_template', 'L2_template']
 OPT_COMPONENT_NP_ATTRS = ['learning_rate_basis', 'L1_basis_vectors', 'L2_basis_vectors', 'L2_basis_weights'] # it's ok if these don't exist
@@ -42,6 +42,7 @@ class Results(object):
             # get other convenient things
             self.bervs = data.bervs
             self.pipeline_rvs = data.pipeline_rvs
+            self.pipeline_sigmas = data.pipeline_sigmas
             self.dates = data.dates
             self.airms = data.airms
             self.drifts = data.drifts

### 2018-12-12 paper/paper.tex
minor figure tweaks and to-dos

@@
 where $v_n$ and $v_r$ are the characteristic \RVs at epoch $n$ and order $r$, $\sigma_{n,r}$ is the estimated measurement uncertainty on $v_{n,r}$, and $\delta_r$ is an additional jitter term specific to order $r$.
-
-\todo{(To do: fill in details: prior on $<v_r>$, optimization over $\delta_r$ with $v_n$ and $v_r$ analytically solved for.)}
+The order-dependent \RV offset should be small (and indeed generally is consistent with zero in the solutions), but is included in the model to capture any color-dependent effects such as non-negligible atmospheric variability during the exposure \citep[see e.g.][]{Blackwell2017}.
 
@@
 
-All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra. 
+All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra.\footnote{Based on observations made with ESO Telescopes at the La Silla Paranal Observatory under programme IDs 091.C-0271, 183.C-0437, 191.C-0505, 072.C-0488, 089.C-0497, 099.C-0880, and 60.A-9700.}
 These data come as extracted 1D spectra, in blocks of 72 orders by 4096 pixels per order.
@@
 This drift is calculated from the simultaneous reference lamp and is therefore only available in certain observing modes. 
-Its contribution to the \RV solution is consistently below 1 \ms, supporting our assumption that the wavelength solution provided by the \HARPS pipeline is generally accurate.
+Its contribution to the \RV solution is usually below 1 \ms, supporting our assumption that the wavelength solution provided by the \HARPS pipeline is generally accurate.
 
@@
 
-We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:329968 and mbedell:331551.}
+We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.
 The 91 publicly available spectra in the \HARPS archive are largely concentrated on a few nights of intensive observing, but these nights are sufficiently spread out throughout the year for a wide enough range in \BERV to disentangle the stellar and telluric spectra.
@@
 
-Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the correct period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}).
+Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}).
 \todo{(numbers here, address the structure in residuals)}
@@
 
-Next, we tested \wobble with 306 epochs of \Mdwarf spectra.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:359445 and mbedell:359501.} 
+Next, we tested \wobble with 306 epochs of \Mdwarf spectra.
 \Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50. 
@@
 We apply \wobble to a data set consisting of a single transit of the hot Jupiter HD 189733b to test its sensitivity to stellar line asymmetries as well as its performance in the regime of a single night's observations.
-The observations in question consist of a consecutive series of 40 spectra, each with an \SNR $\sim 90$, taken on the night of August 28, 2007.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers \todo{x}.}
+The observations in question consist of a consecutive series of 40 spectra, each with an \SNR $\sim 90$, taken on the night of August 28, 2007.
 
@@
 The upper panels show the \wobble best-fit telluric spectrum (blue line), best-fit stellar spectrum (red line), data (black points), and the sum of the telluric and stellar model predictions (black line).
-Residuals are shown in the lower panels.
-Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.
-\todo{(TO DO: add y-label, fix centering)}}
+Residuals to the fit are shown in the lower panels.
+Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
 \label{fig:telluric_comparison}
@@
 
-As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{HR 3090 (+details)}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:385782 and \todo{xxx}.}
+As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star HR 3090 taken by \HARPS at an \SNR of 130. 
 As shown in Figure \ref{fig:telluric_standard}, the spectra agree well. 
@@
 Another key assumption made in this work that there is no gas cell represented in the data. 
-While this is true for \HARPS, other instruments like \acronym{HIRES}, \acronym{APF}, and \acronym{PFS} include a gas cell set in optical path to imprint its absorption spectrum on the observed spectra \todo{(cite)}.
+While this is true for \HARPS, other instruments such as \acronym{HIRES}, \acronym{PFS}, and \acronym{APF} include a gas cell set in optical path to imprint its absorption spectrum on the observed spectra \citep{Butler1996, Crane2010, Vogt2014}.
 Applying \wobble to spectra that include the imprint of a gas cell is technically trivial. 
@@
 This effective spectrograph \RV may vary in different spectral orders or smaller wavelength regions.
-The most extreme extension would be to include the full wavelength solution as model parameters, and optimize them along with the spectral components and radial velocities.
-\todo{(TO DO: add some citations to using tellurics for this)}
+The most extreme extension would be to include corrections to the wavelength solution as model parameters, and optimize them along with the spectral components and radial velocities. 
+This would rely on the use of telluric features as fixed calibration sources \citep{Seifahrt2010}.
 
@@
 
-Another choice we made which is almost certainly suboptimal is the treatment of each spectral order in the echelle data as an independent spectrum with its own stellar \RV. 
-In reality, every \RV should be informed by all spectral orders simultaneously. 
+Another choice we made which is likely suboptimal is the treatment of each spectral order in the echelle data as an independent spectrum with its own stellar \RV. 
+In reality, every stellar \RV should be informed by all spectral orders simultaneously. 
 Implementing this would take on some extra computational cost, but being in \TF will help to mitigate this. 
+However, the reality of color-dependent atmospheric dispersion effects means that in practice each order may have a slightly different effective \RV, since the flux-weighted average observation times will be different \citep{Blackman2017}.
+Thus for some situations allowing the orders to have independent \RV estimates may be the best choice.
 
@@
 
-\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
+\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
 We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.

### 2018-12-14 paper/paper.tex
attempting to implement exoplanet

@@
 %While it has primarily been used in the context of deep learning in the astronomical literature, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
-The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 spectra of 51 Peg runs in \todo{30} minutes on a standard Mac desktop.
+The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 spectra of 51 Peg runs in 60 minutes on a standard Mac desktop.
 
@@
 This has the side effect of flattening out weak features if no strong lines are present in a given order. 
-As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the top panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum} \todo{-- To do: check if there's a better example order}).
+As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the middle panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum}). 
 In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}). 

### 2018-12-18 paper/paper.tex

@@
 \includegraphics[width=5in]{hd189_rm}
-\caption{\RVs measured by \wobble for a single night of observations of the Hot Jupiter host HD 189733. The \RM is clearly seen as the planet transits the star. \todo{[blah blah starry blah]}}
+\caption{\RVs measured by \wobble for a single night of observations of the Hot 
+         Jupiter host HD 189733. The \RM is clearly seen as the planet transits 
+         the star. The top panel shows the data (black points), the maximum
+         likelihood fit using \starry (black curve), and 500 posterior samples
+         (orange lines). The bottom panel shows the residuals of the maximum
+         likelihood model fit.}
 \label{fig:hd189_rm}
@@
 The resulting \RV signal retrieved by \wobble is shown in Figure \ref{fig:hd189_rm}. 
-We fit the signal using \starry \citep{Luger2018}. \todo{(RODRIGO: fill in a bit about methods)} 
-\todo{(comparison of retrieved parameters to previous work here)}
+We fit the signal using \starry \citep{Luger2018}, which computes analytic occultation
+light curves for bodies whose surfaces can be decomposed into sums
+of spherical harmonics. Since the radial component of the velocity field of a 
+differentially rotating star can be expressed in terms of polynomials in 
+$x = \sin\theta\cos\phi$, $y = \sin\theta\sin\phi$, and $z=\cos\theta$, where
+$\theta$ is the polar angle and $\phi$ is the azimuthal angle \citep[c.f. Equation 91 in][]{Short2018},
+this velocity field may be expressed exactly in terms of spherical harmonics
+\citep{LugerBedell2019}.
+We therefore use \starry to fit for the equatorial rotational velocity, the inclination,
+the obliquity, and the shear due to differential rotation of the star, closely
+following the analysis performed in \citep{Cegla2016}.
+We infer a projected stellar obliquity $\lambda = -0.43 \pm 0.34^\circ$, in close agreement
+with the value reported in \citet{Cegla2016}. Our inferred values for the other parameters
+are broadly consistent with the results in \citet{Cegla2016}, except with significantly
+higher uncertainty due to the fact that we are unable to constrain the stellar inclination
+due to the $v\sin i$ degeneracy. We attribute the narrower posteriors in \citet{Cegla2016}
+as a difference in the choice of prior.
+\footnote{See \href{https://github.com/megbedell/wobble/tree/master/scripts/DifferentialRotationWithSphericalHarmonics.ipynb}{this interactive notebook} 
+          for a derivation of the spherical harmonic decomposition of the stellar velocity field. 
+          The notebook used to perform our analysis and produce Figure~\ref{fig:hd189_rm} can be found 
+          \href{https://github.com/megbedell/wobble/tree/master/scripts/HD189733bWithStarry.ipynb}{here}.}
 

### 2018-12-18 paper/paper.tex
minor updates to RV fits

--- 2018-12-14 paper/paper.tex
@@
 These orders are often treated as independent spectra when extracting \RVs, and the \RVs for each order are then combined in some manner to get a final time series. 
-
 We follow this precedent and optimize the \wobble model individually for each order. 
-While some \RV precision gain may be possible by inferring a global Doppler shift from all orders simultaneously, it is not clear that the gain would be worth the added computational expense, so we leave this for future work.
+%While some \RV precision gain may be possible by inferring a global Doppler shift from all orders simultaneously, it is not clear that the gain would be worth the added computational expense, so we leave this for future work.
 
@@
 where $v_n$ and $v_r$ are the characteristic \RVs at epoch $n$ and order $r$, $\sigma_{n,r}$ is the estimated measurement uncertainty on $v_{n,r}$, and $\delta_r$ is an additional jitter term specific to order $r$.
-The order-dependent \RV offset should be small (and indeed generally is consistent with zero in the solutions), but is included in the model to capture any color-dependent effects such as non-negligible atmospheric variability during the exposure \citep[see e.g.][]{Blackwell2017}.
+The order-dependent \RV offset should be small and indeed generally is consistent with zero in the solutions.%, but is included in the model to capture any color-dependent effects such as non-negligible atmospheric variability during the exposure \citep[see e.g.][]{Blackman2017}.
 
@@
 %While it has primarily been used in the context of deep learning in the astronomical literature, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
-The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 spectra of 51 Peg runs in 60 minutes on a standard Mac desktop.
+The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 \HARPS spectra of 51 Peg, including the optimization of 72 spectral orders, runs in 60 minutes on a standard Mac desktop.
 
@@
 Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). 
-\todo{(numbers here, address the structure in residuals)}
+We fit a Keplerian signal with six free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_p$, and \RV offset $c$) using a simple $\chi^2$ optimization with \code{scipy.optimize.leastsq}.
+The resulting best-fit parameters for the \wobble \RVs are consistent with the literature values: $P = 4.2308$ d, $K = 55.9 \ms$, $e = 0.05$
 
@@
 \includegraphics[width=4in]{51peg_planet}
-\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered in the \wobble\ \RVs.
-While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns \todo{(add numbers)}.}
+\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered from the \wobble\ \RVs (left) and from the \HARPS standard pipeline.
+While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns.
+Best-fit Keplerians for each data set are overplotted.}
 \label{fig:51peg_planet}
@@
 
-Next, we tested \wobble with 306 epochs of \Mdwarf spectra.
+Next, we tested \wobble with 237 epochs of \Mdwarf spectra. 
 \Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50. 
@@
 After subtracting the secular \RV trend, the residuals have low dispersion, as expected for a quiet star with no planetary signals above $K \sim 1.2~\ms$ \citep{Choi2013, Ribas2018}. 
-The \acronym{RMS} scatter among \wobble\ \RVs is \todo{2.0} \ms. 
-This compares favorably to the scatter of \todo{2.5} \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates.
+The \acronym{RMS} scatter among \wobble\ \RVs is 2.0 \ms. 
+This compares favorably to the scatter of 2.5 \ms among \RVs produced by the standard closed-source \HARPS pipeline.
+
+We note that many of the residuals have similar non-zero values in both independently produced RV estimates. 
 It is possible that these deviations are a real physical effect in the stellar spectrum, but we caution that approximately \ms-level errors are likely being introduced by the barycentric correction. 

### 2018-12-20 paper/paper.tex
fitting 51peg with exoplanet

@@
-\documentclass[modern]{aastex62}
+\documentclass[twocolumn]{aastex62}
 
@@
 Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
-In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself. 
+In both cases, this precision-limiting reliance on external information can be sidestepped using the data itself. 
 % Aims
@@
 Such data-driven templates have the benefit of being both customized to the star in question and high-\SNR without requiring observing overhead.
-%\todo{(explain how this differs from what we'll do - v similar in spirit although ours is a full forward model with stellar spectrum as a latent variable)}
+
+Common to the issues of telluric correction and stellar template building is the fact that all of the necessary information about the unknown spectra is encoded in the data. 
+With a simple data-driven model, one could learn both stellar and telluric spectra simultaneously with the stellar \RVs.
 
@@
 
-Here we develop a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch. 
+Here we develop and implement such a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch. 
 The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
-Our model requires no prior knowledge of the star or the Earth's atmosphere. 
+Our model requires no prior knowledge of spectral features for the star or for the Earth's atmosphere. 
 As such, it does not yield absolute measurements of \RVs, only highly precise relative measurements between epochs.
@@
 In Section \ref{s:methods}, we outline the model and its key underlying assumptions. 
-We present an open-source implementation of this method in \python and \TF
-\citep{Abadi15} called \wobble.
+We present an open-source implementation of this method in \python and \TF called \wobble.
 In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the canonical planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
 We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}. 
-We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation. 
-We outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data. 
+We revisit many of the assumptions underlying \wobble in Section \ref{s:future} and outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data.
 Finally, we conclude with a brief summary in Section \ref{s:conclusion}.
@@
 
-For practical applications, the number of epochs $N$ must be large ($N \gg N_{components}$). 
-This is a fundamental restriction rooted in the fact that every spectrum being modeled introduces a large number of free parameters to the model. 
-In most cases, these epochs will also need to span a significant fraction of the observing season. 
+To make our method practically feasible, we assume that the number of observations $N$ in the data set is large ($N \gg N_{components}$) and spread out across the observing season. 
+The quantity of spectra needed is a fundamental restriction rooted in the fact that every spectrum being modeled introduces a large number of free parameters to the model. 
+In most cases, the epochs of these observations will also need to span a significant fraction of the observing season. 
 In order to disentangle the stellar and telluric spectral components, the spectra must undergo Doppler shifts with respect to each other that are at least as large as a resolution element of the spectrograph.
 For stars which do not physically undergo large \RV shifts over very short timescales, this means that it is necessary to observe the star over a significant fraction of the year to take advantage of the changes in the Earth's projected motion.
-We will comment on ways to overcome this restriction in Section \ref{s:future}.
+We will comment on ways to overcome these restrictions in Section \ref{s:future}.
 
@@
 
-We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
+We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. orange
 We also have a corresponding $M \times N$ matrix of wavelength solutions. 
@@
 \begin{equation}
-y_{n}^\star = P(v_{obs, n}) \mu_{\star},
+y_{n}^\star = P(v^{obs}_{n}) \mu^{\star},
 \label{eqn:star}
 \end{equation}
-where $\mu_{\star}$ is a spectral template with length $M'$ corresponding to
+where $\mu^{\star}$ is a spectral template with length $M'$ corresponding to
 the length of an arbitrarily-chosen wavelength grid of control points, and
 $P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
-interpolates $\mu_\star$ into the observatory rest frame.
+interpolates $\mu^\star$ into the observatory rest frame.
 Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
@@
 
-The apparent stellar \RV, $v_{obs, n}$, is a sum of the star's actual velocity about its center of mass ($v_{\star, n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
+The apparent stellar \RV, $v^{obs}_{n}$, is a sum of the star's actual velocity about its center of mass ($v^{\star}_{n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
 
@@
 \begin{equation}
-y_{n}^t =  A_n(\mu_{t} + W_{t} z_n).
+y_{n}^t =  A_n(\mu^{t} + W^{t} z_n).
 \end{equation}
-In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
+In addition to its mean spectrum $\mu^t$, the telluric component also includes a time-dependent term assembled from two variables: $W^t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
 $W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
@@
 \end{equation}
-where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.
+where $p$ is the vector of parameters to be normalized (in this case $\mu^{\star}$, $\mu^{t}$, or $W^{t}$) and $\lambda$ is the regularization amplitude.
 
@@
 \begin{split}
-\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}} \\
- &  + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}}  + \Lone{\lambda_5}{W_t}  + \Ltwo{\lambda_6}{W_t} + \Lone{1.0}{z_t} ,
+\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  \\
+ & + \Lone{\lambda_1}{\mu^{\star}} + \Ltwo{\lambda_2}{\mu^{\star}}   + \Lone{\lambda_3}{\mu^{t}} + \Ltwo{\lambda_4}{\mu^{t}} \\
+ &  + \Lone{\lambda_5}{W^t}  + \Ltwo{\lambda_6}{W^t} + \Lone{1.0}{z^t} ,
 \end{split}
@@
 
-As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs. 
-We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
+As an initial guess, we set the star to be stationary, e.g. $v^{obs} = $ \BERV at all epochs. 
+We may then initialize the stellar spectrum $\mu^{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
 The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
@@
 We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
-This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iteration to convergence.
+This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iterations to convergence.
 
@@
 
-The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF. 
+The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF \citep{Abadi15}. 
 \TF is a model building framework that has been primarily designed for machine
@@
 
+\begin{figure*}[ht!]
+\centering
+\includegraphics[width=5in]{51peg_planet}
+\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered from the \wobble\ \RVs (left) and from the \HARPS standard pipeline.
+While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns.
+Maximum-likelihood Keplerians (solid orange lines) and 1-$\sigma$ posterior constraints (shaded orange regions) for each data set are overplotted.}
+\label{fig:51peg_planet}
+\end{figure*}
+
+\begin{deluxetable*}{cccc}
+\tablehead{
+\colhead{Parameter} & \colhead{Units} & \colhead{\wobble Fit} & \colhead{\HARPS Pipeline Fit}}
+\tablecaption{Orbital Parameters of 51 Pegasi b}
+\label{tbl:51peg}
+\startdata
+$K$ & m s$^{-1}$ & $55.57^{+2.28}_{-2.04}$ & $53.84^{+1.96}_{-2.20}$ \\
+$P$ & days & $4.2292^{+0.0003}_{-0.0003}$ & $4.2294^{+0.0003}_{-0.0003}$ \\
+$t_0$ & JD & $2456546.89^{+0.02}_{-0.02}$ & $2456546.94^{+0.01}_{-0.02}$ \\
+$e$ & $-$  & $0.03^{+0.02}_{-0.02}$ & $0.04^{+0.02}_{-0.02}$ \\
+$\omega$ & rad & $0.45^{+0.62}_{-1.00}$ & $-1.79^{+0.34}_{-0.32}$ \\
+$s$ & m s$^{-1}$ & $0.01^{+0.08}_{-0.01}$ & $0.74^{+0.06}_{-0.07}$ \\
+$m_p$ sin$i$ & $M_J$ & $0.41^{+0.02}_{-0.02}$ & $0.40^{+0.01}_{-0.02}$ \\
+\enddata
+\end{deluxetable*}
+
 For \wobble's first test, we chose the first known exoplanet host: 51 Pegasi. 
@@
 Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). 
-We fit a Keplerian signal with six free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_p$, and \RV offset $c$) using a simple $\chi^2$ optimization with \code{scipy.optimize.leastsq}.
-The resulting best-fit parameters for the \wobble \RVs are consistent with the literature values: $P = 4.2308$ d, $K = 55.9 \ms$, $e = 0.05$
+We fit a Keplerian signal with seven free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_0$, \RV offset $c$, and \RV jitter $s$) using the \code{exoplanet} package \citep{exoplanet}.
+The resulting best-fit parameters for the \wobble\ \RVs are generally consistent with expectations based on literature values (Table \ref{tbl:51peg}). 
+The \RV uncertainties derived by \wobble appear accurate based on the negligible jitter in the fit, while the photon-noise-based \RV errors provided by the \HARPS pipeline are much smaller and require a significant jitter to achieve a good fit.
 
-Fits to an individual spectrum are shown in Figure \ref{fig:51peg_spectrum}. 
-We emphasize that no \textit{a priori} information on e.g. expected spectral line positions and depths were used. 
-The current realization of the \wobble algorithm treats each control point of the template spectra as independent and has no line shape parameterization included. 
-Nevertheless, the optimized stellar templates clearly reproduce the expected appearance of the spectra in a variety of regimes, from a crowded-line region to a sparser, continuum-dominated region to an extremely strong absorption line like the H$\alpha$ feature.
-
-\begin{figure}
-\centering
-\includegraphics[width=4in]{51peg_planet}
-\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered from the \wobble\ \RVs (left) and from the \HARPS standard pipeline.
-While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns.
-Best-fit Keplerians for each data set are overplotted.}
-\label{fig:51peg_planet}
-\end{figure}
-
-\begin{figure}
+\begin{figure*}[t]
 \centering
@@
 \label{fig:51peg_spectrum}
-\end{figure}
+\end{figure*}
+
+Fits to an individual spectrum are shown in Figure \ref{fig:51peg_spectrum}. 
+We emphasize that no \textit{a priori} information on e.g. expected spectral line positions and depths were used. 
+The \wobble algorithm as it is currently implemented treats each control point of the template spectra as independent and has no line shape parameterization included. 
+Nevertheless, the optimized stellar templates clearly reproduce the expected appearance of the spectra in a variety of regimes, from a crowded-line region to a sparser, continuum-dominated region to an extremely strong absorption line like the H$\alpha$ feature.
+Moreover, telluric absorption features are recovered down to a very small amplitude.
 
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=5in]{barnards_rvs}
+\caption{Radial velocity measurements for \Mdwarf from \wobble (black circles) and from the standard \HARPS pipeline (blue squares).
+The median \RV has been subtracted from each data set.
+The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid orange line (top panel).
+Residuals away from the predicted trend are plotted in the lower panel.}
+\label{fig:barnards_rvs}
+\end{figure*}
+
 Next, we tested \wobble with 237 epochs of \Mdwarf spectra. 
+We did not utilize the additional spectra taken after the \HARPS 2015 optical fiber upgrade, as these would need to be treated as an independent data set \citep{LoCurto2015}. 
 \Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50. 
@@
 Aside from this linear trend, \Mdwarf is commonly used as an \RV standard M dwarf because little stellar activity has been observed and no planets discovered until very recently despite considerable \RV monitoring \citep{Ribas2018}.
-
-\begin{figure}
-\centering
-\includegraphics[width=5in]{barnards_rvs}
-\caption{Radial velocity measurements for \Mdwarf from \wobble (black circles) and from the standard \HARPS pipeline (purple squares).
-The median \RV has been subtracted from each data set.
-The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid blue line (top panel).
-Residuals away from the predicted trend are plotted in the lower panel.}
-\label{fig:barnards_rvs}
-\end{figure}
 
@@
 
-Testing \wobble on the \Mdwarf data allows us to evaluate its performance in a significantly different regime: the mid-M dwarf spectra are far denser in spectral features than a Sun-like star, and individual observations are at a much lower \SNR than the 51 Peg observations used above, although the total number of spectra is greater.
-The resulting spectral fit is shown in Figure \ref{fig:barnards_spectrum}.
-
-\begin{figure}
+\begin{figure*}[ht!]
 \centering
@@
 \label{fig:barnards_spectrum}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \centering
@@
 Data for the highest-\SNR single observation are shown as black points, while the spectral template derived by \wobble from the full data set is shown as a red line.
-The theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as a purple line.
+The theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as an orange line.
 The \wobble model and data have been Doppler-shifted by eye to match the rest-frame line positions, since \wobble does not deliver an absolute \RV.
@@
 \label{fig:barnards_model}
-\end{figure}
+\end{figure*}
+
+Testing \wobble on the \Mdwarf data allows us to evaluate its performance in a significantly different regime: the mid-M dwarf spectra are far denser in spectral features than a Sun-like star, and individual observations are at a much lower \SNR than the 51 Peg observations used above, although the total number of spectra is greater.
+The resulting spectral fit is shown in Figure \ref{fig:barnards_spectrum}.
 
@@
 The \wobble model for \Mdwarf diverges from the theoretical models in the smaller absorption lines and in the degree of broadening for strong lines.  
-\todo{(TO DO: are these known model weaknesses? ask Travis Barman about opacity binning affecting the detailed features)}
+%\todo{(TO DO: are these known model weaknesses? ask Travis Barman about opacity binning affecting the detailed features)}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
@@
 \label{fig:hd189_rm}
-\end{figure}
+\end{figure*}
 
@@
 These results confirm that the \wobble method measures line asymmetries as \RV shifts in the same manner as traditional \RV analysis techniques. 
-While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{}.
-\todo{citep}
+While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{Queloz2001}. 
 The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified to fit line asymmetries and true Doppler shifts as separate parameters. 
@@
 
-\begin{figure}
+\begin{figure*}[h!]
 \centering
@@
 \label{fig:telluric_comparison}
-\end{figure}
+\end{figure*}
 
@@
 \label{fig:telluric_standard}
-\end{figure}
+\end{figure*}
 
@@
 \label{fig:telluric_basis}
-\end{figure}
+\end{figure*}
 
@@
     scipy \citep{Jones:2001},
-    tensorflow \citep{Abadi15}
+    tensorflow \citep{Abadi15}, 
+    exoplanet \citep{exoplanet}
 }

### 2018-12-20 paper/paper.tex
minor edits, two-column
Former-commit-id: fb302b07e950c4aefd8076a9a750ce27f1d8575a

@@
 the obliquity, and the shear due to differential rotation of the star, closely
-following the analysis performed in \citep{Cegla2016}.
+following the analysis performed in \citet{Cegla2016}.
 We infer a projected stellar obliquity $\lambda = -0.43 \pm 0.34^\circ$, in close agreement
@@
 due to the $v\sin i$ degeneracy. We attribute the narrower posteriors in \citet{Cegla2016}
-as a difference in the choice of prior.
+to a difference in the choice of prior.
 \footnote{See \href{https://github.com/megbedell/wobble/tree/master/scripts/DifferentialRotationWithSphericalHarmonics.ipynb}{this interactive notebook} 
@@
 
-Even working within \wobble's current assumptions and running on \HARPS data, there are potential changes to be made to the implementation of \wobble that may improve its performance. 
+Even working within \wobble's current assumptions and running on \HARPS-like data, there are potential changes to be made to the implementation of \wobble that may improve its performance. 
 Many of these are quite basic and straightforward changes to be made within the existing framework of \wobble. 
-We discuss some of potential adjustments here.
+We discuss some potential adjustments here.
 
@@
 However, setting the regularization amplitudes optimally is by far the computationally slowest and most unwieldy part of applying \wobble to a new data set. 
-In principle, as data sets get larger, regularization becomes less necessary. 
-One way to remove the need for telluric regularization might be to run on a data set that is composed of many different stars, each with their own individual spectra and \RVs but with a shared telluric component, so that the data informing the telluric lines is sufficiently large.
-This could be done with relatively trivial changes to the \wobble code, although it would require strong computational resources.
+Dealing with the possibility of overfitting is an inevitable requirement of any model, especially one that relies on as many free parameters as \wobble. 
+That being said, in principle as data sets get larger, regularization becomes less crucial. 
+One way to lessen the need for telluric regularization might be to run on a data set that is composed of many different stars, each with their own individual spectra and \RVs but with a shared telluric component, so that the data informing the telluric lines is sufficiently large.
+This could be done with relatively trivial changes to the \wobble code, although it would require considerable computational resources.
 
 One choice that \wobble makes differently from other data-driven \RV codes is the way we represent the underlying spectral components. 
-We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated by the $P$ operator. 
+We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated. 
 One possible improvement would be to go to a higher-order interpolation scheme. 

### 2018-12-21 scripts/ylm_rot.py
fixed starry normalization issue

@@ -1,7 +1,20 @@
 import numpy as np
 
 
-def get_ylm_coeffs(inc=90, obl=0, alpha=0):
+def rv_field(x, y, z, veq=1, inc=90, obl=0, alpha=0):
+    """Return the radial velocity at a point on the stellar surface."""
+    cosi = np.cos(inc * np.pi / 180)
+    sini = np.sin(inc * np.pi / 180)
+    cosl = np.cos(obl * np.pi / 180)
+    sinl = np.sin(obl * np.pi / 180)
+    A = sini * cosl
+    B = sini * sinl
+    C = cosi
+    return veq * (A * x + B * y) * (1 - alpha * (-B * x + A * y + C * z) ** 2)
+
+
+def get_ylm_coeffs(veq=1, inc=90, obl=0, alpha=0):
+    """Return the normalized spherical harmonic coefficients for the RV field of a star."""
     cosi = np.cos(inc * np.pi / 180)
     sini = np.sin(inc * np.pi / 180)
     cosl = np.cos(obl * np.pi / 180)
@@ -10,18 +23,18 @@ def get_ylm_coeffs(inc=90, obl=0, alpha=0):
     B = sini * sinl
     C = cosi
     return np.array([0,
-        np.sqrt(3)*np.pi*B*(-A**2*alpha - B**2*alpha - C**2*alpha + 5)/15,
-        0,
-        np.sqrt(3)*np.pi*A*(-A**2*alpha - B**2*alpha - C**2*alpha + 5)/15,
-        0,
-        0,
-        0,
-        0,
-        0,
-        np.sqrt(70)*np.pi*B*alpha*(3*A**2 - B**2)/70,
-        2*np.sqrt(105)*np.pi*C*alpha*(-A**2 + B**2)/105,
-        np.sqrt(42)*np.pi*B*alpha*(A**2 + B**2 - 4*C**2)/210,
-        0,
-        np.sqrt(42)*np.pi*A*alpha*(A**2 + B**2 - 4*C**2)/210,
-        4*np.sqrt(105)*np.pi*A*B*C*alpha/105,
-        np.sqrt(70)*np.pi*A*alpha*(A**2 - 3*B**2)/70])
+                     np.sqrt(3)*np.pi*B*(-A**2*alpha - B**2*alpha - C**2*alpha + 5)/15,
+                     0,
+                     np.sqrt(3)*np.pi*A*(-A**2*alpha - B**2*alpha - C**2*alpha + 5)/15,
+                     0,
[…]

### 2018-12-21 paper/paper.tex
HD189 figure tweaks
Former-commit-id: 331d656716689373cfb02513a674aa70f96b03fe

@@
 $s$ & m s$^{-1}$ & $0.01^{+0.08}_{-0.01}$ & $0.74^{+0.06}_{-0.07}$ \\
-$m_p$ sin$i$ & $M_J$ & $0.41^{+0.02}_{-0.02}$ & $0.40^{+0.01}_{-0.02}$ \\
 \enddata
@@
 We fit a Keplerian signal with seven free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_0$, \RV offset $c$, and \RV jitter $s$) using the \code{exoplanet} package \citep{exoplanet}.
-The resulting best-fit parameters for the \wobble\ \RVs are generally consistent with expectations based on literature values (Table \ref{tbl:51peg}). 
+The resulting best-fit parameters for the two data sets are generally consistent (Table \ref{tbl:51peg}), and both are comparable to literature values \citep{Mayor1995, Naef2004, Butler2006}. 
 The \RV uncertainties derived by \wobble appear accurate based on the negligible jitter in the fit, while the photon-noise-based \RV errors provided by the \HARPS pipeline are much smaller and require a significant jitter to achieve a good fit.
+These results confirm that \wobble is able to extract \RVs with similar precision to the closed-source \HARPS pipeline for Sun-like stars.
 
@@
 \centering
-\includegraphics[width=5in]{hd189_rm}
+\includegraphics[width=6in]{hd189_rm}
 \caption{\RVs measured by \wobble for a single night of observations of the Hot 
@@
           for a derivation of the spherical harmonic decomposition of the stellar velocity field. 
-          The notebook used to perform our analysis and produce Figure~\ref{fig:hd189_rm} can be found 
+          The notebook used to perform our analysis and produce Figure~\ref{fig:hd189_rm}, along with the full posterior constraints we obtain, can be found 
           \href{https://github.com/megbedell/wobble/tree/master/scripts/HD189733bWithStarry.ipynb}{here}.}

### 2018-12-24 paper/paper.tex
pymc3 + theano references
Former-commit-id: 6e69f649f178d1cded6cfc3890ecf96e04a34928

@@
     tensorflow \citep{Abadi15}, 
-    exoplanet \citep{exoplanet}
+    exoplanet \citep{exoplanet},
+    pymc3 \citep{pymc3},
+    theano \citep{theano}
 }

### 2019-01-02 paper/paper.tex
minor edits
Former-commit-id: 86b55d216bc3795ee8c319b02702d7a1adbba712

--- 2018-12-21 paper/paper.tex
@@
 
-We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. orange
+We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
 We also have a corresponding $M \times N$ matrix of wavelength solutions. 
@@
          the star. The top panel shows the data (black points), the maximum
-         likelihood fit using \starry (black curve), and 500 posterior samples
+         likelihood fit using \starry (black line), and 500 posterior samples
          (orange lines). The bottom panel shows the residuals of the maximum
@@
     tensorflow \citep{Abadi15}, 
-    exoplanet \citep{exoplanet}
+    exoplanet \citep{exoplanet},
+    starry \citep{Luger2018}
 }

### 2019-01-02 paper/paper.tex

@@
     exoplanet \citep{exoplanet},
-    starry \citep{Luger2018}
+    starry \citep{Luger2018},
+    pymc3 \citep{pymc3},
+    theano \citep{theano}
 }

### 2019-01-02 paper/paper.tex
words

--- 2019-01-02 paper/paper.tex
@@
-\documentclass[twocolumn]{aastex62}
+\documentclass[modern]{aastex62}
 
@@
 Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
-In both cases, this precision-limiting reliance on external information can be sidestepped using the data itself. 
+In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself. 
 % Aims
@@
 Such data-driven templates have the benefit of being both customized to the star in question and high-\SNR without requiring observing overhead. 
-
-Common to the issues of telluric correction and stellar template building is the fact that all of the necessary information about the unknown spectra is encoded in the data. 
-With a simple data-driven model, one could learn both stellar and telluric spectra simultaneously with the stellar \RVs.
+%\todo{(explain how this differs from what we'll do - v similar in spirit although ours is a full forward model with stellar spectrum as a latent variable)}
 
@@
 
-Here we develop and implement such a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch. 
+Here we develop a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch. 
 The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
-Our model requires no prior knowledge of spectral features for the star or for the Earth's atmosphere. 
+Our model requires no prior knowledge of the star or the Earth's atmosphere. 
 As such, it does not yield absolute measurements of \RVs, only highly precise relative measurements between epochs.
@@
 In Section \ref{s:methods}, we outline the model and its key underlying assumptions. 
-We present an open-source implementation of this method in \python and \TF called \wobble.
+We present an open-source implementation of this method in \python and \TF
+\citep{Abadi15} called \wobble.
 In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the canonical planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
 We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}.
-We revisit many of the assumptions underlying \wobble in Section \ref{s:future} and outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data.
+We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation. 
+We outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data. 
 Finally, we conclude with a brief summary in Section \ref{s:conclusion}.
@@
 
-To make our method practically feasible, we assume that the number of observations $N$ in the data set is large ($N \gg N_{components}$) and spread out across the observing season. 
-The quantity of spectra needed is a fundamental restriction rooted in the fact that every spectrum being modeled introduces a large number of free parameters to the model. 
-In most cases, the epochs of these observations will also need to span a significant fraction of the observing season. 
+For practical applications, the number of epochs $N$ must be large ($N \gg N_{components}$). 
+This is a fundamental restriction rooted in the fact that every spectrum being modeled introduces a large number of free parameters to the model. 
+In most cases, these epochs will also need to span a significant fraction of the observing season. 
 In order to disentangle the stellar and telluric spectral components, the spectra must undergo Doppler shifts with respect to each other that are at least as large as a resolution element of the spectrograph.
 For stars which do not physically undergo large \RV shifts over very short timescales, this means that it is necessary to observe the star over a significant fraction of the year to take advantage of the changes in the Earth's projected motion.
-We will comment on ways to overcome these restrictions in Section \ref{s:future}.
+We will comment on ways to overcome this restriction in Section \ref{s:future}.
 
@@
 \begin{equation}
-y_{n}^\star = P(v^{obs}_{n}) \mu^{\star},
+y_{n}^\star = P(v_{obs, n}) \mu_{\star},
 \label{eqn:star}
 \end{equation}
-where $\mu^{\star}$ is a spectral template with length $M'$ corresponding to
+where $\mu_{\star}$ is a spectral template with length $M'$ corresponding to
 the length of an arbitrarily-chosen wavelength grid of control points, and
 $P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
-interpolates $\mu^\star$ into the observatory rest frame.
+interpolates $\mu_\star$ into the observatory rest frame.
 Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
@@
 
-The apparent stellar \RV, $v^{obs}_{n}$, is a sum of the star's actual velocity about its center of mass ($v^{\star}_{n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
+The apparent stellar \RV, $v_{obs, n}$, is a sum of the star's actual velocity about its center of mass ($v_{\star, n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
 
@@
 \begin{equation}
-y_{n}^t =  A_n(\mu^{t} + W^{t} z_n).
+y_{n}^t =  A_n(\mu_{t} + W_{t} z_n).
 \end{equation}
-In addition to its mean spectrum $\mu^t$, the telluric component also includes a time-dependent term assembled from two variables: $W^t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
+In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
 $W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
@@
 \end{equation}
-where $p$ is the vector of parameters to be normalized (in this case $\mu^{\star}$, $\mu^{t}$, or $W^{t}$) and $\lambda$ is the regularization amplitude.
+where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.
 
@@
 \begin{split}
-\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  \\
- & + \Lone{\lambda_1}{\mu^{\star}} + \Ltwo{\lambda_2}{\mu^{\star}}   + \Lone{\lambda_3}{\mu^{t}} + \Ltwo{\lambda_4}{\mu^{t}} \\
- &  + \Lone{\lambda_5}{W^t}  + \Ltwo{\lambda_6}{W^t} + \Lone{1.0}{z^t} ,
+\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}} \\
+ &  + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}}  + \Lone{\lambda_5}{W_t}  + \Ltwo{\lambda_6}{W_t} + \Lone{1.0}{z_t} ,
 \end{split}
 \end{equation}
-%\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
+\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
 where the regularization amplitude on the basis weights is arbitrarily set to unity, and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
@@
 This regularization choice breaks that perfect degeneracy.
-The basis weights are regularized to encourage our desired outcome where the mean telluric spectrum contains as much telluric information as possible, while the variable basis picks up only the necessary time-variable changes.
+The basis weights are regularized because we want the mean telluric spectrum to do most of the telluric work, and the variable basis to only pick up what it needs to,
+
+%In practice, the most effective regularization amplitudes can vary by orders of magnitude depending on the characteristics of the spectral region being used. 
+%For instance, both $\lambda_{L1, \mu_{t}}$ and $\lambda_{L2, \mu_{t}}$ need to be much higher when the spectra in question span relatively blue wavelengths with few actual telluric features to fit than they do for the near-infrared wavelength regions with strong telluric features.
+%Similarly, the regularization amplitudes for the star tend to be higher for the G dwarf than they are for the M dwarf, which has far more absorption features in its spectra. 
+%Setting the regularization amplitudes is therefore a crucial first step when applying the \wobble model to data.
 
@@
 
-As an initial guess, we set the star to be stationary, e.g. $v^{obs} = $ \BERV at all epochs. 
-We may then initialize the stellar spectrum $\mu^{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
+As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs. 
+We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
 The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
@@
 We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
-This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iterations to convergence.
+This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iteration to convergence.
 
@@
 These orders are often treated as independent spectra when extracting \RVs, and the \RVs for each order are then combined in some manner to get a final time series.
+
 We follow this precedent and optimize the \wobble model individually for each order. 
-%While some \RV precision gain may be possible by inferring a global Doppler shift from all orders simultaneously, it is not clear that the gain would be worth the added computational expense, so we leave this for future work.
+While some \RV precision gain may be possible by inferring a global Doppler shift from all orders simultaneously, it is not clear that the gain would be worth the added computational expense, so we leave this for future work.
 
@@
 where $v_n$ and $v_r$ are the characteristic \RVs at epoch $n$ and order $r$, $\sigma_{n,r}$ is the estimated measurement uncertainty on $v_{n,r}$, and $\delta_r$ is an additional jitter term specific to order $r$.
-The order-dependent \RV offset should be small and indeed generally is consistent with zero in the solutions.%, but is included in the model to capture any color-dependent effects such as non-negligible atmospheric variability during the exposure \citep[see e.g.][]{Blackman2017}.
+
+\todo{(To do: fill in details: prior on $<v_r>$, optimization over $\delta_r$ with $v_n$ and $v_r$ analytically solved for.)}
 
@@
 
-The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF \citep{Abadi15}. 
+The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF. 
 \TF is a model building framework that has been primarily designed for machine
@@
 %While it has primarily been used in the context of deep learning in the astronomical literature, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
-The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 \HARPS spectra of 51 Peg, including the optimization of 72 spectral orders, runs in 60 minutes on a standard Mac desktop.
+The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 spectra of 51 Peg runs in \todo{30} minutes on a standard Mac desktop.
 
@@
 
-All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra.\footnote{Based on observations made with ESO Telescopes at the La Silla Paranal Observatory under programme IDs 091.C-0271, 183.C-0437, 191.C-0505, 072.C-0488, 089.C-0497, 099.C-0880, and 60.A-9700.}
+All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra. 
 These data come as extracted 1D spectra, in blocks of 72 orders by 4096 pixels per order.
@@
 Technically this uncertainty is assumed to be Gaussian in the log space, which is an incorrect approximation, but not very far off for high signal-to-noise data.
+\todo{BEDELL: Check this:}
 For each order we assign to each log flux an uncertainty variance that is the inverse of the mean signal-to-noise-squared of the pixels in that order (as reported by the \HARPS pipelines), scaled up or down by the raw linear flux observed in that pixel in the data prior to logging and normalizing.
@@
 
-If possible, the final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off an \RV drift term specified in the data header, consistent with the standard \HARPS pipeline.
-This drift is calculated from the simultaneous reference lamp and is therefore only available in certain observing modes. 
-Its contribution to the \RV solution is usually below 1 \ms, supporting our assumption that the wavelength solution provided by the \HARPS pipeline is generally accurate.
+The final stellar \RV retrieved by \wobble is corrected for intra-night drift in the wavelength solution by subtracting off a drift term specified in the data header, consistent with the standard \HARPS pipeline.
+\todo{(TO DO: make sure this is implemented!)}
 
@@
 \subsection{51 Pegasi}
-
-\begin{figure*}[ht!]
-\centering
-\includegraphics[width=5in]{51peg_planet}
-\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered from the \wobble\ \RVs (left) and from the \HARPS standard pipeline.
-While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns.
-Maximum-likelihood Keplerians (solid orange lines) and 1-$\sigma$ posterior constraints (shaded orange regions) for each data set are overplotted.}
-\label{fig:51peg_planet}
-\end{figure*}
-
-\begin{deluxetable*}{cccc}
-\tablehead{
-\colhead{Parameter} & \colhead{Units} & \colhead{\wobble Fit} & \colhead{\HARPS Pipeline Fit}}
-\tablecaption{Orbital Parameters of 51 Pegasi b}
-\label{tbl:51peg}
-\startdata
-$K$ & m s$^{-1}$ & $55.57^{+2.28}_{-2.04}$ & $53.84^{+1.96}_{-2.20}$ \\
-$P$ & days & $4.2292^{+0.0003}_{-0.0003}$ & $4.2294^{+0.0003}_{-0.0003}$ \\
-$t_0$ & JD & $2456546.89^{+0.02}_{-0.02}$ & $2456546.94^{+0.01}_{-0.02}$ \\
-$e$ & $-$  & $0.03^{+0.02}_{-0.02}$ & $0.04^{+0.02}_{-0.02}$ \\
-$\omega$ & rad & $0.45^{+0.62}_{-1.00}$ & $-1.79^{+0.34}_{-0.32}$ \\
-$s$ & m s$^{-1}$ & $0.01^{+0.08}_{-0.01}$ & $0.74^{+0.06}_{-0.07}$ \\
-\enddata
-\end{deluxetable*}
 
@@
 
-We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.
+We ran \wobble on these archival data to test its performance on recovering a planetary signal with well-known orbital characteristics.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:329968 and mbedell:331551.}
 The 91 publicly available spectra in the \HARPS archive are largely concentrated on a few nights of intensive observing, but these nights are sufficiently spread out throughout the year for a wide enough range in \BERV to disentangle the stellar and telluric spectra.
@@
 
-Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). 
-We fit a Keplerian signal with seven free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_0$, \RV offset $c$, and \RV jitter $s$) using the \code{exoplanet} package \citep{exoplanet}.
-The resulting best-fit parameters for the two data sets are generally consistent (Table \ref{tbl:51peg}), and both are comparable to literature values \citep{Mayor1995, Naef2004, Butler2006}. 
-The \RV uncertainties derived by \wobble appear accurate based on the negligible jitter in the fit, while the photon-noise-based \RV errors provided by the \HARPS pipeline are much smaller and require a significant jitter to achieve a good fit.
-These results confirm that \wobble is able to extract \RVs with similar precision to the closed-source \HARPS pipeline for Sun-like stars.
+Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the correct period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}).
+\todo{(numbers here, address the structure in residuals)}
 
-\begin{figure*}[t]
+Fits to an individual spectrum are shown in Figure \ref{fig:51peg_spectrum}. 
+We emphasize that no \textit{a priori} information on e.g. expected spectral line positions and depths were used. 
+The current realization of the \wobble algorithm treats each control point of the template spectra as independent and has no line shape parameterization included. 
+Nevertheless, the optimized stellar templates clearly reproduce the expected appearance of the spectra in a variety of regimes, from a crowded-line region to a sparser, continuum-dominated region to an extremely strong absorption line like the H$\alpha$ feature.
+
+\begin{figure}
+\centering
+\includegraphics[width=4in]{51peg_planet}
+\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered in the \wobble\ \RVs.
+While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns \todo{(add numbers)}.
+\todo{check on: The subtle trends seen near secondary eclipse (phase 0.75) are suggestive of contributions by reflected light from the planetary dayside.}}
+\label{fig:51peg_planet}
+\end{figure}
+
+\begin{figure}
 \centering
@@
 \label{fig:51peg_spectrum}
-\end{figure*}
-
-Fits to an individual spectrum are shown in Figure \ref{fig:51peg_spectrum}. 
-We emphasize that no \textit{a priori} information on e.g. expected spectral line positions and depths were used. 
-The \wobble algorithm as it is currently implemented treats each control point of the template spectra as independent and has no line shape parameterization included. 
-Nevertheless, the optimized stellar templates clearly reproduce the expected appearance of the spectra in a variety of regimes, from a crowded-line region to a sparser, continuum-dominated region to an extremely strong absorption line like the H$\alpha$ feature.
-Moreover, telluric absorption features are recovered down to a very small amplitude.
+\end{figure}
 
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=5in]{barnards_rvs}
-\caption{Radial velocity measurements for \Mdwarf from \wobble (black circles) and from the standard \HARPS pipeline (blue squares).
-The median \RV has been subtracted from each data set.
-The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid orange line (top panel).
-Residuals away from the predicted trend are plotted in the lower panel.}
-\label{fig:barnards_rvs}
-\end{figure*}
-
-Next, we tested \wobble with 237 epochs of \Mdwarf spectra. 
-We did not utilize the additional spectra taken after the \HARPS 2015 optical fiber upgrade, as these would need to be treated as an independent data set \citep{LoCurto2015}. 
+Next, we tested \wobble with 306 epochs of \Mdwarf spectra.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:359445 and mbedell:359501.} 
 \Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50. 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=5in]{barnards_rvs}
+\caption{Radial velocity measurements for \Mdwarf from \wobble (black) and from the standard \HARPS pipeline (red).
+The median \RV has been subtracted from each data set.
+The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid blue line (top panel).
+Residuals away from the predicted trend are plotted in the lower panel.
+\todo{(TO DO: add error bars and change color scheme)}}
+\label{fig:barnards_rvs}
+\end{figure}
+
 The \RVs found by \wobble are in excellent agreement with the predicted secular motion (Figure \ref{fig:barnards_rvs}). 
@@
 The \acronym{RMS} scatter among \wobble\ \RVs is 2.0 \ms. 
-This compares favorably to the scatter of 2.5 \ms among \RVs produced by the standard closed-source \HARPS pipeline.
-
-We note that many of the residuals have similar non-zero values in both independently produced RV estimates. 
+This compares favorably to the scatter of 2.5 \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates.
 It is possible that these deviations are a real physical effect in the stellar spectrum, but we caution that approximately \ms-level errors are likely being introduced by the barycentric correction. 
@@
 Indeed, we do note a strong peak at year-long periods in a Lomb-Scargle periodogram of the residual \RVs. 
-Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble.
+Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble. \todo{(TO DO: recalculate \BERV ourselves)} 
+%\todo{How does HARPS calculate the BERVs?
+If there is something like cirrus, where the apparent brightness of the star varies over the observation, it is plausible that the flux-weighted observation midpoint and the actual time midpoint are different enough to cause a m/s difference in the observations, too.
+Does HARPS have a way of accounting for this?
+(HIRES does estimate this, and they have shorter exposures on average).}
 
-\begin{figure*}[ht!]
+Testing \wobble on the \Mdwarf data allows us to evaluate its performance in a significantly different regime: the mid-M dwarf spectra are far denser in spectral features than a Sun-like star, and individual observations are at a much lower \SNR than the 51 Peg observations used above, although the total number of spectra is greater.
+The resulting spectral fit is shown in Figure \ref{fig:barnards_spectrum}.
+
+\begin{figure}
 \centering
@@
 \label{fig:barnards_spectrum}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}
+\begin{figure}
 \centering
@@
 \caption{A subset of the \Mdwarf spectrum including the sodium doublet at 4456-7 \ang.
-Data for the highest-\SNR single observation are shown as black points, while the spectral template derived by \wobble from the full data set is shown as a red line.
-The theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as an orange line.
-The \wobble model and data have been Doppler-shifted by eye to match the rest-frame line positions, since \wobble does not deliver an absolute \RV.
-The data-driven \wobble model of the stellar spectrum shares general strong line locations and relative strengths with the \PHOENIX model, but in detail it delivers a much more accurate fit to the data than the theoretical M dwarf model achieves.}
+The spectral template derived by \wobble is shown as a red line, while a theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as a blue line.
+The \wobble model has been Doppler-shifted to match the rest-frame line positions.
+\todo{(FIX COLORS!)}}
 \label{fig:barnards_model}
-\end{figure*}
-
-Testing \wobble on the \Mdwarf data allows us to evaluate its performance in a significantly different regime: the mid-M dwarf spectra are far denser in spectral features than a Sun-like star, and individual observations are at a much lower \SNR than the 51 Peg observations used above, although the total number of spectra is greater.
-The resulting spectral fit is shown in Figure \ref{fig:barnards_spectrum}.
+\end{figure}
 
@@
 This has the side effect of flattening out weak features if no strong lines are present in a given order. 
-As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the middle panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum}). 
+As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the top panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum} \todo{-- To do: check if there's a better example order}).
 In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}). 
@@
 The \wobble model for \Mdwarf diverges from the theoretical models in the smaller absorption lines and in the degree of broadening for strong lines.  
-%\todo{(TO DO: are these known model weaknesses? ask Travis Barman about opacity binning affecting the detailed features)}
+\todo{(TO DO: are these known model weaknesses? ask Travis Barman about opacity binning affecting the detailed features)}
 
@@
 
-\begin{figure*}
+\begin{figure}
 \centering
-\includegraphics[width=6in]{hd189_rm}
-\caption{\RVs measured by \wobble for a single night of observations of the Hot 
-         Jupiter host HD 189733. The \RM is clearly seen as the planet transits 
-         the star. The top panel shows the data (black points), the maximum
-         likelihood fit using \starry (black line), and 500 posterior samples
-         (orange lines). The bottom panel shows the residuals of the maximum
-         likelihood model fit.}
+\includegraphics[width=5in]{hd189_rm}
+\caption{\RVs measured by \wobble for a single night of observations of the Hot Jupiter host HD 189733. The \RM is clearly seen as the planet transits the star. \todo{[blah blah starry blah]}}
 \label{fig:hd189_rm}
-\end{figure*}
+\end{figure}
 
@@
 We apply \wobble to a data set consisting of a single transit of the hot Jupiter HD 189733b to test its sensitivity to stellar line asymmetries as well as its performance in the regime of a single night's observations.
-The observations in question consist of a consecutive series of 40 spectra, each with an \SNR $\sim 90$, taken on the night of August 28, 2007.
+The observations in question consist of a consecutive series of 40 spectra, each with an \SNR $\sim 90$, taken on the night of August 28, 2007.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers \todo{x}.}
 
@@
 The resulting \RV signal retrieved by \wobble is shown in Figure \ref{fig:hd189_rm}. 
-We fit the signal using \starry \citep{Luger2018}, which computes analytic occultation
-light curves for bodies whose surfaces can be decomposed into sums
-of spherical harmonics. Since the radial component of the velocity field of a 
-differentially rotating star can be expressed in terms of polynomials in 
-$x = \sin\theta\cos\phi$, $y = \sin\theta\sin\phi$, and $z=\cos\theta$, where
-$\theta$ is the polar angle and $\phi$ is the azimuthal angle \citep[c.f. Equation 91 in][]{Short2018},
-this velocity field may be expressed exactly in terms of spherical harmonics
-\citep{LugerBedell2019}.
-We therefore use \starry to fit for the equatorial rotational velocity, the inclination,
-the obliquity, and the shear due to differential rotation of the star, closely
-following the analysis performed in \citet{Cegla2016}.
-We infer a projected stellar obliquity $\lambda = -0.43 \pm 0.34^\circ$, in close agreement
-with the value reported in \citet{Cegla2016}. Our inferred values for the other parameters
-are broadly consistent with the results in \citet{Cegla2016}, except with significantly
-higher uncertainty due to the fact that we are unable to constrain the stellar inclination
-due to the $v\sin i$ degeneracy. We attribute the narrower posteriors in \citet{Cegla2016}
-to a difference in the choice of prior.
-\footnote{See \href{https://github.com/megbedell/wobble/tree/master/scripts/DifferentialRotationWithSphericalHarmonics.ipynb}{this interactive notebook} 
-          for a derivation of the spherical harmonic decomposition of the stellar velocity field. 
-          The notebook used to perform our analysis and produce Figure~\ref{fig:hd189_rm}, along with the full posterior constraints we obtain, can be found 
-          \href{https://github.com/megbedell/wobble/tree/master/scripts/HD189733bWithStarry.ipynb}{here}.}
+We fit the signal using \starry \citep{Luger2018}. \todo{(RODRIGO: fill in a bit about methods)} 
+\todo{(comparison of retrieved parameters to previous work here)}
 
 These results confirm that the \wobble method measures line asymmetries as \RV shifts in the same manner as traditional \RV analysis techniques. 
-While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{Queloz2001}. 
+While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{}.
+\todo{citep}
 The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified to fit line asymmetries and true Doppler shifts as separate parameters. 
@@
 
-\begin{figure*}[h!]
+\begin{figure}
 \centering
@@
 The upper panels show the \wobble best-fit telluric spectrum (blue line), best-fit stellar spectrum (red line), data (black points), and the sum of the telluric and stellar model predictions (black line).
-Residuals to the fit are shown in the lower panels.
-Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.}
+Residuals are shown in the lower panels.
+Although the fits to each data set were performed entirely independently, the optimized telluric spectra are nearly identical, demonstrating that \wobble successfully finds an accurate representation of the common telluric absorption spectrum.
+\todo{(TO DO: add y-label, fix centering)}}
 \label{fig:telluric_comparison}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
+\begin{figure}
 \centering
@@
 \label{fig:telluric_standard}
-\end{figure*}
+\end{figure}
 
-As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star HR 3090 taken by \HARPS at an \SNR of 130. 
+As another test of our derived telluric spectra, we compared the mean template from 51 Peg to an observation of the telluric standard star \todo{HR 3090 (+details)}.\footnote{Based on data obtained from the ESO Science Archive Facility under request numbers mbedell:385782 and \todo{xxx}.}
 As shown in Figure \ref{fig:telluric_standard}, the spectra agree well. 
@@
 \label{fig:telluric_basis}
-\end{figure*}
+\end{figure}
 
@@
 Another key assumption made in this work that there is no gas cell represented in the data. 
-While this is true for \HARPS, other instruments such as \acronym{HIRES}, \acronym{PFS}, and \acronym{APF} include a gas cell set in optical path to imprint its absorption spectrum on the observed spectra \citep{Butler1996, Crane2010, Vogt2014}.
+While this is true for \HARPS, other instruments like \acronym{HIRES}, \acronym{APF}, and \acronym{PFS} include a gas cell set in optical path to imprint its absorption spectrum on the observed spectra \todo{(cite)}.
 Applying \wobble to spectra that include the imprint of a gas cell is technically trivial. 
@@
 This effective spectrograph \RV may vary in different spectral orders or smaller wavelength regions.
-The most extreme extension would be to include corrections to the wavelength solution as model parameters, and optimize them along with the spectral components and radial velocities. 
-This would rely on the use of telluric features as fixed calibration sources \citep{Seifahrt2010}.
+The most extreme extension would be to include the full wavelength solution as model parameters, and optimize them along with the spectral components and radial velocities.
+\todo{(TO DO: add some citations to using tellurics for this)}
 
+%Spectral variability due to LSF changing?
+Additional parameters to the P operators, this would go into the non-convex optimization part (ask Hogg to throw in some words earlier on about location vs linear parameters)
 Another way a less well-calibrated instrument may violate \wobble's assumptions is in the non-negligible variability of all spectral lines due to changes in the instrumental line spread function (\LSF).
@@
 One important feature of \wobble is that it models all fluxes in the log-space. 
-When the model consists solely of multiplicative components, as is the case for a single star + telluric absorption, this choice makes the optimization convex at fixed \RV. 
+When the model consists solely of multiplicative components, as is the case for a single star + telluric absorption, this choice makes the optimization convex at fixed \RV and therefore computationally tractable; it makes possible our iterative optimization scheme.
 If we chose instead to convert the $y$ data vectors to linear fluxes and use a model consisting only of additive components, the same equations and algorithms would apply.
@@
 components as free variables, larger structural changes would be needed in the code and the convexity of the template optimization would be lost. 
-Further tests would be needed to determine whether this loss of convexity would slow down the optimization.
+Further tests would be needed to determine whether this loss of convexity would slow down the optimization or make finding good optima hard or impossible; it seems likely that it will slow the code but not be impossible.
 
@@
 
+Finally, \wobble, while it optimizes a sensible objective function with good properties, it has no good bayesian interpretation:
+The regularizations (especially the L1) do not make much sense as prior beliefs, and the output is a point estimate, not a probability or probability density.
+Furthermore, we may at some point want \RV measurements that are the result of \emph{marginalizing out} the spectral ant telluric models.
+This would be the principled Bayesian thing to do!
+Right now we don't expect a full subjective Bayesian treatment to be computationally possible.
+However, there certainly has been interesting work in this area, and it might become possible in the near future \todo{BEDELL: CITE CZEKALA, maybe others?}.
+
 \subsection{Basic Adjustments}
@@
 
-Even working within \wobble's current assumptions and running on \HARPS-like data, there are potential changes to be made to the implementation of \wobble that may improve its performance. 
+Even working within \wobble's current assumptions and running on \HARPS data, there are potential changes to be made to the implementation of \wobble that may improve its performance. 
 Many of these are quite basic and straightforward changes to be made within the existing framework of \wobble. 
-We discuss some potential adjustments here.
+We discuss some of potential adjustments here.
 
@@
 However, setting the regularization amplitudes optimally is by far the computationally slowest and most unwieldy part of applying \wobble to a new data set. 
-Dealing with the possibility of overfitting is an inevitable requirement of any model, especially one that relies on as many free parameters as \wobble. 
-That being said, in principle as data sets get larger, regularization becomes less crucial. 
-One way to lessen the need for telluric regularization might be to run on a data set that is composed of many different stars, each with their own individual spectra and \RVs but with a shared telluric component, so that the data informing the telluric lines is sufficiently large.
-This could be done with relatively trivial changes to the \wobble code, although it would require considerable computational resources.
+In principle, as data sets get larger, regularization becomes less necessary. 
+One way to remove the need for telluric regularization might be to run on a data set that is composed of many different stars, each with their own individual spectra and \RVs but with a shared telluric component, so that the data informing the telluric lines is sufficiently large.
+This could be done with relatively trivial changes to the \wobble code, although it would require strong computational resources.
 
 One choice that \wobble makes differently from other data-driven \RV codes is the way we represent the underlying spectral components. 
-We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated. 
+We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated by the $P$ operator. 
 One possible improvement would be to go to a higher-order interpolation scheme. 
@@
 
-Another choice we made which is likely suboptimal is the treatment of each spectral order in the echelle data as an independent spectrum with its own stellar \RV. 
-In reality, every stellar \RV should be informed by all spectral orders simultaneously. 
+Another choice we made which is almost certainly suboptimal is the treatment of each spectral order in the echelle data as an independent spectrum with its own stellar \RV. 
+In reality, every \RV should be informed by all spectral orders simultaneously. 
 Implementing this would take on some extra computational cost, but being in \TF will help to mitigate this.
-However, the reality of color-dependent atmospheric dispersion effects means that in practice each order may have a slightly different effective \RV, since the flux-weighted average observation times will be different \citep{Blackman2017}.
-Thus for some situations allowing the orders to have independent \RV estimates may be the best choice.
 
@@
     scipy \citep{Jones:2001},
-    tensorflow \citep{Abadi15}, 
-    exoplanet \citep{exoplanet},
-    starry \citep{Luger2018}
+    tensorflow \citep{Abadi15}
 }
 
-\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
+\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
 We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.

### 2019-01-02 paper/paper.tex

--- 2019-01-02 paper/paper.tex
@@
 To avoid this degeneracy, we might instead represent the stellar spectrum with an explicit parameterization of the line profile and restrict variability to the line depths, widths, and skews; such a change would fall along the same lines as the global \LSF parameterization proposed in Section \ref{s:data-changes}.
+
+Finally, \wobble, while it optimizes a sensible objective function with good properties, it has no good bayesian interpretation:
+The regularizations (especially the L1) do not make much sense as prior beliefs, and the output is a point estimate, not a probability or probability density.
+Furthermore, we may at some point want \RV measurements that are the result of \emph{marginalizing out} the spectral ant telluric models.
+This would be the principled Bayesian thing to do!
+Right now we don't expect a full subjective Bayesian treatment to be computationally possible.
+However, there certainly has been interesting work in this area, and it might become possible in the near future \todo{BEDELL: CITE CZEKALA, maybe others?}.
 

### 2019-01-02 scripts/regularization.py
more docs
Former-commit-id: 9e7a4f64e89fc60fc8233801ad9c715cf4bea7e2

@@ -63,8 +63,8 @@ def improve_order_regularization(r, o, star_filename, tellurics_filename,
     tensors_to_tune = [training_model.components[1].L2_template_tensor, training_model.components[0].L2_template_tensor,
                        training_model.components[1].L1_template_tensor, training_model.components[0].L1_template_tensor]
     tensor_names = ['L2_template', 'L2_template', 'L1_template',
-                     'L1_template'] # HACK - this is needed bc TF appends garbage to the end of the tensor name
-    tensor_components = ['tellurics', 'star', 'tellurics', 'star'] # HACK
+                     'L1_template'] # this isonly  needed bc TF appends garbage to the end of the tensor name
+    tensor_components = ['tellurics', 'star', 'tellurics', 'star'] # ^ same
     if K_star > 0:
         tensors_to_tune = np.append(tensors_to_tune, [training_model.components[0].L2_basis_vectors_tensor, 
                                                     training_model.components[0].L1_basis_vectors_tensor])

### 2019-01-02 paper/paper.tex
more docs

@@
 \DeclareGraphicsExtensions{.pdf,.eps,.png}
+
+%@arxiver{telluric_comparison.pdf}
 
@@
 Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
-In both cases, this precision-limiting reliance on external information can be sidestepped using the data itself. 
+In both cases, precision-limiting reliance on external information can be sidestepped using the data directly. 
 % Aims
@@
 We recover the canonical exoplanet 51 Pegasi b, detect the secular \RV evolution of the M dwarf \Mdwarf, and retrieve the \RM for the Hot Jupiter HD 189733b. 
-We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data. 
+The method additionally produces extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra, which we also present here. 
 \end{abstract}
@@
 We assume that the stellar spectrum is invariant with time. 
-This assumption does not hold true in detail and will be revisited in Section \ref{s:future}.
+This assumption does not hold true in detail and we comment on this in Section \ref{s:model-changes}.
 
@@
 
-Finally, we assume that, when the stellar and telluric components are properly optimized, only Gaussian noise remains.
+Finally, we assume that, when the stellar and telluric components are properly optimized, any remaining noise may be approximated as Gaussian.
 
@@
 The $\chi^2$ for the validation epochs can then be adopted as a goodness-of-fit measurement for the $\lambda$ value, and the procedure is repeated for different $\lambda$s to choose the best regularization amplitude.
-Because we have multiple regularization terms and multiple amplitudes to set, we begin by hand-setting all amplitudes to a reasonable starting guess and optimize each amplitude sequentially with cross-validation.
+In theory, we might wish to regularize using an optimization metric based on \RV accuracy rather than $\chi^2$ over pixels; however, the lack of known ``ground truth'' in the \RV behavior of stars makes this not currently feasible.
+
+Since we have multiple regularization amplitudes to set, we begin by hand-setting all amplitudes to a reasonable starting guess and optimize each amplitude sequentially with cross-validation. 
+Generally speaking, the L2 regularization tends to be stricter than L1 and the telluric components are more strongly regularized than the stellar components, so we aim to roughly go from most to least sensitive regularization component when tuning the amplitudes.
+We found that good performance came from tuning the L2 regularization amplitudes for the mean telluric and stellar spectra, followed by the L1 regularization amplitudes for the same, followed by L2 and L1 for the time-variable telluric basis vectors.
 
@@
 
+It is worthy of note that the stellar spectral model shown in Figure \ref{51peg_multispectrum} is an extremely high-\SNR spectrum of 51 Peg. 
+This template spectrum is essentially a time-averaged composite spectrum. 
+Both it and its residuals as a function of time are scientifically valuable outputs of the \wobble method.
+
 \subsection{\Mdwarf}
@@
 These results emphasize the potential value of \wobble data products for spectral characterization and model testing, particularly in the case of faint stars and M dwarfs, for which many spectra must be combined to get a reasonably detailed composite spectrum.
-The \wobble algorithm is a simple yet robust method for doing such multi-spectra stacking while preserving telluric-contaminated regions and marginalizing over unknown \RV shifts.
+The \wobble algorithm is a simple yet robust method for doing such multi-spectra stacking while preserving telluric-contaminated regions and accounting for unknown \RV shifts.
 
@@
 
-Finally, \wobble, while it optimizes a sensible objective function with good properties, it has no good bayesian interpretation:
-The regularizations (especially the L1) do not make much sense as prior beliefs, and the output is a point estimate, not a probability or probability density.
-Furthermore, we may at some point want \RV measurements that are the result of \emph{marginalizing out} the spectral ant telluric models.
-This would be the principled Bayesian thing to do!
-Right now we don't expect a full subjective Bayesian treatment to be computationally possible.
-However, there certainly has been interesting work in this area, and it might become possible in the near future \todo{BEDELL: CITE CZEKALA, maybe others?}.
-
 \subsection{Basic Adjustments}
@@
 Thus for some situations allowing the orders to have independent \RV estimates may be the best choice.
+
+While \wobble optimizes a sensible objective function with good properties, it has no direct Bayesian interpretation. 
+The regularizations (especially the L1) do not make much sense as prior beliefs, and the output is a point estimate, not a probability or probability density. 
+Furthermore, we may at some point want \RV measurements that are the result of \emph{marginalizing out} the spectral and telluric models. 
+Right now we don't expect such a full subjective Bayesian treatment to be computationally possible.
+However, there certainly has been interesting work in this area, and it might become possible in the near future \citep{Czekala2015}.
 
@@
 We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.
+Special thanks go to Geert Barentsen, Michael Gully-Santiago, Christina Hedges, Susan Mullaly, and Z\'{e} Vin\'{i}cius for their vital contribution to the most important part of \wobble's documentation (the pronunciation guide).
 

### 2019-01-02 paper/paper.tex
moving Rodrigo's shit
Former-commit-id: f5f09de419572b53514d6875baa2fa1007e62cae

@@
 to a difference in the choice of prior.
-\footnote{See \href{https://github.com/megbedell/wobble/tree/master/scripts/DifferentialRotationWithSphericalHarmonics.ipynb}{this interactive notebook} 
+\footnote{See \href{https://github.com/megbedell/wobble/tree/master/paper/figures/HD189733/DifferentialRotationWithSphericalHarmonics.ipynb}{this interactive notebook} 
           for a derivation of the spherical harmonic decomposition of the stellar velocity field. 
           The notebook used to perform our analysis and produce Figure~\ref{fig:hd189_rm}, along with the full posterior constraints we obtain, can be found 
-          \href{https://github.com/megbedell/wobble/tree/master/scripts/HD189733bWithStarry.ipynb}{here}.}
+          \href{https://github.com/megbedell/wobble/tree/master/paper/figures/HD189733/HD189733bWithStarry.ipynb}{here}.}
 

### 2019-07-01 paper/paper.tex
submission version
changing model specification

@@
 We also have a corresponding $M \times N$ matrix of wavelength solutions. 
-Again, the entries of this matrix are in logarithmic form ($\ln \lambda$). 
-This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.
+Again, the entries of this matrix are in logarithmic form ($\ln \lambda_{m,n}$). 
+\deleted{This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.}
 
@@
 \begin{equation}
-y_n = y_{n}^\star + y_{n}^t + noise.
+y_n = y_{n}^\star + y_{n}^t + \textrm{noise}.
 \end{equation}
@@
 \begin{equation}
- \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
+ \ln \lambda(v) = \ln \lambda_{0} + D(v) \equiv \ln \lambda_{0} - \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
 \end{equation}
-The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
+The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space, \added{so that each predicted value $y_{n,m}^\star$ at data wavelength $\ln \lambda_{m,n}$ is determined by:} \todo{(this is not complete)}
+\begin{equation}
+y_{m,n}^\star(\ln \lambda_{m,n}) = \mu^{\star}(\ln \lambda_{m'} + D(v^{obs}_{n})).
+\label{eqn:Pop}
+\end{equation}
 $P$ could be represented as a rectangular $M \times M'$ sparse matrix, but the dot product can be implemented efficiently using linear interpolation instead of instantiating the full matrix.
 
-The apparent stellar \RV, $v^{obs}_{n}$, is a sum of the star's actual velocity about its center of mass ($v^{\star}_{n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
+The apparent stellar \RV, $v^{obs}_{n}$, is a \replaced{sum}{combination} of the star's actual velocity \replaced{about its center of mass}{in the barycentric reference frame} ($v^{\star}_{n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
+\todo{(add Wright equation!)}
 
@@
 
-\subsection{51 Pegasi}
-
 \begin{figure*}[ht!]
@@
 
+\subsection{51 Pegasi}
+
 For \wobble's first test, we chose the first known exoplanet host: 51 Pegasi. 
@@
 Their \SNR is generally high, ranging from 100 up to 300 pix$^{-1}$ at the central wavelength regions of \HARPS.
-
-Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). 
-We fit a Keplerian signal with seven free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_0$, \RV offset $c$, and \RV jitter $s$) using the \code{exoplanet} package \citep{exoplanet}.
-The resulting best-fit parameters for the two data sets are generally consistent (Table \ref{tbl:51peg}), and both are comparable to literature values \citep{Mayor1995, Naef2004, Butler2006}. 
-The \RV uncertainties derived by \wobble appear accurate based on the negligible jitter in the fit, while the photon-noise-based \RV errors provided by the \HARPS pipeline are much smaller and require a significant jitter to achieve a good fit.
-These results confirm that \wobble is able to extract \RVs with similar precision to the closed-source \HARPS pipeline for Sun-like stars.
 
@@
 
+Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). 
+We fit a Keplerian signal with seven free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_0$, \RV offset $c$, and \RV jitter $s$) using the \code{exoplanet} package \citep{exoplanet}.
+The resulting best-fit parameters for the two data sets are generally consistent (Table \ref{tbl:51peg}), and both are comparable to literature values \citep{Mayor1995, Naef2004, Butler2006}. 
+The \RV uncertainties derived by \wobble appear accurate based on the negligible jitter in the fit, while the photon-noise-based \RV errors provided by the \HARPS pipeline are much smaller and require a significant jitter to achieve a good fit.
+These results confirm that \wobble is able to extract \RVs with similar precision to the closed-source \HARPS pipeline for Sun-like stars.
+
 Fits to an individual spectrum are shown in Figure \ref{fig:51peg_spectrum}. 
@@
 
-It is worthy of note that the stellar spectral model shown in Figure \ref{51peg_multispectrum} is an extremely high-\SNR spectrum of 51 Peg. 
+It is worthy of note that the stellar spectral model shown in Figure \ref{fig:51peg_spectrum} is an extremely high-\SNR spectrum of 51 Peg. 
 This template spectrum is essentially a time-averaged composite spectrum. 
@@
 
-\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
+\acknowledgements{\added{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful, constructive, and helpful review.}
+It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
 We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.

### 2019-07-02 paper/paper.tex
more minor edits

@@
 
-Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is a single observation taken for the purpose \citep[e.g.][]{Butler1996}.
+Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is \added{derived from} a single observation taken for this purpose \citep[e.g.][]{Butler1996}.
 Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
@@
 With a simple data-driven model, one could learn both stellar and telluric spectra simultaneously with the stellar \RVs. 
-
-%\todo{(talk about lack of open-source code and difficulty of modifying \RV pipelines here?)}
-
 Here we develop and implement such a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch. 
@@
 
-Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
-%In this sense, ours is a true forward model.  -- this content-less (DFM)
+Our approach \added{to the \EPRV problem} is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
+\added{Similar approaches of learning template spectra for stars and tellurics from the data directly have been used in the literature, although they have not previously been demonstrated to achieve precise stellar \RVs \citep[e.g.]{Hadrava2004, Hadrava2006}.}
 
@@
 First, we assume that the wavelength calibration and spectral extraction of the instrument are perfect: that is, we begin at the stage of having 1D extracted spectra and corresponding wavelength grids in hand and we do not model any corrections to the wavelength solution.
+\added{Similarly, we assume that the line spread function of the instrument remains perfectly constant from one exposure to the next.
+This assumption is needed because it allows us to extract instrumentally-broadened template spectra rather than explicitly modeling and solving for instrumental broadening effects.
+However, \todo{as we discuss in Section \ref{s:future}}, even within the limits of this assumption extracting already-broadened templates is not strictly correct.
+We do this for simplicity only and leave the general case to future work.}
 
@@
 The apparent stellar \RV, $v^{obs}_{n}$, is a \replaced{sum}{combination} of the star's actual velocity \replaced{about its center of mass}{in the barycentric reference frame} ($v^{\star}_{n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
-\todo{(add Wright equation!)}
+\added{For the purposes of this work, we use the \BERV furnished by the \HARPS pipeline. 
+We also make the approximation that the observed velocity is a simple sum of the actual velocity and the \BERV. 
+In detail, this approximation is incorrect; see \citet{Wright2014} and \citet{Wright2019} for further discussion.}
 
@@
 
-\acknowledgements{\added{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful, constructive, and helpful review.}
+\acknowledgements{\added{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful and constructive review.}
 It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 

### 2019-07-05 paper/paper.tex
more notation

@@
 \include{preamble}
+\usepackage{ bbold }
 \newcommand{\Mdwarf}{Barnard's Star\xspace} % or should it be Barnard's star? or GJ 699?
@@
 
-We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
-We also have a corresponding $M \times N$ matrix of wavelength solutions. 
-Again, the entries of this matrix are in logarithmic form ($\ln \lambda_{m,n}$). 
-\deleted{This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.}
+We take the data to be the $M \times N$ matrix y, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
+We also have a corresponding $M \times N$ matrix of wavelength solutions\added{ which we call $\xi$, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.
+\deleted{Again, the entries of this matrix are in logarithmic form ($\ln \lambda_{m,n}$). This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.}
 
-We model each data column $y_n$ as the sum of the stellar and telluric spectra:
+\replaced{We model each data column $y_n$}{For each data column $y_n$, our model prediction $f_n$ can be treated} as the sum of stellar and telluric contributions:
 \begin{equation}
-y_n = y_{n}^\star + y_{n}^t + \textrm{noise}.
+f_n = f_{\star,n} + f_{t,n} + \textrm{noise}.
 \end{equation}
@@
 \begin{equation}
-y_{n}^\star = P(v^{obs}_{n}) \mu^{\star},
+f_{\star,n} = P(\xi_n, \xi_{\star}, v_{obs,n}) \cdot \mu_{\star},
 \label{eqn:star}
 \end{equation}
-where $\mu^{\star}$ is a spectral template with length $M'$ corresponding to
-the length of an arbitrarily-chosen wavelength grid of control points, and
-$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
-interpolates $\mu^\star$ into the observatory rest frame.
-Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
+where $\mu_{\star}$ is a spectral template of log-fluxes and 
+$\xi_{\star}$ is the corresponding vector of template log-wavelengths. 
+The exact values of $\xi_{\star}$ can be chosen somewhat arbitrarily under the conditions that the grid has uniform spacing $\Delta\xi_{\star}$, covers the entire wavelength range of the data, and is over-sampled with regards to the observed spectrum's wavelength grid.
+$P$ is a linear operator whose function is to apply a Doppler shift by observed velocity $v_{obs,n}$ and
+interpolate $\mu_{\star}$ from the $\xi_{\star}$ template grid to the $\xi_n$ data grid.
+Each entry of the $P$ matrix can be defined by a sum of weighted indicator function as follows:
 \begin{equation}
- \ln \lambda(v) = \ln \lambda_{0} + D(v) \equiv \ln \lambda_{0} - \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
+\begin{split}
+ P_{i,j} =  &\ \Big(\frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big(0 < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 1\big) \\
+  &+ \Big(1 - \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big({-1} < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 0\big), \\
+ \end{split}
 \end{equation}
-The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space, \added{so that each predicted value $y_{n,m}^\star$ at data wavelength $\ln \lambda_{m,n}$ is determined by:} \todo{(this is not complete)}
+where $\xi'_{\star}$ is the Doppler-shifted template grid:
 \begin{equation}
-y_{m,n}^\star(\ln \lambda_{m,n}) = \mu^{\star}(\ln \lambda_{m'} + D(v^{obs}_{n})).
-\label{eqn:Pop}
+ \xi'_{\star}(v) = \xi_{\star} + D(v) \equiv \xi_{\star} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
 \end{equation}
-$P$ could be represented as a rectangular $M \times M'$ sparse matrix, but the dot product can be implemented efficiently using linear interpolation instead of instantiating the full matrix.
+We note that because $P$ is quite sparse, in practice we do not instantiate the full matrix when performing calculations.
 
-The apparent stellar \RV, $v^{obs}_{n}$, is a \replaced{sum}{combination} of the star's actual velocity \replaced{about its center of mass}{in the barycentric reference frame} ($v^{\star}_{n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
+The apparent stellar \RV, $v_{obs,n}$, is a \replaced{sum}{combination} of the star's actual velocity \replaced{about its center of mass}{in the barycentric reference frame} $v_{\star,n}$ and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
 \added{For the purposes of this work, we use the \BERV furnished by the \HARPS pipeline. 
@@
 \begin{equation}
-y_{n}^t =  A_n(\mu^{t} + W^{t} z_n).
+f_{t,n} =  A_n(\mu_{t} + W_{t} \cdot z_n).
 \end{equation}
-In addition to its mean spectrum $\mu^t$, the telluric component also includes a time-dependent term assembled from two variables: $W^t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
-$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
+In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
+$W_t$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $M'$ is the length of some vector of template wavelengths $\xi_{t}$ and $K$ is the number of basis vectors used.\footnote{While we refer to these as basis vectors for simplicity, note that they are not actually constrained to be orthogonal.}
 For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
-The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
+The template wavelength grid $\xi_{t}$ should have similar properties to $\xi_{\star}$ and may in practice be the same grid. 
+Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
 
@@
 \begin{equation}
-\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n}),
+\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - f_{n})^T C_{n}^{-1} (y_{n}-f_{n}),
 \label{eqn:lnlike}
 \end{equation}
-comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.
+comparing our model $f$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.
 
@@
 \begin{split}
-\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  \\
- & + \Lone{\lambda_1}{\mu^{\star}} + \Ltwo{\lambda_2}{\mu^{\star}}   + \Lone{\lambda_3}{\mu^{t}} + \Ltwo{\lambda_4}{\mu^{t}} \\
- &  + \Lone{\lambda_5}{W^t}  + \Ltwo{\lambda_6}{W^t} + \Lone{1.0}{z^t} ,
+\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - f_{n})^T C_{n}^{-1} (y_{n}-f_{n})  \\
+ & + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}}   + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}} \\
+ &  + \Lone{\lambda_5}{W_t}  + \Ltwo{\lambda_6}{W_t} + \Lone{1.0}{z} ,
 \end{split}
@@
 The basis-weight regularization amplitude can be set arbitrarily to unity because there is a perfect degeneracy between amplifying the basis vectors and attenuating the basis weights.
-This regularization choice breaks that perfect degeneracy.
+This regularization choice breaks that degeneracy.
 The basis weights are regularized to encourage our desired outcome where the mean telluric spectrum contains as much telluric information as possible, while the variable basis picks up only the necessary time-variable changes.

### 2019-07-18 paper/paper.tex
minor edits

@@
 
-We take the data to be the $M \times N$ matrix y, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
+We take the data to be the $M \times N$ matrix $y$, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
 We also have a corresponding $M \times N$ matrix of wavelength solutions\added{ which we call $\xi$, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.
@@
 interpolate $\mu_{\star}$ from the $\xi_{\star}$ template grid to the $\xi_n$ data grid.
-Each entry of the $P$ matrix can be defined by a sum of weighted indicator function as follows:
+Each entry of the $P$ matrix can be defined by a sum of weighted indicator functions as follows:
 \begin{equation}
@@
 \end{equation}
-We note that because $P$ is quite sparse, in practice we do not instantiate the full matrix when performing calculations.
+Because $P$ is quite sparse, in practice we do not instantiate the full matrix when performing calculations. 
+At this point, it is also useful to note that $P$ could, in principle, encode an instrumental line spread function (\LSF). 
+We return to this point in Section \label{s:future}.
 
@@
 We also make the approximation that the observed velocity is a simple sum of the actual velocity and the \BERV. 
-In detail, this approximation is incorrect; see \citet{Wright2014} and \citet{Wright2019} for further discussion.}
+In detail, this approximation is incorrect, but it should not have a large effect on our results; see \citet{Wright2014} and \citet{Wright2019} for further discussion.}
 
@@
 For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
-The template wavelength grid $\xi_{t}$ should have similar properties to $\xi_{\star}$ and may in practice be the same grid. 
+The template wavelength grid $\xi_{t}$ should have similar properties to $\xi_{\star}$. 
 Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
@@
 \end{equation}
-where $p$ is the vector of parameters to be normalized (in this case $\mu^{\star}$, $\mu^{t}$, or $W^{t}$) and $\lambda$ is the regularization amplitude.
+where $p$ is the vector of parameters to be normalized (in this case $\mu_{\star}$, $\mu_{t}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.
 
@@
 Since we have multiple regularization amplitudes to set, we begin by hand-setting all amplitudes to a reasonable starting guess and optimize each amplitude sequentially with cross-validation. 
-Generally speaking, the L2 regularization tends to be stricter than L1 and the telluric components are more strongly regularized than the stellar components, so we aim to roughly go from most to least sensitive regularization component when tuning the amplitudes.
+Generally speaking, the L2 regularization tends to be stricter than L1 and the telluric components are more strongly regularized than the stellar components, so we aim to go roughly from most to least sensitive regularization component when tuning the amplitudes.
 We found that good performance came from tuning the L2 regularization amplitudes for the mean telluric and stellar spectra, followed by the L1 regularization amplitudes for the same, followed by L2 and L1 for the time-variable telluric basis vectors.
@@
 
-As an initial guess, we set the star to be stationary, e.g. $v^{obs} = $ \BERV at all epochs. 
-We may then initialize the stellar spectrum $\mu^{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
+As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs. 
+We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\xi_{\star}$. 
 The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
 Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
-The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors.
+The $K$ highest eigenweights and their corresponding eigenvectors are taken as the starting guess for the basis weights and vectors.
 
@@
 Technically the velocities are location parameters and their optimization is not convex.
-However, once the velocities are known to a small fraction of a pixel (which they usually are, in practice, in advance of running \wobble), the linearized problem becomes convex at each iteration.
+However, once the velocities are known to a small fraction of a pixel (which they usually are in practice, given an accurate estimation of the \BERV), the linearized problem becomes convex at each iteration.
 We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
@@
 All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra.\footnote{Based on observations made with ESO Telescopes at the La Silla Paranal Observatory under programme IDs 091.C-0271, 183.C-0437, 191.C-0505, 072.C-0488, 089.C-0497, 099.C-0880, and 60.A-9700.}
-These data come as extracted 1D spectra, in blocks of 72 orders by 4096 pixels per order.
+These data come as extracted order-by-order 1D spectra in blocks of 72 orders by 4096 pixels per order.
 The airmass of the observation and the calculated \BERV are both provided in the \acronym{FITS} header of the spectral data file. 
@@
 After masking these data, we convert the flux to log space and continuum normalize by fitting and subtracting a polynomial to an asymmetrically clipped subset of the data. 
+\added{We found good results from clipping all pixels outside of the range [$-$0.3$\sigma$, $+$3$\sigma$] as a way of effectively removing the absorption features; fitting a sixth-order polynomial; and iteratively repeating with the pixel clipping set by the residuals to the previous fit until the selection of clipped pixels is stable.}
 
@@
 It is worthy of note that the stellar spectral model shown in Figure \ref{fig:51peg_spectrum} is an extremely high-\SNR spectrum of 51 Peg. 
-This template spectrum is essentially a time-averaged composite spectrum. 
+\added{The uncertainties on this spectrum can be estimated from the negative inverse Hessian, similarly to the \RVs.
+Doing this calculation for a subset of the template in the continuum around 6000 \ang indicates an approximate \SNR of \todo{xxx}, which is consistent with the expected \SNR from co-adding each individual spectrum.
+}
+\replaced{This template spectrum is essentially a time-averaged composite spectrum.}{In essence, this template is a time-averaged and telluric-cleaned composite stellar spectrum.} 
 Both it and its residuals as a function of time are scientifically valuable outputs of the \wobble method.
@@
 In principle, if the changes to the \LSF were well-understood they could be hard-coded into linear operators like the $P$ operator used in Equation \ref{eqn:star}, leaving the essential equations of \wobble largely unchanged.
+\added{In this case the \LSF operator should be applied to both the stellar and telluric component models after combining them.} 
 Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.

### 2019-07-26 paper/paper.tex
edits and new script

@@
 At this point, it is also useful to note that $P$ could, in principle, encode an instrumental line spread function (\LSF). 
-We return to this point in Section \label{s:future}.
+We return to this point in Section \ref{s:future}.
 
@@
 In principle, if the changes to the \LSF were well-understood they could be hard-coded into linear operators like the $P$ operator used in Equation \ref{eqn:star}, leaving the essential equations of \wobble largely unchanged.
-\added{In this case the \LSF operator should be applied to both the stellar and telluric component models after combining them.} 
-Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.
+\added{In this case the \LSF operator should be applied to both the stellar and telluric component models after combining them.
+Indeed, technically the current \wobble model does not deal correctly with line broadening: the templates we infer for star and tellurics are post-broadening spectra, but this goes against the correct order of operations if line broadening is dominated by instrumental effects that are applied to the \textit{combined} spectra.
+We have assumed that this effect is negligible for the purposes of this work, but it should certainly be considered more carefully in the future.}
+
+Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to \replaced{take this approach}{take a prescriptive approach to their behavior and hard-code them into the form of the $P$ operator}; instead, we might want to fit these variations \replaced{as a data-driven model component}{using a data-driven approach, which would make the \wobble model significantly non-linear}.
 This leads us into the territory of making more fundamental changes to the \wobble model.
@@
 \acknowledgements{\added{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful and constructive review.}
-It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
+It is a pleasure to acknowledge Jacob L.
+Bean, John M.
+Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, Julian St\"{u}rmer, \added{and Sharon Xuesong Wang} for helpful discussions.
 We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.

### 2019-07-26 scripts/missing_wavelength_files.py

@@ -1,0 +1,23 @@
+import numpy as np
+from astropy.io import fits
+import os
+import glob
+
+def missing_wavelength_files(filelist):
+    missing_files = []
+    for f in filelist:
+        path = f[0:str.rfind(f,'/')+1]
+        sp = fits.open(f)
+        header = sp[0].header
+        wave_file = header['HIERARCH ESO DRS CAL TH FILE']
+        if os.path.isfile(path+wave_file):
+            continue
+        else:
+            missing_files = np.append(missing_files, wave_file)
+            
+    return np.unique(missing_files)
+    
+    
+if __name__ == '__main__':
+    filelist = glob.glob('/Users/mbedell/python/wobble/data/51peg/HARPS*ccf_G2_A.fits')
+    np.savetxt('missing_files.txt', missing_files, fmt='%s')

### 2019-07-29 scripts/script_51peg.py
cleaning up and fixing barnards bug

@@ -9,7 +9,7 @@ import os
 if __name__ == "__main__":
     starname = '51peg'
     K_star = 0
-    K_t = 0    
+    K_t = 3    
     niter = 150 # for optimization
     plots = True
     epochs = [0, 50] # to plot
@@ -17,9 +17,9 @@ if __name__ == "__main__":
     
     star_reg_file = '../wobble/regularization/{0}_star_K{1}.hdf5'.format(starname, K_star)
     tellurics_reg_file = '../wobble/regularization/{0}_t_K{1}.hdf5'.format(starname, K_t)
-    plot_dir = '../results/plots_{0}_Kstar{1}_Kt{2}/'.format(starname, K_star, K_t)
+    plot_dir = '../results/plots_{0}_Kstar{1}_Kt{2}-keep/'.format(starname, K_star, K_t)
     
-    if True:
+    if False:
         # quick test on single order
         data = wobble.Data(starname+'_e2ds.hdf5', filepath='../data/', orders=[30, 56])
         results = wobble.Results(data=data)
@@ -40,7 +40,7 @@ if __name__ == "__main__":
     print("running wobble on star {0} with K_star = {1}, K_t = {2}".format(starname, K_star, K_t))
     start_time = time()
     orders = np.arange(72)
-    data = wobble.Data(starname+'_e2ds.hdf5', filepath='../data/', orders=orders)
+    data = wobble.Data(starname+'_e2ds-keep.hdf5', filepath='../data/', orders=orders)
     results = wobble.Results(data=data)
     
     print("data loaded")
@@ -111,7 +111,7 @@ if __name__ == "__main__":
     print("final RVs calculated.")
     print("time elapsed: {0:.2f} minutes".format((time() - start_time)/60.0))
         
-    results_file = '../results/results_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t)
+    results_file = '../results/results_{0}_Kstar{1}_Kt{2}-keep.hdf5'.format(starname, K_star, K_t)
     results.write(results_file)
         
     print("results saved as: {0}".format(results_file))

### 2019-07-29 scripts/script_51peg.py
miscellaneous keyword and attr fixes

@@ -125,14 +125,14 @@ if __name__ == "__main__":
     print("time elapsed: {0:.2f} minutes".format((time() - start_time)/60.0))
         
     # save output:
-    results_file = 'results/results_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t)
+    results_file = '../results/results_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t)
     results.write(results_file)
-    star_rvs_file = 'results/rvs_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t)
+    star_rvs_file = '../results/rvs_{0}_Kstar{1}_Kt{2}.txt'.format(starname, K_star, K_t)
     results.write_rvs('star', star_rvs_file, all_orders=True)
     
         
     print("results saved as: {0} & {1}".format(results_file, star_rvs_file))
     print("-----------------------------")    
-    print("HARPS pipeline std = {0:.3f} m/s".format(np.std(data.pipeline_rvs + data.bervs)))
+    print("HARPS pipeline std = {0:.3f} m/s".format(np.std(data.pipeline_rvs)))
     print("wobble std = {0:.3f} m/s".format(np.std(results.star_time_rvs)))        
     print("total runtime:{0:.2f} minutes".format((time() - start_time)/60.0))

### 2019-07-29 scripts/script_barnards.py

@@ -27,6 +27,8 @@ if __name__ == "__main__":
         upgrade = 2457174.5 # June 2015
         e = data.epochs[data.dates < upgrade]
         data = wobble.Data(filename='../data/'+starname+'_e2ds.hdf5', orders=orders, epochs=e)
+    data.drop_bad_orders()
+    data.drop_bad_epochs()
     orders = np.copy(data.orders)
     results = wobble.Results(data=data)
     
@@ -98,14 +100,14 @@ if __name__ == "__main__":
     print("time elapsed: {0:.2f} minutes".format((time() - start_time)/60.0))
         
     # save output:
-    results_file = 'results/results_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t)
+    results_file = '../results/results_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t)
     results.write(results_file)
-    star_rvs_file = 'results/rvs_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t)
+    star_rvs_file = '../results/rvs_{0}_Kstar{1}_Kt{2}.txt'.format(starname, K_star, K_t)
     results.write_rvs('star', star_rvs_file, all_orders=True)
     
         
     print("results saved as: {0} & {1}".format(results_file, star_rvs_file))
     print("-----------------------------")    
-    print("HARPS pipeline std = {0:.3f} m/s".format(np.std(data.pipeline_rvs + data.bervs)))
+    print("HARPS pipeline std = {0:.3f} m/s".format(np.std(data.pipeline_rvs)))
     print("wobble std = {0:.3f} m/s".format(np.std(results.star_time_rvs)))        
     print("total runtime:{0:.2f} minutes".format((time() - start_time)/60.0))

### 2019-07-29 wobble/data.py

@@ -57,6 +57,10 @@ class Data(object):
         Optional N-epoch array of instrumental drifts.
     filelist : `np.ndarray`, optional
         Optional N-epoch array of file locations for each observation.
+    epochs : `np.ndarray`
+        N-epoch array of indices. In most cases this will be equivalent
+        to np.arange(N), but differences will arise when individual epochs have been
+        dropped from the original data file in post-read-in processing.
     orders : `list`
         R-order list of echelle orders contained in this Data object. 
         Represents a mapping from the order indices in this object to
@@ -69,6 +73,7 @@ class Data(object):
             setattr(self, attr, [])
         for attr in np.append(REQUIRED_1D, OPTIONAL_1D):
             setattr(self, attr, np.array([]))
+        self.epochs = []
         self.N = 0 # number of epochs
         self.R = 0 # number of echelle orders
         
@@ -111,7 +116,8 @@ class Data(object):
                 new = getattr(sp,attr)
             except: # fail silently
                 new = 0.
-            setattr(self, attr, np.append(getattr(self,attr), new))            
+            setattr(self, attr, np.append(getattr(self,attr), new))     
+        self.epochs = np.append(self.epochs, self.N)       
         self.N += 1
         self.empty = False
         
@@ -138,7 +144,7 @@ class Data(object):
             epochs_to_keep = [np.delete(r, i, axis=0) for r in getattr(self, attr)]
             setattr(sp, attr, epoch_to_split)
             setattr(self, attr, epochs_to_keep)
-        for attr in np.append(REQUIRED_1D, OPTIONAL_1D):
+        for attr in mp.append(np.append(REQUIRED_1D, OPTIONAL_1D), ['epochs']):
             all_epochs = getattr(self, attr)
             setattr(sp, attr, all_epochs[i])
             setattr(self, attr, np.delete(all_epochs, i))

### 2019-07-29 wobble/model.py

@@ -172,8 +172,8 @@ class Model(object):
         session.run(tf.global_variables_initializer())
 
     def optimize(self, niter=100, save_history=False, basename='wobble',
-                 verbose=True, rv_uncertainties=True, 
-                 template_uncertainties=False, **kwargs):
+                 movies=False, epochs_to_plot=[0,1,2], verbose=True, 
+                 rv_uncertainties=True, template_uncertainties=False, **kwargs):
         """Optimize the model!
             
         Parameters
@@ -183,8 +183,14 @@ class Model(object):
         save_history : `bool` (default `False`)
             If `True`, create a wobble History object to track progress across 
             iterations and generate plots.
+        movies : `bool` (default `False`)
+            Use with `save_history`; if `True`, will generate animations of 
+            optimization progress.
+        epochs_to_plot : `list` (default `[0,1,2]`)
+            Use with `save_history`; indices of epochs to plot fits for. Each
+            epoch will generate its own plot/movies.
         basename : `str` (default `wobble`)
-            Path/name to use when saving plots. Only accessed if save_history = `True`.
+            Use with `save_history`; path/name stem to use when saving plots.
         verbose : `bool` (default `True`)
             Toggle print statements and progress bars.
         rv_uncertainties : `bool` (default `True`)
@@ -215,7 +221,7 @@ class Model(object):
         self.results.ys_predicted[self.r] = session.run(self.synth)
         # save optimization plots:
         if save_history:
-            history.save_plots(basename)
+            history.save_plots(basename, movies=movies, epochs_to_plot=epochs_to_plot)
             return history
             
     def estimate_uncertainties(self, verbose=True, rvs=True, templates=False):

### 2019-07-29 wobble/results.py

@@ -9,7 +9,7 @@ T = tf.float64
 from .utils import get_session
 
 COMMON_ATTRS = ['R', 'N', 'orders', 'component_names', 'bervs', 'pipeline_rvs', 
-                'pipeline_sigmas', 'drifts', 'dates', 'airms'] # common across all orders & components
+                'pipeline_sigmas', 'drifts', 'dates', 'airms', 'epochs'] # common across all orders & components
 COMPONENT_NP_ATTRS = ['K', 'r', 'rvs_fixed', 'ivars_rvs', 'template_ivars', 'scale_by_airmass', 'learning_rate_rvs', 
                       'learning_rate_template', 'L1_template', 'L2_template']
 OPT_COMPONENT_NP_ATTRS = ['learning_rate_basis', 'L1_basis_vectors', 'L2_basis_vectors', 'L2_basis_weights'] # it's ok if these don't exist
@@ -173,7 +173,8 @@ class Results(object):
                         continue
             self.component_names = [a.encode('utf8') for a in self.component_names] # h5py workaround
             for attr in COMMON_ATTRS:
-                f.create_dataset(attr, data=getattr(self, attr))                    
+                f.create_dataset(attr, data=getattr(self, attr))  
+            self.component_names = [a.decode('utf8') for a in self.component_names] # h5py workaround                  
                 
     def combine_orders(self, component_name):
         """Calculate and save final time-series RVs for a given component after all 

### 2019-07-29 wobble/results.py
reverting to definitely working version

@@ -115,7 +115,11 @@ class Results(object):
         print("Results: reading from {0}".format(filename))
         with h5py.File(filename,'r') as f:
             for attr in COMMON_ATTRS:
-                setattr(self, attr, np.copy(f[attr]))
+                try:
+                    setattr(self, attr, np.copy(f[attr]))
+                except KeyError:
+                    setattr(self, attr, 0)
+                    print("WARNING: attribute {0} could not be read; filling with placeholder.".format(attr))
             self.component_names = np.copy(f['component_names'])
             self.component_names = [a.decode('utf8') for a in self.component_names] # h5py workaround
             all_order_attrs = []

### 2019-07-29 paper/paper.tex

--- 2019-07-05 paper/paper.tex
@@
 \include{preamble}
-\usepackage{ bbold }
 \newcommand{\Mdwarf}{Barnard's Star\xspace} % or should it be Barnard's star? or GJ 699?
@@
 
-\graphicspath{ {figures/} }
+%\graphicspath{ {figures/} }
 \DeclareGraphicsExtensions{.pdf,.eps,.png}
@@
 
-Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is \added{derived from} a single observation taken for this purpose \citep[e.g.][]{Butler1996}.
+Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is a single observation taken for the purpose \citep[e.g.][]{Butler1996}.
 Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
@@
 With a simple data-driven model, one could learn both stellar and telluric spectra simultaneously with the stellar \RVs.
+
+%\todo{(talk about lack of open-source code and difficulty of modifying \RV pipelines here?)}
+
 Here we develop and implement such a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch. 
@@
 
-Our approach \added{to the \EPRV problem} is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
-\added{Similar approaches of learning template spectra for stars and tellurics from the data directly have been used in the literature, although they have not previously been demonstrated to achieve precise stellar \RVs \citep[e.g.]{Hadrava2004, Hadrava2006}.}
+Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
+%In this sense, ours is a true forward model.  -- this content-less (DFM)
 
@@
 First, we assume that the wavelength calibration and spectral extraction of the instrument are perfect: that is, we begin at the stage of having 1D extracted spectra and corresponding wavelength grids in hand and we do not model any corrections to the wavelength solution.
-\added{Similarly, we assume that the line spread function of the instrument remains perfectly constant from one exposure to the next.
-This assumption is needed because it allows us to extract instrumentally-broadened template spectra rather than explicitly modeling and solving for instrumental broadening effects.
-However, \todo{as we discuss in Section \ref{s:future}}, even within the limits of this assumption extracting already-broadened templates is not strictly correct.
-We do this for simplicity only and leave the general case to future work.}
 
@@
 
-We take the data to be the $M \times N$ matrix y, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
-We also have a corresponding $M \times N$ matrix of wavelength solutions\added{ which we call $\xi$, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.
-\deleted{Again, the entries of this matrix are in logarithmic form ($\ln \lambda_{m,n}$). This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.}
+We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
+We also have a corresponding $M \times N$ matrix of wavelength solutions. 
+Again, the entries of this matrix are in logarithmic form ($\ln \lambda$). 
+This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.
 
-\replaced{We model each data column $y_n$}{For each data column $y_n$, our model prediction $f_n$ can be treated} as the sum of stellar and telluric contributions:
+We model each data column $y_n$ as the sum of the stellar and telluric spectra:
 \begin{equation}
-f_n = f_{\star,n} + f_{t,n} + \textrm{noise}.
+y_n = y_{n}^\star + y_{n}^t + noise.
 \end{equation}
@@
 \begin{equation}
-f_{\star,n} = P(\xi_n, \xi_{\star}, v_{obs,n}) \cdot \mu_{\star},
+y_{n}^\star = P(v^{obs}_{n}) \mu^{\star},
 \label{eqn:star}
 \end{equation}
-where $\mu_{\star}$ is a spectral template of log-fluxes and 
-$\xi_{\star}$ is the corresponding vector of template log-wavelengths. 
-The exact values of $\xi_{\star}$ can be chosen somewhat arbitrarily under the conditions that the grid has uniform spacing $\Delta\xi_{\star}$, covers the entire wavelength range of the data, and is over-sampled with regards to the observed spectrum's wavelength grid.
-$P$ is a linear operator whose function is to apply a Doppler shift by observed velocity $v_{obs,n}$ and
-interpolate $\mu_{\star}$ from the $\xi_{\star}$ template grid to the $\xi_n$ data grid.
-Each entry of the $P$ matrix can be defined by a sum of weighted indicator function as follows:
+where $\mu^{\star}$ is a spectral template with length $M'$ corresponding to
+the length of an arbitrarily-chosen wavelength grid of control points, and
+$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
+interpolates $\mu^\star$ into the observatory rest frame.
+Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
 \begin{equation}
-\begin{split}
- P_{i,j} =  &\ \Big(\frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big(0 < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 1\big) \\
-  &+ \Big(1 - \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big({-1} < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 0\big), \\
- \end{split}
+ \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
 \end{equation}
-where $\xi'_{\star}$ is the Doppler-shifted template grid:
-\begin{equation}
- \xi'_{\star}(v) = \xi_{\star} + D(v) \equiv \xi_{\star} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
-\end{equation}
-We note that because $P$ is quite sparse, in practice we do not instantiate the full matrix when performing calculations.
+The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
+$P$ could be represented as a rectangular $M \times M'$ sparse matrix, but the dot product can be implemented efficiently using linear interpolation instead of instantiating the full matrix.
 
-The apparent stellar \RV, $v_{obs,n}$, is a \replaced{sum}{combination} of the star's actual velocity \replaced{about its center of mass}{in the barycentric reference frame} $v_{\star,n}$ and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
-\added{For the purposes of this work, we use the \BERV furnished by the \HARPS pipeline. 
-We also make the approximation that the observed velocity is a simple sum of the actual velocity and the \BERV. 
-In detail, this approximation is incorrect; see \citet{Wright2014} and \citet{Wright2019} for further discussion.}
+The apparent stellar \RV, $v^{obs}_{n}$, is a sum of the star's actual velocity about its center of mass ($v^{\star}_{n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
 
@@
 \begin{equation}
-f_{t,n} =  A_n(\mu_{t} + W_{t} \cdot z_n).
+y_{n}^t =  A_n(\mu^{t} + W^{t} z_n).
 \end{equation}
-In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
-$W_t$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $M'$ is the length of some vector of template wavelengths $\xi_{t}$ and $K$ is the number of basis vectors used.\footnote{While we refer to these as basis vectors for simplicity, note that they are not actually constrained to be orthogonal.}
+In addition to its mean spectrum $\mu^t$, the telluric component also includes a time-dependent term assembled from two variables: $W^t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
+$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
 For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
-The template wavelength grid $\xi_{t}$ should have similar properties to $\xi_{\star}$ and may in practice be the same grid. 
-Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
+The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
 
@@
 \begin{equation}
-\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - f_{n})^T C_{n}^{-1} (y_{n}-f_{n}),
+\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n}),
 \label{eqn:lnlike}
 \end{equation}
-comparing our model $f$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.
+comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.
 
@@
 \begin{split}
-\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - f_{n})^T C_{n}^{-1} (y_{n}-f_{n})  \\
- & + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}}   + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}} \\
- &  + \Lone{\lambda_5}{W_t}  + \Ltwo{\lambda_6}{W_t} + \Lone{1.0}{z} ,
+\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  \\
+ & + \Lone{\lambda_1}{\mu^{\star}} + \Ltwo{\lambda_2}{\mu^{\star}}   + \Lone{\lambda_3}{\mu^{t}} + \Ltwo{\lambda_4}{\mu^{t}} \\
+ &  + \Lone{\lambda_5}{W^t}  + \Ltwo{\lambda_6}{W^t} + \Lone{1.0}{z^t} ,
 \end{split}
@@
 The basis-weight regularization amplitude can be set arbitrarily to unity because there is a perfect degeneracy between amplifying the basis vectors and attenuating the basis weights.
-This regularization choice breaks that degeneracy.
+This regularization choice breaks that perfect degeneracy.
 The basis weights are regularized to encourage our desired outcome where the mean telluric spectrum contains as much telluric information as possible, while the variable basis picks up only the necessary time-variable changes.
@@
 
-\acknowledgements{\added{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful and constructive review.}
-It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
+\acknowledgements{It is a pleasure to acknowledge Jacob L. Bean, John M. Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, and Julian St\"{u}rmer for helpful discussions. 
 We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.

### 2019-07-29 wobble/results.py
Revert "reverting to definitely working version"
This reverts commit 9fce3f8b77edffb5356cd88bd60534344f1e9b39.

@@ -117,11 +117,7 @@ class Results(object):
         print("Results: reading from {0}".format(filename))
         with h5py.File(filename,'r') as f:
             for attr in COMMON_ATTRS:
-                try:
-                    setattr(self, attr, np.copy(f[attr]))
-                except KeyError:
-                    setattr(self, attr, 0)
-                    print("WARNING: attribute {0} could not be read; filling with placeholder.".format(attr))
+                setattr(self, attr, np.copy(f[attr]))
             self.component_names = np.copy(f['component_names'])
             self.component_names = [a.decode('utf8') for a in self.component_names] # h5py workaround
             self.ys_predicted = [0 for r in range(self.R)]

### 2019-08-12 paper/paper.tex
Revert "submission version"
This reverts commit 3067c7736e0f0a8902027f94205e8490ca28d83f.
updated figures

@@
 
-%\graphicspath{ {figures/} }
+\graphicspath{ {figures/} }
 \DeclareGraphicsExtensions{.pdf,.eps,.png}
@@
 \includegraphics[width=5in]{51peg_planet}
-\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered from the \wobble\ \RVs (left) and from the \HARPS standard pipeline.
+\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered from the \wobble\ \RVs (left) and from the \HARPS standard pipeline (right).
 While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns.
@@
 \startdata
-$K$ & m s$^{-1}$ & $55.57^{+2.28}_{-2.04}$ & $53.84^{+1.96}_{-2.20}$ \\
-$P$ & days & $4.2292^{+0.0003}_{-0.0003}$ & $4.2294^{+0.0003}_{-0.0003}$ \\
-$t_0$ & JD & $2456546.89^{+0.02}_{-0.02}$ & $2456546.94^{+0.01}_{-0.02}$ \\
+$K$ & m s$^{-1}$ & $56.30^{+2.42}_{-2.05}$ & $54.02^{+1.84}_{-1.89}$ \\
+$P$ & days & $4.2289^{+0.0004}_{-0.0004}$ & $4.2293^{+0.0003}_{-0.0003}$ \\
+$t_0$ & JD & $2456543.85^{+0.44}_{-0.47}$ & $2456546.41^{+0.27}_{-3.78}$ \\
 $e$ & $-$  & $0.03^{+0.02}_{-0.02}$ & $0.04^{+0.02}_{-0.02}$ \\
-$\omega$ & rad & $0.45^{+0.62}_{-1.00}$ & $-1.79^{+0.34}_{-0.32}$ \\
-$s$ & m s$^{-1}$ & $0.01^{+0.08}_{-0.01}$ & $0.74^{+0.06}_{-0.07}$ \\
+$\omega$ & rad & $0.91^{+4.99}_{-0.64}$ & $4.51^{+0.32}_{-0.35}$ \\
+$s$ & m s$^{-1}$ & $0.01^{+0.10}_{-0.01}$ & $0.72^{+0.07}_{-0.06}$ \\
 \enddata
@@
 \caption{Three example echelle orders at a randomly chosen epoch from the \HARPS observations of 51 Pegasi.
-Data are shown as dark circles, while the \wobble predictions for the stellar and telluric spectral contributions at this epoch are red and blue lines, respectively.
-Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered.
+\replaced{Data are shown as dark circles, while the \wobble predictions for the stellar and telluric spectral contributions at this epoch are red and blue lines, respectively.}{Data and best-fit models are plotted in black, while the predicted stellar and telluric spectral contributions from the model fit are plotted in red and blue with arbitrary flux offsets for clarity.} Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered.
 Residuals after subtracting both star and telluric models are shown below each spectrum.}
@@
 \centering
-\includegraphics[width=5in]{telluric_basis2}
+\includegraphics[width=5in]{telluric_basis}
 \caption{Example of the mean telluric spectral template in a small wavelength region (upper panel) and the three basis vectors used to capture the time-variable contributions to the spectrum, staggered by an arbitrary offset for display (lower panel).}
@@
 As we might expect, many of the lines co-vary with each other: as all known telluric lines in this region arise from H$_2$O, this likely corresponds to time-variable levels of atmospheric water vapor content.
-In general, the basis vectors are suggestive of Doppler shifts, perhaps due to wind speeds in different layers of the atmosphere contributing individually to the overall line shapes. 
-These hypotheses could be tested in the future by seeking correlations between basis weights and externally-measured weather diagnostics. 
+\deleted{In general, the basis vectors are suggestive of Doppler shifts, perhaps due to wind speeds in different layers of the atmosphere contributing individually to the overall line shapes.}
+\replaced{These hypotheses}{This hypothesis} could be tested in the future by seeking correlations between basis weights and externally-measured weather diagnostics. 
 %\todo{(is this enough? also show that basis weights change seasonally-ish?)}
@@
 In principle, if the changes to the \LSF were well-understood they could be hard-coded into linear operators like the $P$ operator used in Equation \ref{eqn:star}, leaving the essential equations of \wobble largely unchanged.
-Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.
+\replaced{Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.}{In fact, it is worth noting at this point that the \wobble method does not deal correctly with the instrumental \LSF even for a perfectly stable spectrograph; the equations presented in this work disentangle \LSF-broadened telluric and stellar templates from each other, whereas in reality those templates should be combined \textit{before} the \LSF is applied.
+For the purposes of this work, where the \LSF is assumed to be time-invariant, the method demonstrably works despite this flaw.
+However, for other applications the \LSF variations might need to be included as a data-driven model component within a second interpolation operator that acts on the sum of the stellar and telluric model predictions.}
 This leads us into the territory of making more fundamental changes to the \wobble model.

### 2019-08-12 scripts/script_51peg.py

@@ -19,7 +19,7 @@ if __name__ == "__main__":
     tellurics_reg_file = '../wobble/regularization/{0}_t_K{1}.hdf5'.format(starname, K_t)
     plot_dir = '../results/plots_{0}_Kstar{1}_Kt{2}/'.format(starname, K_star, K_t)
     
-    if True:
+    if False:
         # quick test on single order
         data = wobble.Data(starname+'_e2ds.hdf5', filepath='../data/', orders=[30, 56])
         results = wobble.Results(data=data)
@@ -40,7 +40,7 @@ if __name__ == "__main__":
     print("running wobble on star {0} with K_star = {1}, K_t = {2}".format(starname, K_star, K_t))
     start_time = time()
     orders = np.arange(72)
-    data = wobble.Data(starname+'_e2ds.hdf5', filepath='../data/', orders=orders)
+    data = wobble.Data(starname+'_e2ds-keep.hdf5', filepath='../data/', orders=orders)
     results = wobble.Results(data=data)
     
     print("data loaded")
@@ -111,7 +111,7 @@ if __name__ == "__main__":
     print("final RVs calculated.")
     print("time elapsed: {0:.2f} minutes".format((time() - start_time)/60.0))
         
-    results_file = '../results/results_{0}_Kstar{1}_Kt{2}.hdf5'.format(starname, K_star, K_t)
+    results_file = '../results/results_{0}_Kstar{1}_Kt{2}-keep.hdf5'.format(starname, K_star, K_t)
     results.write(results_file)
         
     print("results saved as: {0}".format(results_file))

### 2019-08-12 paper/paper.tex
overwriting with updated paper

--- 2019-07-26 paper/paper.tex
@@
 As we might expect, many of the lines co-vary with each other: as all known telluric lines in this region arise from H$_2$O, this likely corresponds to time-variable levels of atmospheric water vapor content.
-In general, the basis vectors are suggestive of Doppler shifts, perhaps due to wind speeds in different layers of the atmosphere contributing individually to the overall line shapes. 
+\deleted{In general, the basis vectors are suggestive of Doppler shifts, perhaps due to wind speeds in different layers of the atmosphere contributing individually to the overall line shapes.} 
 These hypotheses could be tested in the future by seeking correlations between basis weights and externally-measured weather diagnostics. 

### 2019-08-13 paper/paper.tex
integrated all changes to date (i think)

@@
 
-One of the primary contributors to the noise budget in \RV measurements is the incomplete treatment of telluric features in the
-Earth's atmosphere \citep{Halverson2016}. 
+One of the primary contributors to the noise budget in \RV measurements is the incomplete treatment of telluric features in the Earth's atmosphere \citep{Halverson2016}. 
 Often, particular sections of a spectrum that are likely to feature telluric features are identified before the velocity shift of the stellar spectrum is inferred. 
@@
 
-Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is \added{derived from} a single observation taken for this purpose \citep[e.g.][]{Butler1996}.
+Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is \changed{derived from} a single observation taken for this purpose \citep[e.g.][]{Butler1996}.
 Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
@@
 
-Our approach \added{to the \EPRV problem} is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
-\added{Similar approaches of learning template spectra for stars and tellurics from the data directly have been used in the literature, although they have not previously been demonstrated to achieve precise stellar \RVs \citep[e.g.]{Hadrava2004, Hadrava2006}.}
+Our approach \changed{to the \EPRV problem} is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
+\changed{Similar approaches of learning template spectra for stars and tellurics from the data directly have been used in the literature, although they have not previously been demonstrated to achieve precise stellar \RVs \citep[e.g.]{Hadrava2004, Hadrava2006}.}
 
@@
 First, we assume that the wavelength calibration and spectral extraction of the instrument are perfect: that is, we begin at the stage of having 1D extracted spectra and corresponding wavelength grids in hand and we do not model any corrections to the wavelength solution.
-\added{Similarly, we assume that the line spread function of the instrument remains perfectly constant from one exposure to the next.
+\changed{Similarly, we assume that the line spread function of the instrument remains perfectly constant from one exposure to the next.
 This assumption is needed because it allows us to extract instrumentally-broadened template spectra rather than explicitly modeling and solving for instrumental broadening effects.
@@
 We take the data to be the $M \times N$ matrix $y$, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
-We also have a corresponding $M \times N$ matrix of wavelength solutions\added{ which we call $\xi$, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.
-\deleted{Again, the entries of this matrix are in logarithmic form ($\ln \lambda_{m,n}$). This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.}
+We also have a corresponding $M \times N$ matrix of wavelength solutions \changed{which we call $\xi$, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.
 
-\replaced{We model each data column $y_n$}{For each data column $y_n$, our model prediction $f_n$ can be treated} as the sum of stellar and telluric contributions:
+\changed{For each data column $y_n$, our model prediction $f_n$ can be treated} as the sum of stellar and telluric contributions:
 \begin{equation}
@@
 
-The stellar spectrum contribution is: 
+\changed{The stellar spectrum contribution is: 
 \begin{equation}
@@
 At this point, it is also useful to note that $P$ could, in principle, encode an instrumental line spread function (\LSF). 
-We return to this point in Section \ref{s:future}.
+We return to this point in Section \ref{s:future}.}
 
-The apparent stellar \RV, $v_{obs,n}$, is a \replaced{sum}{combination} of the star's actual velocity \replaced{about its center of mass}{in the barycentric reference frame} $v_{\star,n}$ and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
-\added{For the purposes of this work, we use the \BERV furnished by the \HARPS pipeline. 
+The apparent stellar \RV, $v_{obs,n}$, is a \changed{combination} of the star's actual velocity \changed{in the barycentric reference frame} $v_{\star,n}$ and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.
+\changed{For the purposes of this work, we use the \BERV furnished by the \HARPS pipeline. 
 We also make the approximation that the observed velocity is a simple sum of the actual velocity and the \BERV. 
@@
 For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
-The template wavelength grid $\xi_{t}$ should have similar properties to $\xi_{\star}$. 
+\changed{The template wavelength grid $\xi_{t}$ should have similar properties to $\xi_{\star}$.} 
 Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
@@
 We set regularization amplitudes for \wobble using a cross-validation scheme. 
-In brief, we randomly select 12.5\% of the total epochs to set aside as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set). 
+In brief, we randomly select $10-15$\% of the total epochs to set aside as a validation set and, using some value of $\lambda$, run the model optimization on the remaining epochs (the training set). 
 The resulting best-fit spectral templates and basis vectors are taken as fixed and the time-dependent terms only (\RVs and basis weights) are optimized for the validation epochs. 
@@
 
-For the purposes of this work, we chose to test the performance of \wobble on archival spectra from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}. 
+For the purposes of this work, we chose to test the performance of the \wobble method on archival spectra from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}.
 \HARPS has operated continuously since 2003 and as such has an extensive catalog of publicly available data. 
@@
 After masking these data, we convert the flux to log space and continuum normalize by fitting and subtracting a polynomial to an asymmetrically clipped subset of the data. 
-\added{We found good results from clipping all pixels outside of the range [$-$0.3$\sigma$, $+$3$\sigma$] as a way of effectively removing the absorption features; fitting a sixth-order polynomial; and iteratively repeating with the pixel clipping set by the residuals to the previous fit until the selection of clipped pixels is stable.}
+\changed{We found good results from clipping all pixels outside of the range [$-$0.3$\sigma$, $+$3$\sigma$] as a way of effectively removing the absorption features; fitting a sixth-order polynomial; and iteratively repeating with the pixel clipping set by the residuals to the previous fit until the selection of clipped pixels is stable.
+This was done independently for each echelle order.}
 
@@
 This is the best Poisson estimate we can make given that we are working in the log space, and we don't have individual-pixel uncertainty estimates. 
-Technically a small bias is introduced by taking the logarithm of the flux, but this bias is very small at the relevant signal-to-noise ratios; furthermore, the small bias acts in the flux direction, not the radial-velocity direction.
+Technically a small bias is introduced by taking the logarithm of the flux, but this bias is very small at the relevant signal-to-noise ratios; furthermore, the small bias \changed{in flux is unlikely to map onto the radial velocities obtained.}%acts in the flux direction, not the radial-velocity direction.
 
@@
 \caption{Three example echelle orders at a randomly chosen epoch from the \HARPS observations of 51 Pegasi.
-Data are shown as dark circles, while the \wobble predictions for the stellar and telluric spectral contributions at this epoch are red and blue lines, respectively.
-Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered.
+\changed{Data and best-fit models are plotted in black, while the predicted stellar and telluric spectral contributions from the model fit are plotted in red and blue with arbitrary flux offsets for clarity.} Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered.
 Residuals after subtracting both star and telluric models are shown below each spectrum.}
+
 \label{fig:51peg_spectrum}
@@
 It is worthy of note that the stellar spectral model shown in Figure \ref{fig:51peg_spectrum} is an extremely high-\SNR spectrum of 51 Peg. 
-\added{The uncertainties on this spectrum can be estimated from the negative inverse Hessian, similarly to the \RVs.
+\changed{The uncertainties on this spectrum can be estimated from the negative inverse Hessian, similarly to the \RVs.
 Doing this calculation for a subset of the template in the continuum around 6000 \ang indicates an approximate \SNR of \todo{xxx}, which is consistent with the expected \SNR from co-adding each individual spectrum.
 }
-\replaced{This template spectrum is essentially a time-averaged composite spectrum.}{In essence, this template is a time-averaged and telluric-cleaned composite stellar spectrum.} 
+\changed{In essence, this template is a time-averaged and telluric-cleaned composite stellar spectrum.} 
 Both it and its residuals as a function of time are scientifically valuable outputs of the \wobble method.
@@
 As we might expect, many of the lines co-vary with each other: as all known telluric lines in this region arise from H$_2$O, this likely corresponds to time-variable levels of atmospheric water vapor content.
-\deleted{In general, the basis vectors are suggestive of Doppler shifts, perhaps due to wind speeds in different layers of the atmosphere contributing individually to the overall line shapes.} 
-These hypotheses could be tested in the future by seeking correlations between basis weights and externally-measured weather diagnostics. 
+%\deleted{In general, the basis vectors are suggestive of Doppler shifts, perhaps due to wind speeds in different layers of the atmosphere contributing individually to the overall line shapes.} 
+This hypothesis could be tested in the future by seeking correlations between basis weights and externally-measured weather diagnostics. 
 %\todo{(is this enough? also show that basis weights change seasonally-ish?)}
@@
 Another way a less well-calibrated instrument may violate \wobble's assumptions is in the non-negligible variability of all spectral lines due to changes in the instrumental line spread function (\LSF).
-In principle, if the changes to the \LSF were well-understood they could be hard-coded into linear operators like the $P$ operator used in Equation \ref{eqn:star}, leaving the essential equations of \wobble largely unchanged.
-\added{In this case the \LSF operator should be applied to both the stellar and telluric component models after combining them.
+In principle, if the changes to the \LSF were well-understood they could be hard-coded into a linear operator like the $P$ operator used in Equation \ref{eqn:star}. 
+\changed{In this case the \LSF operator should be applied to both the stellar and telluric component models after combining them.
 Indeed, technically the current \wobble model does not deal correctly with line broadening: the templates we infer for star and tellurics are post-broadening spectra, but this goes against the correct order of operations if line broadening is dominated by instrumental effects that are applied to the \textit{combined} spectra.
@@
 
-Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to \replaced{take this approach}{take a prescriptive approach to their behavior and hard-code them into the form of the $P$ operator}; instead, we might want to fit these variations \replaced{as a data-driven model component}{using a data-driven approach, which would make the \wobble model significantly non-linear}.
+Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to \changed{take a prescriptive approach to their behavior and hard-code them into the form of the $P$ operator}; instead, we might want to fit these variations \changed{using a data-driven approach, which would make the \wobble model significantly non-linear}.
 This leads us into the territory of making more fundamental changes to the \wobble model.
@@
 
-\acknowledgements{\added{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful and constructive review.}
+\acknowledgements{\changed{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful and constructive review.}
 It is a pleasure to acknowledge Jacob L.
 Bean, John M.
-Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, Julian St\"{u}rmer, \added{and Sharon Xuesong Wang} for helpful discussions.
+Brewer, Heather Knutson, Timothy Morton, Melissa Ness, Andreas Seifahrt, Julian St\"{u}rmer, \changed{and Sharon Xuesong Wang} for helpful discussions.
 We are grateful to Hans-Walter Rix and the Max-Planck-Institut f\"ur Astronomie for their hospitality while developing key parts of this project.

### 2019-08-13 paper/paper.tex
clearly marked all changes

@@
 
-\changed{For each data column $y_n$, our model prediction $f_n$ can be treated} as the sum of stellar and telluric contributions:
+\changed{For each data column $y_n$, our model prediction $f_n$ can be treated as the sum of stellar and telluric contributions:
 \begin{equation}
 f_n = f_{\star,n} + f_{t,n} + \textrm{noise}.
-\end{equation}
+\end{equation}}
 
@@
  \xi'_{\star}(v) = \xi_{\star} + D(v) \equiv \xi_{\star} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
-\end{equation}
+\end{equation}}
 Because $P$ is quite sparse, in practice we do not instantiate the full matrix when performing calculations. 
-At this point, it is also useful to note that $P$ could, in principle, encode an instrumental line spread function (\LSF). 
+\changed{At this point, it is also useful to note that $P$ could, in principle, encode an instrumental line spread function (\LSF). 
 We return to this point in Section \ref{s:future}.}
@@
 The telluric spectrum contribution is:
-\begin{equation}
+\changed{\begin{equation}
 f_{t,n} =  A_n(\mu_{t} + W_{t} \cdot z_n).
-\end{equation}
+\end{equation}}
 In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
-$W_t$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $M'$ is the length of some vector of template wavelengths $\xi_{t}$ and $K$ is the number of basis vectors used.\footnote{While we refer to these as basis vectors for simplicity, note that they are not actually constrained to be orthogonal.}
+\changed{$W_t$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $M'$ is the length of some vector of template wavelengths $\xi_{t}$ and $K$ is the number of basis vectors used.\footnote{While we refer to these as basis vectors for simplicity, note that they are not actually constrained to be orthogonal.}}
 For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
@@
 We evaluate a log-likelihood for each epoch
+\changed{
 \begin{equation}
@@
 \label{eqn:lnlike}
-\end{equation}
-comparing our model $f$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.
+\end{equation}}
+comparing our model \changed{$f$} to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.
 
@@
 As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs. 
-We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\xi_{\star}$. 
+We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength \changed{$\xi_{\star}$}.
 The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
-Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
-The $K$ highest eigenweights and their corresponding eigenvectors are taken as the starting guess for the basis weights and vectors.
+Finally, we initialize the telluric spectrum's basis vectors \changed{$W_t$} and weights \changed{$z_t$} by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
+The $K$ highest eigenweights and their corresponding eigenvectors are taken as \changed{the starting guess for} the basis weights and vectors.
 
@@
 
-For the purposes of this work, we chose to test the performance of the \wobble method on archival spectra from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}.
+For the purposes of this work, we chose to test the performance of \changed{the \wobble method} on archival spectra from the High Accuracy Radial Velocity Planet Searcher (\HARPS) spectrograph \citep{Mayor2003}.
 \HARPS has operated continuously since 2003 and as such has an extensive catalog of publicly available data. 
@@
 All data were obtained from the \acronym{ESO} public data archive in the form of ``\code{e2ds}'' spectra.\footnote{Based on observations made with ESO Telescopes at the La Silla Paranal Observatory under programme IDs 091.C-0271, 183.C-0437, 191.C-0505, 072.C-0488, 089.C-0497, 099.C-0880, and 60.A-9700.}
-These data come as extracted order-by-order 1D spectra in blocks of 72 orders by 4096 pixels per order.
+These data come as extracted \changed{order-by-order} 1D spectra in blocks of 72 orders by 4096 pixels per order.
 The airmass of the observation and the calculated \BERV are both provided in the \acronym{FITS} header of the spectral data file. 

### 2019-08-14 paper/paper.tex
edits to title, abstract, conclusion

@@
 
-\title{\textsc{\wobble: a data-driven method for precision radial velocities}}
+%\title{\textsc{\wobble: a data-driven method for precision radial velocities}}
+\title{\textsc{\wobble: a data-driven \changed{analysis technique for time-series stellar spectra}}}
 
@@
 \begin{abstract}
+
+\changed{In recent years, dedicated extreme-precision radial velocity (\EPRV) spectrographs have produced vast quantities of high-resolution, high-signal-to-noise time-series spectra for bright stars.
+These data contain valuable information for the dual purposes of planet detection via the measured \RVs and stellar characterization via the co-added spectra. 
+However, considerable data analysis challenges exist in extracting these data products from the observed spectra at the highest possible precision, including the issue of poorly-characterized telluric absorption features and the common use of an assumed stellar spectral template.
+In both of these examples, precision-limiting reliance on external information can be sidestepped using the data directly. } 
+%The most widely-used methods for \RV extraction neglect to make use of the full information content of the spectral time series.}
 % Context
-Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra. 
-Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features which are not perfectly known. 
-Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
-In both cases, precision-limiting reliance on external information can be sidestepped using the data directly. 
+%Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra. 
+%Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features which are not perfectly known. 
+%Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
+%In both cases, precision-limiting reliance on external information can be sidestepped using the data directly. 
 % Aims
@@
 
-One of the primary contributors to the noise budget in \RV measurements is the incomplete treatment of telluric features in the Earth's atmosphere \citep{Halverson2016}. 
+One \changed{major contributor} to the noise budget in \RV measurements is the incomplete treatment of telluric features in the Earth's atmosphere \citep{Halverson2016}. 
 Often, particular sections of a spectrum that are likely to feature telluric features are identified before the velocity shift of the stellar spectrum is inferred. 
@@
 Such data-driven templates have the benefit of being both customized to the star in question and high-\SNR without requiring observing overhead. 
+\changed{Moreover, these templates contain valuable scientific information, as an optimally-combined stellar spectral template will yield the most precise possible constraints on the spectroscopic parameters and abundances for the star.}
 
@@
 Our approach \changed{to the \EPRV problem} is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
-\changed{Similar approaches of learning template spectra for stars and tellurics from the data directly have been used in the literature, although they have not previously been demonstrated to achieve precise stellar \RVs \citep[e.g.]{Hadrava2004, Hadrava2006}.}
+\changed{Similar approaches of learning template spectra for stars and tellurics from the data directly have been used in the literature, although they have not previously been demonstrated to achieve precise stellar \RVs \citep[e.g.][]{Hadrava2004, Hadrava2006}.}
 
@@
 This assumption is needed because it allows us to extract instrumentally-broadened template spectra rather than explicitly modeling and solving for instrumental broadening effects.
-However, \todo{as we discuss in Section \ref{s:future}}, even within the limits of this assumption extracting already-broadened templates is not strictly correct.
+However, as we discuss in Section \ref{s:data-changes}, even within the limits of this assumption extracting already-broadened templates is not strictly correct.
 We do this for simplicity only and leave the general case to future work.}
@@
 This is the best Poisson estimate we can make given that we are working in the log space, and we don't have individual-pixel uncertainty estimates. 
-Technically a small bias is introduced by taking the logarithm of the flux, but this bias is very small at the relevant signal-to-noise ratios; furthermore, the small bias \changed{in flux is unlikely to map onto the radial velocities obtained.}%acts in the flux direction, not the radial-velocity direction.
+Technically a small bias is introduced by taking the logarithm of the flux, but this bias is very small at the relevant signal-to-noise ratios; furthermore, the small bias \changed{in flux is unlikely to map strongly onto the radial velocities obtained.}%acts in the flux direction, not the radial-velocity direction.
 
@@
 \changed{The uncertainties on this spectrum can be estimated from the negative inverse Hessian, similarly to the \RVs.
-Doing this calculation for a subset of the template in the continuum around 6000 \ang indicates an approximate \SNR of \todo{xxx}, which is consistent with the expected \SNR from co-adding each individual spectrum.
+Doing this calculation for a subset of the template in the continuum around 6000 \ang indicates an approximate \SNR of 3000 per 0.02-\ang template ``pixel,'' which is consistent with the expected net \SNR from co-adding each individual \HARPS spectrum.
 }
@@
 Both it and its residuals as a function of time are scientifically valuable outputs of the \wobble method. 
+\changed{The template, with its exquisite \SNR, can be used for very precise stellar characterization and abundance analysis.
+Meanwhile, the time-series residuals contain information about the stellar spectrum's non-Doppler-shift variations in time, a matter of key importance to both stellar physics and the effort to mitigate stellar ``noise'' in \EPRV measurements.}
 
@@
 Moreover, most existing \EPRV pipelines are closed-source, proprietary, and built for use with a specific instrument in mind. 
-While traditional \RV extraction methods are demonstrably suboptimal when it comes to treatment of telluric features or use of a fixed stellar template spectrum, experimenting with changes to the existing pipelines is generally not possible.
+While traditional \RV extraction methods are demonstrably suboptimal when it comes to treatment of telluric features and use of a fixed stellar template spectrum, experimenting with changes to the existing pipelines is generally not possible.
+
+\changed{The data from these \EPRV campaigns are also useful for the purpose of precise stellar characterization using the co-added spectra. 
+Since such a co-added spectrum is not a primary intended data product of \EPRV instrument pipelines, this is generally left to the user, who is again limited by the closed-source nature of most pipelines in their ability to take properly into account the true uncertainties in producing a net spectrum, or in removing the telluric absorption features that introduce noise to the stellar spectrum.}
 
@@
 Our model does not rely on physical knowledge of the star or the Earth's atmosphere. 
+\changed{It is designed with the dual aims of producing precise \RV estimates and robust stellar spectral templates from the data.}
 We implement this method in \wobble, an open-source code designed to be extensible and adaptable for a variety of data sets.
@@
 We highlight the versatility of the algorithm in Section \ref{s:future} with suggested changes and improvements for applications to different regimes. 
-It is our hope that openly shared development of next-generation \RV analysis techniques will enable the community as a whole to make full use of the information-rich data being produced by current and future \EPRV surveys.
+It is our hope that openly shared development of next-generation \changed{spectral} analysis techniques will enable the community as a whole to make full use of the information-rich data being produced by current and future \EPRV surveys.
 
@@
 
-\acknowledgements{\changed{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful and constructive review.}
+\acknowledgements{\changed{We gratefully acknowledge the referee, Ian Czekala, for a thoughtful review that substantially improved this work.}
 It is a pleasure to acknowledge Jacob L.

### 2019-08-19 paper/paper.tex
slightly better math notation

@@
 
-We take the data to be the $M \times N$ matrix $y$, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
-We also have a corresponding $M \times N$ matrix of wavelength solutions \changed{which we call $\xi$, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.
+We take the data to be the $M \times N$ matrix \mmatrix{y}, where each entry $y_{m,n}$ is the logarithm of the observed flux for pixel $m$ of $M$ at epoch $n$ of $N$. 
+We also have a corresponding $M \times N$ matrix of wavelength solutions \changed{which we call \mmatrix{\xi}, where each entry $\xi_{m,n}$ is the logarithm of the wavelength for pixel $m$ at epoch $n$}.
 
-\changed{For each data column $y_n$, our model prediction $f_n$ can be treated as the sum of stellar and telluric contributions:
+\changed{For each data column \mvec{y_n}, our model prediction \mvec{f_n} can be treated as the sum of stellar and telluric contributions:
 \begin{equation}
-f_n = f_{\star,n} + f_{t,n} + \textrm{noise}.
+\mvec{f_n} = \mvec{f_{\star,n}} + \mvec{f_{t,n}} + \textrm{noise}.
 \end{equation}}
@@
 \begin{equation}
-f_{\star,n} = P(\xi_n, \xi_{\star}, v_{obs,n}) \cdot \mu_{\star},
+\mvec{f_{\star,n}} = \mmatrix{P}(\mvec{\xi_n}, \mvec{\xi_{\star}}, v_{obs,n}) \cdot \mvec{\mu_{\star}},
 \label{eqn:star}
 \end{equation}
-where $\mu_{\star}$ is a spectral template of log-fluxes and 
-$\xi_{\star}$ is the corresponding vector of template log-wavelengths. 
-The exact values of $\xi_{\star}$ can be chosen somewhat arbitrarily under the conditions that the grid has uniform spacing $\Delta\xi_{\star}$, covers the entire wavelength range of the data, and is over-sampled with regards to the observed spectrum's wavelength grid.
-$P$ is a linear operator whose function is to apply a Doppler shift by observed velocity $v_{obs,n}$ and
-interpolate $\mu_{\star}$ from the $\xi_{\star}$ template grid to the $\xi_n$ data grid.
-Each entry of the $P$ matrix can be defined by a sum of weighted indicator functions as follows:
+where \mvec{\mu_{\star}} is a spectral template of log-fluxes and 
+\mvec{\xi_{\star}} is the corresponding vector of template log-wavelengths. 
+The exact values of \mvec{\xi_{\star}} can be chosen somewhat arbitrarily under the conditions that the grid has uniform spacing $\Delta\xi_{\star}$, covers the entire wavelength range of the data, and is over-sampled with regards to the observed spectrum's wavelength grid.
+\mmatrix{P} is a linear operator whose function is to apply a Doppler shift by observed velocity $v_{obs,n}$ and
+interpolate \mvec{\mu_{\star}} from the \mvec{\xi_{\star}} template grid to the \mvec{\xi_n} data grid.
+Each entry of the \mmatrix{P} matrix can be defined by a sum of weighted indicator functions (where an indicator function is denoted here as $\mathbb{1}(x)$ and is defined to have value 1 when condition $x$ is fulfilled and zero otherwise):
 \begin{equation}
@@
 \end{equation}
-where $\xi'_{\star}$ is the Doppler-shifted template grid:
+where \mvec{\xi'_{\star}} is the Doppler-shifted template grid:
 \begin{equation}
- \xi'_{\star}(v) = \xi_{\star} + D(v) \equiv \xi_{\star} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
+ \mvec{\xi'_{\star}}(v) = \mvec{\xi_{\star}} + D(v) \equiv \mvec{\xi_{\star}} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
 \end{equation}}
-Because $P$ is quite sparse, in practice we do not instantiate the full matrix when performing calculations. 
-\changed{At this point, it is also useful to note that $P$ could, in principle, encode an instrumental line spread function (\LSF). 
+Because \mmatrix{P} is quite sparse, in practice we do not instantiate the full matrix when performing calculations. 
+\changed{At this point, it is also useful to note that \mmatrix{P} could, in principle, encode an instrumental line spread function (\LSF). 
 We return to this point in Section \ref{s:future}.}
@@
 \changed{\begin{equation}
-f_{t,n} =  A_n(\mu_{t} + W_{t} \cdot z_n).
+\mvec{f_{t,n}} =  A_n(\mvec{\mu_{t}} + \mmatrix{W_{t}} \cdot \mvec{z_n}).
 \end{equation}}
-In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
-\changed{$W_t$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $M'$ is the length of some vector of template wavelengths $\xi_{t}$ and $K$ is the number of basis vectors used.\footnote{While we refer to these as basis vectors for simplicity, note that they are not actually constrained to be orthogonal.}}
+In addition to its mean spectrum \mvec{\mu_t}, the telluric component also includes a time-dependent term assembled from two variables: \mmatrix{W_t}, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by \mvec{z_n} to form the spectral contribution at epoch $n$.
+\changed{\mmatrix{W_t} has the shape $M' \times K$ and \mvec{z_n} is a $K$-vector, where $M'$ is the length of some vector of template wavelengths \mvec{\xi_{t}} and $K$ is the number of basis vectors used.\footnote{While we refer to these as basis vectors for simplicity, note that they are not actually constrained to be orthogonal.}}
 For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
-\changed{The template wavelength grid $\xi_{t}$ should have similar properties to $\xi_{\star}$.} 
+\changed{The template wavelength grid \mvec{\xi_{t}} should have similar properties to \mvec{\xi_{\star}}.} 
 Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
@@
 \begin{equation}
-\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - f_{n})^T C_{n}^{-1} (y_{n}-f_{n}),
+\ln \mathcal{L}_n = -\frac{1}{2} (\mvec{y_{n}} - \mvec{f_{n}})^T \mvec{C_{n}}^{-1} (\mvec{y_{n}} - \mvec{f_{n}}),
 \label{eqn:lnlike}
 \end{equation}}
-comparing our model \changed{$f$} to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.
+comparing our model \changed{\mmatrix{f}} to the data \mmatrix{y}, with \mmatrix{C} representing the covariance matrix of uncertainties for the data.
 
@@
 \end{equation}
-where $p$ is the vector of parameters to be normalized (in this case $\mu_{\star}$, $\mu_{t}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.
+where \mvec{p} is the vector of parameters to be normalized (in this case \mvec{\mu_{\star}}, \mvec{\mu_{t}}, or \mmatrix{W_{t}}) and $\lambda$ is the regularization amplitude.
 
@@
 The effectiveness of the regularization depends sensitively on the value of $\lambda$ used: if $\lambda$ is too high, real features will be lost as the parameters are forced to zero, whereas setting $\lambda$ too low will make the regularization ineffective, leaving the model vulnerable to overfitting.
-
 We set regularization amplitudes for \wobble using a cross-validation scheme. 
@@
 \begin{split}
-\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - f_{n})^T C_{n}^{-1} (y_{n}-f_{n})  \\
+\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (\mvec{y_{n}} - \mvec{f_{n}})^T \mvec{C_{n}}^{-1} (\mvec{y_{n}} - \mvec{f_{n}})  \\
  & + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}}   + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}} \\
@@
 As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs. 
-We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength \changed{$\xi_{\star}$}.
+We may then initialize the stellar spectrum \mvec{\mu_{\star}} by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength \changed{\mvec{\xi_{\star}}}.
 The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin.
-Finally, we initialize the telluric spectrum's basis vectors \changed{$W_t$} and weights \changed{$z_t$} by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
+Finally, we initialize the telluric spectrum's basis vectors \changed{\mmatrix{W_t}} and weights \changed{\mvec{z_t}} by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed.
 The $K$ highest eigenweights and their corresponding eigenvectors are taken as \changed{the starting guess for} the basis weights and vectors.
@@
 This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iterations to convergence.
-
 After convergence, we estimate the uncertainties on the parameters by
@@
 However, in principle the telluric features should be common to all \HARPS spectra, an assumption that we used for the case of HD 189733. 
-
-We show a comparison between 51 Peg and \Mdwarf telluric fits in Figure \ref{fig:telluric_comparison}. 
+\changed{As a test of this assumption, we made a comparison between the telluric template spectra derived from the 51 Peg and \Mdwarf data (Figure \ref{fig:telluric_comparison}).} 
 Although they were fit using entirely different data sets with dissimilar stellar spectra, the resulting telluric fits are indeed extremely similar. 
@@
 Another way in which the \wobble model may fail for some regimes is in its treatment of spectral variability.
-
 We have assumed that the telluric spectrum is variable in a low-dimensional sense only. 
@@
 has been provided by national institutions, in particular the institutions
-participating in the {\it Gaia} Multilateral Agreement.}
+participating in the {\it Gaia} Multilateral Agreement.
+
+Work by B.T.M. was performed under contract with the Jet
+Propulsion Laboratory (JPL) funded by NASA through
+the Sagan Fellowship Program executed by the NASA
+Exoplanet Science Institute.}
 

### 2019-08-20 paper/paper.tex
changes in re Ben's comments

@@
 We also make the approximation that the observed velocity is a simple sum of the actual velocity and the \BERV. 
-In detail, this approximation is incorrect, but it should not have a large effect on our results; see \citet{Wright2014} and \citet{Wright2019} for further discussion.}
+In detail, this approximation incorrectly assumes that the transverse component of the Earth's motion with respect to the target star is negligible, which may affect our resulting \RV precision by up to a few \ms \citep{Wright2014, Wright2019}.
+Nevertheless, because only the 1D \BERV is supplied by (and presumably used by) the \HARPS pipeline, we adopt this approximation to ensure that our results are comparable to those reported by the pipeline.}
 
@@
 \changed{\begin{equation}
-\mvec{f_{t,n}} =  A_n(\mvec{\mu_{t}} + \mmatrix{W_{t}} \cdot \mvec{z_n}).
+\mvec{f_{t,n}} =  a_n(\mvec{\mu_{t}} + \mmatrix{W_{t}} \cdot \mvec{z_n}).
 \end{equation}}
@@
 \changed{The template wavelength grid \mvec{\xi_{t}} should have similar properties to \mvec{\xi_{\star}}.} 
-Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.
+Finally, the net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $a_n$, a known quantity.
 
@@
 This is the best Poisson estimate we can make given that we are working in the log space, and we don't have individual-pixel uncertainty estimates. 
-Technically a small bias is introduced by taking the logarithm of the flux, but this bias is very small at the relevant signal-to-noise ratios; furthermore, the small bias \changed{in flux is unlikely to map strongly onto the radial velocities obtained.}%acts in the flux direction, not the radial-velocity direction.
+Technically a small bias is introduced by taking the logarithm of the flux, but this bias is very small at the relevant signal-to-noise ratios; furthermore, this small bias \changed{is in the flux direction and will not necessarily map onto the radial velocities obtained.}%acts in the flux direction, not the radial-velocity direction.
 

### 2019-08-29 paper/paper.tex
math tweaks

@@
 \changed{Similarly, we assume that the line spread function of the instrument remains perfectly constant from one exposure to the next.
-This assumption is needed because it allows us to extract instrumentally-broadened template spectra rather than explicitly modeling and solving for instrumental broadening effects.
+This assumption is needed because it allows us to extract constant instrumentally-broadened template spectra rather than explicitly modeling and solving for time-variable broadening effects.
 However, as we discuss in Section \ref{s:data-changes}, even within the limits of this assumption extracting already-broadened templates is not strictly correct.
@@
 
-\changed{For each data column \mvec{y_n}, our model prediction \mvec{f_n} can be treated as the sum of stellar and telluric contributions:
+\changed{For each data column \mvec{y_n}, our model prediction \mvec{f_n} can be treated as the sum of stellar and telluric contributions at time $n$:
 \begin{equation}
@@
 \begin{split}
- P_{i,j} =  &\ \Big(\frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big(0 < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 1\big) \\
-  &+ \Big(1 - \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\Big) \mathbb{1}\big({-1} < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 0\big), \\
+ P_{i,j} =  &\ \bigg(\frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\bigg) \cdot \mathbb{1}\bigg(0 \leq \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} < 1\bigg) \\
+  &+ \bigg(1 - \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}}\bigg) \cdot \mathbb{1}\bigg({-1} < \frac{\xi_{n,i} - \xi'_{\star,j}}{\Delta\xi_{\star}} \leq 0\bigg), \\
  \end{split}
@@
 
-We evaluate a log-likelihood for each epoch
-\changed{
+\changed{The contribution of each data epoch to the net log-likelihood may be evaluated as
 \begin{equation}
@@
 \label{eqn:lnlike}
-\end{equation}}
-comparing our model \changed{\mmatrix{f}} to the data \mmatrix{y}, with \mmatrix{C} representing the covariance matrix of uncertainties for the data.
+\end{equation}
+}with \mmatrix{C_n} representing the covariance matrix of uncertainties on the data.
 
@@
 where \mvec{p} is the vector of parameters to be normalized (in this case \mvec{\mu_{\star}}, \mvec{\mu_{t}}, or \mmatrix{W_{t}}) and $\lambda$ is the regularization amplitude.
-
 Similarly, L2 normalization adds a term of the form:

### 2019-09-17 paper/paper.tex
added zenodo reference

@@
 
-Our code is made open-source under the \acronym{MIT} license and is publicly available on GitHub.\footnote{\url{https://www.github.com/megbedell/wobble}}
+Our code is made open-source under the \acronym{MIT} license and is publicly available on GitHub.\footnote{\url{https://www.github.com/megbedell/wobble}; in this work, we use the version released as Zenodo v0.1.0 \citep{wobble-code}}
 
